 ---
title: "STA258: Statistics with Applied Probability"
author: "Nishan Mudalige, Masoud Ataei, Nurlana Alili, Xi (Bryan) Su"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output:
  bookdown::gitbook
knit: bookdown::render_book
documentclass: book
description: "Bookdown hosted on github using bookdown::gitbook."
header-includes: 
  - \usepackage{tikz}
  - \usepackage{pgfplots}
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{graphicx}
  - \usepackage{booktabs}
  - \usepackage{hyperref}
  - \usepackage{fancyvrb}
  - \usepackage{listings}
  - \usepackage{tikz}
  - \usepackage{tikz-cd}
  - \usepackage{tikzsymbols}
  - \usepackage{tikz-3dplot}
  - \usepackage{tikz-feynman}
  - \usepackage{tikz-qtree}
  - \usepackage{tikz-uml}
  - \usepackage{tikz-dependency}
  - \usepackage{tikz-graph}
  - \usepackage{color}
  - \usepackage{xcolor}
  - \usepackage[T1]{fontenc}
  - \usepackage[oldstyle,expert,lf]{ebgaramond}
---

# STA258: Statistics with Applied Probability <br> University of Toronto Mississauga {-}

<!--
```{r setup, include=FALSE}
library(rgl)                 # load first
# disable auto‑printing of every rgl scene
rgl::setupKnitr(autoprint = FALSE)   # ← key change :contentReference[oaicite:0]{index=0}

# no chunk is treated as “WebGL” by default:
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  webgl   = FALSE,   # <- OFF globally :contentReference[oaicite:1]{index=1}
  rgl     = FALSE    # <- OFF globally; rgl scenes won’t print :contentReference[oaicite:2]{index=2}
)
```
-->

<center>
```{r,echo=FALSE,warning=FALSE,webgl=TRUE,fig.show="hide"}
library(rgl); library(png); library(RCurl)

# 1. load & flip
myurl <- "https://raw.githubusercontent.com/nishanmudalige/STA258/main/Hex%20Badges/STA258_no_bg.png"
img   <- readPNG(getURLContent(myurl))
img   <- img[nrow(img):1, , ]

# 2. flatten over white
alpha      <- img[,,4]
img[,,1]   <- img[,,1] * alpha + (1 - alpha)
img[,,2]   <- img[,,2] * alpha + (1 - alpha)
img[,,3]   <- img[,,3] * alpha + (1 - alpha)
img[,,4]   <- 1

# 3. build quad
h  <- dim(img)[1]; w <- dim(img)[2]; sf <- max(w, h)/2
front <- matrix(c(
  -w/2, -h/2, 0,
  w/2, -h/2, 0,
  w/2,  h/2, 0,
  -w/2,  h/2, 0
), byrow=TRUE, ncol=3) / sf
front <- rotate3d(front, angle=pi/2, x=1, y=0, z=0)

# 4. texture coords
tex <- matrix(c(0,0, 1,0, 1,1, 0,1), byrow=TRUE, ncol=2)

# 5. draw & embed
invisible(open3d())                # no "glX 3" printed

bg3d("white")
view3d(theta=0, phi=-90, zoom=0.675) # smaller values means zoom in
invisible(quads3d(
  front,
  texture   = img,
  texcoords = tex,
  lit       = FALSE,
  col       = "white"
))
```
</center>


```{r setup2, include=FALSE}
library(rgl)                 # load first
# disable auto‑printing of every rgl scene
rgl::setupKnitr(autoprint = FALSE)   # ← key change :contentReference[oaicite:0]{index=0}

# no chunk is treated as “WebGL” by default:
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  webgl   = FALSE,   # <- OFF globally :contentReference[oaicite:1]{index=1}
  rgl     = FALSE    # <- OFF globally; rgl scenes won’t print :contentReference[oaicite:2]{index=2}
)

rgl.close() 
```


<h2 style="text-align: center; color: #FF2C21;"> Work in Progress</h1>

---

### Nishan Mudalige, Masoud Ataei, Nurlana Alili, Xi (Bryan) Su {.unlisted .unnumbered}

This is a customized e-book for the course STA258: Statistics with Applied Probability at the University of Toronto at Mississauga.

Please select a chapter in the table of contents.

---

<!-- Script for GitHub buttons to work -->
<script async defer src="https://buttons.github.io/buttons.js"></script>

<!-- Star -->
<a class="github-button" href="https://github.com/nishanmudalige/STA258" data-color-scheme="no-preference: light; light: light; dark: dark;" data-icon="octicon-star" data-size="large" aria-label="Star nishanmudalige/STA258 on GitHub">Star</a>
<!-- Watch -->
<a class="github-button" href="https://github.com/nishanmudalige/STA258/subscription" data-color-scheme="no-preference: dark; light: light; dark: dark;" data-icon="octicon-eye" data-size="large" aria-label="Watch nishanmudalige/STA258 on GitHub">Watch</a>
<!-- Fork -->
<a class="github-button" href="https://github.com/nishanmudalige/STA258/fork" data-color-scheme="no-preference: light; light: light; dark: dark;" data-icon="octicon-repo-forked" data-size="large" aria-label="Fork nishanmudalige/STA258 on GitHub">Fork</a>
<!-- Issues -->
<a class="github-button" href="https://github.com/nishanmudalige/STA258/issues" data-color-scheme="no-preference: light; light: light; dark: dark;" data-icon="octicon-issue-opened" data-size="large" aria-label="Issue nishanmudalige/STA258 on GitHub">Issue</a>
<!-- Discuss -->
<a class="github-button" href="https://github.com/nishanmudalige/STA258/discussions" data-color-scheme="no-preference: light; light: light; dark: dark;" data-icon="octicon-comment-discussion" data-size="large" aria-label="Discuss nishanmudalige/STA258 on GitHub">Discuss</a>







# About the Authors {.unnumbered}

<!-- Script for GitHub buttons to work -->
<script async defer src="https://buttons.github.io/buttons.js"></script>

### Nishan Mudalige {.unlisted .unnumbered}

**To be completed by Nishan**

<a class="github-button" href="https://github.com/nishanmudalige" data-color-scheme="no-preference: light; light: light; dark: dark;" data-size="large" aria-label="@nishanmudalige on GitHub">Follow @nishanmudalige</a>

---

### Masoud Ataei {.unlisted .unnumbered}

**To be completed by Masoud**

<a class="github-button" href="https://github.com/chimera1001" data-color-scheme="no-preference: light; light: light; dark: dark;" data-size="large" aria-label="@chimera1001 on GitHub">Follow @chimera1001</a>

---

### Nurlana Alili {.unlisted .unnumbered}

**To be completed by Nurlana**

<a class="github-button" href="https://github.com/nurlanaalili" data-color-scheme="no-preference: light; light: light; dark: dark;" data-size="large" aria-label="@nurlanaalili on GitHub">Follow @nurlanaalili</a>

---

### Xi (Bryan) Su {.unlisted .unnumbered}

**To be completed by Xi (Bryan) Su**

<a class="github-button" href="https://github.com/BryanSu426" data-color-scheme="no-preference: light; light: light; dark: dark;" data-size="large" aria-label="@BryanSu426 on GitHub">Follow @BryanSu426</a>






# Acknowledgements {-}

The authors would like to thank the following people for their feedback and review of this e-book:

- Dr. Steve Tully, Okanagan College
- Ms. Avanti Coonghe, United Nations (World Health Organization)


# Introduction and Overview {-}

This e-book is written for students in STA258 – Statistics with Applied Probability at the University of Toronto Mississauga. Its purpose is to give you a practical feel for the ideas you meet in lectures and tutorials, showing how the mathematics of probability lets us talk sensibly about uncertainty in everyday life as well as in scientific work.

Uncertainty surrounds almost everything we care about. Will tomorrow’s lecture be moved online because of a snowstorm? Can a spacecraft launch if cross-winds rise above a certain threshold? Will a new medication actually shorten recovery time once it leaves the controlled environment of a clinical trial? Each question feels different on the surface, yet the same statistical machinery helps us tackle them all. By framing an uncertain outcome as a random variable we can describe its typical behaviour, calculate the odds of unusual events and, most importantly, update those odds when new information arrives.

The human mind does a rough version of this calculation all the time. We decide to carry an umbrella because dark clouds look threatening, or we leave extra time for a commute after hearing a traffic report. But gut instinct is easily misled: streaks appear in gambling even when the dice are fair, and vivid anecdotes can drown out a mountain of quieter data. Formal probability offers a sturdier foundation. If meteorologists combine radar images, satellite readings and decades of storm records, they can say not just “it might snow,” but “there is a 70 % chance of more than ten centimetres by morning.” If engineers measure how often a particular alloy fractures under stress, they can quote the lifetime of a bridge girder with a clear margin of error rather than a hopeful guess. In each case the numbers come from a sampling distribution—an imagined catalogue of results we would see if we could repeat the experiment or observation indefinitely.

Throughout this book we will revisit that catalogue many times. We start by defining random variables and their distributions, then learn how to simulate, visualise and summarise those distributions with real data. From there we build the tools needed for inference: confidence intervals that put a reasonable bound around an unknown parameter; hypothesis tests that weigh the evidence for competing explanations; and regression models that let us tease out the influence of several factors at once. The tone is informal and example-driven—you will see as much reference to snowfall totals, pulse rates and exam scores as to abstract notation—yet every result rests on rigorous mathematics. By the end you should feel at home translating a messy, uncertain situation into symbols, pushing those symbols through the machinery of probability and translating the answer back into plain language someone with no statistical background can understand.

That ability—to move comfortably between raw experience and quantitative insight—is what STA258 is all about, and what this e-book aims to nurture.