# Descriptive Statistics

## Representing Data

Data comes to us in the form of \textit{observations} (i.e. measurements) which we write symbolically as a lower case letter along with a subscript. For example, suppose we have taken a total of $n$ observations. We have observation 1, observation 2, observation 3, $\ldots$, observation $n-1$ and observation $n$. By letting $x$ represent an observation, all $n$ observations can be represented as:

$$x_1,~x_2,~ \ldots ,~ x_n$$ 

where $x_i$ is an individual observation, and the index $i=1, \ldots , n$. Here $n$ is referred to as the \textit{sample size}. We can also represent this data in condensed form as:

$$x_i:\ i=1, \ldots , n$$

or in tabular form as:

::: {#smalltbl .smalltbl}
| Index       | 1     | 2     | $\ldots$  | $n$   |
| :---------- | :-:   | :-:   | :-:       | :-:   |
| Observation | $x_1$ | $x_2$ | $\ldots$  | $x_n$ |
:::



## Numerical Measures 

The most common numerical measures of central tendency are the mean, median, and mode. These measures summarize a set of data by identifying the central point within that data.

The \textit{mean} (or more precisely the arithmetic mean) of a data set is obtained by adding up all of the observations and dividing by the sample size (number of observations). The mean is the typical average that we are all familiar with. 
We use the symbol $\bar{x}$ to represent the sample mean.


### Mean

::: {.definition name="Mean" #Mean}
Let $x_{1}, ~x_{2}, ~x_{3}, ~\ldots , ~x_{n}$ represent a sample of $n$ observations.
The sample mean is defined as:
	$$\bar{x} = \frac{ \displaystyle\sum_{i = 1}^{n} x_{i} }{n} = \frac{ x_{1} + x_{2} + \dots + x _{n} }{n}$$
:::

### Median

::: {.definition name="Median" #Median}
The median is the middle value of a data set when the observations are arranged in ascending order.
:::

The calculation of the median depends on whether the sample size is odd or even.
If the sample size is odd, the median is the middle observation. If the sample size is even, the median is the average of the two middle observations.

Consider $n$ ordered observations $x_{(1)},x_{(2)},~\cdots~,x_{(n)}$.

- If $n$ is odd
    $$\text{Median} = \text{observation} \> \left( \frac{n + 1}{2} \right) $$
- If $n$ is even,
    $$\text{Median} = \text{average of observation} ~ \left(\frac{n}{2} \right) ~\text{and observation} ~ \left( \frac{n}{2} + 1 \right)$$
    
### Mode

::: {.definition name="Mode" #Mode}
The mode is the most frequently occurring observation in a data set.
:::

We are also interested in how the data is spread out or dispersed.
The most common numerical measures of the dispersion are the range, variance, and standard deviation. These measures summarize a set of data by identifying the spread or variability within that data.


### Variance

::: {.definition name="Variance" #Var}
The sample variance is the average squared deviation of each observation from the sample mean.

Let $x_{1}, ~x_{2}, ~x_{3}, ~\ldots , ~x_{n}$ represent a sample of $n$ observations.
The sample Variance is defined as:
	$$s^{2} = \frac{ \displaystyle\sum_{i=1}^{n} (x_{i} - \bar{x})^{2} }{n - 1}  = \frac{ (x_{1} - \bar{x})^{2} +  (x_{2} - \bar{x})^{2} + \ldots + (x_{n} - \bar{x})^{2} }{n-1}$$
where $\bar{x}$ is the sample mean in Definition \@ref(def:Mean).
:::


In the Definition \@ref(def:Var), we divide by $n-1$ instead of $n$ since we are estimating the
value of the population variance using sample data, and this calculation already includes utilizing
the sample mean $\bar{x}$ which itself is an estimate of the population mean $\mu$. 
This division by $n-1$ is known as Bessel's correction.

Another reason for dividing by $n-1$ is because the sample variance is an unbiased estimator of the population variance, which is a topic covered in [STA260](https://utm.calendar.utoronto.ca/course/sta260h5) and beyond the scope of this course.

In the calculation of the sample variance, notice each term is squared, which means that the variance is always a non-negative number. 
When this occurs, the units of the variance are the square of the units of the original data.
To return to the original units, we take the square root of the variance to obtain the standard deviation.


### Standard Deviation

::: {.definition name="Standard Deviation" #StDev}
Let $x_{1}, x_{2}, ~x_{3}, ~\ldots , ~x_{n}$ represent a sample of $n$ observations.
The sample standard deviation is defined as:
$$s = + \sqrt{s^{2}}$$
where $s^{2}$ is the sample variance in Definition \@ref(def:Var).
:::


When we analyze data we typically describe the data in terms of the mean and standard deviation
as the units are the same. However, the calculation of the sample variance is an important
intermediate step.

One of the measures of central tendency and dispersion on their own will not give as a complete picture
of the data, however when analyzed collectively, we are able to get a better overall understanding of the data.


### Range

The range is the difference between the largest and smallest observations in a data set.

::: {.definition name="Range" #Range}
Let $x_{1}, ~x_{2}, ~x_{3}, ~\ldots , ~x_{n}$ represent a sample of $n$ observations.
The range is
$$\text{Range} = \max(x_{1}, x_{2}, ~x_{3}, ~\ldots , ~x_{n}) - \min(x_{1}, x_{2}, ~x_{3}, ~\ldots , ~x_{n}) $$
:::

### Percentiles

::: {.definition name="Percentile" #Percentile}
In an ordered data set, the $p^{th}$ percentile is the value such that $p\%$ of all observations lie below it.
:::

Percentiles are a value that are relative to the rest of the data.

::: {.example}
A class writes a difficult test and a student obtained a mark of 70\% on this test.
Although 72\% is a B-, since the test was difficult, this may be a good score. 
In fact, the student who scored 72\% is in the $90^{th}$ percentile for this test. 
This means that they scored better than 90\% of the rest of the class.
:::


### Quartiles

Quartiles are special cases of percentiles.

::: {.definition name="Quartiles" #Quartiles}
In an ordered data set, quartiles are three values which divide the data into four groups such that each group consists of one fourth of the data. The three quartiles are the:

-	First Quartile (Q$_{1}$)	:	A value such that 25\% (i.e. a quarter) of all observations lie below it.
-	Second Quartile (Q$_{2}$)	:	A value such that 50\% (i.e. two quarters) of all observations lie below it.
-	Third Quartile (Q$_{3}$)	:	A value such that 75\% (i.e. three quarters) of all observations lie below it.
:::


::: {.remark}
The second quartile ($Q_2$) in Definition \@ref(def:Quartiles) is the same as the *median* in Definition \@ref(def:Median)
:::

<!-- ::: {#note:skewnote .note name="Skewness Note"} -->
<!-- **Note:** The second quartile ($Q_2$) in Definition \@ref(def:Quartiles) is the same as the *median* in Definition \@ref(def:Median). -->
<!-- ::: -->

We work with quartiles since they provide an intuitive way to partition data into quarters, and quartiles are also used in graphical summaries discussed in Section \@ref(sec:GraphicalTechniques).

A measure of dispersion involving quartiles is the interquartile range (IQR).

::: {.definition name="Interquartile Range" #IQR}
The interquartile range (IQR) is the difference between the third quartile and the first quartile.
$$\text{IQR} = Q_{3} - Q_{1}$$
where $Q_{1}$ and $Q_{3}$ are the first and third quartiles respectively in Definition \@ref(def:Quartiles).
:::

## Graphical Techniques {#sec:GraphicalTechniques} 

Raw values on their own can be difficult to interpret, especially with very large data sets. 
Graphical representations can be very useful for representing information. 
When implemented correctly, graphical plots can provide more intuitive and direct method to interpret information being analyzed.

We begin this section by introducing a term to describe the shape of a distribution of data.

::: {.definition name="Skewness" }
Skewness refers to such a measure of symmetry or lack of symmetry in the distribution of data.
:::

We can describe data as being left skewed, right skewed or symmetric. If the data is
described as symmetric, this implies that most observations are concentrated around the
mean and tail off fairly evenly on both sides of the mean. If the data is described as right
skewed, this implies that more observations are concentrated on smaller
values and we observe a longer tail to the right side of the mean. If the data is described
as left skewed, this implies that more observations are concentrated on
large values and we observe a longer tail to the left side of the mean. 
We also may have data that is bimodal, which means that there are two distinct peaks in the distribution of the data.


::: {.remark}
*Right* skewed data is also referred to as *positively* skewed and *left* skewed data is also referred to as *negatively* skewed.
:::

Figures \@ref(fig:LeftSkewExample), \@ref(fig:RightSkewExample) and \@ref(fig:SymmetryExample) illustrate types of skewness.

```{r LeftSkewExample, fig.cap="An illustration of a left (negative) skewed distribution.", echo=FALSE, fig.align='center', warning=FALSE}
library(ggplot2)
# style‑guide colors
# fill_blue   <- "#529EFF"
# line_blue   <- "#1F4E8C"
# median_blue <- "#00C1AA"
# mean_pink   <- "#FF62BC"
fill_blue   <- "#619CFF"
line_blue   <- "#0064c8"
median_blue <- "#0064c8"
mean_pink   <- "#FF68A1"

# simulate a true left‑skewed sample (Beta(5,1))
set.seed(2025)
df <- data.frame(x = rbeta(1000, 5, 1))

# density estimate (for annotation positions)
dens     <- density(df$x, adjust = 4)    # smoother density now
med_val  <- median(df$x)
mean_val <- mean(df$x) - 0.15

# y‑positions for labels
y_tail <- approx(dens$x, dens$y, xout = 0.2)$y + 0.005
y_med  <- approx(dens$x, dens$y, xout = med_val)$y + 0.3
y_mean <- approx(dens$x, dens$y, xout = mean_val)$y + 0.3

# determine top of plot to clip exactly
y_max <- max(y_tail, y_med, y_mean)

ggplot(df, aes(x = x)) +
  # shaded density
  geom_density(
    fill   = fill_blue,
    color  = line_blue,
    size   = 1.2,
    adjust = 4    # increased from 2 for extra smoothness
  ) +
  # median & mean lines
  geom_vline(xintercept = med_val,  color = median_blue, size = 1) +
  geom_vline(xintercept = mean_val, color = mean_pink,   size = 1, linetype = "dashed") +
  # tail label (left tail)
  annotate(
    "text",
    x      = 0.2,
    y      = y_tail + 0.1,
    label  = "Left Tail",
    size    = 5,
    # fontface= "bold",
    hjust   = 0.5,
    vjust   = -1.5
  ) +
  # median label to left of line
  annotate(
    "text",
    x      = med_val - 0.025,
    y      = y_med - 0.1,
    label  = "median",
    # fontface= "bold",
    hjust   = -0.5,
    size    = 5,
    color   = median_blue
  ) +
  # mean label to right of line
  annotate(
    "text",
    x      = mean_val - 0.020,
    y      = y_mean - 0.1,
    label  = "mean",
    size    = 5,
    # fontface= "bold",
    hjust   = 1.1,
    color   = mean_pink
  ) +
  # exact axis limits, no extra padding
  scale_x_continuous(limits = c(-0.0, 1.35), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, y_max), expand = c(0, 0)) +
  # clean theme, no fixed aspect
  theme_minimal(base_size = 14) +
  theme(
    panel.grid       = element_blank(),
    panel.background = element_blank(),
    plot.title       = element_text(face = "bold", hjust = 0.5),
    axis.title       = element_text(face = "bold")
  ) +
  labs(
    title = "Left-Skewed Density",
    x     = "Value",
    y     = "Density"
  )
```


```{r RightSkewExample, fig.cap="An illustration of a right (positive) skewed distribution.", echo=FALSE, fig.align='center', warning=FALSE}
library(ggplot2)

# style‑guide colors
fill_blue   <- "#619CFF"
line_blue   <- "#0064c8"
median_blue <- "#0064c8"
mean_pink   <- "#FF68A1"

# simulate right‑skewed data
set.seed(2025)
df <- data.frame(x = rbeta(1000, 2, 5))

# density estimate for positioning & y‑limit
dens     <- density(df$x, adjust = 2)
med_val  <- median(df$x)
mean_val <- mean(df$x) + 0.15

# compute annotation y‑positions
y_tail <- approx(dens$x, dens$y, xout = 0.8)$y + 0.01
y_med  <- approx(dens$x, dens$y, xout = med_val)$y + 0.3
y_mean <- approx(dens$x, dens$y, xout = mean_val)$y + 0.3

# find top of plot
y_max <- max(y_tail, y_med, y_mean)

ggplot(df, aes(x = x)) +
  # shaded density
  geom_density(
    fill   = fill_blue,
    color  = line_blue,
    size   = 1.2,
    adjust = 2
  ) +
  # median & mean lines
  geom_vline(xintercept = med_val,  color = median_blue, size = 1) +
  geom_vline(xintercept = mean_val, color = mean_pink,   size = 1, linetype = "dashed") +
  # annotations
  annotate("text",
           x      = 0.8, y      = y_tail,
           label  = "Right Tail",
           # fontface= "bold", 
           size    = 5,
           hjust = 0.25, 
           vjust = -1.5) +
  annotate("text",
           x      = med_val - 0.025, y = y_med - 0.1,
           label  = "median",
           # fontface= "bold", 
           hjust = 1.1, 
           size    = 5,
           color = median_blue) +
  annotate("text",
           x      = mean_val + 0.025, y = y_mean,
           label  = "mean",
           # fontface= "bold",
           size    = 5,
           hjust = -0.1, color = mean_pink) +
  # exact axis limits, no extra padding
  scale_x_continuous(limits = c(-0.15, 1),   expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, y_max),   expand = c(0, 0)) +
  # clean theme
  theme_minimal(base_size = 14) +
  theme(
    panel.grid       = element_blank(),
    panel.background = element_blank(),
    plot.title       = element_text(face = "bold", hjust = 0.5),
    axis.title       = element_text(face = "bold")
  ) +
  labs(
    title = "Right-Skewed Density",
    x     = "Value",
    y     = "Density"
  )
```

<!-- rgba(0,100,200,1) #00C19A -->

```{r SymmetryExample, fig.cap="An illustration of an approximately symmetric distribution.", echo=FALSE, fig.align='center', warning=FALSE}
library(ggplot2)

# style‑guide colors
fill_blue   <- "#619CFF"
line_blue   <- "#0064c8"
median_blue <- "#0064c8"
mean_pink   <- "#FF68A1"

# simulate a symmetric sample (Beta(5,5) is symmetric on [0,1])
set.seed(2025)
df <- data.frame(x = rbeta(1000, 5, 5))

# density estimate (for annotation positions)
dens     <- density(df$x, adjust = 2)
med_val  <- median(df$x) - 0.01
mean_val <- mean(df$x) + 0.01

# y‑positions for labels
y_med  <- approx(dens$x, dens$y, xout = med_val)$y + 0.3
y_mean <- approx(dens$x, dens$y, xout = mean_val)$y + 0.3

# determine top of plot to clip exactly
y_max <- max(y_med, y_mean)

ggplot(df, aes(x = x)) +
  # shaded density
  geom_density(
    fill   = fill_blue,
    color  = line_blue,
    size   = 1.2,
    adjust = 2
  ) +
  # median & mean lines
  geom_vline(xintercept = med_val,  color = median_blue, size = 1) +
  geom_vline(xintercept = mean_val, color = mean_pink,   size = 1, linetype = "dashed") +
  # median label to left of line
  annotate(
    "text",
    x       = med_val - 0.025,
    y       = y_med - 0.1,
    label   = "median",
    size    = 5,
    # fontface= "bold",
    hjust   = 1.1,
    color   = median_blue
  ) +
  # mean label to right of line
  annotate(
    "text",
    x       = mean_val + 0.025,
    y       = y_mean - 0.1,
    label   = "mean",
    size    = 5,
    # fontface= "bold",
    hjust   = -0.1,
    color   = mean_pink
  ) +
  # exact axis limits, no extra padding
  scale_x_continuous(limits = c(0, 1),    expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, y_max), expand = c(0, 0)) +
  # clean theme, no fixed aspect ratio
  theme_minimal(base_size = 14) +
  theme(
    panel.grid       = element_blank(),
    panel.background = element_blank(),
    plot.title       = element_text(face = "bold", hjust = 0.5),
    axis.title       = element_text(face = "bold")
  ) +
  labs(
    title = "Symmetric Density",
    x     = "Value",
    y     = "Density"
  )
```

<!-- ::: {#note:skewnote .note name="Skewness Note"} -->
<!-- **Note:** *Right* skewed data is also referred to as *positively* skewed and *left* skewed data is also referred to as *negatively* skewed. -->
<!-- ::: -->

Data sets can have points in it which may be unusual or unexpected. 
These terms are called *outliers*.

::: {.definition name="Outlier" }
An outlier is an unusual data point which appears to lie outside the overall pattern of the
rest of the data.
:::

In other words, an outlier falls outside the range in which we would expect to see “typical”
data values. Outliers may be present in data for a variety of reasons such as transcription
error, measurement error, or we may have just measured a rare and unusual observation. In
any case it is not a good practice to simply ignore outliers. All outliers should be investigated
before deciding whether the observation should be included in data analysis or discarded.


### Histograms

Histograms and boxplots are useful tools which can allow us to visually determine
the skewness of a distribution. By noting skewness, we get even more information about
our data. Note however that we may not always be able to determine skewness by visually
observing a histogram. A histogram is similar to a bar chart with the distinction that his-
tograms can only be created for quantitative data.

The width of the bars does not have any numerical meaning for bar charts, however bar
width matters for histograms. They are an important factor in the creation of histograms.
Interval width selection (or bin size selection) is an advanced topic that can be studied ex-
tensively and there are several rules available to select interval widths. However for the scope
of this course we will keep things simple and allow the reader to intuitively choose bin size.
After constructing a histogram, we can change the interval widths until we feel that we have
a picture that is a good representation of our data.

A histogram is constructed by dividing the range of the data into intervals (or bins) and counting how many observations fall into each interval. The intervals are usually of equal width, but this is not a requirement. 

<!-- | Class Interval       | Frequency | Relative Freq.           |        | Cumulative Freq.             | Cumulative Relative Freq.    | -->
<!-- |----------------------|:---------:|--------------------------|--------|------------------------------|------------------------------| -->
<!-- | $[a_{1}, b_{1})$     | $f_{1}$   | $r_{1} = f_{1} / F$      |        | $f_{1}$                      | $r_{1}$                      | -->
<!-- | $[a_{2}, b_{2})$     | $f_{2}$   | $r_{2} = f_{2} / F$      |        | $f_{1} + f_{2}$              | $r_{1} + r_{2}$              | -->
<!-- | $[a_{3}, b_{3})$     | $f_{3}$   | $r_{3} = f_{3} / F$      |        | $f_{1} + f_{2} + f_{3}$      | $r_{1} + r_{2} + r_{3}$      | -->
<!-- | $\vdots$             | $\vdots$  | $\vdots$                 |        | $\vdots$                     | $\vdots$                     | -->
<!-- | $[a_{m}, b_{m}]$     | $f_{m}$   | $r_{m} = f_{m} / F$      |        | $f_{1} + \ldots + f_{m} = F$ | $r_{1} + \ldots + r_{m} = 1$ | -->
<!-- |                      | $F = \displaystyle\sum_{i=1}^{m} f_{i}$       | $1$ | |                      |                              | -->

```{r histconstruct, echo=FALSE, results='asis', warning = FALSE}
library(knitr)
library(kableExtra)
class_int = c("$[a_{1}, b_{1})$", 
              "$[a_{2}, b_{2})$", 
              "$[a_{3}, b_{3})$",
              "$\\vdots$",
              "$[a_{m}, b_{m}]$",
              "")

Frequency = c("$f_{1}$",
              "$f_{2}$",
              "$f_{3}$",
              "$\\vdots$",
              "$f_{m}$",
              "$F = \\displaystyle\\sum_{i=1}^{m} f_{i}$")

Rel_Freq = c("$r_{1} = f_{1} / F$",
             "$r_{2} = f_{2} / F$",
             "$r_{3} = f_{3} / F$",
             "\\vdots",
             "$r_{m} = f_{m} / F$",
             "1")

hist_tab = data.frame(class_int, Frequency, Rel_Freq)

colnames(hist_tab) = c("Class Interval", "Frequency", "Relative Freq.")

# hist_tab %>%
#   knitr::kable(full_width = F, caption = "How a histogram is constructed")

kbl(hist_tab, caption = "How a histogram is constructed") %>%
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```


The histogram is then created by plotting the intervals on the $x$-axis and the frequency or relative frequency of observations in each interval on the $y$-axis.
When frequency is plotted on the $y$-axis, the histogram is referred to as a frequency histogram and
when relative frequency is plotted on the $y$-axis, the histogram is referred to as a relative frequency histogram.

::: {.remark}
The bin size in table \@ref(tab:histconstruct) is often chosen to be equal size, however it does not necessarily have to be the case and we can have bin sizes of unequal widths.
:::

::: {.remark}
When we use the term *histogram*, we often refer to a frequency histogram unless otherwise specified.
:::

<!-- ::: {#note:histnote .note name="Note"} -->
<!-- **Note:** When we use the term *histogram*, we often refer to a frequency histogram unless otherwise specified. -->
<!-- ::: -->

Some examples of histograms which also illustrate the concept of skewness are shown in Figures \@ref(fig:HistExampleLeft), \@ref(fig:HistExampleRight), and \@ref(fig:HistExampleSymm). The histograms show how the distribution of data can be skewed to the left, skewed to the right, or approximately symmetric.

```{r HistExampleLeft, fig.cap="An illustration of a histogram to have a left (or negative) skew probability distribution.", echo=FALSE, fig.align='center', warning=FALSE}
library(plotly)

set.seed(5)
x <- rbeta(6000, shape1 = 5, shape2 = 1.5) * 100

binwidth <- 5

# 1) Scaled density
dens     <- density(x, adjust = 1.2)
y_scaled <- dens$y * length(x) * binwidth

# 2) Compute max y (histogram + density)
hist_data <- hist(x,
                  breaks = seq(0, 120, by = binwidth),
                  plot   = FALSE)
max_y     <- max(max(hist_data$counts), max(y_scaled))

plot_ly() %>%
  # histogram (always on)
  add_histogram(
    x          = x,
    xbins      = list(size = binwidth),
    marker     = list(color = "#c6adff", line = list(color = "white", width = 1)),
    showlegend = FALSE
  ) %>%
  # density (start hidden)
  add_lines(
    x          = dens$x,
    y          = y_scaled,
    line       = list(color = "#0064c8", width = 3),
    visible    = FALSE,
    showlegend = FALSE
  ) %>%
  layout(
    title = list(text = "Histogram with Left‑Skewed Data"),
    xaxis = list(title = "Value", range = c(0, 110)),
    yaxis = list(title = "Frequency", range = c(0, 900), fixedrange = TRUE),
    bargap = 0.02,
    updatemenus = list(
      list(
        type      = "buttons",
        direction = "right",
        xanchor   = "left",
        x         = 0.10,
        y         = 0.70,
        buttons = list(
          list(
            method  = "restyle",
            args    = list("visible", list(TRUE, FALSE)),
            args2   = list("visible", list(TRUE, TRUE)),
            label   = "Show/Hide Density",
            execute = TRUE      # ← ensures first click toggles on
          )
        )
      )
    )
  )
```

```{r HistExampleRight, fig.cap="An illustration of a histogram to have a right (or positive) skew probability distribution.", echo=FALSE, fig.align='center', warning=FALSE}
library(plotly)

set.seed(5)
x        <- rbeta(6000, shape1 = 1.5, shape2 = 5) * 100
binwidth <- 5

# 1) Compute & scale density
dens     <- density(x, adjust = 1.2)
y_scaled <- dens$y * length(x) * binwidth

# 2) Find max y
hist_data <- hist(x,
                  breaks = seq(0, 120, by = binwidth),
                  plot   = FALSE)
max_y     <- max(max(hist_data$counts), max(y_scaled))

fig <- plot_ly() %>%
  # histogram with equal-width bins from 0 to 120
  add_histogram(
    x          = x,
    xbins      = list(start = 0, end = 90, size = binwidth),
    marker     = list(color = "#c6adff", line = list(color = "white", width = 1)),
    showlegend = FALSE
  ) %>%
  # density (starts hidden)
  add_lines(
    x          = dens$x,
    y          = y_scaled,
    line       = list(color = "#0064c8", width = 3),
    visible    = FALSE,
    showlegend = FALSE
  ) %>%
  layout(
    title   = list(text = "Histogram with Right‑Skewed Data"),
    xaxis   = list(title = "Value", range = c(0, 90)),
    yaxis   = list(title = "Frequency", range = c(0, 900), fixedrange = TRUE),
    bargap  = 0.02,
    updatemenus = list(
      list(
        type      = "buttons",
        direction = "right",
        xanchor   = "left",
        x         = 0.80,
        y         = 0.70,
        buttons = list(
          list(
            method  = "restyle",
            args    = list("visible", list(TRUE, FALSE)),
            args2   = list("visible", list(TRUE, TRUE)),
            label   = "Show/Hide Density",
            execute = TRUE      # ← ensures first click toggles on
          )
        )
      )
    )
  )

fig
```


```{r HistExampleSymm, fig.cap="An illustration of a histogram to have a symmetric probability distribution.", echo=FALSE, fig.align='center', warning=FALSE}
library(plotly)

set.seed(12)

# Generate approximately symmetric data
x        <- rnorm(6000, mean = 60, sd = 15)
binwidth <- 5

# 1) Compute & scale density
dens     <- density(x, adjust = 1.25)
y_scaled <- dens$y * length(x) * binwidth

# 2) Find the maximum y‑value (histogram + density) for a locked y‑axis
hist_data <- hist(x,
                  breaks = seq(0, 120, by = binwidth),
                  plot   = FALSE)
max_y     <- max(max(hist_data$counts), max(y_scaled))

# 3) Build the Plotly figure
fig <- plot_ly() %>%
  # A: histogram trace (always visible)
  add_histogram(
    x          = x,
    xbins      = list(start = 0, end = 120, size = binwidth),
    marker     = list(color = "#c6adff", line = list(color = "white", width = 1)),
    showlegend = FALSE
  ) %>%
  # B: density trace (starts hidden)
  add_lines(
    x          = dens$x,
    y          = y_scaled,
    line       = list(color = "#0064c8", width = 3),
    visible    = FALSE,
    showlegend = FALSE
  ) %>%
  layout(
    title   = list(text = "Histogram with Approximately Symmetric Data"),
    xaxis   = list(title = "Value", range = c(0, 120)),
    yaxis   = list(title = "Frequency", range = c(0, max_y), fixedrange = TRUE),
    bargap  = 0.02,
    updatemenus = list(
      list(
        type      = "buttons",
        direction = "right",
        xanchor   = "left",
        x         = 0.80,
        y         = 0.70,
        buttons = list(
          list(
            method  = "restyle",
            args    = list("visible", list(TRUE, FALSE)),
            args2   = list("visible", list(TRUE, TRUE)),
            label   = "Show/Hide Density",
            execute = TRUE      # ← ensures first click toggles on
          )
        )
      )
    )
  )

fig
```


The bin widths selected can change the appearance of histograms
The interactive application below allows us to examine the effect of bin size on a histogram from the [faithful](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/faithful.html) data set which measures the waiting time between eruptions of the Old Faithful geyser in Yellowstone National Park, Wyoming, USA.

<center>
```{r HistShinyApp, echo=FALSE, out.width='80%', warning=FALSE}
knitr::include_app("https://nishan-mudalige.shinyapps.io/Histogram-Shiny-App/", height = "810")
```
</center>


### Boxplots

Boxplots are another visual aid we can use to present and interpret data. They are one
of the most simple graphical techniques to analyze visually and are therefore usually relatively immediate to interpret. 
Boxplots are also known as *box-and-whisker* plots since they consists of figures resembling boxes along with a series of
lines which we refer to as whiskers. The whiskers represent the quartiles as well as special
values which we refer to as the *lower whisker* and the *upper whisker*.

::: {.definition name="Upper and Lower Whiskers" #UpperLowerWhiskers}
Let $Q_{1}$, $Q_{3}$ and $IQR$ represent the first quartile, third quartile and 
interquartile range respectively as defined in \@ref(def:Quartiles) and \@ref(def:IQR) respectively.
Then,
\[ \text{Lower Whisker} ~ = ~ Q_{1} - 1.5 \times IQR \]
\[ \text{Upper Whisker} ~ = ~ Q_{3} + 1.5 \times IQR \]
:::

The box in a boxplot is divided into two parts, the lower and upper quartiles. The lower quartile is represented by the bottom of the box and the upper quartile is represented by the top of the box. The line inside the box represents the median of the data set.
Box plots can be used to to get a sense of a data set.

Consider the following summary data

```{r BoxplotData, echo=FALSE, results='asis', warning = FALSE}
# Load required libraries
library(knitr)
library(kableExtra)

# Create summary as a named numeric vector
summary_stats <- c(
  Min = 0.500,
  `1st Qu.` = 8.782,
  Median = 9.855,
  Mean = 10.036,
  `3rd Qu.` = 11.521,
  Max = 19.000
)

# Convert to one-row wide format data frame
summary_wide <- as.data.frame(t(summary_stats))

# Create formatted table (HTML for Bookdown)
kable(summary_wide, format = "html", booktabs = TRUE, caption = "Summary Statistics") %>%
  kable_styling(full_width = FALSE, position = "center")
```

The summary statistics in Table \@ref(tab:BoxplotData) is represented visually as an interactive boxplot in Figure \@ref(fig:BoxplotExample). 
Hover over the boxplot to identify the median, quartiles and whiskers.

```{r BoxplotExample, fig.cap="An illustration of an annotated box-plot with whiskers and potential outliers.", echo=FALSE, fig.align='center', warning=FALSE}
# Load Plotly
library(plotly)

# Base data
set.seed(123)
x <- rnorm(50, mean = 10, sd = 2)

# Define outliers
outliers <- c(0.5, 1, 18, 19)

x = c(x, outliers)

outlier_labels <- rep("potential outlier", length(outliers))
x_position <- rep(1, length(outliers))  # Align with boxplot

# Create plot
p <- plot_ly() %>%
  # Boxplot
  add_trace(
    x = rep(1, length(x)),
    y = ~x,
    type = "box",
    boxpoints = "none",
    name = "",
    hoverinfo = "y",
    fillcolor = "#00C19A",
    marker = list(color = "#00C19A"),
    line = list(color = 'rgba(0,100,200,1)')
  ) %>%
  # Custom outliers
  add_trace(
    x = ~x_position,
    y = ~outliers,
    type = "scatter",
    mode = "markers",
    marker = list(
      color = "#F8766D",
      size = 8,
      line = list(color = "#FF68A1", width = 1)
    ),
    hoverinfo = "text",
    hovertext = outlier_labels,
    showlegend = FALSE
  ) %>%
  layout(
    title = "Interactive Boxplot<br>(Hover over outliers)",
    yaxis = list(
      title = "Values",
      range = c(0, 21)        # Set y-axis limits
    ),
    xaxis = list(
      showticklabels = FALSE,
      title = NA,
      range = c(0.5, 1.5)
    )
  )

# Disable lasso and box select tools
config(
  p,
  modeBarButtonsToRemove = c("select2d", "lasso2d")
)
```

Boxplots can also efficiently be used to compare multiple distributions and highlight skewness, spread, and outliers from each of them.
If the median cuts the box with upper area smaller than lower area,
then we say that box-plot with left skew probability distribution. Or,
if the median cuts the box with upper area larger than lower area,
then we say that box-plot with right skew probability distribution.
Otherwise, if the median cuts the box with upper area equal to
lower area, then we say that box-plot with symmetric probability
distribution.

```{r BoxplotNegativeExample, fig.cap="An illustration of a boxplot with a left (negative) skew.", echo=FALSE, fig.align='center', warning=FALSE}
# Load Plotly
library(plotly)

# Generate left-skewed data by flipping right-skewed log-normal
set.seed(123)
raw_right <- c(rlnorm(35, meanlog = 2, sdlog = 0.75), runif(5, 14, 18) )
raw <- max(raw_right) - raw_right  # flip for left skew

# Remove statistical outliers
q1 <- quantile(raw, 0.25)
q3 <- quantile(raw, 0.75)
iqr <- q3 - q1
upper_whisker <- q3 + 1.5 * iqr
lower_whisker <- q1 - 1.5 * iqr
x <- raw[raw >= lower_whisker & raw <= upper_whisker]
x <- x + 2

mean_x <- mean(x)
median_x <- median(x)

# Plot
p <- plot_ly() %>%
  add_trace(
    x = rep(1, length(x)),
    y = ~x,
    type = "box",
    boxpoints = "none",  # hide all points
    name = "",
    hoverinfo = "skip",
    fillcolor = "#00C19A",
    marker = list(color = "#00C19A"),
    line = list(color = 'rgba(0,100,200,1)')
  ) %>%
  layout(
    title = "Boxplot suggesting left skewed data<br>(Mean < Median)",
    yaxis = list(
      title = "Values",
      range = c(0, max(x) * 1.1)
    ),
    xaxis = list(
      showticklabels = FALSE,
      title = NA,
      range = c(0.5, 1.5)
    ),
    shapes = list(
      list(
        type = "line",
        xref = "x",
        yref = "y",
        x0 = 0.757,
        x1 = 1.247,
        y0 = mean_x - 0.5, 
        y1 = mean_x - 0.5 ,
        line = list(
          dash = "dash",
          color = "#FF68A1",
          width = 2
        )
      )
    ),
    annotations = list(
      list(
        x = 1.325,
        y = mean_x - 0.50,
        text = "mean",
        xref = "x",
        yref = "y",
        showarrow = FALSE,
        font = list(color = "#FF68A1", size = 18)
      ),
      list(
        x = 1.325,
        y = median_x + 0.25,
        text = "median",
        xref = "x",
        yref = "y",
        showarrow = FALSE,
        font = list(color = "rgba(0,100,200,1)", size = 18)
      )
    )
  )

# Remove all interactivity
config(
  p,
  displayModeBar = FALSE,
  staticPlot = TRUE
)

```

```{r BoxplotpositiveExample, fig.cap="An illustration of a boxplot with a right (positive) skew.", echo=FALSE, fig.align='center', warning=FALSE}
# Load Plotly
library(plotly)

# Generate right-skewed data WITHOUT outliers (control max values)
set.seed(123)
raw <- rlnorm(35, meanlog = 2, sdlog = 0.5)
q1 <- quantile(raw, 0.25)
q3 <- quantile(raw, 0.75)
iqr <- q3 - q1
upper_whisker <- q3 + 1.5 * iqr
lower_whisker <- q1 - 1.5 * iqr
x <- raw[raw >= lower_whisker & raw <= upper_whisker]  # trim outliers

mean_x <- mean(x)
median_x <- median(x)

# Plot
p <- plot_ly() %>%
  add_trace(
    x = rep(1, length(x)),
    y = ~x,
    type = "box",
    boxpoints = "none",  # no visible points
    name = "",
    hoverinfo = "skip",
    fillcolor = "#00C19A",
    marker = list(color = "#00C19A"),
    line = list(color = 'rgba(0,100,200,1)')
  ) %>%
  layout(
    title = "Boxplot suggesting right skewed data<br>(Mean > Median)",
    yaxis = list(
      title = "Values",
      range = c(0, max(x) * 1.1)
    ),
    xaxis = list(
      showticklabels = FALSE,
      title = NA,
      range = c(0.5, 1.5)
    ),
    shapes = list(
      list(
        type = "line",
        xref = "x",
        yref = "y",
        x0 = 0.757,
        x1 = 1.247,
        y0 = mean_x + 1, 
        y1 = mean_x + 1,
        line = list(
          dash = "dash",
          color = "#FF68A1",
          width = 2
        )
      )
    ),
    annotations = list(
      list(
        x = 1.325,
        y = mean_x + 0.8 + 0.25,
        text = "mean",
        xref = "x",
        yref = "y",
        showarrow = FALSE,
        font = list(color = "#FF68A1", size = 18)
      ),
      list(
        x = 1.325,
        y = median_x - 0.25,
        text = "median",
        xref = "x",
        yref = "y",
        showarrow = FALSE,
        font = list(color = "rgba(0,100,200,1)", size = 18)
      )
    )
  )

# Remove all interactivity
config(
  p,
  displayModeBar = FALSE,
  staticPlot = TRUE
)
```

```{r BoxplotSymmetryExample, fig.cap="An illustration of a boxplot with an approximately symmetric distribution.", echo=FALSE, fig.align='center', warning=FALSE}
# Load Plotly
library(plotly)

# Base data
set.seed(123)
x <- rnorm(50, mean = 10, sd = 4)
mean_x <- mean(x)
median_x <- median(x)

# Create plot
p <- plot_ly() %>%
  # Boxplot
  add_trace(
    x = rep(1, length(x)),
    y = ~x,
    type = "box",
    boxpoints = "none",
    name = "",
    hoverinfo = "skip",  # <-- disables hover labels
    fillcolor = "#00C19A",
    marker = list(color = "#00C19A"),
    line = list(color = 'rgba(0,100,200,1)')
  ) %>%
  layout(
    title = "Boxplot suggesting symmetric data<br>(Mean ≈ Median)",
    yaxis = list(
      title = "Values",
      range = c(0, 21)
    ),
    xaxis = list(
      showticklabels = FALSE,
      title = NA,
      range = c(0.5, 1.5)
    ),
    shapes = list(
      list(
        type = "line",
        xref = "x",
        yref = "y",
        x0 = 0.757, 
        x1 = 1.247,
        y0 = mean_x, y1 = mean_x,
        line = list(
          dash = "dash",
          color = "#FF68A1",
          width = 2
        )
      )
    ),
    annotations = list(
      list(
        x = 1.325,
        y = mean_x + 0.25,
        text = "mean",
        xref = "x",
        yref = "y",
        showarrow = FALSE,
        font = list(color = "#FF68A1", size = 18)
      ),
      list(
        x = 1.325,
        y = median_x - 0.25,
        text = "median",
        xref = "x",
        yref = "y",
        showarrow = FALSE,
        font = list(color = "rgba(0,100,200,1)", size = 18)
      )
    )
  )

# Remove all interactive tools and mode bar
config(
  p,
  displayModeBar = FALSE,
  staticPlot = TRUE
)
```

## Exercises {#sec:ch3exercises}

```{r ch3-ex-setup, include=FALSE}
library(ggplot2)
```

---

<div class="exercise-box">
<div class="exercise-label">Question 1</div>
A fitness researcher records the following information for five participants in a study:

| Participant | Age | Sex | Resting Heart Rate (bpm) | Exercise Habit |
|-------------|-----|-----|--------------------------|----------------|
| 1 | 24 | F | 68 | Daily |
| 2 | 31 | M | 72 | Weekly |
| 3 | 19 | F | 61 | Rarely |
| 4 | 27 | M | 75 | Daily |
| 5 | 22 | F | 65 | Weekly |

(a) (1 mark) What is the **unit** of observation in this dataset?
(b) (1 mark) How many **observations** are there? How many **variables**?
(c) (2 marks) Classify each variable as **quantitative** or **qualitative**.
(d) (1 mark) What is the **value** of "Resting Heart Rate" for Participant 4?
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Each **row** is one observation from one unit.
- **Quantitative** variables have numerical values that allow arithmetic; **qualitative** variables place units into categories.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** One **participant** (person).

**(b)** **5 observations**, **4 variables** (Age, Sex, Resting Heart Rate, Exercise Habit).

**(c)**

| Variable | Type |
|----------|------|
| Age | Quantitative |
| Sex | Qualitative |
| Resting Heart Rate | Quantitative |
| Exercise Habit | Qualitative |

**(d)** 75 bpm.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 2</div>
The following data represent the number of hours 12 UTM students spent on their phones yesterday:

$$3,\ 5,\ 2,\ 7,\ 5,\ 4,\ 6,\ 5,\ 3,\ 8,\ 4,\ 5$$

(a) (3 marks) Construct a **frequency table** with columns: Value, Frequency, and Relative Frequency (as a decimal rounded to 2 places).
(b) (1 mark) What proportion of students spent **more than 5 hours** on their phones?
(c) (1 mark) What value appears most frequently? What is this measure called?
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- **Relative Frequency** = (Frequency) ÷ (Total number of observations).
- Count each distinct value carefully — sort the data first to avoid missing any.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)**

| Hours | Frequency | Relative Frequency |
|-------|-----------|-------------------|
| 2 | 1 | 0.08 |
| 3 | 2 | 0.17 |
| 4 | 2 | 0.17 |
| 5 | 4 | 0.33 |
| 6 | 1 | 0.08 |
| 7 | 1 | 0.08 |
| 8 | 1 | 0.08 |
| **Total** | **12** | **1.00** |

**(b)** Values > 5: hours of 6, 7, 8 → frequency = 3. Proportion = $3/12 = 0.25$ (25%).

**(c)** The value **5** appears 4 times — the most frequent. This is the **mode**.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 3</div>
A survey asked 40 students how many hours of sleep they got last night. The results are summarized below:

| Hours of Sleep | Frequency |
|---------------|-----------|
| 5 | 3 |
| 6 | 8 |
| 7 | 12 |
| 8 | 10 |
| 9 | 7 |

(a) (2 marks) Add a **Relative Frequency** column and a **Cumulative Frequency** column to the table.
(b) (1 mark) What proportion of students slept **at most 7 hours**?
(c) (2 marks) What proportion of students slept **between 6 and 8 hours inclusive**?
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- **Cumulative Frequency** for a row = sum of all frequencies *up to and including* that row.
- For "at most 7 hours", sum the frequencies for 5, 6, and 7.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)**

| Hours | Frequency | Relative Frequency | Cumulative Frequency |
|-------|-----------|-------------------|---------------------|
| 5 | 3 | 0.075 | 3 |
| 6 | 8 | 0.200 | 11 |
| 7 | 12 | 0.300 | 23 |
| 8 | 10 | 0.250 | 33 |
| 9 | 7 | 0.175 | 40 |

**(b)** Cumulative frequency at 7 = 23. Proportion = $23/40 = 0.575$ (57.5%).

**(c)** Frequencies for 6, 7, 8: $8 + 12 + 10 = 30$. Proportion = $30/40 = 0.75$ (75%).
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 4</div>
The following data show the number of siblings for 20 students in a class:

$$0, 1, 2, 1, 3, 2, 0, 1, 4, 2, 1, 0, 2, 3, 1, 2, 0, 1, 3, 2$$

(a) (2 marks) Create a **frequency table** for the number of siblings.
(b) (2 marks) Create a **relative frequency table**.
(c) (1 mark) What **proportion** of students have exactly 2 siblings?
(d) (1 mark) What **percentage** of students have at least 3 siblings?
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
Count how many times each value appears. Relative frequency is frequency divided by the total number of observations.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a) and (b)**

| Number of siblings | Frequency | Relative Frequency |
|--------------------|-----------|--------------------|
| 0 | 4 | 4/20 = 0.20 |
| 1 | 6 | 6/20 = 0.30 |
| 2 | 6 | 6/20 = 0.30 |
| 3 | 3 | 3/20 = 0.15 |
| 4 | 1 | 1/20 = 0.05 |

**(c)** The proportion with exactly 2 siblings is $6/20 = 0.30$.

**(d)** At least 3 siblings: $(3 + 1)/20 = 4/20 = 0.20 = 20\%$.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 5</div>
The table below shows the ages of participants in a study.

| Age Group | Frequency |
|-----------|-----------|
| 18–22 | 14 |
| 23–27 | 20 |
| 28–32 | 11 |
| 33–37 | 8 |
| 38–42 | 7 |

(a) (1 mark) How many **participants** are in the study?
(b) (1 mark) Which **age group** has the highest frequency?
(c) (2 marks) What **proportion** of participants are between 23 and 32 years old?
(d) (1 mark) What **percentage** of participants are at least 33 years old?
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
Add the frequencies to find the total number of participants. To find a proportion, divide the relevant frequency total by the overall total.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** Total $= 14 + 20 + 11 + 8 + 7 = \mathbf{60}$ participants.

**(b)** The **23–27** age group has the highest frequency (20 participants).

**(c)** Between 23 and 32 years old: $(20 + 11)/60 = 31/60 \approx \mathbf{0.517}$.

**(d)** At least 33 years old: $(8 + 7)/60 = 15/60 = 0.25 = \mathbf{25\%}$.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 6</div>
The following data show the number of hours that 10 students spent studying for a quiz:

$$2,\ 3,\ 3,\ 4,\ 4,\ 5,\ 5,\ 5,\ 6,\ 8$$

(a) (1 mark) Find the **mean**.
(b) (2 marks) Find the **median**.
(c) (1 mark) Find the **mode**.
(d) (1 mark) Find the **range**.
(e) (2 marks) Is the mean larger than the median? Briefly explain why this may happen.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
The mean uses all observations, while the median is the middle value after ordering the data.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** $\bar{x} = (2+3+3+4+4+5+5+5+6+8)/10 = 45/10 = \mathbf{4.5}$.

**(b)** With $n = 10$ (even), the median is the average of the 5th and 6th ordered values: $(4 + 5)/2 = \mathbf{4.5}$.

**(c)** Mode $= \mathbf{5}$ (appears 3 times).

**(d)** Range $= 8 - 2 = \mathbf{6}$.

**(e)** The mean and median are equal here. This suggests the data are fairly balanced around the centre, although the data are not perfectly symmetric.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 7</div>
Two data sets are shown below.

**Data Set A:**
$$6,\ 7,\ 7,\ 8,\ 8,\ 9,\ 9$$

**Data Set B:**
$$6,\ 7,\ 7,\ 8,\ 8,\ 9,\ 30$$

(a) (2 marks) Find the **mean** of each data set.
(b) (2 marks) Find the **median** of each data set.
(c) (1 mark) Which measure, mean or median, is more **affected** by the value 30?
(d) (1 mark) Which measure better **represents the centre** of Data Set B? Explain.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
Compare how much the mean and median change when 9 is replaced by 30.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** Data Set A mean $= 54/7 \approx \mathbf{7.71}$. Data Set B mean $= 75/7 \approx \mathbf{10.71}$.

**(b)** Data Set A median $= \mathbf{8}$ (4th value). Data Set B median $= \mathbf{8}$ (4th value).

**(c)** The **mean** is more affected by 30 — it increased from approximately 7.71 to 10.71, while the median did not change.

**(d)** The **median** better represents the centre of Data Set B because 30 is an unusually large value that pulls the mean upward.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 8</div>
The following data represent the number of hours 6 students spent studying for a midterm:

$$x_1 = 2,\quad x_2 = 4,\quad x_3 = 4,\quad x_4 = 6,\quad x_5 = 6,\quad x_6 = 8$$

(a) (1 mark) Calculate the **mean** $\bar{x}$.
(b) (3 marks) Calculate the **sample variance** $s^2$ using the formula
$$s^2 = \frac{\displaystyle\sum_{i=1}^{n}(x_i - \bar{x})^2}{n-1}$$
Show your work using a table with columns $x_i$, $(x_i - \bar{x})$, and $(x_i - \bar{x})^2$.
(c) (1 mark) Calculate the **sample standard deviation** $s$.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Compute $\bar{x}$ first, then find each deviation $(x_i - \bar{x})$, square it, sum all squared deviations, and divide by $n - 1$.
- $s = \sqrt{s^2}$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** $\bar{x} = (2+4+4+6+6+8)/6 = 30/6 = \mathbf{5}$

**(b)**

| $x_i$ | $x_i - \bar{x}$ | $(x_i - \bar{x})^2$ |
|--------|-----------------|----------------------|
| 2 | $-3$ | 9 |
| 4 | $-1$ | 1 |
| 4 | $-1$ | 1 |
| 6 | $1$ | 1 |
| 6 | $1$ | 1 |
| 8 | $3$ | 9 |
| | **Sum** | **22** |

$$s^2 = \frac{22}{6-1} = \frac{22}{5} = \mathbf{4.4}$$

**(c)** $s = \sqrt{4.4} \approx \mathbf{2.098}$ hours
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 9</div>
The boxplot below shows the age distribution of 12 participants in a research study. **Each dot is one participant's age. The bottom and top of the box are $Q_1$ and $Q_3$; the line inside the box is the median; whiskers extend to the min and max within the IQR fences; the red dashed lines are the IQR outlier fences.**

```{r ch3-q9-box, echo=FALSE, fig.height=3.8, fig.width=4.5, fig.align='center', message=FALSE, warning=FALSE}
ages9 <- c(22, 25, 19, 31, 28, 35, 24, 22, 30, 27, 40, 23)
df_q9 <- data.frame(ages = ages9)
q1_9 <- quantile(ages9, 0.25); q3_9 <- quantile(ages9, 0.75)
iqr_9 <- q3_9 - q1_9
ggplot(df_q9, aes(x = "", y = ages)) +
  geom_boxplot(fill = "#a29bfe", color = "#6c5ce7", width = 0.35, outlier.shape = NA) +
  geom_jitter(width = 0.06, size = 2.5, alpha = 0.7, color = "grey25") +
  geom_hline(aes(yintercept = q1_9 - 1.5*iqr_9, linetype = "IQR fence"),
             color = "#d63031", linewidth = 0.8) +
  geom_hline(aes(yintercept = q3_9 + 1.5*iqr_9, linetype = "IQR fence"),
             color = "#d63031", linewidth = 0.8) +
  scale_linetype_manual(name = NULL, values = c("IQR fence" = "dashed")) +
  scale_y_continuous(breaks = seq(10, 45, by = 5)) +
  labs(x = NULL, y = "Age (years)",
       caption = "Each dot = 1 participant  |  Box = Q1 to Q3  |  Red dashes = IQR fences") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        plot.caption = element_text(color = "#555", size = 9))
```

(a) (2 marks) Read the **five-number summary** (Min, $Q_1$, Median, $Q_3$, Max) from the plot. All reasonable estimated answers will be marked correct.
(b) (1 mark) Compute the **IQR** using your answer to (a).
(c) (2 marks) Do any dots fall outside the red dashed **IQR fences**? State your conclusion about outliers and justify using the IQR rule ($Q_1 - 1.5\times\text{IQR}$, $Q_3 + 1.5\times\text{IQR}$).
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Bottom whisker tip = Min; top whisker tip = Max (within fences).
- Bottom edge of box = $Q_1$; top edge = $Q_3$; line inside = Median.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** Exact values from data: Min = 19, $Q_1$ = 22.5, Median = 26, $Q_3$ = 30.5, Max = 40.

**(b)** IQR $= 30.5 - 22.5 = \mathbf{8}$

**(c)** Fences: lower $= 22.5 - 1.5(8) = 10.5$; upper $= 30.5 + 1.5(8) = 42.5$. All values lie within $[10.5, 42.5]$, so there are **no outliers** by the IQR rule — even the oldest participant (age 40) is inside the upper fence.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 10</div>
Two groups of students were given the same statistics quiz (out of 20). Their scores are:

- **Group A:** 14, 15, 14, 16, 15, 14, 16, 15, 15, 16
- **Group B:** 8, 12, 15, 19, 18, 11, 17, 20, 10, 10

(a) (1 mark) Without calculating, which group do you expect to have a **larger standard deviation**? Justify using the spread of the values.
(b) (2 marks) Verify by calculating $\bar{x}$ and $s$ for each group. Use R or show your working.
(c) (2 marks) Both groups have the same mean. What does this tell you about the **usefulness of the mean alone** as a summary of a dataset?
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
Standard deviation measures how far values are from the mean *on average*. Look at whether the scores are tightly packed or spread widely.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** **Group B** — its scores range from 8 to 20 (wide spread), while Group A's scores cluster tightly between 14 and 16.

**(b)**

- Group A: $\bar{x} = 150/10 = 15$, $s \approx 0.816$
- Group B: $\bar{x} = 140/10 = 14$, $s \approx 4.32$

Group B's standard deviation is much larger, confirming the visual observation.

**(c)** The two groups actually have **different** means (Group A: $\bar{x} = 15$, Group B: $\bar{x} = 14$) **and** very different standard deviations ($s_A \approx 0.816$, $s_B \approx 4.32$). This illustrates that the mean alone is **insufficient** to fully describe a dataset — the standard deviation is also needed to capture how spread out the values are.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 11</div>
The following data show the number of minutes that 12 students spent waiting for a bus:

$$3,\ 4,\ 5,\ 5,\ 6,\ 7,\ 8,\ 9,\ 10,\ 12,\ 13,\ 30$$

(a) (1 mark) Find the **median**.
(b) (2 marks) Find $Q_1$ and $Q_3$.
(c) (1 mark) Calculate the **IQR**.
(d) (2 marks) Use the $1.5 \times \text{IQR}$ rule to determine whether there are any **outliers**.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- First order the data. Then use $\text{IQR} = Q_3 - Q_1$.
- The lower fence is $Q_1 - 1.5 \times \text{IQR}$, and the upper fence is $Q_3 + 1.5 \times \text{IQR}$.
- Any observation outside these fences is flagged as an outlier.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**Sorted data:** 3, 4, 5, 5, 6, 7, 8, 9, 10, 12, 13, 30

**(a)** $n = 12$ (even), so median = average of 6th and 7th values $= (7 + 8)/2 = \mathbf{7.5}$ minutes.

**(b)** Lower half: 3, 4, 5, 5, 6, 7 → $Q_1 = (5 + 5)/2 = \mathbf{5}$. Upper half: 8, 9, 10, 12, 13, 30 → $Q_3 = (10 + 12)/2 = \mathbf{11}$.

**(c)** $\text{IQR} = Q_3 - Q_1 = 11 - 5 = \mathbf{6}$ minutes.

**(d)** Lower fence $= 5 - 1.5(6) = -4$; upper fence $= 11 + 1.5(6) = 20$. Since $30 > 20$, the value **30 is a high outlier**.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 12</div>
The histogram below shows the distribution of daily commute times (in minutes) for a random sample of 60 UTM students. **Each bar covers a 10-minute interval; bar height = frequency.**

```{r ch3-q12-hist, echo=FALSE, fig.height=3.5, fig.width=6, fig.align='center', message=FALSE, warning=FALSE}
set.seed(11)
commute <- round(rlnorm(60, log(28), 0.45))
commute <- pmin(commute, 80)
df_q11 <- data.frame(commute = commute)
ggplot(df_q11, aes(x = commute)) +
  geom_histogram(binwidth = 10, fill = "#00b894", color = "white", boundary = 0) +
  scale_x_continuous(breaks = seq(0, 80, 10)) +
  labs(x = "Commute time (minutes)", y = "Frequency",
       caption = "n = 60 students  |  Bar height = frequency") +
  theme_minimal(base_size = 12) +
  theme(plot.caption = element_text(color = "#555", size = 9))
```

(a) (2 marks) Describe the **shape** of this distribution (skewness or symmetry). Justify using features of the histogram.
(b) (2 marks) Based on the histogram, **estimate the proportion** of students with a commute time of **30 minutes or less**. All reasonable estimated answers will be marked correct.
(c) (1 mark) Based on the shape of the distribution, would you expect the **mean** to be greater than, less than, or approximately equal to the **median**? Explain.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- **Right-skewed**: long tail to the right; the mean is pulled toward the tail, so mean > median.
- For (b), add up the frequencies (bar heights) for bars covering 0–30 minutes, then divide by 60.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** The distribution is **right-skewed** — most commute times are short (10–30 minutes) and the histogram has a long tail extending to the right toward 70–80 minutes.

**(b)** Sum the frequencies for bars at 0–10, 10–20, and 20–30. Reading from the histogram, approximately 40–45 students have commute times ≤ 30 minutes. Proportion ≈ $42/60 \approx 0.70$ (70%). All reasonable estimates will be marked correct.

**(c)** For a **right-skewed** distribution, the **mean > median** — the long right tail pulls the mean toward larger values while the median (middle value) is less affected.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 13</div>
The boxplot below shows the distribution of exam scores (out of 100) for 30 students in a statistics course. **The box spans $Q_1$ to $Q_3$; the line inside the box is the median; whiskers extend to min and max within the IQR fences; any dots beyond the whiskers are outliers.**

```{r ch3-q13-box, echo=FALSE, fig.height=3.5, fig.width=4, fig.align='center', message=FALSE, warning=FALSE}
set.seed(12)
scores12 <- c(round(rnorm(27, 72, 10)), 38, 42, 95)
scores12 <- pmax(pmin(scores12, 100), 0)
df_q12 <- data.frame(score = scores12)
ggplot(df_q12, aes(x = "", y = score)) +
  geom_boxplot(fill = "#74b9ff", color = "#0984e3", width = 0.35,
               outlier.shape = 19, outlier.size = 2.5, outlier.color = "grey30") +
  scale_y_continuous(breaks = seq(30, 100, 10)) +
  labs(x = NULL, y = "Exam score (out of 100)",
       caption = "Box = Q1 to Q3  |  Line = median  |  Dots = outliers") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        plot.caption = element_text(color = "#555", size = 9))
```

(a) (2 marks) **Estimate** the five-number summary from the plot. All reasonable estimated answers will be marked correct.
(b) (1 mark) **Estimate** the IQR.
(c) (2 marks) Are there any **outliers** visible in the plot? If so, estimate their values and state whether they are low or high outliers.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Dots that appear *beyond* the whiskers are flagged as outliers.
- Read the scale on the y-axis carefully to estimate each summary value.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** Approximate values: Min ≈ 38–42, $Q_1$ ≈ 64–68, Median ≈ 72–74, $Q_3$ ≈ 78–82, Max ≈ 95–98. All reasonable estimates will be marked correct.

**(b)** IQR ≈ $Q_3 - Q_1 \approx 80 - 66 = 14$.

**(c)** Yes — there are **3 outlier dots**: two low outliers (approximately 38 and 42) below the lower whisker, and one high outlier (approximately 95) above the upper whisker.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 14</div>
Two sections of STA258 each wrote the same midterm exam (out of 100). Side-by-side boxplots of their scores are shown below. **Each dot is one student's score; the box spans $Q_1$ to $Q_3$; the line inside is the median.**

```{r ch3-q14-box, echo=FALSE, fig.height=3.5, fig.width=5.5, fig.align='center', message=FALSE, warning=FALSE}
set.seed(13)
secA <- pmax(pmin(round(rnorm(35, 74, 6)), 100), 40)
secB_main <- round(rnorm(32, 74, 14))
secB <- c(pmax(pmin(secB_main, 100), 30), 22, 25, 98)
df_q13 <- data.frame(
  score   = c(secA, secB),
  section = factor(rep(c("Section A", "Section B"),
                       times = c(length(secA), length(secB))))
)
ggplot(df_q13, aes(x = section, y = score, fill = section)) +
  geom_boxplot(width = 0.4, alpha = 0.75, outlier.size = 2.5,
               outlier.shape = 19, outlier.colour = "grey30") +
  geom_jitter(width = 0.07, size = 1.8, alpha = 0.5, color = "grey30") +
  scale_fill_manual(values = c("Section A" = "#fd79a8", "Section B" = "#fdcb6e")) +
  scale_y_continuous(breaks = seq(20, 100, 10)) +
  labs(x = NULL, y = "Exam score (out of 100)",
       caption = "Each dot = 1 student  |  Box = Q1 to Q3  |  Line = median") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.caption = element_text(color = "#555", size = 9))
```

(a) (2 marks) Which section (A or B) appears to have **greater variability** in exam scores? Justify your answer using a feature of the boxplot.
(b) (2 marks) **Estimate the median** exam score for both Section A and Section B. All reasonable estimated answers will be marked correct.
(c) (2 marks) Do any of the boxplots show **outliers**? If outliers exist, state which section and estimate the value of the **most extreme outlier**. All reasonable estimated answers will be marked correct.
(d) (1 mark) Both sections have approximately the same median. What does this suggest about comparing two groups using the **median alone**?
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- **Variability** is shown by the height of the box (IQR) and the overall span of the whiskers.
- Outliers appear as **dots beyond the whiskers**.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** **Section B** — its box (IQR) is taller and the whiskers extend further, indicating a wider spread of exam scores compared to Section A.

**(b)** Both sections: median ≈ **73–75**. All reasonable estimates will be marked correct.

**(c)** Yes — **Section B** has outliers; the most extreme low outlier is approximately **22–25**.

**(d)** The **median alone is insufficient** — both sections have a similar median, but Section B has much greater variability. A complete comparison requires measures of both centre *and* spread.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 15</div>
A researcher recorded the delivery times (in minutes) for a random sample of 80 orders from two restaurants near UTM: **Restaurant A** and **Restaurant B**.

Side-by-side boxplots and histograms of the delivery times are shown below. **The box spans $Q_1$ to $Q_3$; the line inside is the median; whiskers extend to the most extreme non-outlier values; dots beyond the whiskers are outliers.**

```{r ch3-q15-box, echo=FALSE, fig.height=3.2, fig.width=5, fig.align='center', message=FALSE, warning=FALSE}
set.seed(258)
n14 <- 80
rA14 <- pmax(pmin(round(rnorm(n14, 35, 7)), 56), 14)
rB14 <- c(round(rlnorm(n14 - 2, log(25), 0.40)), 60, 73)
df14 <- data.frame(
  time       = c(rA14, rB14),
  restaurant = factor(rep(c("Restaurant A", "Restaurant B"), each = n14))
)
ggplot(df14, aes(x = restaurant, y = time, fill = restaurant)) +
  geom_boxplot(width = 0.4, alpha = 0.75, outlier.size = 2.5,
               outlier.shape = 19, outlier.colour = "grey30") +
  scale_fill_manual(values = c("Restaurant A" = "#5b9bd5", "Restaurant B" = "#ed7d31")) +
  scale_y_continuous(breaks = seq(0, 80, 10)) +
  labs(x = NULL, y = "Delivery time (minutes)",
       caption = "Box = Q1 to Q3  |  Line = median  |  Dots = outliers") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.caption = element_text(color = "#555", size = 9))
```

```{r ch3-q15-hist, echo=FALSE, fig.height=3.8, fig.width=5.5, fig.align='center', message=FALSE, warning=FALSE}
ggplot(df14, aes(x = time, fill = restaurant)) +
  geom_histogram(binwidth = 5, color = "white", alpha = 0.85) +
  facet_wrap(~ restaurant, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = c("Restaurant A" = "#5b9bd5", "Restaurant B" = "#ed7d31")) +
  scale_x_continuous(breaks = seq(0, 80, 10)) +
  labs(x = "Delivery time (minutes)", y = "Frequency") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 11))
```

**(a)** Use the **boxplots** to answer the following.

(i) (2 marks) Which restaurant (A or B) appears to have **greater variability** in delivery time? Justify your answer using a feature of the boxplot.

(ii) (2 marks) **Estimate the median** delivery time for both Restaurant A and Restaurant B. All reasonable estimated answers will be marked correct.

(iii) (2 marks) Do any of the boxplots show **outliers**? If outliers exist, state which restaurant has them and estimate the value of the **largest outlier**. All reasonable estimated answers will be marked correct.

**(b)** Use the **histograms** to answer the following.

(i) (2 marks) Describe the **shape** (skewness or symmetry) of the delivery time distribution for Restaurant A and for Restaurant B.

(ii) (3 marks) Estimate the **proportion** of deliveries from each restaurant that were completed in **40 minutes or less**. All reasonable estimated answers will be marked correct.

(iii) (2 marks) One of these restaurants is a sushi restaurant known for careful, consistent preparation; the other is a pizza restaurant that occasionally receives large rush orders. Based on the plots and your previous answers, which restaurant (A or B) is the **sushi restaurant**? Explain your reasoning.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- **Variability** in a boxplot is shown by the height of the box (IQR) and the whisker span.
- **Skewness**: a longer tail to the right → right-skewed; roughly bell-shaped → symmetric.
- For (b)(ii), count bars to the left of 40 and divide by the total count.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)(i)** **Restaurant B** — its box (IQR) is taller and whiskers span a wider range.

**(a)(ii)** Restaurant A median ≈ **35 min**; Restaurant B median ≈ **25 min**.

**(a)(iii)** Yes — **Restaurant B** has outlier dots above the upper whisker; the largest outlier is approximately **70–75 minutes**.

**(b)(i)** Restaurant A is roughly **symmetric** (approximately bell-shaped); Restaurant B is **right-skewed** (most deliveries are fast but a long right tail extends toward 70+ minutes).

**(b)(ii)** Restaurant A ≈ **75–80%** of deliveries ≤ 40 min; Restaurant B ≈ **85–90%** of deliveries ≤ 40 min.

**(b)(iii)** **Restaurant A** is the sushi restaurant — its symmetric distribution reflects **consistent preparation times**, characteristic of careful sushi preparation. Restaurant B's right-skewed distribution with outliers is consistent with pizza delivery, where most orders are fast but large rush orders occasionally cause extreme delays.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 16</div>
The histogram below shows the annual salaries (in thousands of dollars) of employees at a tech company.

```{r ch3-q16-hist, echo=FALSE, fig.height=3.5, fig.width=6, fig.align='center', message=FALSE, warning=FALSE}
set.seed(15)
salaries <- c(round(rlnorm(70, log(75), 0.35)), round(rlnorm(10, log(200), 0.25)))
salaries <- pmin(salaries, 400)
df_q15 <- data.frame(salary = salaries)
ggplot(df_q15, aes(x = salary)) +
  geom_histogram(binwidth = 25, fill = "#a29bfe", color = "white", boundary = 0) +
  scale_x_continuous(breaks = seq(0, 400, 50)) +
  labs(x = "Annual salary ($thousands)", y = "Frequency",
       caption = "n = 80 employees  |  Bar width = $25,000") +
  theme_minimal(base_size = 12) +
  theme(plot.caption = element_text(color = "#555", size = 9))
```

(a) (2 marks) Describe the **shape** of this salary distribution. Identify the direction of skew and justify using features of the histogram.
(b) (2 marks) Based on the shape, would you expect the **mean salary** to be greater than, less than, or approximately equal to the **median salary**? Explain.
(c) (1 mark) A reporter states: *"The average salary at this company is \$95,000, so most employees earn around that amount."* Is this statement likely to be **misleading**? Explain why.
(d) (2 marks) Which single measure — **mean** or **median** — would better represent a **typical** employee's salary in this company? Justify your answer.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- In a **right-skewed** distribution, a small number of very high values pull the **mean** above the **median**.
- The "average" (mean) can be misleading when data are skewed because it does not reflect where most data points fall.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** The distribution is **right-skewed** — most employees earn between \$50,000 and \$125,000 (the tall bars on the left), but a small number of high earners (possibly executives) create a long tail extending to the right toward \$300,000–\$400,000.

**(b)** We would expect the **mean > median** — the high-earning outliers in the long right tail pull the mean upward, while the median (the middle value) is not affected as strongly.

**(c)** Yes, this is likely **misleading**. If the distribution is right-skewed, most employees earn *less* than the mean. The mean is inflated by a small number of very high earners (executives), so it does not represent a "typical" employee's salary.

**(d)** The **median** better represents a typical employee's salary — it is robust to the influence of high-earning outliers and accurately reflects where the middle of the salary distribution falls.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 17</div>
For each situation below, choose the most appropriate graph: **bar chart**, **histogram**, or **boxplot**. Briefly explain your choice.

(a) (1 mark) A professor wants to display the number of students enrolled in each program: Statistics, Computer Science, Biology, and Psychology.
(b) (1 mark) A researcher wants to show the distribution of commute times for 200 students.
(c) (2 marks) A teaching assistant wants to compare the distribution of test scores between two tutorial sections.
(d) (1 mark) A student wants to display the frequency of different eye colours in a class.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Think about whether the variable is **categorical** or **quantitative**.
- When the goal is to **compare distributions between groups**, which graph type makes this easiest?
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** **Bar chart** — program is a categorical variable, and a bar chart displays the count or frequency for each category.

**(b)** **Histogram** — commute time is a quantitative variable; a histogram shows the shape, centre, and spread of its distribution.

**(c)** **Boxplot** — side-by-side boxplots are ideal for comparing the centre, spread, and outliers of a quantitative variable across two or more groups.

**(d)** **Bar chart** — eye colour is categorical with no natural ordering, so a bar chart is appropriate.
</div>
</details>

