# Descriptive Statistics

## Representing Data

Data comes to us in the form of \textit{observations} (i.e. measurements) which we write symbolically as a lower case letter along with a subscript. For example, suppose we have taken a total of $n$ observations. We have observation 1, observation 2, observation 3, $\ldots$, observation $n-1$ and observation $n$. By letting $x$ represent an observation, all $n$ observations can be represented as:

$$x_1,~x_2,~ \ldots ,~ x_n$$ 

where $x_i$ is an individual observation, and the index $i=1, \ldots , n$. Here $n$ is referred to as the \textit{sample size}. We can also represent this data in condensed form as:

$$x_i:\ i=1, \ldots , n$$

or in tabular form as:

::: {#smalltbl .smalltbl}
| Index       | 1     | 2     | $\ldots$  | $n$   |
| :---------- | :-:   | :-:   | :-:       | :-:   |
| Observation | $x_1$ | $x_2$ | $\ldots$  | $x_2$ |
:::



## Numerical Measures 

The most common numerical measures of central tendency are the mean, median, and mode. These measures summarize a set of data by identifying the central point within that data.

The \textit{mean} (or more precisely the arithmetic mean) of a data set is obtained by adding up all of the observations and dividing by the sample size (number of observations). The mean is the typical average that we are all familiar with. 
We use the symbol $\bar{x}$ to represent the sample mean.


### Mean

::: {.definition name="Mean" #Mean}
Let $x_{1}, ~x_{2}, ~x_{3}, ~\ldots , ~x_{n}$ represent a sample of $n$ observations.
The sample mean is defined as:
	$$\bar{x} = \frac{ \displaystyle\sum_{i = 1}^{n} x_{i} }{n} = \frac{ x_{1} + x_{2} + \dots + x _{n} }{n}$$
:::

### Median

::: {.definition name="Median" #Median}
The median is the middle value of a data set when the observations are arranged in ascending order.
:::

The calculation of the median depends on whether the sample size is odd or even.
If the sample size is odd, the median is the middle observation. If the sample size is even, the median is the average of the two middle observations.

Consider $n$ ordered observations $x_{(1)},x_{(2)},~\cdots~,x_{(n)}$.

- If $n$ is odd
    $$\text{Median} = \text{observation} \> \left( \frac{n + 1}{2} \right) $$
- If $n$ is even,
    $$\text{Median} = \text{average of observation} ~ \left(\frac{n}{2} \right) ~\text{and observation} ~ \left( \frac{n}{2} + 1 \right)$$
    
### Mode

::: {.definition name="Mode" #Mode}
The mode is the most frequently occurring observation in a data set.
:::

We are als interested in how the data is spread out or dispersed.
The most common numerical measures of the dispersion are the range, variance, and standard deviation. These measures summarize a set of data by identifying the spread or variability within that data.


### Variance

::: {.definition name="Variance" #Var}
The sample variance is the average squared deviation of each observation from the sample mean.

Let $x_{1}, ~x_{2}, ~x_{3}, ~\ldots , ~x_{n}$ represent a sample of $n$ observations.
The sample Variance is defined as:
	$$s^{2} = \frac{ \displaystyle\sum_{i=1}^{n} (x_{i} - \bar{x})^{2} }{n - 1}  = \frac{ (x_{1} - \bar{x})^{2} +  (x_{2} - \bar{x})^{2} + \ldots + (x_{n} - \bar{x})^{2} }{n-1}$$
where $\bar{x}$ is the sample mean in Definition \@ref(def:Mean).
:::


In the Definition \@ref(def:Var), we divide by $n-1$ instead of $n$ since we are estimating the
value of the population variance using sample data, and this calculation already includes utilizing
the sample mean $\bar{x}$ which itself is an estimate of the population mean $\mu$. 
This division by $n-1$ is known as Bessel's correction.

Another reason for dividing by $n-1$ is because the sample variance is an unbiased estimator of the population variance, which is a topic covered in [STA260](https://utm.calendar.utoronto.ca/course/sta260h5) and beyond the scope of this course.

In the calculation of the sample variance, notice each term is squared, which means that the variance is always a non-negative number. 
When this occurs, the units of the variance are the square of the units of the original data.
To return to the original units, we take the square root of the variance to obtain the standard deviation.


### Standard Deviation

::: {.definition name="Standard Deviation" #StDev}
Let $x_{1}, x_{2}, ~x_{3}, ~\ldots , ~x_{n}$ represent a sample of $n$ observations.
The sample standard deviation is defined as:
$$s = + \sqrt{s^{2}}$$
where $s^{2}$ is the sample variance in Definition \@ref(def:Var).
:::


When we analyze data we typically describe the data in terms of the mean and standard deviation
as the units are the same. However, the calculation of the sample variance is an important
intermediate step.

One of the measures of central tendency and dispersion on their own will not give as a complete picture
of the data, however when analyzed collectively, we are able to get a better overall understanding of the data.


### Range

The range is the difference between the largest and smallest observations in a data set. 

::: {.definition name="Standard Deviation" #StDev}
Let $x_{1}, ~x_{2}, ~x_{3}, ~\ldots , ~x_{n}$ represent a sample of $n$ observations.
The range is 
$$\text{Range} = \max(x_{1}, x_{2}, ~x_{3}, ~\ldots , ~x_{n}) - \min(x_{1}, x_{2}, ~x_{3}, ~\ldots , ~x_{n}) $$
:::

### Percentiles

::: {.definition name="Percentile" #Percentile}
In an ordered data set, the $p^{th}$ percentile is the value such that $p\%$ of all observations lie below it.
:::

Percentiles are a value that are relative to the rest of the data.

::: {.example}
A class writes a difficult test and a student obtained a mark of 70\% on this test.
Although 72\% is a B-, since the test was difficult, this may be a good score. 
In fact, the student who scored 72\% is in the $90^{th}$ percentile for this test. 
This means that they scored better than 90\% of the rest of the class.
:::


### Quartiles

Quartiles are special cases of percentiles.

::: {.definition name="Quartiles" #Quartiles}
In an ordered data set, quartiles are three values which divide the data into four groups such that each group consists of one fourth of the data. There three quartiles are the:

-	First Quartile (Q$_{1}$)	:	A value such that 25\% (i.e. a quarter) of all observations lie below it.
-	Second Quartile (Q$_{2}$)	:	A value such that 50\% (i.e. two quarters) of all observations lie below it.
-	Third Quartile (Q$_{3}$)	:	A value such that 75\% (i.e. three quarters) of all observations lie below it.
:::


::: {.remark}
The second quartile ($Q_2$) in Definition \@ref(def:Quartiles) is the same as the *median* in Definition \@ref(def:Median)
:::

<!-- ::: {#note:skewnote .note name="Skewness Note"} -->
<!-- **Note:** The second quartile ($Q_2$) in Definition \@ref(def:Quartiles) is the same as the *median* in Definition \@ref(def:Median). -->
<!-- ::: -->

We work with quartiles since they are provide an intutitive way to partition data into quarters, and quartiles are also used in graphical summaries discussed in Section \@ref(sec:GraphicalTechniques).

A measure of dispersion involving quartiles is the inter-quartils range (IQR).

::: {.definition name="Interquartile Range" #IQR}
The interquartile range (IQR) is the difference between the third quartile and the first quartile.
$$\text{IQR} = Q_{1} - Q_{3}$$
where $Q_{1}$ and $Q_{3}$ are the first and third quartiles respectively in Definition \@ref(def:Quartiles).
:::

## Graphical Techniques {#sec:GraphicalTechniques} 

Raw values on their own can be difficult to interpret, especially with very large data sets. 
Graphical representations can be very useful for representing information. 
When implemented correctly, graphical plots can provide more intuitive and direct method to interpret information being analyzed.

We begin this section by introducing a term to describe the shape of a distribution of data.

::: {.definition name="Skewness" }
Skewness refers to such a measure of symmetry or lack of symmetry in the distribution of data.
:::

We can describe data as being left skewed, right skewed or symmetric. If the data is
described as symmetric, this implies that most observations are concentrated around the
mean and tail off fairly evenly on both sides of the mean. If the data is described as right
skewed, this implies that more observations are concentrated on smaller
values and we observe a longer tail to the right side of the mean. If the data is described
as left skewed, this implies that more observations are concentrated on
large values and we observe a longer tail to the left side of the mean. 
We also may have data that is bimodal, which means that there are two distinct peaks in the distribution of the data.


::: {.remark}
*Right* skewed data is is also referred to as *positively* skewed and *left* skewed data is is also referred to as *negatively* skewed.
:::

Figures \@ref(fig:LeftSkewExample), \@ref(fig:RightSkewExample) and \@ref(fig:SymmetryExample) illustrate types of skewness.

```{r LeftSkewExample, fig.cap="An illustration of a left (negative) skewed distribution.", echo=FALSE, fig.align='center', warning=FALSE}
library(ggplot2)

# style‑guide colors
# fill_blue   <- "#529EFF"
# line_blue   <- "#1F4E8C"
# median_blue <- "#00C1AA"
# mean_pink   <- "#FF62BC"
fill_blue   <- "#619CFF"
line_blue   <- "#0064c8"
median_blue <- "#0064c8"
mean_pink   <- "#FF68A1"

# simulate a true left‑skewed sample (Beta(5,2))
set.seed(2025)
df <- data.frame(x = rbeta(1000, 5, 2))

# density estimate (for annotation positions)
dens     <- density(df$x, adjust = 2)
med_val  <- median(df$x)
mean_val <- mean(df$x) - 0.15

# y‑positions for labels
y_tail <- approx(dens$x, dens$y, xout = 0.2)$y + 0.005
y_med  <- approx(dens$x, dens$y, xout = med_val)$y + 0.3
y_mean <- approx(dens$x, dens$y, xout = mean_val)$y + 0.3

# determine top of plot to clip exactly
y_max <- max(y_tail, y_med, y_mean)

ggplot(df, aes(x = x)) +
  # shaded density
  geom_density(
    fill   = fill_blue,
    color  = line_blue,
    size   = 1.2,
    adjust = 2
  ) +
  # median & mean lines
  geom_vline(xintercept = med_val,  color = median_blue, size = 1) +
  geom_vline(xintercept = mean_val, color = mean_pink,   size = 1, linetype = "dashed") +
  # tail label (left tail)
  annotate(
    "text",
    x      = 0.2,
    y      = y_tail + 0.1,
    label  = "Left Tail",
    fontface= "bold",
    hjust   = 0.5,
    vjust   = -1.5
  ) +
  # median label to left of line
  annotate(
    "text",
    x      = med_val - 0.025,
    y      = y_med - 0.1,
    label  = "median",
    fontface= "bold",
    hjust   = -0.5,
    color   = median_blue
  ) +
  # mean label to right of line
  annotate(
    "text",
    x      = mean_val - 0.020,
    y      = y_mean - 0.1,
    label  = "mean",
    fontface= "bold",
    hjust   = 1.1,
    color   = mean_pink
  ) +
  # exact axis limits, no extra padding
  scale_x_continuous(limits = c(-0.0, 1.25), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, y_max), expand = c(0, 0)) +
  # clean theme, no fixed aspect
  theme_minimal(base_size = 14) +
  theme(
    panel.grid       = element_blank(),
    panel.background = element_blank(),
    plot.title       = element_text(face = "bold", hjust = 0.5),
    axis.title       = element_text(face = "bold")
  ) +
  labs(
    title = "Left-Skewed Density",
    x     = "Value",
    y     = "Density"
  )
```


```{r RightSkewExample, fig.cap="An illustration of a right (positive) skewed distribution.", echo=FALSE, fig.align='center', warning=FALSE}
library(ggplot2)

# style‑guide colors
fill_blue   <- "#619CFF"
line_blue   <- "#0064c8"
median_blue <- "#0064c8"
mean_pink   <- "#FF68A1"

# simulate right‑skewed data
set.seed(2025)
df <- data.frame(x = rbeta(1000, 2, 5))

# density estimate for positioning & y‑limit
dens     <- density(df$x, adjust = 2)
med_val  <- median(df$x)
mean_val <- mean(df$x) + 0.15

# compute annotation y‑positions
y_tail <- approx(dens$x, dens$y, xout = 0.8)$y + 0.01
y_med  <- approx(dens$x, dens$y, xout = med_val)$y + 0.3
y_mean <- approx(dens$x, dens$y, xout = mean_val)$y + 0.3

# find top of plot
y_max <- max(y_tail, y_med, y_mean)

ggplot(df, aes(x = x)) +
  # shaded density
  geom_density(
    fill   = fill_blue,
    color  = line_blue,
    size   = 1.2,
    adjust = 2
  ) +
  # median & mean lines
  geom_vline(xintercept = med_val,  color = median_blue, size = 1) +
  geom_vline(xintercept = mean_val, color = mean_pink,   size = 1, linetype = "dashed") +
  # annotations
  annotate("text",
           x      = 0.8, y      = y_tail,
           label  = "Right Tail",
           fontface= "bold", hjust = 0.25, vjust = -1.5) +
  annotate("text",
           x      = med_val - 0.025, y = y_med - 0.1,
           label  = "median",
           fontface= "bold", hjust = 1.1, color = median_blue) +
  annotate("text",
           x      = mean_val + 0.025, y = y_mean,
           label  = "mean",
           fontface= "bold", hjust = -0.1, color = mean_pink) +
  # exact axis limits, no extra padding
  scale_x_continuous(limits = c(-0.15, 1),   expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, y_max),   expand = c(0, 0)) +
  # clean theme
  theme_minimal(base_size = 14) +
  theme(
    panel.grid       = element_blank(),
    panel.background = element_blank(),
    plot.title       = element_text(face = "bold", hjust = 0.5),
    axis.title       = element_text(face = "bold")
  ) +
  labs(
    title = "Right-Skewed Density",
    x     = "Value",
    y     = "Density"
  )
```

<!-- rgba(0,100,200,1) #00C19A -->

```{r SymmetryExample, fig.cap="An illustration of an approximately symmetric distribution.", echo=FALSE, fig.align='center', warning=FALSE}
library(ggplot2)

# style‑guide colors
fill_blue   <- "#619CFF"
line_blue   <- "#0064c8"
median_blue <- "#0064c8"
mean_pink   <- "#FF68A1"

# simulate a symmetric sample (Beta(5,5) is symmetric on [0,1])
set.seed(2025)
df <- data.frame(x = rbeta(1000, 5, 5))

# density estimate (for annotation positions)
dens     <- density(df$x, adjust = 2)
med_val  <- median(df$x) - 0.01
mean_val <- mean(df$x) + 0.01

# y‑positions for labels
y_med  <- approx(dens$x, dens$y, xout = med_val)$y + 0.3
y_mean <- approx(dens$x, dens$y, xout = mean_val)$y + 0.3

# determine top of plot to clip exactly
y_max <- max(y_med, y_mean)

ggplot(df, aes(x = x)) +
  # shaded density
  geom_density(
    fill   = fill_blue,
    color  = line_blue,
    size   = 1.2,
    adjust = 2
  ) +
  # median & mean lines
  geom_vline(xintercept = med_val,  color = median_blue, size = 1) +
  geom_vline(xintercept = mean_val, color = mean_pink,   size = 1, linetype = "dashed") +
  # median label to left of line
  annotate(
    "text",
    x       = med_val - 0.025,
    y       = y_med - 0.1,
    label   = "median",
    fontface= "bold",
    hjust   = 1.1,
    color   = median_blue
  ) +
  # mean label to right of line
  annotate(
    "text",
    x       = mean_val + 0.025,
    y       = y_mean - 0.1,
    label   = "mean",
    fontface= "bold",
    hjust   = -0.1,
    color   = mean_pink
  ) +
  # exact axis limits, no extra padding
  scale_x_continuous(limits = c(0, 1),    expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, y_max), expand = c(0, 0)) +
  # clean theme, no fixed aspect ratio
  theme_minimal(base_size = 14) +
  theme(
    panel.grid       = element_blank(),
    panel.background = element_blank(),
    plot.title       = element_text(face = "bold", hjust = 0.5),
    axis.title       = element_text(face = "bold")
  ) +
  labs(
    title = "Symmetric Density",
    x     = "Value",
    y     = "Density"
  )
```

<!-- ::: {#note:skewnote .note name="Skewness Note"} -->
<!-- **Note:** *Right* skewed data is is also referred to as *positively* skewed and *left* skewed data is is also referred to as *negatively* skewed. -->
<!-- ::: -->

Data sets can have points in it which may be unusual or unexpected. 
These terms are called *outliers*.

::: {.definition name="Outlier" }
An outlier is an unusual data point which appears to lies outside the overall pattern of the
rest of the data.
:::

In other words, an outlier falls outside the range in which we would expect to see “typical”
data values. Outliers may be present in data for a variety of reasons such as transcription
error, measurement error, or we may have just measured a rare and unusual observation. In
any case it is not a good practice to simply ignore outliers. All outliers should be investigated
before deciding whether the observation should be included in data analysis or discarded.


### Histograms

Histograms and boxplots are useful tools which can allow us to visually determine
the skewness of a distribution. By noting skewness, we get even more information about
our data. Note however that we may not always be able to determine skewness by visually
observing a histogram. A histogram is similar to a bar chart with the distinction that his-
tograms can only be created for quantitative data.

The width of the bars does not have any numerical meaning for bar charts, however bar
width matters for histograms. They are an important factor in the creation of histograms.
Interval width selection (or bin size selection) is an advanced topic that can be studied ex-
tensively and there are several rules available to select interval widths. However for the scope
of this course we will keep things simple and allow the reader to intuitively choose bin size.
After constructing a histogram, we can change the interval widths until we feel that we have
a picture that is a good representation of our data.

A histogram is constructed by dividing the range of the data into intervals (or bins) and counting how many observations fall into each interval. The intervals are usually of equal width, but this is not a requirement. 

<!-- | Class Interval       | Frequency | Relative Freq.           |        | Cumulative Freq.             | Cumulative Relative Freq.    | -->
<!-- |----------------------|:---------:|--------------------------|--------|------------------------------|------------------------------| -->
<!-- | $[a_{1}, b_{1})$     | $f_{1}$   | $r_{1} = f_{1} / F$      |        | $f_{1}$                      | $r_{1}$                      | -->
<!-- | $[a_{2}, b_{2})$     | $f_{2}$   | $r_{2} = f_{2} / F$      |        | $f_{1} + f_{2}$              | $r_{1} + r_{2}$              | -->
<!-- | $[a_{3}, b_{3})$     | $f_{3}$   | $r_{3} = f_{3} / F$      |        | $f_{1} + f_{2} + f_{3}$      | $r_{1} + r_{2} + r_{3}$      | -->
<!-- | $\vdots$             | $\vdots$  | $\vdots$                 |        | $\vdots$                     | $\vdots$                     | -->
<!-- | $[a_{m}, b_{m}]$     | $f_{m}$   | $r_{m} = f_{m} / F$      |        | $f_{1} + \ldots + f_{m} = F$ | $r_{1} + \ldots + r_{m} = 1$ | -->
<!-- |                      | $F = \displaystyle\sum_{i=1}^{m} f_{i}$       | $1$ | |                      |                              | -->

```{r histconstruct, echo=FALSE, results='asis', warning = FALSE}
library(knitr)
library(kableExtra)
class_int = c("$[a_{1}, b_{1})$", 
              "$[a_{2}, b_{2})$", 
              "$[a_{3}, b_{3})$",
              "$\\vdots$",
              "$[a_{m}, b_{m}]$",
              "")

Frequency = c("$f_{1}$",
              "$f_{2}$",
              "$f_{3}$",
              "$\\vdots$",
              "$f_{m}$",
              "$F = \\displaystyle\\sum_{i=1}^{m} f_{i}$")

Rel_Freq = c("$r_{1} = f_{1} / F$",
             "$r_{2} = f_{2} / F$",
             "$r_{3} = f_{3} / F$",
             "\\vdots",
             "$r_{m} = f_{m} / F$",
             "1")

hist_tab = data.frame(class_int, Frequency, Rel_Freq)

colnames(hist_tab) = c("Class Interval", "Frequency", "Relative Freq.")

# hist_tab %>%
#   knitr::kable(full_width = F, caption = "How a histogram is constructed")

kbl(hist_tab, caption = "How a histogram is constructed") %>%
  kable_styling(full_width = F, bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```


The histogram is then created by plotting the intervals on the $x$-axis and the frequency or relative frequency of observations in each interval on the $y$-axis.
When frequency is plotted on the $y$-axis, the histogram is referred to as a frequency histogram and
when relative frequency is plotted on the $y$-axis, the histogram is referred to as a relative frequency histogram.

::: {.remark}
The bin size in table \@ref(tab:histconstruct) is often chosen to be equal size, however it does not necessarily have to be the case and we can have bin sizes of unequal widths.
:::

::: {.remark}
When we use the term *histogram*, we often refer to a frequency histogram unless otherwise specified.
:::

<!-- ::: {#note:histnote .note name="Note"} -->
<!-- **Note:** When we use the term *histogram*, we often refer to a frequency histogram unless otherwise specified. -->
<!-- ::: -->

Some examples of histograms which also illustrate the concept of skewness are shown in Figures \@ref(fig:HistExampleLeft), \@ref(fig:HistExampleRight), and \@ref(fig:HistExampleSymm). The histograms show how the distribution of data can be skewed to the left, skewed to the right, or approximately symmetric.

```{r HistExampleLeft, fig.cap="An illustration of a histogram to have a left (or negative) skew probability distribution.", echo=FALSE, fig.align='center', warning=FALSE}
library(plotly)

set.seed(5)
x <- rbeta(6000, shape1 = 5, shape2 = 1.5) * 100

binwidth <- 5

# 1) Scaled density
dens     <- density(x, adjust = 1.2)
y_scaled <- dens$y * length(x) * binwidth

# 2) Compute max y (histogram + density)
hist_data <- hist(x,
                  breaks = seq(0, 120, by = binwidth),
                  plot   = FALSE)
max_y     <- max(max(hist_data$counts), max(y_scaled))

plot_ly() %>%
  # histogram (always on)
  add_histogram(
    x          = x,
    xbins      = list(size = binwidth),
    marker     = list(color = "#c6adff", line = list(color = "white", width = 1)),
    showlegend = FALSE
  ) %>%
  # density (start hidden)
  add_lines(
    x          = dens$x,
    y          = y_scaled,
    line       = list(color = "#0064c8", width = 3),
    visible    = FALSE,
    showlegend = FALSE
  ) %>%
  layout(
    title = list(text = "Histogram with Left‑Skewed Data"),
    xaxis = list(title = "Value", range = c(0, 110)),
    yaxis = list(title = "Frequency", range = c(0, 900), fixedrange = TRUE),
    bargap = 0.02,
    updatemenus = list(
      list(
        type      = "buttons",
        direction = "right",
        xanchor   = "left",
        x         = 0.10,
        y         = 0.70,
        buttons = list(
          list(
            method  = "restyle",
            args    = list("visible", list(TRUE, FALSE)),
            args2   = list("visible", list(TRUE, TRUE)),
            label   = "Show/Hide Density",
            execute = TRUE      # ← ensures first click toggles on
          )
        )
      )
    )
  )
```

```{r HistExampleRight, fig.cap="An illustration of a histogram to have a right (or positive) skew probability distribution.", echo=FALSE, fig.align='center', warning=FALSE}
library(plotly)

set.seed(5)
x        <- rbeta(6000, shape1 = 1.5, shape2 = 5) * 100
binwidth <- 5

# 1) Compute & scale density
dens     <- density(x, adjust = 1.2)
y_scaled <- dens$y * length(x) * binwidth

# 2) Find max y
hist_data <- hist(x,
                  breaks = seq(0, 120, by = binwidth),
                  plot   = FALSE)
max_y     <- max(max(hist_data$counts), max(y_scaled))

fig <- plot_ly() %>%
  # histogram with equal-width bins from 0 to 120
  add_histogram(
    x          = x,
    xbins      = list(start = 0, end = 90, size = binwidth),
    marker     = list(color = "#c6adff", line = list(color = "white", width = 1)),
    showlegend = FALSE
  ) %>%
  # density (starts hidden)
  add_lines(
    x          = dens$x,
    y          = y_scaled,
    line       = list(color = "#0064c8", width = 3),
    visible    = FALSE,
    showlegend = FALSE
  ) %>%
  layout(
    title   = list(text = "Histogram with Right‑Skewed Data"),
    xaxis   = list(title = "Value", range = c(0, 90)),
    yaxis   = list(title = "Frequency", range = c(0, 900), fixedrange = TRUE),
    bargap  = 0.02,
    updatemenus = list(
      list(
        type      = "buttons",
        direction = "right",
        xanchor   = "left",
        x         = 0.80,
        y         = 0.70,
        buttons = list(
          list(
            method  = "restyle",
            args    = list("visible", list(TRUE, FALSE)),
            args2   = list("visible", list(TRUE, TRUE)),
            label   = "Show/Hide Density",
            execute = TRUE      # ← ensures first click toggles on
          )
        )
      )
    )
  )

fig
```


```{r HistExampleSymm, fig.cap="An illustration of a histogram to have a symmetric probability distribution.", echo=FALSE, fig.align='center', warning=FALSE}
library(plotly)

set.seed(12)

# Generate approximately symmetric data
x        <- rnorm(6000, mean = 60, sd = 15)
binwidth <- 5

# 1) Compute & scale density
dens     <- density(x, adjust = 1.25)
y_scaled <- dens$y * length(x) * binwidth

# 2) Find the maximum y‑value (histogram + density) for a locked y‑axis
hist_data <- hist(x,
                  breaks = seq(0, 120, by = binwidth),
                  plot   = FALSE)
max_y     <- max(max(hist_data$counts), max(y_scaled))

# 3) Build the Plotly figure
fig <- plot_ly() %>%
  # A: histogram trace (always visible)
  add_histogram(
    x          = x,
    xbins      = list(start = 0, end = 120, size = binwidth),
    marker     = list(color = "#c6adff", line = list(color = "white", width = 1)),
    showlegend = FALSE
  ) %>%
  # B: density trace (starts hidden)
  add_lines(
    x          = dens$x,
    y          = y_scaled,
    line       = list(color = "#0064c8", width = 3),
    visible    = FALSE,
    showlegend = FALSE
  ) %>%
  layout(
    title   = list(text = "Histogram with Approximaetly Symmetric Data"),
    xaxis   = list(title = "Value", range = c(0, 120)),
    yaxis   = list(title = "Frequency", range = c(0, max_y), fixedrange = TRUE),
    bargap  = 0.02,
    updatemenus = list(
      list(
        type      = "buttons",
        direction = "right",
        xanchor   = "left",
        x         = 0.80,
        y         = 0.70,
        buttons = list(
          list(
            method  = "restyle",
            args    = list("visible", list(TRUE, FALSE)),
            args2   = list("visible", list(TRUE, TRUE)),
            label   = "Show/Hide Density",
            execute = TRUE      # ← ensures first click toggles on
          )
        )
      )
    )
  )

fig
```


The bin widths selected can change the appearance of histograms
The interactive application below allows us to examine the effect of bin size on a histogram from the [faithful](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/faithful.html) data set which measures the waiting time between eruptions of the Old Faithful geyser in Yellowstone National Park, Wyoming, USA.

<center>
```{r HistShinyApp, echo=FALSE, out.width='100%', warning=FALSE}
knitr::include_app("https://nishan-mudalige.shinyapps.io/Histogram-Shiny-App/", height = "743")
```
</center>


### Boxplots

Boxplots are another visual aid we can use to present and interpret data. They are one
of the most simple graphical techniques to analyze visually and are therefore usually relatively immediate to interpret. 
Boxplots are also known as *box-and-whisker* plots since they consists of figures resembling boxes along with a series of
lines which we refer to as whiskers. The whiskers represent the quartiles as well as special
values which we refer to as the *lower whisker* and the *upper whisker*.

::: {.definition name="Upper and Lower Whiskers" #UpperLowerWhiskers}
Let $Q_{1}$, $Q_{3}$ and $IQR$ represent the first quartile, third quartile and 
interquartile range respectively as defined in \@ref(def:Quartiles) and \@ref(def:IQR) respectively.
Then,
\[ \text{Lower Whisker} ~ = ~ Q_{1} - 1.5 \times IQR \]
\[ \text{Upper Whisker} ~ = ~ Q_{3} + 1.5 \times IQR \]
:::

The box in a boxplot is divided into two parts, the lower and upper quartiles. The lower quartile is represented by the bottom of the box and the upper quartile is represented by the top of the box. The line inside the box represents the median of the data set.
Box plots can be used to to get a sense of a data set.

Consider the following summary data

```{r BoxplotData, echo=FALSE, results='asis', warning = FALSE}
# Load required libraries
library(knitr)
library(kableExtra)

# Create summary as a named numeric vector
summary_stats <- c(
  Min = 0.500,
  `1st Qu.` = 8.782,
  Median = 9.855,
  Mean = 10.036,
  `3rd Qu.` = 11.521,
  Max = 19.000
)

# Convert to one-row wide format data frame
summary_wide <- as.data.frame(t(summary_stats))

# Create formatted table (HTML for Bookdown)
kable(summary_wide, format = "html", booktabs = TRUE, caption = "Summary Statistics") %>%
  kable_styling(full_width = FALSE, position = "center")
```

The summary statistics in Table \@ref(tab:BoxplotData) is represented visually as an interactive boxplot in Figure \@ref(fig:BoxplotExample). 
Hover over the boxplot to identify the median, quartiles and whiskers.

```{r BoxplotExample, fig.cap="An illustration of an annotated box-plot with whiskers and potential outliers.", echo=FALSE, fig.align='center', warning=FALSE}
# Load Plotly
library(plotly)

# Base data
set.seed(123)
x <- rnorm(50, mean = 10, sd = 2)

# Define outliers
outliers <- c(0.5, 1, 18, 19)

x = c(x, outliers)

outlier_labels <- rep("potential outlier", length(outliers))
x_position <- rep(1, length(outliers))  # Align with boxplot

# Create plot
p <- plot_ly() %>%
  # Boxplot
  add_trace(
    x = rep(1, length(x)),
    y = ~x,
    type = "box",
    boxpoints = "none",
    name = "",
    hoverinfo = "y",
    fillcolor = "#00C19A",
    marker = list(color = "#00C19A"),
    line = list(color = 'rgba(0,100,200,1)')
  ) %>%
  # Custom outliers
  add_trace(
    x = ~x_position,
    y = ~outliers,
    type = "scatter",
    mode = "markers",
    marker = list(
      color = "#F8766D",
      size = 8,
      line = list(color = "#FF68A1", width = 1)
    ),
    hoverinfo = "text",
    hovertext = outlier_labels,
    showlegend = FALSE
  ) %>%
  layout(
    title = "Interactive Boxplot<br>(Hover over outliers)",
    yaxis = list(
      title = "Values",
      range = c(0, 21)        # Set y-axis limits
    ),
    xaxis = list(
      showticklabels = FALSE,
      title = NA,
      range = c(0.5, 1.5)
    )
  )

# Disable lasso and box select tools
config(
  p,
  modeBarButtonsToRemove = c("select2d", "lasso2d")
)
```

Boxplots can also efficiently be used to compare multiple distributions and highlight skewness, spread, and outliers from each of them.
If the median cuts the box with upper area smaller than lower area,
then we say that box-plot with left skew probability distribution. Or,
if the median cuts the box with upper area larger than lower area,
then we say that box-plot with right skew probability distribution.
Otherwise, if the median cuts the box with upper area equal to
lower area, then we say that box-plot with symmetric probability
distribution.

```{r BoxplotNegativeExample, fig.cap="An illustration of an boxplot with left (negative) skewed.", echo=FALSE, fig.align='center', warning=FALSE}
# Load Plotly
library(plotly)

# Generate left-skewed data by flipping right-skewed log-normal
set.seed(123)
raw_right <- c(rlnorm(35, meanlog = 2, sdlog = 0.75), runif(5, 14, 18) )
raw <- max(raw_right) - raw_right  # flip for left skew

# Remove statistical outliers
q1 <- quantile(raw, 0.25)
q3 <- quantile(raw, 0.75)
iqr <- q3 - q1
upper_whisker <- q3 + 1.5 * iqr
lower_whisker <- q1 - 1.5 * iqr
x <- raw[raw >= lower_whisker & raw <= upper_whisker]
x <- x + 2

mean_x <- mean(x)
median_x <- median(x)

# Plot
p <- plot_ly() %>%
  add_trace(
    x = rep(1, length(x)),
    y = ~x,
    type = "box",
    boxpoints = "none",  # hide all points
    name = "",
    hoverinfo = "skip",
    fillcolor = "#00C19A",
    marker = list(color = "#00C19A"),
    line = list(color = 'rgba(0,100,200,1)')
  ) %>%
  layout(
    title = "Boxplot suggesting left skewed data<br>(Mean < Median)",
    yaxis = list(
      title = "Values",
      range = c(0, max(x) * 1.1)
    ),
    xaxis = list(
      showticklabels = FALSE,
      title = NA,
      range = c(0.5, 1.5)
    ),
    shapes = list(
      list(
        type = "line",
        xref = "x",
        yref = "y",
        x0 = 0.757,
        x1 = 1.247,
        y0 = mean_x - 0.5, 
        y1 = mean_x - 0.5 ,
        line = list(
          dash = "dash",
          color = "#FF68A1",
          width = 2
        )
      )
    ),
    annotations = list(
      list(
        x = 1.325,
        y = mean_x - 0.50,
        text = "mean",
        xref = "x",
        yref = "y",
        showarrow = FALSE,
        font = list(color = "#FF68A1", size = 18)
      ),
      list(
        x = 1.325,
        y = median_x + 0.25,
        text = "median",
        xref = "x",
        yref = "y",
        showarrow = FALSE,
        font = list(color = "rgba(0,100,200,1)", size = 18)
      )
    )
  )

# Remove all interactivity
config(
  p,
  displayModeBar = FALSE,
  staticPlot = TRUE
)

```

```{r BoxplotpositiveExample, fig.cap="An illustration of an boxplot with right (positive) skewed.", echo=FALSE, fig.align='center', warning=FALSE}
# Load Plotly
library(plotly)

# Generate right-skewed data WITHOUT outliers (control max values)
set.seed(123)
raw <- rlnorm(35, meanlog = 2, sdlog = 0.5)
q1 <- quantile(raw, 0.25)
q3 <- quantile(raw, 0.75)
iqr <- q3 - q1
upper_whisker <- q3 + 1.5 * iqr
lower_whisker <- q1 - 1.5 * iqr
x <- raw[raw >= lower_whisker & raw <= upper_whisker]  # trim outliers

mean_x <- mean(x)
median_x <- median(x)

# Plot
p <- plot_ly() %>%
  add_trace(
    x = rep(1, length(x)),
    y = ~x,
    type = "box",
    boxpoints = "none",  # no visible points
    name = "",
    hoverinfo = "skip",
    fillcolor = "#00C19A",
    marker = list(color = "#00C19A"),
    line = list(color = 'rgba(0,100,200,1)')
  ) %>%
  layout(
    title = "Boxplot suggesting right skewed data<br>(Mean > Median)",
    yaxis = list(
      title = "Values",
      range = c(0, max(x) * 1.1)
    ),
    xaxis = list(
      showticklabels = FALSE,
      title = NA,
      range = c(0.5, 1.5)
    ),
    shapes = list(
      list(
        type = "line",
        xref = "x",
        yref = "y",
        x0 = 0.757,
        x1 = 1.247,
        y0 = mean_x + 1, 
        y1 = mean_x + 1,
        line = list(
          dash = "dash",
          color = "#FF68A1",
          width = 2
        )
      )
    ),
    annotations = list(
      list(
        x = 1.325,
        y = mean_x + 0.8 + 0.25,
        text = "mean",
        xref = "x",
        yref = "y",
        showarrow = FALSE,
        font = list(color = "#FF68A1", size = 18)
      ),
      list(
        x = 1.325,
        y = median_x - 0.25,
        text = "median",
        xref = "x",
        yref = "y",
        showarrow = FALSE,
        font = list(color = "rgba(0,100,200,1)", size = 18)
      )
    )
  )

# Remove all interactivity
config(
  p,
  displayModeBar = FALSE,
  staticPlot = TRUE
)
```

```{r BoxplotSymmetryExample, fig.cap="An illustration of an boxplot with symetric skewed.", echo=FALSE, fig.align='center', warning=FALSE}
# Load Plotly
library(plotly)

# Base data
set.seed(123)
x <- rnorm(50, mean = 10, sd = 4)
mean_x <- mean(x)
median_x <- median(x)

# Create plot
p <- plot_ly() %>%
  # Boxplot
  add_trace(
    x = rep(1, length(x)),
    y = ~x,
    type = "box",
    boxpoints = "none",
    name = "",
    hoverinfo = "skip",  # <-- disables hover labels
    fillcolor = "#00C19A",
    marker = list(color = "#00C19A"),
    line = list(color = 'rgba(0,100,200,1)')
  ) %>%
  layout(
    title = "Boxplot suggesting symmetic data<br>(Mean ≈ Median)",
    yaxis = list(
      title = "Values",
      range = c(0, 21)
    ),
    xaxis = list(
      showticklabels = FALSE,
      title = NA,
      range = c(0.5, 1.5)
    ),
    shapes = list(
      list(
        type = "line",
        xref = "x",
        yref = "y",
        x0 = 0.757, 
        x1 = 1.247,
        y0 = mean_x, y1 = mean_x,
        line = list(
          dash = "dash",
          color = "#FF68A1",
          width = 2
        )
      )
    ),
    annotations = list(
      list(
        x = 1.325,
        y = mean_x + 0.25,
        text = "mean",
        xref = "x",
        yref = "y",
        showarrow = FALSE,
        font = list(color = "#FF68A1", size = 18)
      ),
      list(
        x = 1.325,
        y = median_x - 0.25,
        text = "median",
        xref = "x",
        yref = "y",
        showarrow = FALSE,
        font = list(color = "rgba(0,100,200,1)", size = 18)
      )
    )
  )

# Remove all interactive tools and mode bar
config(
  p,
  displayModeBar = FALSE,
  staticPlot = TRUE
)
```
