# Foundations of Inference

The aim of statistical inference is to produce estimators of the population parameters
based on a smaller sample and to quantify the accuracy of these estimators in terms of a
probability statement. In other words we are interested in the values of unknown parameters
for a population thus we collect data from a sample and calculate statistics to estimate the
parameters of interest. We then quantify our confidence that the statistic is representative
of the parameter.

## Some Important Statistical Distributions {#sec:inference}

### Introduction to the Normal Distribution

In probability theory and statistics, the normal distribution is one of the most
important distribution used to model many populations and processes which occur in reality.

::: {.definition name="Normal distribution" #NormalDist}
Let $X$ be a random variable with mean $-\infty < \mu < +\infty$ and variance $\sigma^2 > 0$.
We say that $X$ folowes a normal distribution if the probability density function of $X$ is:
$$f(x) = \frac{1}{\sqrt{2 \pi \sigma^2}} \cdot e^{-\displaystyle\frac{(x - \mu)^2}{2\sigma^2}}, \text{ for $-\infty < x < \infty$.}$$
:::

::: {.remark}
The normal distribution is also called Gaussian distribution since it was 
discovered Johann Carl Friedrich Gauss in 1809. 
Gauss is one of the most influential and important mathematicians in history.
:::

The normal distribution is symmetric about its mean $\mu$ and has a bell-shaped curve.
It is completely characterized by two parameters which are the the mean $\mu$ and the standard deviation $\sigma$.
Some examples are shown in Figure \@ref(fig:NormalExample).

```{r NormalExample, fig.cap="Some examples of normal distrubution curves with different means and variances", echo=FALSE, fig.align='center', warning=FALSE}
library(ggplot2)
library(grid)

# 1) Define your parameters
mu1  <- -2; sd1  <- 0.5; col1 <- "#00BA38"; lab1 <- "N(-2, 0.5²)"
mu2  <-  0; sd2  <- 1.0; col2 <- "#F8766D"; lab2 <- "N(0, 1²)"
mu3  <-  3; sd3  <- 1.5; col3 <- "#619CFF"; lab3 <- "N(3, 1.5²)"

# 2) Base plot with no default axis lines or grid
p <- ggplot(data.frame(x = c(-6, 10)), aes(x = x)) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid      = element_blank(),
    axis.line       = element_blank(),
    axis.ticks      = element_blank(),
    axis.text       = element_blank(),
    axis.title      = element_text(face = "bold")
  ) +
  labs(x = "x", y = "Density")

# 3) Add arrow‐tipped axes via geom_segment
p <- p +
  # x-axis from x=-6→10
  geom_segment(
    aes(x = -6, y = 0, xend = 10, yend = 0),
    arrow = arrow(length = unit(0.2, "cm")), inherit.aes = FALSE
  ) +
  # y-axis from y=0→max density (~0.8)
  geom_segment(
    aes(x = -6, y = 0, xend = -6, yend = 0.85),
    arrow = arrow(length = unit(0.2, "cm")), inherit.aes = FALSE
  )

# 4) Shade area and draw first curve + label
p <- p +
  stat_function(
    fun       = dnorm,
    args      = list(mean = mu1, sd = sd1),
    geom      = "area",
    fill      = col1,
    alpha     = 0.61,
    n         = 801,
    inherit.aes = TRUE
  ) +
  stat_function(
    fun       = dnorm,
    args      = list(mean = mu1, sd = sd1),
    colour    = col1,
    size      = 1.2,
    n         = 801,
    inherit.aes = TRUE
  ) +
  annotate(
    "text",
    x        = mu1,
    y        = dnorm(mu1, mu1, sd1) * 1.05,
    label    = lab1,
    colour   = col1,
    fontface = "bold",
    hjust    = 0.5
  )

# 5) Shade area and draw second curve + label
p <- p +
  stat_function(
    fun       = dnorm,
    args      = list(mean = mu2, sd = sd2),
    geom      = "area",
    fill      = col2,
    alpha     = 0.61,
    n         = 801
  ) +
  stat_function(
    fun       = dnorm,
    args      = list(mean = mu2, sd = sd2),
    colour    = col2,
    size      = 1.2,
    n         = 801
  ) +
  annotate(
    "text",
    x        = mu2,
    y        = dnorm(mu2, mu2, sd2) * 1.05,
    label    = lab2,
    colour   = col2,
    fontface = "bold",
    hjust    = 0.50,
    vjust    = -0.15
  )

# 6) Shade area and draw third curve + label
p <- p +
  stat_function(
    fun       = dnorm,
    args      = list(mean = mu3, sd = sd3),
    geom      = "area",
    fill      = col3,
    alpha     = 0.61,
    n         = 801
  ) +
  stat_function(
    fun       = dnorm,
    args      = list(mean = mu3, sd = sd3),
    colour    = col3,
    size      = 1.2,
    n         = 801
  ) +
  annotate(
    "text",
    x        = mu3,
    y        = dnorm(mu3, mu3, sd3) * 1.05,
    label    = lab3,
    colour   = col3,
    fontface = "bold",
    hjust    = 0.5,
    vjust    = -0.25
  )

# 7) Render the final plot
p
```

A special case of normal distribution is standard normal distribution which is explained in Definition \@ref(def:StandardNormalDist).

::: {.definition name="Standard normal distribution" #StandardNormalDist}
Let $Z \sim N( \mu = 0, \sigma^2 = 1)$, then $Z$ has
probability density function as:
$$f(y) = \frac{1}{\sqrt{2\pi}} \cdot e^{-\displaystyle\frac{z^2}{2}}.$$
which is called the standard normal distribution.
:::

::: {.remark}
Standard normal random variables are often denoted by $Z$.
:::


A useful property of the normal distribution is that it is symmetric about its mean $\mu$.

::: {.remark}
For any symmetric (bell-shaped) curve, let $\mu$ be its mean and
$\sigma$ be its standard deviation, the following probability set
function is true:

- $1.$ $P(\mu - \sigma < X < \mu + \sigma) = 68.27\%;$

- $2.$ $P(\mu - 2\sigma < X < \mu + 2\sigma) = 95.45\%;$

- $3.$ $P(\mu - 3\sigma < X < \mu + 3\sigma) = 99.73\%.$
:::

```{r fig.align='center', echo=FALSE, engine='tikz', out.width='90%', fig.ext=if (knitr::is_latex_output()) 'pdf' else 'png', fig.cap='An illustration of the empirical rule.'}
\usetikzlibrary{decorations.pathmorphing, arrows, shapes, trees, positioning, matrix, calc, backgrounds}
\begin{tikzpicture}[xscale=1.8, yscale=6]
  \draw[->, thick] (-3.5,0) -- (3.5,0) node[right] {$x$};
  \draw[->, thick] (0,0) -- (0,0.45) node[above] {$f(x) = \text{peak}$};

  \draw[domain=-3.2:3.2, smooth, variable=\x, thick, blue] 
    plot ({\x}, {1/sqrt(2*pi) * exp(-0.5*\x*\x)});

  \fill[green!40, opacity=0.7] 
    plot[domain=-1:1] (\x, {1/sqrt(2*pi) * exp(-0.5*\x*\x)}) -- (1,0) -- (-1,0) -- cycle;

  \fill[orange!40, opacity=0.7] 
    plot[domain=-2:-1] (\x, {1/sqrt(2*pi) * exp(-0.5*\x*\x)}) -- (-1,0) -- (-2,0) -- cycle;
  \fill[orange!40, opacity=0.7] 
    plot[domain=1:2] (\x, {1/sqrt(2*pi) * exp(-0.5*\x*\x)}) -- (2,0) -- (1,0) -- cycle;

  \fill[cyan!40, opacity=0.7] 
    plot[domain=-3:-2] (\x, {1/sqrt(2*pi) * exp(-0.5*\x*\x)}) -- (-2,0) -- (-3,0) -- cycle;
  \fill[cyan!40, opacity=0.7] 
    plot[domain=2:3] (\x, {1/sqrt(2*pi) * exp(-0.5*\x*\x)}) -- (3,0) -- (2,0) -- cycle;

  \foreach \x in {-3,-2,-1,0,1,2,3} {
    \draw[dashed] (\x,0) -- (\x,{1/sqrt(2*pi) * exp(-0.5*\x*\x)});
    \node[below] at (\x,0) {\x$\sigma$};
  }

  \node at (0,0.22) {\textbf{68\%}};
  \node at (-1.5,0.22) {13.5\%};
  \node at (1.5,0.22) {13.5\%};
  \node at (-2.5,0.22) {2.35\%};
  \node at (2.5,0.22) {2.35\%};
  \node at (-3.25,0.22) {0.15\%};
  \node at (3.25,0.22) {0.15\%};
\end{tikzpicture}
```

### Introduction to the Chi-Square distribution

Before we introduce the chi-Square square distribution, we recall the gamma distribution.

::: {.definition name="Gamma distribution" #GammaFunction}
A random variable \(X\) is said to follow a **Gamma distribution** with shape parameter \(\alpha > 0\) and scale parameter \(\beta > 0\) (denoted \(X \sim \mathrm{Gamma}(\alpha,\beta)\)) if its probability density function is
\[
f_{X}(x; \alpha, \beta)
\;=\;
\begin{cases}
\dfrac{1}{\Gamma(\alpha)\,\beta^{\alpha}}\,x^{\alpha - 1}\,e^{-x/\beta}, 
& x > 0, \\[1em]
0, 
& \text{otherwise},
\end{cases}
\]
where 
\(\Gamma(\alpha) = \displaystyle\int_{0}^{\infty} t^{\alpha - 1} e^{-t}\,dt\)
is the Gamma function.
:::


The Chi-square random variable follows a Chi-square distribution with $k$ degrees of freedom if it is defined as the sum of the squares of $k$ independent standard normal random variables.

::: {.definition name="Chi-square random variable" #ChiSquareRV}
A chi-square random variable with \(k\) degrees of freedom is defined as the sum of the squares of \(k\) independent standard normal random variables.  Formally, if
\[
Z_{1}, Z_{2}, \dots, Z_{k} \;\overset{\mathrm{iid}}{\sim}\; N(0,1),
\]
then
\[
X \;=\;\sum_{i=1}^{k} Z_{i}^{2}
\]
follows a chi-square distribution with \(k\) degrees of freedom, denoted $X \sim \chi^{2}_{k}$ and has the probability density function 
\[
f_{X}(x; k)
\;=\;
\begin{cases}
\dfrac{1}{2^{\,k/2}\,\Gamma\!\bigl(\tfrac{k}{2}\bigr)}\,x^{\,\tfrac{k}{2}-1}\,e^{-x/2}, 
& x > 0, \\[1em]
0, 
& \text{otherwise}.
\end{cases}
\]
Where \(\Gamma(\cdot)\) is the Gamma function in Definition \@ref(def:GammaFunction).
:::

The definition of a chi-square random variable can be proven using moment generating functions.

::: {.remark}
The Chi-square distribution is also a special case of the gamma distribution,
however we will not be covering the gamma distribution in this course.
:::


The shape of the chi-square distribution is determined by its degrees of freedom.
Some examples of chi-square distributions are shown in Figure \@ref(fig:ChiSqExample).

```{r ChiSqExample, fig.cap="Some examples of chi-square distrubution curves with different degrees of freedom", echo=FALSE, fig.align='center', warning=FALSE}
# 1) Define your parameters for chi-square dfs
df1   <- 2; df2   <- 5; df3   <- 10
col1  <- "#00BA38"; lab1 <- expression(chi[2]^2)
col2  <- "#F8766D"; lab2 <- expression(chi[5]^2)
col3  <- "#619CFF"; lab3 <- expression(chi[10]^2)

# 2) Base plot with no default grid or axes
p <- ggplot(data.frame(x = c(0, 30)), aes(x = x)) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid   = element_blank(),
    axis.line    = element_blank(),
    axis.ticks   = element_blank(),
    axis.text    = element_blank(),
    axis.title   = element_text(face = "bold")
  ) +
  labs(x = expression(x), y = "Density") +
  # arrow-tipped axes
  geom_segment(aes(x = 0, y = 0, xend = 30, yend = 0),
               arrow = arrow(length = unit(0.2, "cm")),
               inherit.aes = FALSE) +
  geom_segment(aes(x = 0, y = 0, xend = 0, yend = 0.18),
               arrow = arrow(length = unit(0.2, "cm")),
               inherit.aes = FALSE)

# 3) Shade & draw χ²(df=2)
p <- p +
  stat_function(fun = dchisq, args = list(df = df1),
                geom = "area", fill = col1, alpha = 0.61, n = 801) +
  stat_function(fun = dchisq, args = list(df = df1),
                colour = col1, size = 1.2, n = 801) +
  annotate("text",
           x       = df1,
           y       = dchisq(df1, df1) * 1.05,
           label   = lab1,
           parse   = TRUE,
           colour  = col1,
           fontface= "bold",
           vjust   = -0.75,
           hjust   = -0.5)

# 4) Shade & draw χ²(df=5)
p <- p +
  stat_function(fun = dchisq, args = list(df = df2),
                geom = "area", fill = col2, alpha = 0.61, n = 801) +
  stat_function(fun = dchisq, args = list(df = df2),
                colour = col2, size = 1.2, n = 801) +
  annotate("text",
           x       = df2,
           y       = dchisq(df2, df2) * 1.05,
           label   = lab2,
           parse   = TRUE,
           colour  = col2,
           fontface= "bold",
           vjust   = -0.75,
           hjust   = 0.5)

# 5) Shade & draw χ²(df=10)
p <- p +
  stat_function(fun = dchisq, args = list(df = df3),
                geom = "area", fill = col3, alpha = 0.61, n = 801) +
  stat_function(fun = dchisq, args = list(df = df3),
                colour = col3, size = 1.2, n = 801) +
  annotate("text",
           x       = df3,
           y       = dchisq(df3, df3) * 1.05,
           label   = lab3,
           parse   = TRUE,
           colour  = col3,
           fontface= "bold",
           vjust   = -0.5,
           hjust   = 0.5)

# 6) Render the plot
p
```

### Introduction to the t-Distribution

### Introduction to the F-Distribution

### Arithmetics of RV's

## Sampling Distributions

## The Central Limit Theorem
