# Foundations of Inference

The aim of statistical inference is to produce estimators of the population parameters
based on a smaller sample and to quantify the accuracy of these estimators in terms of a
probability statement. In other words we are interested in the values of unknown parameters
for a population thus we collect data from a sample and calculate statistics to estimate the
parameters of interest. We then quantify our confidence that the statistic is representative
of the parameter.

Statistical inference is concerned primarily with understanding the
quality of parameter estimates. For example, a classic inferential
question is, "How sure are we that the estimated mean, $\bar{x}$, is
near the true population mean, $\mu$?" While the equations and details
change depending on the setting, the foundations for inference are the
same throughout all of statistics. 

## Some Important Statistical Distributions {#sec:inference}

### Introduction to the Normal Distribution {#sec:IntroNormalDist}

In probability theory and statistics, the normal distribution is one of the most
important distribution used to model many populations and processes which occur in reality.

::: {.definition name="Normal distribution" #NormalDist}
Let $X$ be a random variable with mean $-\infty < \mu < +\infty$ and variance $\sigma^2 > 0$.
We say that $X$ folowes a normal distribution if the probability density function of $X$ is:
$$f(x) = \frac{1}{\sqrt{2 \pi \sigma^2}} \cdot e^{-\displaystyle\frac{(x - \mu)^2}{2\sigma^2}}, \text{ for $-\infty < x < \infty$.}$$
:::

::: {.remark}
The normal distribution is also called Gaussian distribution since it was 
discovered Johann Carl Friedrich Gauss in 1809. 
Gauss is one of the most influential and important mathematicians in history.
:::

The normal distribution is symmetric about its mean $\mu$ and has a bell-shaped curve.
It is completely characterized by two parameters which are the the mean $\mu$ and the standard deviation $\sigma$.
Some examples are shown in Figure \@ref(fig:NormalExample).

```{r NormalExample, fig.cap="Some examples of normal distrubution curves with different means and variances", echo=FALSE, fig.align='center', warning=FALSE}
library(ggplot2)
library(grid)

# 1) Define your parameters
mu1  <- -2; sd1  <- 0.5; col1 <- "#00BA38"; lab1 <- "N(-2, 0.5²)"
mu2  <-  0; sd2  <- 1.0; col2 <- "#F8766D"; lab2 <- "N(0, 1²)"
mu3  <-  3; sd3  <- 1.5; col3 <- "#619CFF"; lab3 <- "N(3, 1.5²)"

# 2) Base plot with no default axis lines or grid
p <- ggplot(data.frame(x = c(-6, 10)), aes(x = x)) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid      = element_blank(),
    axis.line       = element_blank(),
    axis.ticks      = element_blank(),
    axis.text       = element_blank(),
    axis.title      = element_text(face = "bold")
  ) +
  labs(x = "x", y = "Density")

# 3) Add arrow‐tipped axes via geom_segment
p <- p +
  # x-axis from x=-6→10
  geom_segment(
    aes(x = -6, y = 0, xend = 10, yend = 0),
    arrow = arrow(length = unit(0.2, "cm")), inherit.aes = FALSE
  ) +
  # y-axis from y=0→max density (~0.8)
  geom_segment(
    aes(x = -6, y = 0, xend = -6, yend = 0.85),
    arrow = arrow(length = unit(0.2, "cm")), inherit.aes = FALSE
  )

# 4) Shade area and draw first curve + label
p <- p +
  stat_function(
    fun       = dnorm,
    args      = list(mean = mu1, sd = sd1),
    geom      = "area",
    fill      = col1,
    alpha     = 0.61,
    n         = 801,
    inherit.aes = TRUE
  ) +
  stat_function(
    fun       = dnorm,
    args      = list(mean = mu1, sd = sd1),
    colour    = col1,
    size      = 1.2,
    n         = 801,
    inherit.aes = TRUE
  ) +
  annotate(
    "text",
    x        = mu1,
    y        = dnorm(mu1, mu1, sd1) * 1.05,
    label    = lab1,
    colour   = col1,
    fontface = "bold",
    hjust    = 0.5
  )

# 5) Shade area and draw second curve + label
p <- p +
  stat_function(
    fun       = dnorm,
    args      = list(mean = mu2, sd = sd2),
    geom      = "area",
    fill      = col2,
    alpha     = 0.61,
    n         = 801
  ) +
  stat_function(
    fun       = dnorm,
    args      = list(mean = mu2, sd = sd2),
    colour    = col2,
    size      = 1.2,
    n         = 801
  ) +
  annotate(
    "text",
    x        = mu2,
    y        = dnorm(mu2, mu2, sd2) * 1.05,
    label    = lab2,
    colour   = col2,
    fontface = "bold",
    hjust    = 0.50,
    vjust    = -0.15
  )

# 6) Shade area and draw third curve + label
p <- p +
  stat_function(
    fun       = dnorm,
    args      = list(mean = mu3, sd = sd3),
    geom      = "area",
    fill      = col3,
    alpha     = 0.61,
    n         = 801
  ) +
  stat_function(
    fun       = dnorm,
    args      = list(mean = mu3, sd = sd3),
    colour    = col3,
    size      = 1.2,
    n         = 801
  ) +
  annotate(
    "text",
    x        = mu3,
    y        = dnorm(mu3, mu3, sd3) * 1.05,
    label    = lab3,
    colour   = col3,
    fontface = "bold",
    hjust    = 0.5,
    vjust    = -0.25
  )

# 7) Render the final plot
p
```

A special case of normal distribution is standard normal distribution which is explained in Definition \@ref(def:StandardNormalDist).

::: {.definition name="Standard normal distribution" #StandardNormalDist}
Let $Z \sim N( \mu = 0, \sigma^2 = 1)$, then $Z$ has
probability density function as:
$$f(y) = \frac{1}{\sqrt{2\pi}} \cdot e^{-\displaystyle\frac{z^2}{2}}.$$
which is called the standard normal distribution.
:::

::: {.remark}
Standard normal random variables are often denoted by $Z$.
:::


A useful property of the normal distribution is that it is symmetric about its mean $\mu$.

::: {.remark}
The Empirical Rule informs us that for any symmetric (bell-shaped) curve, let $\mu$ be its mean and
$\sigma$ be its standard deviation, the following probability set
function is true:

- $1.$ $P(\mu - \sigma < X < \mu + \sigma) = 68.27\%;$

- $2.$ $P(\mu - 2\sigma < X < \mu + 2\sigma) = 95.45\%;$

- $3.$ $P(\mu - 3\sigma < X < \mu + 3\sigma) = 99.73\%.$
:::




<!--
```{r EmpiricalRuleExample, fig.cap="An illustration of the empirical rule.", echo=FALSE, fig.align='center', warning=FALSE}
library(plotly)
library(plotly)

# Standard normal distribution
x <- seq(-4, 4, length.out = 1000)
y <- dnorm(x)

# Shaded segments under curve
segments <- list(
  list(start = -4, end = -3, color = "#F066EA"),
  list(start = -3, end = -2, color = "#7997FF"),
  list(start = -2, end = -1, color = "#00B0F6"),
  list(start = -1, end = 1,  color = "#00C0B8"),
  list(start = 1,  end = 2,  color = "#00B0F6"),
  list(start = 2,  end = 3,  color = "#7997FF"),
  list(start = 3,  end = 4,  color = "#F066EA")
)

# Base plot
p <- plot_ly(showlegend = FALSE) %>%
  add_trace(
    x = x,
    y = y,
    type = "scatter",
    mode = "lines",
    line = list(color = "black")
  )

# Add filled segments
for (s in segments) {
  xs <- seq(s$start, s$end, length.out = 200)
  ys <- dnorm(xs)
  p <- add_trace(
    p,
    x = c(xs, rev(xs)),
    y = c(ys, rep(0, length(ys))),
    type = "scatter",
    mode = "none",
    fill = "toself",
    fillcolor = s$color,
    hoverinfo = "skip"
  )
}

# Vertical dashed lines ending at the curve
segment_bounds <- c(-3, -2, -1, 0, 1, 2, 3)
dashed_lines <- lapply(segment_bounds, function(xpos) {
  list(
    type = "line",
    x0 = xpos, x1 = xpos,
    y0 = 0, y1 = dnorm(xpos),
    xref = "x", yref = "y",
    line = list(color = "black", dash = "dash", width = 1)
  )
})

# Annotations
annotations <- list(
  list(x = -2.35, y = 0.01, text = "0.15%", font = list(size = 12, color = "black"), showarrow = FALSE),
  list(x = -1.5, y = 0.05, text = "13.5%", font = list(size = 12, color = "black"), showarrow = FALSE),
  list(x =  0,   y = 0.1, text = "<b>68%</b>", font = list(size = 16, color = "black"), showarrow = FALSE),
  list(x =  1.5, y = 0.05, text = "13.5%", font = list(size = 12, color = "black"), showarrow = FALSE),
  list(x =  2.35, y = 0.01, text = "0.15%", font = list(size = 12, color = "black"), showarrow = FALSE),
  list(x = -3, y = -0.01, text = "-3σ", font = list(size = 12), showarrow = FALSE),
  list(x = -2, y = -0.01, text = "-2σ", font = list(size = 12), showarrow = FALSE),
  list(x = -1, y = -0.01, text = "-1σ", font = list(size = 12), showarrow = FALSE),
  list(x =  0, y = -0.01, text = "0", font = list(size = 12), showarrow = FALSE),
  list(x =  1, y = -0.01, text = "1σ", font = list(size = 12), showarrow = FALSE),
  list(x =  2, y = -0.01, text = "2σ", font = list(size = 12), showarrow = FALSE),
  list(x =  3, y = -0.01, text = "3σ", font = list(size = 12), showarrow = FALSE),
  list(x = 0, y = max(y) + 0.01, text = "f(x)", font = list(size = 14), showarrow = FALSE)
)

# Final layout with lowered title
p <- layout(
  p,
  title = list(
    text = "Empirical Rule: 68%–95%–99.7%",
    y = 0.90
  ),
  xaxis = list(
    title = NULL,
    showticklabels = FALSE,
    showgrid = FALSE,
    zeroline = FALSE
  ),
  yaxis = list(
    title = NULL,
    showticklabels = FALSE,
    showgrid = FALSE,
    zeroline = FALSE,
    range = c(-0.02, max(y) + 0.10)
  ),
  shapes = dashed_lines,
  annotations = annotations
)

# Interactivity options
config(p, displayModeBar = TRUE)
```
-->


The interactive application below illustrates the empirical rule.


<center>
```{r HistEmpRuleApp, echo=FALSE, out.width='100%'}
knitr::include_app("https://nishan-mudalige.shinyapps.io/Empirical_Rule_Shiny_App/", height = "950")
```
</center>




::: {.remark}
A consequence of the Empirical Rule from \@ref(sec:IntroNormalDist) is that for a sample drawn from a population that is 
approximately normal, we can roughly estimate the standard deviation using the _crude_ approximate :

$$\hat{\sigma} \approx \frac{\text{Sample Range}}{4}.$$
:::



### Introduction to the Chi-Square distribution

Before we introduce the chi-Square square distribution, we recall the gamma distribution.

::: {.definition name="Gamma distribution" #GammaFunction}
A random variable \(X\) is said to follow a Gamma distribution with shape parameter \(\alpha > 0\) and scale parameter \(\beta > 0\) (denoted \(X \sim \mathrm{Gamma}(\alpha,\beta)\)) if its probability density function is
\[
f_{X}(x; \alpha, \beta)
\;=\;
\begin{cases}
\dfrac{1}{\Gamma(\alpha)\,\beta^{\alpha}}\,x^{\alpha - 1}\,e^{-x/\beta}, 
& x > 0, \\[1em]
0, 
& \text{otherwise},
\end{cases}
\]
where 
\(\Gamma(\alpha) = \displaystyle\int_{0}^{\infty} t^{\alpha - 1} e^{-t}\,dt\)
is the Gamma function.
:::


The Chi-square random variable follows a Chi-square distribution with $k$ degrees of freedom if it is defined as the sum of the squares of $k$ independent standard normal random variables.

::: {.definition name="Chi-square random variable" #ChiSquareRV}
A chi-square random variable with \(k\) degrees of freedom is defined as the sum of the squares of \(k\) independent standard normal random variables.  Formally, if
\[
Z_{1}, Z_{2}, \dots, Z_{k} \;\overset{\mathrm{iid}}{\sim}\; N(0,1),
\]
then
\[
X \;=\;\sum_{i=1}^{k} Z_{i}^{2}
\]
follows a chi-square distribution with \(k\) degrees of freedom, denoted $X \sim \chi^{2}_{k}$ and has the probability density function 
\[
f_{X}(x; k)
\;=\;
\begin{cases}
\dfrac{1}{2^{\,k/2}\,\Gamma\!\bigl(\tfrac{k}{2}\bigr)}\,x^{\,\tfrac{k}{2}-1}\,e^{-x/2}, 
& x > 0, \\[1em]
0, 
& \text{otherwise}.
\end{cases}
\]
Where \(\Gamma(\cdot)\) is the Gamma function in Definition \@ref(def:GammaFunction).
:::

The definition of a chi-square random variable can be proven using moment generating functions.

::: {.remark}
The Chi-square distribution is also a special case of the gamma distribution,
however we will not be covering the gamma distribution in this course.
:::


The shape of the chi-square distribution is determined by its degrees of freedom.
Some examples of chi-square distributions are shown in Figure \@ref(fig:ChiSqExample).

```{r ChiSqExample, fig.cap="Some examples of chi-square distrubution curves with different degrees of freedom", echo=FALSE, fig.align='center', warning=FALSE}
# 1) Define your parameters for chi-square dfs
df1   <- 2; df2   <- 5; df3   <- 10
col1  <- "#00BA38"; lab1 <- expression(chi[2]^2)
col2  <- "#F8766D"; lab2 <- expression(chi[5]^2)
col3  <- "#619CFF"; lab3 <- expression(chi[10]^2)

# 2) Base plot with no default grid or axes
p <- ggplot(data.frame(x = c(0, 30)), aes(x = x)) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid   = element_blank(),
    axis.line    = element_blank(),
    axis.ticks   = element_blank(),
    axis.text    = element_blank(),
    axis.title   = element_text(face = "bold")
  ) +
  labs(x = expression(x), y = "Density") +
  # arrow-tipped axes
  geom_segment(aes(x = 0, y = 0, xend = 30, yend = 0),
               arrow = arrow(length = unit(0.2, "cm")),
               inherit.aes = FALSE) +
  geom_segment(aes(x = 0, y = 0, xend = 0, yend = 0.55),
               arrow = arrow(length = unit(0.2, "cm")),
               inherit.aes = FALSE)

# 3) Shade & draw χ²(df=2)
p <- p +
  stat_function(fun = dchisq, args = list(df = df1),
                geom = "area", fill = col1, alpha = 0.61, n = 801) +
  stat_function(fun = dchisq, args = list(df = df1),
                colour = col1, size = 1.2, n = 801) +
  annotate("text",
           x       = df1,
           y       = dchisq(df1, df1) * 1.05,
           label   = lab1,
           parse   = TRUE,
           colour  = col1,
           fontface= "bold",
           vjust   = -0.75,
           hjust   = -0.5)

# 4) Shade & draw χ²(df=5)
p <- p +
  stat_function(fun = dchisq, args = list(df = df2),
                geom = "area", fill = col2, alpha = 0.61, n = 801) +
  stat_function(fun = dchisq, args = list(df = df2),
                colour = col2, size = 1.2, n = 801) +
  annotate("text",
           x       = df2,
           y       = dchisq(df2, df2) * 1.05,
           label   = lab2,
           parse   = TRUE,
           colour  = col2,
           fontface= "bold",
           vjust   = -0.75,
           hjust   = 0.5)

# 5) Shade & draw χ²(df=10)
p <- p +
  stat_function(fun = dchisq, args = list(df = df3),
                geom = "area", fill = col3, alpha = 0.61, n = 801) +
  stat_function(fun = dchisq, args = list(df = df3),
                colour = col3, size = 1.2, n = 801) +
  annotate("text",
           x       = df3,
           y       = dchisq(df3, df3) * 1.05,
           label   = lab3,
           parse   = TRUE,
           colour  = col3,
           fontface= "bold",
           vjust   = -0.5,
           hjust   = 0.5)

# 6) Render the plot
p
```

### Introduction to the t-Distribution

::: {.definition name="Student’s \(t\) Distribution" #TDist}
A **Student’s \(t\) random variable** with \(\nu\) degrees of freedom is defined as the ratio of a standard normal variate and the square root of a scaled chi-square variate.  Formally, if
\[
Z \;\overset{\mathrm{iid}}{\sim}\; N(0,1)
\quad\text{and}\quad
U \;\overset{\mathrm{iid}}{\sim}\; \chi^{2}_{\nu}
\]
are independent, then
\[
T \;=\;\frac{Z}{\sqrt{\,U / \nu\,}}
\]
follows a Student’s \(t\) distribution with \(\nu\) degrees of freedom, denoted
\[
T \sim t_{\nu},
\]
and has the probability density function
\[
f_{T}(t; \nu)
\;=\;
\frac{\Gamma\!\bigl(\tfrac{\nu+1}{2}\bigr)}{\sqrt{\pi\,\nu}\,\Gamma\!\bigl(\tfrac{\nu}{2}\bigr)}
\Bigl(1 + \tfrac{t^{2}}{\nu}\Bigr)^{-\tfrac{\nu+1}{2}},
\quad
-\infty < t < \infty.
\]
Here \(\Gamma(\cdot)\) is the Gamma function as in Definition \@ref(def:GammaFunction).
:::

The t-distribution resembles the normal distribution but with heavier tails.

```{r t-densities, echo=FALSE, fig.align='center', fig.cap='The $t-$Distributions at different degrees of freedom.', out.width='90%', warning=FALSE}
library(ggplot2)
library(grid)   # for unit()

## 1) Degrees of freedom, colours, labels ------------------------------------
df1  <-  2;  col1 <- "#00BA38"; lab1 <- expression(t[2])
df2  <-  5;  col2 <- "#F8766D"; lab2 <- expression(t[5])
df3  <- 10;  col3 <- "#619CFF"; lab3 <- expression(t[10])

## 2) Base plot (no default grid or axes) ------------------------------------
p <- ggplot(data.frame(x = c(-6,  6)), aes(x = x)) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid   = element_blank(),
    axis.line    = element_blank(),
    axis.ticks   = element_blank(),
    axis.text    = element_blank(),
    axis.title   = element_text(face = "bold")
  ) +
  labs(x = expression(x), y = "Density") +
  # arrow-tipped axes
  geom_segment(aes(x = -6, y = 0, xend =  6, yend = 0),
               arrow = arrow(length = unit(0.2, "cm")),
               inherit.aes = FALSE) +
  geom_segment(aes(x =  0, y = 0, xend =  0, yend = 0.40),
               arrow = arrow(length = unit(0.2, "cm")),
               inherit.aes = FALSE)

## 3) Shade & draw t(df = 2) --------------------------------------------------
p <- p +
  stat_function(fun = dt, args = list(df = df1),
                geom = "area", fill = col1, alpha = 0.61, n = 801) +
  stat_function(fun = dt, args = list(df = df1),
                colour = col1, size = 1.2, n = 801) +
  annotate("text",
           x       = 0,
           y       = dt(0, df1) * 1.05,
           label   = lab1,      # t[2]
           parse   = TRUE,
           colour  = col1,
           fontface = "bold",
           hjust   = 0.5) +
  annotate("text",
           x       = 2.5,       # a little to the right of the curve
           y       = dt(2.5, df1),
           label   = "df = 2",
           colour  = col1,
           fontface = "bold",
           hjust   = 0)         # left-aligned for readability

## 4) Shade & draw t(df = 5) --------------------------------------------------
p <- p +
  stat_function(fun = dt, args = list(df = df2),
                geom = "area", fill = col2, alpha = 0.34, n = 801) +
  stat_function(fun = dt, args = list(df = df2),
                colour = col2, size = 1.2, n = 801) +
  annotate("text",
           x       = 0,
           y       = dt(0, df2) * 1.05,
           label   = lab2,      # t[5]
           parse   = TRUE,
           colour  = col2,
           fontface = "bold",
           hjust   = 0.5,
           vjust   = -0.35) +
  annotate("text",
           x       = 3,
           y       = dt(3, df2),
           label   = "df = 5",
           colour  = col2,
           fontface = "bold",
           hjust   = 0)

## 5) Shade & draw t(df = 10) -------------------------------------------------
p <- p +
  stat_function(fun = dt, args = list(df = df3),
                geom = "area", fill = col3, alpha = 0.25, n = 801) +
  stat_function(fun = dt, args = list(df = df3),
                colour = col3, size = 1.2, n = 801) +
  annotate("text",
           x       = 0,
           y       = dt(0, df3) * 1.05,
           label   = lab3,      # t[10]
           parse   = TRUE,
           colour  = col3,
           fontface = "bold",
           hjust   = 0.5,
           vjust   = -0.70) +
  annotate("text",
           x       = 3.5,
           y       = dt(3.5, df3),
           label   = "df = 10",
           colour  = col3,
           fontface = "bold",
           hjust   = 0)

## 6) Render the combined plot -----------------------------------------------
print(p)
```

### Introduction to the F-Distribution

::: {.definition name="Beta function" #BetaFunction}
The **Beta function**, also called the Euler integral of the first kind, is defined for complex numbers \(\;z_{1}, z_{2}\) with \(\Re(z_{1}), \Re(z_{2})>0\) by the integral
$$
B(z_{1}, z_{2})
=
\int_{0}^{1} t^{\,z_{1}-1}\,(1 - t)^{\,z_{2}-1}\,\mathrm{d}t.
$$

<!-- It is symmetric in its arguments, i.e. -->
<!-- $$ -->
<!-- B(z_{1}, z_{2}) = B(z_{2}, z_{1}), -->
<!-- $$ -->
<!-- and satisfies the relationship with the Gamma function: -->
<!-- $$ -->
<!-- B(z_{1}, z_{2}) -->
<!-- = -->
<!-- \frac{\Gamma(z_{1})\,\Gamma(z_{2})}{\Gamma(z_{1} + z_{2})}. -->
<!-- $$ -->
:::


::: {.definition name="F random variable" #dist}
An \(F\) random variable with \(\nu_{1}\) and \(\nu_{2}\) degrees of freedom is defined as the ratio of two independent chi-square variates, each divided by its degrees of freedom.  Formally, if  
$$
U \;\overset{\mathrm{iid}}{\sim}\;\chi^{2}_{\nu_{1}}
\quad\text{and}\quad
V \;\overset{\mathrm{iid}}{\sim}\;\chi^{2}_{\nu_{2}}
$$
are independent, then
$$
F \;=\;\frac{\,U / \nu_{1}\,}{\,V / \nu_{2}\,}
$$
follows an \(F\) distribution with \(\nu_{1}\) and \(\nu_{2}\) degrees of freedom, denoted
$$
F \sim F_{\nu_{1},\nu_{2}},
$$
and has probability density function
$$
f_{F}(x; \nu_{1}, \nu_{2})
=
\begin{cases}
\displaystyle
\frac{\bigl(\tfrac{\nu_{1}}{\nu_{2}}\bigr)^{\!\nu_{1}/2}
      \,x^{\,\tfrac{\nu_{1}}{2}-1}}
     {B\!\bigl(\tfrac{\nu_{1}}{2},\,\tfrac{\nu_{2}}{2}\bigr)}
\,
\Bigl(1 + \tfrac{\nu_{1}}{\nu_{2}}\,x\Bigr)^{-\tfrac{\nu_{1}+\nu_{2}}{2}},
& x > 0 \\[1em]
0,
& \text{otherwise}.
\end{cases}
$$
where $B(\cdot,\cdot)$ is the Beta function (see Definition \@ref(def:BetaFunction)).
:::


```{r f-densities, echo=FALSE, fig.align='center', fig.cap='The F-Distribution at different degrees of freedom.', out.width='90%', warning=FALSE}
library(ggplot2)
library(grid)   # for arrow()

# 1) Parameters for the three F distributions
nu1a <-  2; nu2a <-  5; colA <- "#00BA38"; labA <- expression(bold(F[2][", "][5]))
nu1b <-  5; nu2b <- 10; colB <- "#F8766D"; labB <- expression(bold(F[5][", "][10]))
nu1c <- 10; nu2c <- 30; colC <- "#619CFF"; labC <- expression(bold(F[10][", "][30]))

# 2) Common x‐axis values
x_vals <- seq(0, 5, length.out = 1000)

# 3) Compute densities
f_dens_a <- df(x_vals, df1 = nu1a, df2 = nu2a)
f_dens_b <- df(x_vals, df1 = nu1b, df2 = nu2b)
f_dens_c <- df(x_vals, df1 = nu1c, df2 = nu2c)

# 4) Single data frame with grouping
f_data <- data.frame(
  x       = rep(x_vals, 3),
  density = c(f_dens_a, f_dens_b, f_dens_c),
  group   = factor(rep(c("2,5", "5,10", "10,20"), each = length(x_vals)))
)

# 5) Base ggplot
f_plot <- ggplot(f_data, aes(x = x, y = density, group = group)) +
  
  # F(2,5)
  geom_area(data = subset(f_data, group == "2,5"),
            fill = colA, alpha = 0.61) +
  geom_line(data = subset(f_data, group == "2,5"),
            colour = colA, size = 1.2) +
  annotate("text",
           x      = nu1a/(nu2a-2),
           y      = df(nu1a/(nu2a-2), df1 = nu1a, df2 = nu2a)*1.05,
           label  = labA,
           parse  = TRUE,
           colour = colA,
           vjust  = -10, 
           hjust  = 2.0) +
  
  # F(5,10)
  geom_area(data = subset(f_data, group == "5,10"),
            fill = colB, alpha = 0.61) +
  geom_line(data = subset(f_data, group == "5,10"),
            colour = colB, size = 1.2) +
  annotate("text",
           x      = nu1b/(nu2b-2),
           y      = df(nu1b/(nu2b-2), df1 = nu1b, df2 = nu2b)*1.05,
           label  = labB,
           parse  = TRUE,
           colour = colB,
           vjust  = -0.5,
           hjust  = 1.0) +
  
  # F(10,20)
  geom_area(data = subset(f_data, group == "10,20"),
            fill = colC, alpha = 0.61) +
  geom_line(data = subset(f_data, group == "10,20"),
            colour = colC, size = 1.2) +
  annotate("text",
           x      = nu1c/(nu2c-2),
           y      = df(nu1c/(nu2c-2), df1 = nu1c, df2 = nu2c)*1.05,
           label  = labC,
           parse  = TRUE,
           colour = colC,
           hjust  = -3.75) +
  
  # Arrow‐tipped axes
  geom_segment(aes(x = 0, y = 0, xend = 5, yend = 0),
               arrow = arrow(length = unit(0.2, "cm")),
               inherit.aes = FALSE) +
  geom_segment(aes(x = 0, y = 0, xend = 0, yend = max(f_dens_a, f_dens_b, f_dens_c)*1.1),
               arrow = arrow(length = unit(0.2, "cm")),
               inherit.aes = FALSE) +
  
  # Axis labels
  labs(x = "x", y = "Density") +
  
  # Clean theme: only axes, no grid/box/ticks/text
  theme_minimal(base_size = 14) +
  theme(
    panel.grid     = element_blank(),
    panel.border   = element_blank(),
    axis.ticks     = element_blank(),
    axis.text      = element_blank(),
    axis.title.x   = element_text(face = "plain"),
    axis.title.y   = element_text(face = "bold"),
    axis.line      = element_blank()
  )

# 6) Render
print(f_plot)
```



<!-- Both student's t-distribution and F-distribution are highly used in -->
<!-- inferential statistics, until confidence interval, testing hypothesis -->
<!-- and ANOVA analysis, these two distributions will come to play a lot. At -->
<!-- this point, just guarantee that you know how to obtain those -->
<!-- distribution from random given information is sufficient. -->


### Arithmetics of RV's

**To be completed**

## Sampling Distributions

Although this may be a new concept, a statistic can follow a distribution. The distribution of a statistic is called its sampling distribution.

::: {.definition name="Sampling distribution" #SamplingDist}
The sampling distribution of a statistic is the probability distribution of that statistic computed from all possible random samples of a given size drawn from the same population.
:::

In other words, Definition \@ref(def:SamplingDist) in informing us that if we repeatedly draw samples of size \(n\) and compute the statistic (e.g.\ the sample mean \(\bar X\)), then the values of \(\bar X\) across those samples form a distribution, which is its sampling distribution.
The sampling distribution is important because it allows us to understand the variability of a statistic from sample to sample, and it forms the basis for inferential statistics, such as confidence intervals and hypothesis tests.




## The Central Limit Theorem

The central limit theorem (CLT) explains why the distribution of a properly
standardized sample mean becomes approximately normal as the sample size
grows—even when the underlying population is not normal.  

Suppose \(X_1,X_2,\dots,X_n\) are independent and identically distributed
(i.i.d.) with finite mean \(\mu\) and variance \(\sigma^{2}\). Define the sample mean \(\bar X_n = \tfrac{1}{n}\sum_{i=1}^{n} X_i\). The Centrel Limit Theorem tells us that the random fluctuation of \(\bar X_n\) around the
population mean \(\mu\) (after appropriate scaling) behaves more and more like
a standard normal variable as \(n\) increases.

::: {.theorem name="Central Limit Theorem" #CLT}
Let \(X_{1},X_{2},\dots,X_{n}\) be i.i.d.\ random variables with  
\(E[X_i]=\mu\) and \(\operatorname{Var}(X_i)=\sigma^{2}<\infty\).
Then, as \(n\to\infty\),

$$
\frac{\sqrt{n}\,(\bar X_{n}-\mu)}{\sigma}
\;\overset{d}{\longrightarrow}\;
N(0,1).
$$

<!-- i.e.\ for every real \(z\), -->

<!-- $$ -->
<!-- \lim_{n\to\infty} -->
<!-- P\!\Bigl(\frac{\sqrt{n}(\bar X_{n}-\mu)}{\sigma}\le z\Bigr) -->
<!-- = -->
<!-- \Phi(z), -->
<!-- $$ -->
<!-- where \(\Phi(\cdot)\) is the standard normal cumulative distribution function. -->
:::

A practical consequence is that for large \(n\) we can use normal‐based
confidence intervals and hypothesis tests for \(\mu\) even when the parent
distribution of the data is unknown or skewed; the accuracy of this normal
approximation improves as the sample size grows.





## Law of Large Numbers

The law of large numbers (LLN) says that sample averages settle down near their common expected value as the sample size grows. This idea is an foundational component statistical inference with for large samples.

### Convergence in Probability

::: definition
The sequence of random variables $X_1, X_2, \ldots$ is
said to converge in probability to the constant $c$, if for every
$\varepsilon > 0$,
$$\lim_{n \to \infty} P\left( |X_n - c| \leq \varepsilon \right) = 1$$ or
equivalently,
$$\lim_{n \to \infty} P\left( |X_n - c| > \varepsilon \right) = 0$$
:::

::: {.remark}
If a sequence of random variables $X_n = \{X_1, X_2, \ldots \}$ converges in probability to $c$, we denote this as $X_n \overset{p}{\longrightarrow} c$.
:::

This concept plays a key role in the Law of Large Numbers, where the
sample mean of independent and identically distributed random variables
converges in probability to the population mean as the sample size
grows.

Recall Chebyshev's Inequality, which provides a bound on the probability that a random variable deviates from its mean.

::: {.theorem Name="Chebyshev's Inequality" #Chebyshev}
Let $X$ be a random variable with finite mean $\mu$ and variance
$\sigma^2$. Then, for any $k > 0$,
$$P\left( |X - \mu| \geq k \right) \leq \frac{\sigma^2}{k^2}$$

*Using complements:*
$$P\left( |X - \mu| < k \right) \geq 1 - \frac{\sigma^2}{k^2}$$
:::

### Weak Law of Large Numbers (WLLN)

::: definition
Let $X_1, X_2, \ldots$ be a sequence of independent and identically
distributed random variables, each having finite mean $E(X_i) = \mu$ and
variance $\mathrm{Var}(X_i) = \sigma^2$. Then, for any $\varepsilon > 0$,
$$P\left( \left| \frac{X_1 + X_2 + \cdots + X_n}{n} - \mu \right| \geq \varepsilon \right) \longrightarrow 0 \quad \text{as } n \longrightarrow \infty.$$

This is denoted as $\bar{X}_n \overset{p}{\longrightarrow} \mu$
:::

#### Proof of the Weak Law of Large Numbers (WLLN) {#proof-of-the-weak-law-of-large-numbers-wlln .unnumbered}

<!-- First recall Chebchev's inequality -->

<!-- ::: {.theorem} -->
<!-- Chebyshev's Inequality states that for any random variable $X$ with mean $\mu$ and variance $\sigma^2$, -->
<!-- $$P\left( |X - \mu| > k \right) \leq \frac{\sigma^2}{k^2} \quad \text{for } k > 0.$$ -->
<!-- ::: -->

We aim to show that for every $\varepsilon > 0$,
$$\lim_{n \to \infty} P\left( \left| \bar{X}_n - \mu \right| > \varepsilon \right) = 0$$
where $\bar{X}_n$ is the sample mean of $n$ independent and identically
distributed (i.i.d.) random variables with
$$E(X_i) = \mu, \quad \text{and} \quad \mathrm{Var}(X_i) = \sigma^2.$$

Let $$\bar{X}_n = \frac{1}{n} \sum_{i=1}^{n} X_i.$$

By the Central Limit Theorem (CLT), we know that
$$\bar{X}_n \sim \mathcal{N} \left( \mu, \frac{\sigma^2}{n} \right).$$

Now, applying Chebyshev's Inequality from Theorem \@ref(thm:Chebyshev) to $\bar{X}_n$, we set $k = \varepsilon$, and obtain:
$$P\left( \left| \bar{X}_n - \mu \right| > \varepsilon \right) \leq \frac{\mathrm{Var}(\bar{X}_n)}{\varepsilon^2} = \frac{\sigma^2 / n}{\varepsilon^2} = \frac{\sigma^2}{n \varepsilon^2}.$$

Taking the limit as $n \to \infty$, we have:
$$\lim_{n \to \infty} P\left( \left| \bar{X}_n - \mu \right| > \varepsilon \right) \leq \lim_{n \to \infty} \frac{\sigma^2}{n \varepsilon^2} = 0.$$

Since probabilities are always non-negative, we conclude:
$$\lim_{n \to \infty} P\left( \left| \bar{X}_n - \mu \right| > \varepsilon \right) = 0.$$

By the definition of convergence in probability,
$$\bar{X}_n \overset{p}{\longrightarrow} \mu.$$

::: example
Let $X_i$, for $i = 1, 2, 3, \ldots$, be independent Poisson random
variables with rate parameter $\lambda = 3$. Prove that:
$$\bar{X}_n \xrightarrow{P} 3$$

**Properties of Poisson Distribution:**
$$E(X_i) = \lambda, \quad \mathrm{Var}(X_i) = \lambda$$ In this case,
$\lambda = 3$, so: $$E(X_i) = \mathrm{Var}(X_i) = 3$$

**Proof:**

We know:
$$E\left( \frac{X_1 + X_2 + \cdots + X_n}{n} \right) = 3, \quad \text{and} \quad
\mathrm{Var}\left( \frac{X_1 + X_2 + \cdots + X_n}{n} \right) = \frac{3}{n}$$

Applying Chebyshev's Inequality:
$$P\left( \left| \frac{X_1 + X_2 + \cdots + X_n}{n} - 3 \right| \geq \varepsilon \right) \leq \frac{3}{n \varepsilon^2}$$

Taking the limit as $n \to \infty$:
$$P\left( \left| \frac{X_1 + X_2 + \cdots + X_n}{n} - 3 \right| \geq \varepsilon \right) \to 0$$

**Conclusion:** $$\bar{X}_n \xrightarrow{P} 3$$
:::

```{r fig.cap="Simulation of running sample mean of Bernoulli \\(p = 0.5\\) trials over time", echo=FALSE, fig.align='center'}
library(ggplot2)

set.seed(123)
n  <- 10; p  <- 0.5
df <- data.frame(
  trial        = 1:n,
  running_mean = cumsum(rbinom(n, 1, p)) / (1:n)
)

# Pick two lovely colors from ColorBrewer Set2
line_col  <- "#619CFF"
point_col <- "#F8766D"

ggplot(df, aes(x = trial, y = running_mean)) +
  geom_hline(yintercept = p, linetype = "dashed", colour = "grey50") +
  geom_line(colour = line_col, size = 1.2) +
  geom_point(colour = point_col, size = 3) +
  scale_x_continuous(breaks = 1:n) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "trial",
    y = "(Cumulative sum of sample)/(Trial number)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major = element_line(colour = "grey90"),
    panel.grid.minor = element_blank(),
    axis.ticks       = element_line(colour = "grey80"),
    axis.text        = element_text(colour = "grey30")
  )
```

**R Simulation Code (Single Sample Path):**

::: {.example}
    n = 10
    trial = seq(1, n, by = 1)
    sample = rbinom(n, 1, 1/2)

    plot(trial, cumsum(sample)/trial, type = "l", ylim = c(0,1), col = "blue")
    points(trial, cumsum(sample)/trial, col = "red")
    abline(h = 0.5, lty = 2, col = "black")
:::

```{r fig.cap="A Simulation of 20 running sample means of Bernoulli \\(p = 0.5\\) trials converging over 100 trials", echo=FALSE, fig.align='center'}
library(ggplot2)
library(tidyr)

set.seed(123)

# 1) Simulate eight independent Bernoulli(½) streams and compute running means
n       <- 100
p       <- 0.5
trial   <- 1:n
samples <- replicate(8, rbinom(n, 1, p))
runs    <- apply(samples, 2, function(x) cumsum(x) / trial)

# 2) Pivot to long format
df        <- as.data.frame(runs)
df$trial  <- trial
df_long   <- pivot_longer(
  df,
  cols      = starts_with("V"),
  names_to  = "series",
  values_to = "running_mean"
)

# 3) Plot with thinner, lighter lines and no legend
ggplot(df_long, aes(x = trial, y = running_mean, color = series)) +
  geom_line(size = 0.75, alpha = 0.61) +             # thinner and semi-transparent :contentReference[oaicite:0]{index=0}
  geom_hline(yintercept = p, linetype = "dashed", color = "black") +
  scale_x_continuous(breaks = seq(0, n, by = 10)) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x     = "Trial",
    y     = "Cumulative Proportion of Ones"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey90"),
    axis.ticks       = element_line(color = "grey80"),
    legend.position  = "none"
  )
```

**R Simulation Code (Multiple Sample Paths):**

::: {.example}
    n = 100
    trial = seq(1, 100, by = 1)

    sample1 = rbinom(n, 1, 1/2)
    sample2 = rbinom(n, 1, 1/2)
    sample3 = rbinom(n, 1, 1/2)
    sample4 = rbinom(n, 1, 1/2)
    sample5 = rbinom(n, 1, 1/2)
    sample6 = rbinom(n, 1, 1/2)
    sample7 = rbinom(n, 1, 1/2)
    sample8 = rbinom(n, 1, 1/2)

    colors = rainbow(8)


    plot(trial, cumsum(sample1)/trial, type = "l", col = colors[1], ylim = c(0,1))
    lines(trial, cumsum(sample2)/trial, col = colors[2])
    lines(trial, cumsum(sample3)/trial, col = colors[3])
    lines(trial, cumsum(sample4)/trial, col = colors[4])
    lines(trial, cumsum(sample5)/trial, col = colors[5])
    lines(trial, cumsum(sample6)/trial, col = colors[6])
    lines(trial, cumsum(sample7)/trial, col = colors[7])
    lines(trial, cumsum(sample8)/trial, col = colors[8])
    abline(h = 0.5, lty = 2, col = "black")
:::

### Empirical Probability Insight {.unlisted .unnumbered #empirical-probability-insight}

The Law of Large Numbers gives us empirical probabilities. Consider
tossing a fair coin. Define the random variable $X$ as:

$$X = \begin{cases}
1 & \text{heads up} \\
0 & \text{tails up}
\end{cases}$$

Then as we sample more and more values of $X$, the sample mean
$\bar{X}_n$ converges in probability to $P(\text{heads up})$, that is:

$$\bar{X}_n \overset{p}{\longrightarrow} P(\text{heads up}).$$

