# Hypothesis Tests

## Introduction 

A hypothesis test is a statistical method used to make inferences about a population based on sample data. It involves formulating a null hypothesis and an alternative hypothesis, calculating a test statistic, and determining the likelihood of observing the data under the null hypothesis. Each of these terms will be explained in more detail below.

::: {.definition}
An inferential procedure to determine whether there is sufficient
evidence to suggest a condition for a population parameter using
statistics from a sample.
:::

The main goal of hypothesis testing is to assess the strength of evidence against a null hypothesis, which is a statement of no effect or no difference. The alternative hypothesis represents the research hypothesis or the condition we are interested in testing.

One of the steps in hypothesis testing is to calculate a p-value, which quantifies the evidence against the null hypothesis. The p-value is the probability of observing a test statistic at least as extreme as the one calculated from the sample data, assuming that the null hypothesis is true.

::: {.definition}
The p-value of a hypothesis test is the probability of observing a test statistic at least as extreme as the one calculated in step 3, assuming that the null hypothesis is true by pure chance alone.
:::


The procedure to conduct any hypothesis often follows a standard sequence of steps.

##### Steps {.unnumbered .unlisted}

**[ [Reorder and start with checking assumptions first?] ]{style="color: red"}**


1.  Decide on a level of significance ($\alpha$).

2.  State the null hypothesis ($H_0$) and the alternative hypothesis ($H_a$).
    
3.  Calculate the appropriate test statistic.

4.  Use the test statistic and a reference distribution to calculate a
    p-value.

5.  Compare p-value to $\alpha$ to make a conclusion.

We will now examine each of the steps above in more detail.

**Step 1: Decide on a Level of Significance ($\alpha$)**

This is a threshold for decision making which the study or researcher is willing to implement.
Depends on tolerance for consequences of errors, sample size, nature of
the study, and variability.

Some common Common values are $\alpha = 0.10$, $\alpha = 0.05$, $\alpha = 0.01$.


**Step 2: State the Null Hypothesis and the Alternative Hypothesis**

We will use the following notation for describing the null and alternative hypotheses:

<table class="param-table">
  <tr>
    <pd class="term">$\Theta$</td>
    <pd class="sep"> : </pd>
    <pd class="alt">Parameter of interest.</pd>
  </tr>
</table>

<table class="param-table">
  <tr>
    <pd class="term">$\Theta_0$</td>
    <pd class="sep"> : </pd>
    <pd class="alt">Numerical value hypothesized under the null hypothesis.</pd>
  </tr>
</table>

::: {.remark}
Note that $\Theta$ can represent a combination of parameters, 
such as $\Theta = \Theta_1 - \Theta_2$ 
or $\Theta = \Theta_1 / \Theta_2$.
:::

<!-- <table class="param-table"> -->
<!--   <tr> -->
<!--     <pd class="term">$\Theta_0$</pd> -->
<!--     <pd class="sep"> : </pd> -->
<!--     <pd class="alt">Numerical value of the parameter of interest hypothesized under the null hypothesis.</pd> -->
<!--   </tr> -->
<!-- </table> -->

<!-- $$\begin{array}{lcl} -->
<!-- \Theta    & : & \text{parameter of interest}\\ -->
<!-- \Theta_0  & : & \text{numerical value of the parameter of interest hypothesized under the}\\ -->
<!--           &   & \text{null hypothesis.} -->
<!-- \end{array}$$ -->

A hypothesis test is a comparison between two mutually exclusive hypotheses which are:



<table class="definition-table">
  <tr>
    <td>Null hypothesis $(H_0)$</td>
    <td>: </td>
    <td>Represents the current belief or the conservative belief or the status quo. It is a statement of no effect or no difference.</td>
  </tr>
  <tr>
    <td>Alternative hypothesis $(H_a)$</td>
    <td>: </td>
    <td>Represents the research hypothesis, or what you are requested to test or trying to prove.</td>
  </tr>
</table>




<!-- ::: {.smalltblfit} -->
<!-- | Null hypothesis        | Represents the current belief or the conservative Represents the current belief or the conservative Represents the current belief or the conservative Represents the current belief or the conservative Represents the current belief or the conservative |  -->
<!-- | :--------------------- | :----------   |  -->
<!-- | Alternative hypothesis | $x_1$ | -->
<!-- ::: -->



<!-- $$\begin{array}{lcl} -->
<!-- \text{Null hypothesis} \, (H_0)         & : & \text{Represents the current belief or the conservative}\\ -->
<!--                                         &   & \text{belief or the status quo. It is a statement of no effect}\\ -->
<!--                                         &   & \text{or no difference.}\\ -->
<!-- \text{Alternative hypothesis} \, (H_a)  & : & \text{Represents the research hypothesis, or what you are}\\ -->
<!--                                         &   & \text{requested to test or trying to prove).} -->
<!-- \end{array}$$ -->

Our choices for the null and alternative hypothesis are:

<!-- $$\begin{array}{lccl} -->
<!-- H_0: \Theta = \Theta_0 & (\Theta \leq \Theta_0) & \quad & H_a: \Theta > \Theta_0 \quad {\text{one-sided (one-tailed)}} \\ -->
<!-- H_0: \Theta = \Theta_0 & (\Theta \geq \Theta_0) &       & H_a: \Theta < \Theta_0 \quad {\text{one-sided (one-tailed)}} \\ -->
<!-- H_0: \Theta = \Theta_0 &                        &       & H_a: \Theta \neq \Theta_0 \quad {\text{two-sided (two-tailed)}} -->
<!-- \end{array}$$ -->

<!-- $$\begin{array}{l lcl} -->
<!-- 1. &  H_0: \Theta = \Theta_0 & \quad  & H_a: \Theta > \Theta_0\\ -->
<!-- 2. &  H_0: \Theta = \Theta_0 &        & H_a: \Theta < \Theta_0\\ -->
<!-- 3. &  H_0: \Theta = \Theta_0 &        & H_a: \Theta \neq \Theta_0 -->
<!-- \end{array}$$ -->

<!-- - $H_0: \Theta = \Theta_0$ vs. $H_a: \Theta > \Theta_0$ -->
<!-- - $H_0: \Theta = \Theta_0$ vs. $H_a: \Theta < \Theta_0$ -->
<!-- - $H_0: \Theta = \Theta_0$ vs. $H_a: \Theta \neq \Theta$ -->


<table class="hypothesis-table">
  <tr>
    <td class="term">$H_0: \Theta = \Theta_0$</td>
    <td class="sep">vs.</td>
    <td class="alt">$H_a: \Theta > \Theta_0$</td>
  </tr>
  <tr>
    <td class="term">$H_0: \Theta = \Theta_0$</td>
    <td class="sep">vs.</td>
    <td class="alt">$H_a: \Theta < \Theta_0$</td>
  </tr>
  <tr>
    <td class="term">$H_0: \Theta = \Theta_0$</td>
    <td class="sep">vs.</td>
    <td class="alt">$H_a: \Theta \neq \Theta_0$</td>
  </tr>
</table>





:::{.remark}
The first two hypothesis tests above are referred to as one-sided or one-tailed tests.
and the third hypothesis test is referred to as a two-sided or two-tailed test.
:::

:::{.remark}
The one-sided hypothesis test 
$$\begin{array}{lcl}
H_0: \Theta = \Theta_0 & \quad  & H_a: \Theta > \Theta_0\\
\end{array}$$
is equivalent to
$$\begin{array}{lcl}
H_0: \Theta \leq \Theta_0 & \quad  & H_a: \Theta > \Theta_0\\
\end{array}$$

and the other one-sided hypothesis test 

$$\begin{array}{lcl}
H_0: \Theta = \Theta_0 & \quad  & H_a: \Theta < \Theta_0\\
\end{array}$$
is equivalent to
$$\begin{array}{lcl}
H_0: \Theta \geq \Theta_0 & \quad  & H_a: \Theta < \Theta_0\\
\end{array}$$
:::


**Step 3: Calculate an appropriate test statistic**

Depends on the hypothesis test conducted and the information available.

::: definition
$$\text{test statistic} = \frac{\text{(a statistic)} - \text{(hypothesized value of parameters under } H_0 \text{)}}{\text{standard error of statistic}}$$
:::

The test statistic follows a reference distribution. 
Common distributions which thetest statistic follows are the standard normal distrubution, $t$-distribution, 
$F$-distribution and $\chi^2$-distribution.
<!-- $Z$, $t$, $F$, $\chi^2$ -->

**Step 4: Calculate the p-value**

<!-- It quantifies the evidence against the null hypothesis. -->

We use the test statistic, reference distribution, and refer back to $H_a$.

```{r fig.cap="Right-tailed test: p-value is the area to the right of the test statistic", fig.align='center', fig.show='hold', out.width='48%', echo=FALSE}

library(ggplot2)

# --- First Plot: Standard blue bracket lower ---
x <- seq(-4, 4, length = 1000)
y <- dnorm(x)
test_stat <- 1.5
p_x <- seq(test_stat, 4, length = 500)
p_y <- dnorm(p_x)
x_left <- test_stat + 0.1
x_right <- 3.4
y_top <- 0.2
y_mid <- 0.185
y_bot <- 0.17

p1 <- ggplot() +
  geom_line(aes(x, y), color = "#000000", size = 1.2) +
  geom_area(aes(x = p_x, y = p_y), fill = "#619CFF", alpha = 0.6) +
  geom_hline(yintercept = 0, color = "#000000", linewidth = 1) +
  geom_segment(aes(x = test_stat, xend = test_stat, y = 0, yend = dnorm(test_stat)),
               color = "#000000", linewidth = 1.3) +
  geom_segment(aes(x = x_left, xend = x_left, y = y_bot, yend = y_mid), color = "#000000") +
  geom_segment(aes(x = x_left, xend = x_right, y = y_mid, yend = y_mid), color = "#000000") +
  geom_segment(aes(x = x_right, xend = x_right, y = y_mid, yend = y_bot), color = "#000000") +
  annotate("text", x = (x_left + x_right)/2, y = y_top, label = "p-value", size = 5) +
  annotate("text", x = -2.7, y = 0.35, label = "H[0]:~theta==theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = -2.7, y = 0.29, label = "H[a]:~theta > theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = test_stat, y = 0.005, label = "test stat", size = 4.5, vjust = 1.5) +
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank())

print(p1)

# --- Second Plot: Blue with widest shading and highest bracket ---
y <- dnorm(x)
test_stat <- -1.75
p_x <- seq(test_stat, 4, length = 500)
p_y <- dnorm(p_x)
x_left <- test_stat + 0.3
x_right <- 3.4
y_top <- 0.45
y_mid <- 0.435
y_bot <- 0.42

p2 <- ggplot() +
  geom_line(aes(x, y), color = "#000000" , size = 1.2) +
  geom_area(aes(x = p_x, y = p_y), fill = "#619CFF", alpha = 0.6) +
  geom_hline(yintercept = 0, color = "#000000", linewidth = 1) +
  geom_segment(aes(x = test_stat, xend = test_stat, y = 0, yend = dnorm(test_stat)),
               color = "#000000", linewidth = 1.3) +
  geom_segment(aes(x = x_left, xend = x_left, y = y_bot, yend = y_mid), color = "#000000") +
  geom_segment(aes(x = x_left, xend = x_right, y = y_mid, yend = y_mid), color = "#000000") +
  geom_segment(aes(x = x_right, xend = x_right, y = y_mid, yend = y_bot), color = "#000000") +
  annotate("text", x = (x_left + x_right)/2, y = y_top, label = "p-value", size = 5) +
  annotate("text", x = -2.7, y = 0.35, label = "H[0]:~theta==theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = -2.7, y = 0.29, label = "H[a]:~theta > theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = test_stat, y = 0.005, label = "test stat", size = 4.5, vjust = 1.5) +
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank())

print(p2)
```

```{r fig.cap="Left-tailed test: p-value is the area to the left of the test statistic", fig.align='center', fig.show='hold', out.width='48%', echo=FALSE}

# First plot (lower pink bracket)
library(ggplot2)
x <- seq(-4, 4, length = 1000)
y <- dnorm(x)
test_stat <- -1.5
p_x <- seq(-4, test_stat, length = 500)
p_y <- dnorm(p_x)
x_left <- -3.4
x_right <- test_stat - 0.1
y_top <- 0.17
y_mid <- 0.155
y_bot <- 0.14
p1 <- ggplot() +
  geom_line(aes(x, y), color = "#000000", linewidth = 1.2) +
  geom_area(aes(x = p_x, y = p_y), fill = "#F8766D", alpha = 0.6) +
  geom_hline(yintercept = 0, color = "#000000", linewidth = 1) +
  geom_segment(aes(x = test_stat, xend = test_stat, y = 0, yend = dnorm(test_stat)), color = "#000000", linewidth = 1.3) +
  geom_segment(aes(x = x_left, xend = x_left, y = y_bot, yend = y_mid), color = "#000000") +
  geom_segment(aes(x = x_left, xend = x_right, y = y_mid, yend = y_mid), color = "#000000") +
  geom_segment(aes(x = x_right, xend = x_right, y = y_mid, yend = y_bot), color = "#000000") +
  annotate("text", x = (x_left + x_right) / 2, y = y_top, label = "p-value", size = 5) +
  annotate("text", x = 2.5, y = 0.35, label = "H[0]:~theta==theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = 2.5, y = 0.29, label = "H[a]:~theta < theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = test_stat, y = 0.005, label = "test stat", size = 4.5, vjust = 1.5) +
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), panel.grid = element_blank())

print(p1)

# Second plot (curve slightly higher, bracket above)
y <- dnorm(x) * 0.95
test_stat <- 1.5
p_x <- seq(-4, test_stat, length = 500)
p_y <- dnorm(p_x) * 0.95
x_left <- -3.4
x_right <- test_stat - 0.1
y_top <- 0.43
y_mid <- 0.3989
y_bot <- 0.39
p2 <- ggplot() +
  geom_line(aes(x, y), color = "#000000", linewidth = 1.2) +
  geom_area(aes(x = p_x, y = p_y), fill = "#F8766D", alpha = 0.6) +
  geom_hline(yintercept = 0, color = "#000000", linewidth = 1) +
  geom_segment(aes(x = test_stat, xend = test_stat, y = 0, yend = dnorm(test_stat) * 0.95), color = "#000000", linewidth = 1.3) +
  geom_segment(aes(x = x_left, xend = x_left, y = y_bot, yend = y_mid), color = "#000000") +
  geom_segment(aes(x = x_left, xend = x_right, y = y_mid, yend = y_mid), color = "#000000") +
  geom_segment(aes(x = x_right, xend = x_right, y = y_mid, yend = y_bot), color = "#000000") +
  annotate("text", x = (x_left + x_right) / 2, y = y_top, label = "p-value", size = 5) +
  annotate("text", x = 2.5, y = 0.34, label = "H[0]:~theta==theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = 2.5, y = 0.28, label = "H[a]:~theta < theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = test_stat, y = 0.005, label = "test stat", size = 4.5, vjust = 1.5) +
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), panel.grid = element_blank())

print(p2)
```


```{r fig.cap="Two-tailed test: p-value is the total area in both tails beyond ±test statistic", echo=FALSE, fig.align='center', out.width='70%'}
library(ggplot2)

# Generate bell curve
x <- seq(-4, 4, length = 1000)
y <- dnorm(x) * 0.95

# Test statistics for two tails
left_stat <- -1.75
right_stat <- 1.75

# Shaded green tail areas
p_x_left <- seq(-4, left_stat, length = 300)
p_y_left <- dnorm(p_x_left) * 0.95
p_x_right <- seq(right_stat, 4, length = 300)
p_y_right <- dnorm(p_x_right) * 0.95

# Bracket height - slightly higher than previous "blue" reference
y_top <- 0.11
y_mid <- 0.095
y_bot <- 0.08

# Bracket horizontal positions
x_l1 <- -3.4
x_l2 <- left_stat - 0.1
x_r1 <- right_stat + 0.1
x_r2 <- 3.4

# Create plot
p <- ggplot() +
  geom_line(aes(x, y), color = "#000000", linewidth = 1.2) +
  geom_area(aes(x = p_x_left, y = p_y_left), fill = "#00BA38", alpha = 0.6) +
  geom_area(aes(x = p_x_right, y = p_y_right), fill = "#00BA38", alpha = 0.6) +
  geom_hline(yintercept = 0, color = "#000000", linewidth = 1) +
  geom_segment(aes(x = left_stat, xend = left_stat, y = 0, yend = dnorm(left_stat) * 0.95),
               color = "#000000", linewidth = 1.3) +
  geom_segment(aes(x = right_stat, xend = right_stat, y = 0, yend = dnorm(right_stat) * 0.95),
               color = "#000000", linewidth = 1.3) +

  # Left bracket
  geom_segment(aes(x = x_l1, xend = x_l1, y = y_bot, yend = y_mid), color = "#000000") +
  geom_segment(aes(x = x_l1, xend = x_l2, y = y_mid, yend = y_mid), color = "#000000") +
  geom_segment(aes(x = x_l2, xend = x_l2, y = y_mid, yend = y_bot), color = "#000000") +
  annotate("text", x = (x_l1 + x_l2)/2, y = y_top, label = "1/2 p-value", size = 5) +

  # Right bracket
  geom_segment(aes(x = x_r1, xend = x_r1, y = y_bot, yend = y_mid), color = "#000000") +
  geom_segment(aes(x = x_r1, xend = x_r2, y = y_mid, yend = y_mid), color = "#000000") +
  geom_segment(aes(x = x_r2, xend = x_r2, y = y_mid, yend = y_bot), color = "#000000") +
  annotate("text", x = (x_r1 + x_r2)/2, y = y_top, label = "1/2 p-value", size = 5) +

  # Hypotheses on the right
  annotate("text", x = 2.5, y = 0.25, label = "H[0]:~theta==theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = 2.5, y = 0.195, label = "H[a]:~theta != theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = left_stat, y = 0.005, label = "test stat", size = 4.5, vjust = 1.5) +
  annotate("text", x = right_stat, y = 0.005, label = "test stat", size = 4.5, vjust = 1.5) +

  # Style
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank()
  )
# PRINT the plot
p
```



**Step 5: Compare *p*-value to level of significance $\alpha$ and make a conclusion**

- If the *p*-value $< \alpha$, we say there is sufficient evidence against $H_0$ and the hypothesis test rejects $H_0$
  in favor of $H_a$.

- If the *p*-value $> \alpha$, thenwe say there is insufficient evidence against $H_0$ and we do not reject $H_0$ 
  (in this situation we can also say fail to reject $H_0$).

<!-- **Note:** It is not good practice to give conclusions in the context of -->
<!-- stating we *accept $H_0$* or *accept $H_a$*. -->


**About the p-value of the test statistics**

- The p-value is a conditional probability.

- It is not the probability that $H_0$ (null hypothesis: current belief)
  is true.

- It is: P(observed statistic value --- $H_0$).
  Given $H_0$ (the null hypothesis), because $H_0$ gives the parameter
  values that we need to find required probability.

- The p-value serves as a measure of the strength of the evidence against
  the null hypothesis (but it should not serve as a hard and fast rule
  for decision).

- If the p-value = 0.03 (for example) all we can say is that there is 3%
  chance of observing the statistic value we actually observed (or one
  even more inconsistent with the null value).

- The p-value is the chance (the proportion) of getting a, for instance,
  $\hat{p}$ as far as or further from $H_0$ than the value observed.

- The p-value is the probability of getting at least something (e.g., sample
  proportion $\hat{p}$) more extreme (e.g., unusual, unlikely, or rare)
  than what we have already found (our observed value of $\hat{p}$) that
  provide even stronger evidence against $H_0$.

- The more extreme the z-score (large in absolute values) are the ones
  that denote farther departure of the observed value (e.g., our
  $\hat{p}$) from the parameter value ($p_0$) in $H_0$.

- In the one-sided test, e.g., $H_a : p > p_0$, p-value is one-tailed
  probability. This is the probability that sample proportion $\hat{p}$
  falls at least as far from $p_0$ in one direction as the observed
  value of $\hat{p}$.

- In the two-sided test, e.g., $H_a : p \ne p_0$, p-value is two-tailed
  probability. This is the probability that sample proportion $\hat{p}$
  falls at least as far from $p_0$ in either direction as the observed
  value of $\hat{p}$.

The probability, computed assuming that $H_0$ is true, that the test
statistic would take a value as extreme or more extreme than that
actually observed is called the **P-value** of the test. The smaller the
P-value, the stronger the evidence against $H_0$ provided by the data.

Small P-values are evidence against $H_0$, because they say that the
observed result is unlikely to occur when $H_0$ is true. Large P-values
fail to give evidence against $H_0$.

**The P-value Scale**

- If P-value $<$ 0.001, we have very strong evidence against $H_0$.

- If 0.001 $\leq$ P-value $<$ 0.01, we have strong evidence against
  $H_0$.

- If 0.01 $\leq$ P-value $<$ 0.05, we have evidence against $H_0$.

- If 0.05 $\leq$ P-value $<$ 0.075, we have some evidence against $H_0$.

- If 0.075 $\leq$ P-value $<$ 0.10, we have slight evidence against
  $H_0$.

**Use p-value Method to Make a Decision (Reject or Fail to Reject
$H_0$)**

But how small is small p-value?\
We would need to choose an $\alpha$-level (significance-level): a number
such that if:

- $P\text{–value} \leq \alpha$-level, we reject $H_0$; We can conclude
  $H_a$ (we have evidence to support our claim). Often we phrase as a
  statistically significant result at that specified $\alpha$-level.

- $P\text{–value} > \alpha$-level, we fail to reject $H_0$; We cannot
  conclude $H_a$ (we have not enough evidence to support our claim;
  thus, $H_0$ is plausible - We do not accept $H_0$). Often we phrase as
  the result is not statistically significant at that specified
  $\alpha$-level.

- The default $\alpha$-level (significance-level) is typically
  $\alpha = 0.05$ (but it can be different based on the context of the
  study - it is usually not higher than 0.10).

The p-value in the previous example was extremely small (less than
0.0001). That is a strong evidence to suggest that more than 5% of
children have genetic abnormalities. However, it does not say that the
percentage of sampled children with genetic abnormalities was "a lot
more than 5%". That is, the p-value by itself says nothing about how
much greater the percentage might be. The confidence interval provides
that information.\
To assess the difference in practical terms, we should also construct a
confidence interval:

$$0.1198 \pm (1.96 \times 0.0166)$$ $$0.1198 \pm 0.0324$$
$$(0.0874,\ 0.1522)$$

Interpretation: We are 95% Confident that the true percentage of
children with genetic abnormalities is between 8.74% and 15.22%.\
95% CI for $p$: (9.1%, 15.6%) -- We are 95% confident that the true
percentage of all children that have genetic abnormalities is between
approximately 9.1% and 15.6%. Since both values of this CI are more than
the hypothesized value of $p = 0.05$ (5%), we can further infer that
this true percentage is more than 5%.\
**Do environmental chemicals cause congenital abnormalities?**

We do not know that environmental chemicals cause genetic abnormalities.
We merely have evidence that suggests that a greater percentage of
children are diagnosed with genetic abnormalities now, compared to the
1980s.

**More About P-values**

- Big p-values just mean that what we have observed is not surprising.
  It means that the results are in line with our assumption that the
  null hypothesis models the world, so we have no reason to reject it.

- A big p-value does not prove that the null hypothesis is true.

- When we see a big p-value, all we can say is: we cannot reject $H_0$
  (we fail to reject $H_0$) -- we cannot conclude $H_a$ (We have no
  evidence to support $H_a$).














## One Sample Hypothesis Tests

### On a Population Mean

#### When $\sigma$ is Known


<!-- Draw a simple random sample of size $n$ from a Normal population that -->
<!-- has unknown mean $\mu$ and known standard deviation $\sigma$.  -->
To test $H_0 : \mu = \mu_0$, in the situation where the population standard deviation $\sigma$ is known,
our options for the null and alternative hypotheses are:

- $H_0: \mu = \mu_0$ vs. $H_a: \mu > \mu_0$

- $H_0: \mu = \mu_0$ vs. $H_a: \mu < \mu_0$

- $H_0: \mu = \mu_0$ vs. $H_a: \mu \neq \mu_0$

<!-- <table class="hypothesis-table"> -->
<!--   <tr> -->
<!--     <td class="term" style="text-align: left !important;">$H_0: \mu = \mu_0$</td> -->
<!--     <td class="sep" style="text-align: left !important;">vs.</td> -->
<!--     <td class="alt" style="text-align: left !important;">$H_a: \mu > \mu_0$</td> -->
<!--   </tr> -->
<!--   <tr> -->
<!--     <td class="term" style="text-align: left !important;">$H_0: \mu = \mu_0$</td> -->
<!--     <td class="sep" style="text-align: left !important;">vs.</td> -->
<!--     <td class="alt" style="text-align: left !important;">$H_a: \mu < \mu_0$</td> -->
<!--   </tr> -->
<!--   <tr> -->
<!--     <td class="term" style="text-align: left !important;">$H_0: \mu = \mu_0$</td> -->
<!--     <td class="sep" style="text-align: left !important;">vs.</td> -->
<!--     <td class="alt" style="text-align: left !important;">$H_a: \mu \neq \mu_0$</td> -->
<!--   </tr> -->
<!-- </table> -->


We calculate the $z$-statistic:

$$z^{*} = \frac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} \sim N(0,1)$$

The reference distribution is the standard normal distribution.

In terms of a variable $Z$ having the standard normal distribution, the
p-value for a test of $H_0$ against:

- $H_a: \mu > \mu_0$ is $P(Z > z^{*})$

- $H_a: \mu < \mu_0$ is $P(Z < z^{*})$

- $H_a: \mu \ne \mu_0$ is $2P(Z > |z^{*}|)$


::::: example
Deer are a common sight on the UTM campus. Suppose an ecologist is
interested in the average mass of adult white-tailed does (female deer)
around the Mississauga campus to determine whether they are healthy for
the upcoming winter. The ecologist captures a sample of 36 adult females
around the UTM and measures the average mass of this sample to be 42.53
kg.

From previous studies conducted in the area, the average mass of healthy
does was reported to be 45 kg. Conduct a hypothesis test at the 5%
significance level to determine whether the mass of does around UTM has
decreased. Assume the standard deviation is known to be 5.25 kg.

**1. Level of significance.** $\alpha = 0.05$

**2. State the null and alternative hypotheses.**
$$H_0: \mu = 45 \qquad H_a: \mu < 45$$

**3. Calculate appropriate test statistic.**

Given: $$n = 36, \quad \bar{x} = 42.53, \quad \sigma = 5.25$$

Since $\sigma$ is known, the test statistic is:
$$z^* = \frac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} = \frac{42.53 - 45}{5.25/\sqrt{36}} = -2.82$$

Reference distribution: standard normal

**4. Calculate p-value**

$$\text{p-value} = P(Z < -2.82) \approx 0.0024$$
```{r fig.cap="Left-tailed p-value for the test statistic \\( z^* = -2.82 \\)", fig.align='center', echo=FALSE, out.width='70%'}
library(ggplot2)

# Normal curve data
x <- seq(-4, 4, length = 1000)
y <- dnorm(x)

# Shaded left-tail region
test_stat <- -2.82
p_x <- seq(-4, test_stat, length = 500)
p_y <- dnorm(p_x)

# Bracket near pink area
x_left <- -3.6
x_right <- test_stat - 0.1
y_bot <- 0.005
y_mid <- 0.028
y_top <- 0.04

# Build plot
p <- ggplot() +
  geom_line(aes(x, y), color = "#000000", linewidth = 1.4) +
  geom_area(aes(x = p_x, y = p_y), fill = "#DC71FA", alpha = 0.65) +
  geom_hline(yintercept = 0, color = "#000000", linewidth = 1.1) +

  # Test statistic line and label
  geom_segment(aes(x = test_stat, xend = test_stat, y = 0, yend = dnorm(test_stat)), 
               color = "#000000", linewidth = 1.2) +
  annotate("text", x = test_stat, y = -0.02, label = "z* = -2.82", size = 6, fontface = "italic") +

  # Bracket and p-value label
  geom_segment(aes(x = x_left, xend = x_left, y = y_bot, yend = y_mid), color = "#000000") +
  geom_segment(aes(x = x_left, xend = x_right, y = y_mid, yend = y_mid), color = "#000000") +
  geom_segment(aes(x = x_right, xend = x_right, y = y_mid, yend = y_bot), color = "#000000") +
  annotate("text", x = (x_left + x_right)/2, y = y_top, label = "p-value", size = 5.2) +

  # Hypotheses
  annotate("text", x = 2.4, y = 0.32, label = "H[0]:~mu==45", parse = TRUE, size = 6.2) +
  annotate("text", x = 2.4, y = 0.26, label = "H[a]:~mu < 45", parse = TRUE, size = 6.2) +

  # Axes
  scale_x_continuous(breaks = c(-4, -2, 0, 2, 4), limits = c(-4, 4), expand = c(0, 0)) +
  scale_y_continuous(limits = c(-0.03, 0.42)) +

  theme_minimal() +
  labs(x = NULL, y = NULL) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )

p
```


**5. Compare p-value with level of significance $\alpha$ and make a
conclusion:**

$$0.0024 < 0.05 \Rightarrow \text{p-value} < \alpha$$

There is sufficient evidence at the 5% level of significance to reject
the null that does this winter weigh the same as in the past and to
conclude the alternative that does this winter weigh less than 45 kg.

**R code:**

```r
    # Find test stat
    z_test_stat = (42.53 - 45) / (5.25 / sqrt(36))
    z_test_stat
    [1] -2.822857

    # Find the p-value
    # Since the alternative is Ha : mu < 45
    p-value = pnorm(z_test_stat)
    [1] 0.00237989
```

*Note:* The `pnorm()` function in R, by default, returns the cumulative
probability (area) to the left of the given value.\

**R code: Using BSDA package**

```r
    # Using the BSDA library. install BSDA if it is not already installed.
    # install.packages("BSDA")
    > library(BSDA)
    > # Conduct the z-test with the zsum.test function
    > zsum.test(mean.x = 42.53, sigma.x = 5.24, n.x = 36, mu = 45, alternative = "less")

            One-sample z-Test

    data:  Summarized x
    z = -2.8282, p-value = 0.00234
    alternative hypothesis: true mean is less than 45
    95 percent confidence interval:
     NA 43.96651
    sample estimates:
    mean of x 
        42.53 
```

**Interpretation:**

There is sufficient evidence at the 5% level of significance to reject
the null hypothesis. We conclude that the average mass of does this
winter is significantly less than 45 kg.
:::::


:::::: example
Diet colas use artificial sweeteners to avoid sugar. These sweeteners
gradually lose their sweetness over time. Manufacturers therefore test
new colas for loss of sweetness before marketing them. Trained tasters
sip the cola along with drinks of standard sweetness and score the cola
on a "sweetness score" of 1 to 10. The cola is then stored for a month
at high temperature to imitate the effect of four months' storage at
room temperature. Each taster scores the cola again after storage. This
is a matched pairs experiment. Our data are the differences (score
before storage minus score after storage) in the tasters' scores. The
bigger these differences, the bigger the loss of sweetness.

Suppose we know that for any cola, the sweetness loss scores vary from
taster to taster according to a Normal distribution with standard
deviation $\sigma = 1$. The mean $\mu$ for all tasters measures loss of
sweetness.

The following are the sweetness losses for a new cola as measured by 10
trained tasters:

$$2.0, 0.4, 0.7, 2.0, -0.4, 2.2, -1.3, 1.2, 1.1, 2.3$$

Are these data good evidence that the cola lost sweetness in storage?\
**Solution**\
$\mu$ = mean sweetness loss for the population of **all** tasters.\
**Step 1:** State hypotheses. $$\begin{aligned}
H_0 &: \mu = 0 \\
H_a &: \mu > 0
\end{aligned}$$ **Step 2:** Test statistic:
$z_\star = \dfrac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} = \dfrac{1.02 - 0}{1 / \sqrt{10}} = 3.23$\
**Step 3:** P-value. $P(Z > z_\star) = P(Z > 3.23) = 0.0006$\
**Step 4:** Conclusion. We would rarely observe a mean as large as 1.02
if $H_0$ were true. The small p-value provides strong evidence against
$H_0$, supporting $H_a: \mu > 0$. That is, the mean sweetness loss is
likely positive.

**R code (Simulation)**

```r
    # n = sample size;
    n<-10;
    mu.zero<-0;
    sigma<-1;
    sigma.xbar<-sigma/sqrt(n);

    # x bar = sample mean with 10 obs;
    x.bar<-rnorm(1,mean=mu.zero,sd=sigma.xbar);
    x.bar;

    ## [1] 0.3265859

    # z.star = test statistic;
    z.star<-(x.bar-mu.zero)/sigma.xbar;
    z.star;

    ## [1] 1.032755
```

**R code (10,000 Simulations)**

```r
    n <- 10;
    mu.zero <- 0;
    sigma <- 1;
    sigma.xbar <- sigma / sqrt(n);
    # x bar = sample mean with 10 obs;
    # m = number of simulations;
    m <- 10000;
    x.bar <- rnorm(m, mean = mu.zero, sd = sigma.xbar);

    # z.star = test statistic;
    z.star <- (x.bar - mu.zero) / sigma.xbar;
    hist(z.star, xlab = "differences", col = "blue");
```


```{r fig.cap="Histogram of \\( z^\\star \\) values from 10,000 simulations under \\( H_0 \\)", fig.align='center', echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)

# Parameters
n <- 10
mu.zero <- 0
sigma <- 1
sigma.xbar <- sigma / sqrt(n)

# Simulations
m <- 10000
x.bar <- rnorm(m, mean = mu.zero, sd = sigma.xbar)

# z.star = test statistic
z.star <- (x.bar - mu.zero) / sigma.xbar
z.df <- data.frame(z.star = z.star)

# ggplot histogram
ggplot(z.df, aes(x = z.star)) +
  geom_histogram(binwidth = 0.2, fill = "#619CFF", color = "#FFFFFF") +
  labs(
    title = " ",
    x = "differences",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 14)
```


**R code (Empirical p-value)**

```r
    ## P-value

    p_value <- length(z.star[z.star > 3.23]) / m;

    p_value

    ## [1] 8e-04
```
::::::


::: example
The National Center for Health Statistics reports that the systolic
blood pressure for males 35 to 44 years of age has mean 128 and standard
deviation 15.

The medical director of a large company looks at the medical records of
72 executives in this age group and finds that the mean systolic blood
pressure in this sample is $\bar{x} = 126.07$. Is this evidence that the
company's executives have a different mean blood pressure from the
general population?

Suppose we know that executives' blood pressures follow a Normal
distribution with standard deviation $\sigma = 15$.

**Solution:** Let $\mu$ be the mean systolic blood pressure of the
executive population.

1.  **State hypotheses:**\
    $$\begin{aligned}
    H_0 &: \mu = 128 \\
    H_a &: \mu \ne 128
    \end{aligned}$$

2.  **Test statistic:**\
    $$z^{*} = \frac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} = \frac{126.07 - 128}{15 / \sqrt{72}} = -1.09$$

3.  **P-value:**\
    $$2P(Z > |z^{*}|) = 2P(Z > 1.09) = 2(1 - 0.8621) = 0.2758$$

4.  **Conclusion:**\
    More than 27% of the time, a simple random sample of size 72 from
    the general male population would have a mean blood pressure at
    least as far from 128 as that of the executive sample. The observed
    $\bar{x} = 126.07$ is therefore not good evidence that executives
    differ from other men.
:::

<!-- ::: tcolorbox -->
<!-- There are four steps in carrying out a significance test: -->

<!-- 1.  State the hypotheses. -->

<!-- 2.  Calculate the test statistic. -->

<!-- 3.  Find the p-value. -->

<!-- 4.  State your conclusion in the context of your specific setting. -->

<!-- Once you have stated your hypotheses and identified the proper test, you -->
<!-- or your computer can do Steps 2 and 3 by following a recipe. -->
<!-- ::: -->


::: example
Consider the following hypothesis test:

$$\begin{aligned}
H_0 &: \mu = 20 \\
H_a &: \mu < 20
\end{aligned}$$

A sample of 50 provided a sample mean of 19.4. The population standard
deviation is 2.

1.  Compute the value of the test statistic.

2.  What is the p-value?

3.  Using $\alpha = 0.05$, what is your conclusion?

**Solution:**

1.  **Test statistic:**
    $$z^{*} = \frac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} = \frac{19.4 - 20}{2 / \sqrt{50}} = -2.1213$$

2.  **P-value:** $$P(Z < z^{*}) = P(Z < -2.1213) = 0.0169$$

3.  **Conclusion:**\
    Since the p-value $= 0.0169 < \alpha = 0.05$, we reject
    $H_0 : \mu = 20$. We conclude that $\mu < 20$.
:::

::: example
Consider the following hypothesis test:

$$\begin{aligned}
H_0 &: \mu = 25 \\
H_a &: \mu > 25
\end{aligned}$$

A sample of 40 provided a sample mean of 26.4. The population standard
deviation is 6.

1.  Compute the value of the test statistic.

2.  What is the p-value?

3.  Using $\alpha = 0.01$, what is your conclusion?

**Solution:**

1.  **Test statistic:**
    $$z^{*} = \frac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} = \frac{26.4 - 25}{6 / \sqrt{40}} = 1.4757$$

2.  **P-value:** $$P(Z > z^{*}) = P(Z > 1.4757) = 0.0700$$

3.  **Conclusion:**\
    Since the p-value $= 0.0700 > \alpha = 0.01$, we **cannot reject**
    $H_0 : \mu = 25$.\
    We conclude that we don't have enough evidence to claim that
    $\mu > 25$.
:::

::: example
Consider the following hypothesis test:

$$\begin{aligned}
H_0 &: \mu = 15 \\ 
H_a &: \mu \ne 15
\end{aligned}$$

A sample of 50 provided a sample mean of 14.15. The population standard
deviation is 3.

1.  Compute the value of the test statistic.

2.  What is the p-value?

3.  Using $\alpha = 0.05$, what is your conclusion?

**Solution:**

1.  **Test statistic:**
    $$z^{*} = \frac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} = \frac{14.15 - 15}{3 / \sqrt{50}} = -2.0034$$

2.  **P-value:**
    $$2P(Z > |z^{*}|) = 2P(Z > |{-2.0034}|) = 2P(Z > 2.0034) = 0.0451$$

3.  **Conclusion:**\
    Since P-value $= 0.0451 < \alpha = 0.05$, we reject
    $H_0 : \mu = 15$.\
    We conclude that $\mu \ne 15$.

**Confidence Interval Interpretation:**

The 95% confidence interval for $\mu$ is:
$$\bar{x} \pm z^{*} \left( \frac{\sigma}{\sqrt{n}} \right)$$
$$14.15 \pm 1.96 \left( \frac{3}{\sqrt{50}} \right)
= (13.3184,\ 14.9815)$$

Since the hypothesized value $\mu_0 = 15$ falls **outside** this
interval, we again reject $H_0 : \mu = 15$.
:::

<!-- ::: {. remark} -->
<!-- *A level $\alpha$ two-sided significance test rejects a hypothesis -->
<!-- $H_0 : \mu = \mu_0$ exactly when the value $\mu_0$ falls outside a level -->
<!-- $1 - \alpha$ confidence interval for $\mu$.* -->
<!-- ::: -->









#### When $\sigma$ is Not Known

To test $H_0 : \mu = \mu_0$, in the situation where the population standard deviation $\sigma$ is not known,
our options for the null and alternative hypotheses are:

- $H_0: \mu = \mu_0$ vs. $H_a: \mu > \mu_0$

- $H_0: \mu = \mu_0$ vs. $H_a: \mu < \mu_0$

- $H_0: \mu = \mu_0$ vs. $H_a: \mu \neq \mu_0$

We calculate the $t$-statistic:

$$t^{*} = \frac{\bar{x} - \mu_0}{s / \sqrt{n}} \sim t_{n-1}$$

In terms of a variable $T$ which follows a $t$-distribution at $n-1$ degrees of freedom, the
p-value for a test of $H_0$ against:

- $H_a: \mu > \mu_0$ is $P(T > t^{*})$

- $H_a: \mu < \mu_0$ is $P(T < t^{*})$

- $H_a: \mu \ne \mu_0$ is $2P(T > |t^{*}|)$


::: example
Researchers studied the physiological effects of laughter. They measured
heart rates (in beats per minute) of **$n = 25$** subjects (ages 18--34)
while they laughed. They obtained:
$$\bar{x} = 73.5, \quad s = 6, \quad \alpha = 0.05$$ It is well known
that the resting heart rate is 71 bpm. Is there evidence that the mean
heart rate during laughter exceeds 71 bpm?

**Step 1: State the hypotheses.** $$\begin{aligned}
H_0 &: \mu = 71 \\
H_a &: \mu > 71
\end{aligned}$$

**Step 2: Check assumptions.**

- The sample is an independent random sample of individuals aged 18--34.

- The population of heart rates during laughter is normally distributed.

**Step 3: Compute the test statistic.**

Since $\sigma$ is unknown, we use the $t$ statistic:
$$t^\ast = \frac{\bar{x} - \mu_0}{s / \sqrt{n}} = \frac{73.5 - 71}{6 / \sqrt{25}} = 2.083$$

Reference distribution: $t$ distribution with $n - 1 = 25 - 1 = 24$
degrees of freedom.

**Step 4: Determine the p-value.**

Using the $t$ distribution with 24 df: $$0.01 < \text{p-value} < 0.025$$

**Step 5: Make a conclusion.**

Since $\text{p-value} < \alpha = 0.05$, we reject $H_0$.

*There is sufficient evidence at the 5% level of significance to reject
the null that the mean is 71 bpm in favor of the alternative that the
mean is greater than 71 bpm for people who are laughing.*
:::

::::: example
A researcher is asked to test the hypothesis that the average price of a
2-star (CAA rating) motel room has decreased since last year. Last year,
a study showed that the prices were Normally distributed with a mean of
\$89.50.

A random sample of twelve 2-star motels produced the following room
prices:

$$\text{\$85.00, 92.50, 87.50, 89.90, 90.00, 82.50, 87.50, 90.00, 85.00, 89.00, 91.50, 87.50}$$

At the 5% level of significance, can we conclude that the mean price has
decreased?

**Solution:**

Let $\mu$ be the true average price of a 2-star motel room.

1.  **State hypotheses.** $$\begin{aligned}
    H_0 &: \mu = 89.5 \\
    H_a &: \mu < 89.5
    \end{aligned}$$

2.  **Compute test statistic.**
    $$t^\ast = \frac{\bar{x} - \mu_0}{s / \sqrt{n}} = \frac{88.1583 - 89.5}{2.9203 / \sqrt{12}} = -1.5915$$

3.  **Find the p-value.**\
    With $df = 11$, the p-value (from the $t$-distribution table) is
    between 0.05 and 0.10.

4.  **Conclusion.**\
    Since the p-value $> 0.05$, we **fail to reject** $H_0$.\
    There is not sufficient evidence to conclude that the average price
    of 2-star motels has decreased this year.

**R code (One Sample t-test)**

```r
    # Step 1. Entering data;
    prices=c(85.00, 92.50, 87.50, 89.90, 90.00, 82.50,
             87.50, 90.00, 85.00, 89.00, 91.50, 87.50);

    # Step 2. Hypothesis test;
    t.test(prices, alternative="less", mu=89.5);
```

**R output**

```r
    ## 
    ##  One Sample t-test
    ## 
    ## data:  prices
    ## t = -1.5915, df = 11, p-value = 0.0699
    ## alternative hypothesis: true mean is less than 89.5
    ## 95 percent confidence interval:
    ##  -Inf 89.67229
    ## sample estimates:
    ## mean of x 
    ##  88.15833 
```
:::::

::: tcolorbox
Draw an SRS of size $n$ from a large population having unknown mean
$\mu$.

To *test the hypothesis* $H_0 : \mu = \mu_0$, compute the *one-sample
$t$ statistic* $$t^\ast = \frac{\bar{x} - \mu_0}{s / \sqrt{n}}$$

In terms of a variable $T$ having the $t_{n - 1}$ distribution, the
p-value for a test of $H_0$ against

$$\begin{aligned}
H_a : \mu > \mu_0 &\quad \text{is} \quad P(T \ge t^\ast) \\
H_a : \mu < \mu_0 &\quad \text{is} \quad P(T \le t^\ast) \\
H_a : \mu \ne \mu_0 &\quad \text{is} \quad 2P(T \ge |t^\ast|)
\end{aligned}$$

These p-values are exact if the population distribution is Normal and
are approximately correct for large $n$ in other cases.
:::

:::: example
We are conducting a two-sided one-sample $t$-test for the hypotheses:
$$\begin{aligned}
H_0 &: \mu = 64 \\
H_a &: \mu \ne 64
\end{aligned}$$

based on a sample of $n = 15$ observations, with test statistic
$t^* = 2.12$.

**a) Degrees of freedom:**

$$df = n - 1 = 15 - 1 = 14$$

**b) Critical values and p-value bounds:**

From the $t$-distribution table for $df = 14$:

- $t = 1.761$ corresponds to a two-tailed probability of 0.10

- $t = 2.145$ corresponds to a two-tailed probability of 0.05

Since $t^* = 2.12$ falls between these values, the two-sided p-value
satisfies: $$0.05 < \text{p-value} < 0.10$$

**c) Significance:**

- At the 10% level: **Yes**, since the p-value $< 0.10$

- At the 5% level: **No**, since the p-value $> 0.05$

**d) Exact two-sided p-value using R:**

```r
    # Compute exact two-sided p-value for t* = 2.12 with df = 14
    2 * (1 - pt(2.12, df = 14))

    ## [1] 0.05235683
```

Thus, the exact two-sided p-value is approximately **0.0524**,
confirming the bracketing result.
::::











### On a Population Proportion

To test $H_0 : p = p_0$, our options for the null and alternative hypotheses are:

- $H_0: p = p_0$ vs. $H_a: p > p_0$

- $H_0: p = p_0$ vs. $H_a: p < p_0$

- $H_0: p - p_0$ vs. $H_a: p \neq p_0$


We calculate the $z$-statistic:

$$z^* = \dfrac{\hat{p} - p_0}{\sqrt{\dfrac{p_0(1 - p_0)}{n}}}$$

In terms of a variable $Z$ having the standard normal distribution, the
p-value for a test of $H_0$ against:

- $H_a: p > \mu_0$ is $P(Z > z^{*})$

- $H_a: p < \mu_0$ is $P(Z < z^{*})$

- $H_a: p \neq \mu_0$ is $2P(Z > |z^{*}|)$


<!-- ::: example -->
<!-- A YouTuber goes to her nearest Tim Hortons and buys 100 empty cups. -->
<!-- After rolling up the rims, she ends up with 12 winning cups out of the -->
<!-- 100 she bought, all of them were food prizes. -->

<!-- If the probability of winning a food prize is supposed to be -->
<!-- $\frac{1}{6}$, does she have evidence to claim that the probability of -->
<!-- winning a food prize is less than $\frac{1}{6}$? -->
<!-- ::: -->

<!-- ::: definition -->
<!-- The **sample space** $\mathbf{S}$ of a random phenomenon is the set of -->
<!-- all possible outcomes. -->

<!-- An **event** is an outcome or a set of outcomes of a random phenomenon. -->
<!-- That is, an event is a subset of the sample space. -->

<!-- A **probability model** is a mathematical description of a random -->
<!-- phenomenon consisting of two parts: a sample space $S$ and a way of -->
<!-- assigning probabilities to events. -->
<!-- ::: -->

<!-- Rolling a fair die (random phenomenon). There are 6 possible outcomes -->
<!-- when we roll a die.\ -->
<!-- The sample space for rolling a die and counting the pips is -->

<!-- $$S = \{1,\, 2,\, 3,\, 4,\, 5,\, 6\}$$ -->

<!-- "Roll a 6" is an event that contains one of these 6 outcomes. -->

<!-- ::: definition -->
<!-- A random variable $X$ has a **discrete uniform distribution** if each of -->
<!-- the $n$ values in its range, say, $x_1, x_2, \ldots, x_n$, has equal -->
<!-- probability. Then, $$f(x_i) = \frac{1}{n}$$ -->
<!-- ::: -->

<!-- **R code:** -->

<!-- ::: tcolorbox -->
<!--     # Define a die with values 1 through 6 -->
<!--     die <- c(1, 2, 3, 4, 5, 6) -->

<!--     # Roll the die once -->
<!--     sample(die, 1, replace = TRUE) -->
<!--     ## [1] 2 -->

<!--     # Roll the die six times -->
<!--     sample(die, 6, replace = TRUE) -->
<!--     ## [1] 1 3 2 6 3 1 -->
<!-- ::: -->

<!-- ```{r fig.cap="Plot of frequencies from 60 simulations of a fair six-sided die", echo=FALSE, fig.align='center', out.width='70%'} -->
<!-- # Simulate rolling a fair die 60 times -->
<!-- set.seed(123) -->
<!-- rolls <- sample(1:6, size = 60, replace = TRUE) -->

<!-- # Tabulate frequencies -->
<!-- counts <- table(rolls) -->

<!-- # Base plot: vertical lines (type = "h") -->
<!-- plot(counts, -->
<!--      type = "h", -->
<!--      col = "blue", -->
<!--      lwd = 3, -->
<!--      ylim = c(0, max(counts) + 2), -->
<!--      ylab = "Frequency", -->
<!--      xlab = "rolls", -->
<!--      main = "60 simulations", -->
<!--      xaxt = "n") -->
<!-- axis(1, at = 1:6) -->

<!-- # Add blue bullets on top of bars -->
<!-- points(1:6, as.numeric(counts), col = "blue", pch = 19, cex = 1.5) -->
<!-- ``` -->


<!-- ::: definition -->
<!-- A **random variable** is a variable whose value is a numerical outcome -->
<!-- of a random phenomenon. -->

<!-- The **probability distribution** of a random variable $X$ tells us what -->
<!-- values $X$ can take and how to assign probabilities to those values. -->
<!-- ::: -->

<!-- ::: nt -->
<!-- The Binomial setting -->

<!-- - There are a fixed number $n$ of observations. -->

<!-- - The $n$ observations are all **independent**. That is, knowing the -->
<!--   result of one observation tells you nothing about the other -->
<!--   observations. -->

<!-- - Each observation falls into one of just two categories, which for -->
<!--   convenience we call "success" and "failure". -->

<!-- - The probability of a success, call it $p$, is the same for each -->
<!--   observation. -->
<!-- ::: -->

<!-- A random variable $Y$ is said to have a **binomial distribution** based -->
<!-- on $n$ trials with success probability $p$ if and only if -->
<!-- $$p(y) = \frac{n!}{y!(n - y)!} \, p^y (1 - p)^{n - y}, \quad y = 0, 1, 2, \ldots, n \quad \text{and} \quad 0 \leq p \leq 1.$$ -->

<!-- ::: example -->
<!-- Think of rolling a die $n$ times as an example of the binomial setting. -->
<!-- Each roll gives either a six or a number different from six. Knowing the -->
<!-- outcome of one roll doesn't tell us anything about other rolls, so the -->
<!-- $n$ rolls are independent. -->

<!-- If we call six a success, then $p$ is the probability of a six and -->
<!-- remains the same as long as we roll the same die. The number of sixes we -->
<!-- count is a random variable $X$. The distribution of $X$ is called a -->
<!-- **binomial distribution**. -->
<!-- ::: -->

<!-- **R code (Binomial Simulations and PMF)** -->

<!-- ::: tcolorbox -->
<!--     ## Simulation: Binomial with n = 10 and p = 1/6. -->
<!--     rbinom(1, size = 10, prob = 1/6); -->
<!--     ## [1] 3 -->

<!--     rbinom(1, size = 10, prob = 1/6); -->
<!--     ## [1] 1 -->

<!--     rbinom(1, size = 10, prob = 1/6); -->
<!--     ## [1] 0 -->

<!--     ## Pmf: Binomial with n = 10 and p = 1/6. -->
<!--     x <- seq(0, 10, by = 1); -->
<!--     y <- dbinom(x, 10, 1/6); -->
<!--     plot(x, y, type = "p", col = "blue", pch = 19); -->
<!-- ::: -->

<!-- ```{r fig.cap="PMF of the Binomial distribution with \\( n = 10 \\) and \\( p = \\frac{1}{6} \\)", echo=FALSE, fig.align='center', out.width='70%'} -->
<!-- # Define x values from 0 to 10 -->
<!-- x <- 0:10 -->

<!-- # Binomial PMF: n = 10, p = 1/6 -->
<!-- y <- dbinom(x, size = 10, prob = 1/6) -->

<!-- # Plot PMF -->
<!-- plot(x, y, -->
<!--      type = "p",        # Points -->
<!--      pch = 19,          # Solid circle -->
<!--      col = "blue",      # Blue color -->
<!--      xlab = "x", -->
<!--      ylab = "y", -->
<!--      main = "Probability Mass Function when n = 10 and p = 1/6") -->

<!-- # Add vertical dashed red line at x = 1 -->
<!-- abline(v = 1, col = "red", lty = 2) -->
<!-- ``` -->

<!-- ```{r fig.cap="PMF of the Binomial distribution with \\( n = 100 \\) and \\( p = \\frac{1}{6} \\)", echo=FALSE, fig.align='center', out.width='70%'} -->
<!-- # Define x values from 0 to 100 -->
<!-- x <- 0:100 -->

<!-- # Compute PMF values -->
<!-- y <- dbinom(x, size = 100, prob = 1/6) -->

<!-- # Plot the PMF -->
<!-- plot(x, y, -->
<!--      type = "p",           # Points -->
<!--      pch = 19,             # Solid circles -->
<!--      col = "blue",         # Blue color -->
<!--      xlab = "x", -->
<!--      ylab = "y", -->
<!--      main = "Pmf when n = 100 and p = 1/6") -->

<!-- # Add vertical dashed line at expected value (mean) -->
<!-- abline(v = 100 * (1/6), col = "red", lty = 2) -->

<!-- # Close the PDF device -->
<!-- ``` -->


<!-- **R code (PMF values for selected $x$ values)** -->

<!-- ::: tcolorbox -->
<!--     dbinom(c(15, 16, 17, 18), size = 100, prob = 1/6); -->
<!--     ## [1] 0.10023663 0.10650142 0.10524847 0.09706247 -->
<!-- ::: -->

<!-- ```{r fig.cap="Distribution of successes from 2000 simulated YouTubers with \\( n = 100 \\), \\( p = \\frac{1}{6} \\)", echo=FALSE, fig.align='center', out.width='70%'} -->
<!-- # Simulate 1000 binomial observations with n = 30, p = 1/6 -->
<!-- set.seed(123)  # for reproducibility -->
<!-- sim_data <- rbinom(1000, size = 30, prob = 1/6) -->

<!-- # Create barplot of frequencies -->
<!-- counts <- table(sim_data) -->
<!-- plot(counts, -->
<!--      type = "h", -->
<!--      col = "blue", -->
<!--      lwd = 3, -->
<!--      ylab = "", -->
<!--      xlab = "", -->
<!--      main = "") -->
<!-- ``` -->


<!-- **R code (A few values from our simulation)** -->

<!-- ::: tcolorbox -->
<!--     ## vec.prop -->
<!--     ##  6  7  8  9 10 11 12  -->
<!--     ##  7  3  8 24 46 72 106  -->
<!--     ## [1] 266 -->
<!--     ## [1] 0.133 -->
<!-- ::: -->

<!-- It turns out that our p-value for this simulation is:\ -->
<!-- 0.133 -->

<!-- ```{r fig.cap="Simulation vs Theoretical pmf", echo=FALSE, fig.align='center', out.width='70%'} -->
<!-- # Simulate 2000 observations from Binomial(n=10, p=0.5) -->
<!-- set.seed(123) -->
<!-- sim <- rbinom(2000, size = 10, prob = 0.5) -->

<!-- # Frequency table -->
<!-- tab <- table(sim) -->
<!-- x_vals <- as.numeric(names(tab)) -->
<!-- y_vals <- as.numeric(tab) -->

<!-- # Plot vertical blue lines -->
<!-- plot(x_vals, y_vals, -->
<!--      type = "h", -->
<!--      col = "blue", -->
<!--      lwd = 2, -->
<!--      ylim = c(0, max(y_vals) + 20), -->
<!--      xlab = "", -->
<!--      ylab = "", -->
<!--      main = "") -->

<!-- # Add red dots at top of lines (perfectly aligned) -->
<!-- points(x_vals, y_vals, -->
<!--        col = "red", -->
<!--        pch = 19, -->
<!--        cex = 1.1) -->
<!-- ``` -->


<!-- ::: tcolorbox -->
<!-- Draw an SRS of size $n$ from a large population that contains proportion -->
<!-- $p$ of "successes". Let $\hat{p}$ be the **sample proportion** of -->
<!-- successes, -->

<!-- $$\hat{p} = \frac{\text{number of successes in the sample}}{n}$$ -->

<!-- Then: -->

<!-- - The **mean** of the sampling distribution of $\hat{p}$ is $p$. -->

<!-- - The **standard deviation** of the sampling distribution is -->
<!--   $$\sqrt{\frac{p(1 - p)}{n}}.$$ -->

<!-- - As the sample size increases, the sampling distribution of $\hat{p}$ -->
<!--   becomes **approximately Normal**. That is, for large $n$, $\hat{p}$ -->
<!--   has approximately the $$N\left(p, \sqrt{\frac{p(1 - p)}{n}}\right)$$ -->
<!--   distribution. -->
<!-- ::: -->

<!-- ```{r fig.cap=" Binomial with Normal Approximation", echo=FALSE, fig.align='center', out.width='70%'} -->
<!-- # Parameters -->
<!-- n <- 100 -->
<!-- p <- 1/6 -->

<!-- # PMF values -->
<!-- x <- 0:n -->
<!-- binom_probs <- dbinom(x, size = n, prob = p) -->

<!-- # Normal approximation -->
<!-- mu <- n * p -->
<!-- sigma <- sqrt(n * p * (1 - p)) -->
<!-- normal_curve <- dnorm(x, mean = mu, sd = sigma) -->

<!-- # Plot binomial PMF (blue points with outline) -->
<!-- plot(x, binom_probs, -->
<!--      type = "p", -->
<!--      pch = 21, -->
<!--      bg = "blue", -->
<!--      col = "darkblue", -->
<!--      cex = 1.2, -->
<!--      ylim = c(0, max(binom_probs) + 0.01), -->
<!--      xlab = "", -->
<!--      ylab = "") -->

<!-- # Overlay normal approximation (red line) -->
<!-- lines(x, normal_curve, col = "red", lwd = 1) -->

<!-- # Add legend -->
<!-- legend("topright", legend = "Normal\napproximation", -->
<!--        col = "red", lwd = 1, bty = "n") -->
<!-- ``` -->


<!-- ::: tcolorbox -->
<!-- To *test the hypothesis* $H_0 : p = p_0$, compute the $z^{*}$ -->
<!-- statistic: -->
<!-- $$z^{*} = \frac{\hat{p} - p_0}{\sqrt{\dfrac{p_0(1 - p_0)}{n}}}$$ -->

<!-- In terms of a variable $Z$ having the standard Normal distribution, the -->
<!-- approximate p-value for a test of $H_0$ against: -->

<!-- $$\begin{aligned} -->
<!-- H_a &: p > p_0 \quad \text{is} \quad P(Z > z^{*}) \\ -->
<!-- H_a &: p < p_0 \quad \text{is} \quad P(Z < z^{*}) \\ -->
<!-- H_a &: p \ne p_0 \quad \text{is} \quad 2P(Z > |z^{*}|) \\ -->
<!-- \end{aligned}$$ -->
<!-- ::: -->

<!-- ## Introduction to Hypothesis Testing (Significance Test) {#introduction-to-hypothesis-testing-significance-test .unnumbered .unlisted} -->

<!-- Consider the following problem: In 1980s, it was generally believed that -->
<!-- congenital abnormalities affect 5% of the nation's children. Some people -->
<!-- believe that the increase in the number of chemicals in the environment -->
<!-- in recent years has led to an increase in the incidence of -->
<!-- abnormalities. A recent study examined 384 children and found that 46 of -->
<!-- them showed signs of abnormality. Is this strong evidence that the risk -->
<!-- has increased? -->

<!-- - The above statement serves as a hypothesis, moreover it is a Research -->
<!--   Hypothesis. -->

<!-- A hypothesis is: -->

<!-- - a statement about a population. -->

<!-- - a prediction that a parameter describing some characteristics of a -->
<!--   variable (e.g., true proportion, $p$) takes a particular numerical -->
<!--   value or falls in a certain range of values. -->

<!-- For conducting a Significance Test: -->

<!-- - Researchers (you) use data to summarize the evidence about a -->
<!--   hypothesis. -->

<!-- - With data, you can compare the point estimates of parameters to the -->
<!--   values predicted by the hypothesis. -->

<!-- **Important Ideas about Hypothesis Testing** -->

<!-- - All the hypothesis tests boil down to the same question: "Is an -->
<!--   observed difference or pattern too large to be attributed to chance?" -->

<!-- - We measure "how large" by putting our sample results in the context of -->
<!--   a sampling distribution model (e.g., Normal model, $t$ distribution). -->

<!-- To plan a statistical hypothesis test, specify the model you will use to -->
<!-- test the null hypothesis and the parameter of interest. -->

<!-- - All models require assumptions, so you will need to state them and -->
<!--   check any corresponding conditions. -->

<!-- - For example, if the conditions are satisfied, we can model the -->
<!--   sampling distribution of the proportion with a Normal model. -->
<!--   Otherwise, we cannot proceed with the test (we need to stop and -->
<!--   reconsider). -->

<!-- ## Steps in conducting Hypothesis Testing {#steps-in-conducting-hypothesis-testing .unnumbered .unlisted} -->

<!-- 1.  State the null and the alternative hypothesis. -->

<!-- 2.  Check the necessary assumptions. -->

<!-- 3.  Identify the test-statistic. Find the value of the test-statistic. -->

<!-- 4.  Find the p-value of the test-statistic. -->

<!-- 5.  State (if any) a conclusion. -->

:::: example
**Example of Hypothesis Testing for a Proportion**

In 1980s, it was generally believed that congenital abnormalities affect
5% of the nation's children. Some people believe that the increase in
the number of chemicals in the environment in recent years has led to an
increase in the incidence of abnormalities. A recent study examined 384
children and found that 46 of them showed signs of abnormality. Is this
strong evidence that the risk has increased?

**Step 1. Set up the null and alternative hypothesis:**

- The null hypothesis is the current belief: $H_0 : p = p_0$

In our example it would have a form: $H_0 : p = 0.05$

- The Alternative hypothesis is what the researcher(s) want to
  prove: $H_a : p > p_0$

In our example it would have a form: $H_a : p > 0.05$

This means a one-sided test.

- The goal here is to provide evidence against $H_0$ (e.g., suggest
  $H_a$).

You want to conclude $H_a$.\
Try a Proof by Contradiction: Assume $H_0$ is true ...and hope your data
contradicts it.

**Step 2. Check the Necessary Assumptions:**

- **Independence Assumption:** There is no reason to think that one
  child having genetic abnormalities would affect the probability that
  other children have them.

- **Randomization Condition:** This sample may not be random, but
  genetic abnormalities are plausibly independent. The sample is
  probably representative of all children, with regards to genetic
  abnormalities.

- **10% Condition:** The sample of 384 children is less than 10% of all
  children.

- **Success/Failure Condition:** $np = (384)(0.05) = 19.2$ and\
  $n(1 - p) = (384)(0.95) = 364.8$ are both greater than 10, so the
  sample is large enough.

**Step 3. Identify the test-statistics. Find the value of the
test-statistic:**

Since the conditions are met, assume $H_0$ is true:\
The sampling distribution of $\hat{p}$ becomes **approximately Normal**.
That is, for large $n$, $\hat{p}$ has approximately the
$$N\left(p_0, \sqrt{\frac{p_0(1 - p_0)}{n}}\right)$$ distribution.

$$z^{*} = \frac{\hat{p} - p_0}{\sqrt{\dfrac{p_0(1 - p_0)}{n}}}
= \frac{0.1198 - 0.05}{\sqrt{\dfrac{(0.05)(0.95)}{384}}}
\approx 6.28$$

Recall that $$\hat{p} = \frac{46}{384} = 0.1198.$$

The value of $z^\ast$ is approximately 6.28, meaning that the observed
proportion of children with genetic abnormalities is over 6 standard
deviations above the hypothesized proportion ($p_0 = 0.05$).

**Step 4.** Find the p-value of the test-statistic.\
The p-value = $P(Z > 6.28) \approx 0.000$ [(better to report
$p\text{-value} < 0.0001$)]{.nodecor}\
*Note:* We find the area above $Z = 6.28$ since $H_a : p > 0.05$.\
**Meaning of this p-value:**\
If 5% of children have genetic abnormalities, the chance of observing 46
children with genetic abnormalities in a random sample of 384 children
is almost 0.

**Step 5.** Give (if any) a conclusion.\
p-value is less than 0.0001, which is less than $\alpha = 0.05$; We
reject $H_0 : p = 0.05$, and conclude $H_a : p > 0.05$. Our result is
statistically significant at $\alpha = 0.05$.\
There is very strong evidence that more than 5% of children have genetic
abnormalities.

**R code (1-sample proportion test)**

```r
    prop.test(x=46, n = 384 ,p=0.05,alternative="greater", correct=FALSE);

    ##
    ## 1-sample proportions test without continuity correction
    ##
    ## data:  46 out of 384, null probability 0.05
    ## X-squared = 39.377, df = 1, p-value = 1.747e-10
    ## alternative hypothesis: true p is greater than 0.05
    ## 95 percent confidence interval:
    ##  0.09516097 1.00000000
    ## sample estimates:
    ##        p 
    ## 0.1197917 
```
::::

::: example
Consider the following hypothesis test:

$$\begin{aligned}
H_0 &: p = 0.75 \\
H_a &: p < 0.75
\end{aligned}$$

A sample of 300 items was selected. Compute the p-value and state your
conclusion for each of the following sample results. Use
$\alpha = 0.05$.

1. $\hat{p} = 0.68$

2. $\hat{p} = 0.72$

3. $\hat{p} = 0.70$

**Solution 1.**

$$z_* = \frac{\hat{p} - p_0}{\sqrt{p_0(1 - p_0)/n}} = \frac{0.68 - 0.75}{\sqrt{0.75(1 - 0.75)/300}} = -2.80$$

Using Normal table, P-value $= P(Z < z_*) = P(Z < -2.80) = 0.0026$\
P-value $< \alpha = 0.05$, reject $H_0$.

**Solution 2.**

$$z_* = \frac{\hat{p} - p_0}{\sqrt{p_0(1 - p_0)/n}} = \frac{0.72 - 0.75}{\sqrt{0.75(1 - 0.75)/300}} = -1.20$$

Using Normal table, P-value $= P(Z < z_*) = P(Z < -1.20) = 0.1151$\
P-value $> \alpha = 0.05$, do not reject $H_0$.

**Solution 3.**

$$z_* = \frac{\hat{p} - p_0}{\sqrt{p_0(1 - p_0)/n}} = \frac{0.70 - 0.75}{\sqrt{0.75(1 - 0.75)/300}} = -2.00$$

Using Normal table, P-value $= P(Z < z_*) = P(Z < -2.00) = 0.0228$\
P-value $< \alpha = 0.05$, reject $H_0$.
:::

::: example
Consider the following hypothesis test:

$$\begin{aligned}
H_0 &: p = 0.20 \\
H_a &: p \ne 0.20
\end{aligned}$$

A sample of 400 provided a sample proportion $\hat{p} = 0.175$.

- Compute the value of the test statistic.

- What is the p-value?

- At the $\alpha = 0.05$, what is your conclusion?

- What is the rejection rule using the critical value? What is your
  conclusion?

**Solution**

- $$z_* = \frac{\hat{p} - p_0}{\sqrt{p_0(1 - p_0)/n}} = \frac{0.175 - 0.20}{\sqrt{(0.20)(0.80)/400}} = -1.25$$

- Using Normal table, P-value =
  $$2P(Z > |z_*|) = 2P(Z > |-1.25|) = 2P(Z > 1.25) = 2(0.1056) = 0.2112$$

- P-value $> \alpha = 0.05$, we CAN'T reject $H_0$.
:::

:::: example
A study found that, in 2005, 12.5% of U.S. workers belonged to unions.
Suppose a sample of 400 U.S. workers is collected in 2006 to determine
whether union efforts to organize have increased union membership.

- Formulate the hypotheses that can be used to determine whether union
  membership increased in 2006.

- If the sample results show that 52 of the workers belonged to unions,
  what is the p-value for your hypothesis test?

- At $\alpha = 0.05$, what is your conclusion?

**Solution**

- $$\begin{aligned}
  H_0 &: p = 0.125 \\
  H_a &: p > 0.125
  \end{aligned}$$

- $$\hat{p} = \frac{52}{400} = 0.13$$
  $$z_* = \frac{\hat{p} - p_0}{\sqrt{p_0(1 - p_0)/n}} = \frac{0.13 - 0.125}{\sqrt{(0.125)(0.875)/400}} = 0.30$$
  Using Normal table, P-value =
  $$P(Z > z_*) = P(Z > 0.30) = 1 - 0.6179 = 0.3821$$

- P-value $> 0.05$, do not reject $H_0$. We cannot conclude that there
  has been an increase in union membership.

**R code**

```r
    prop.test(52, 400, p=0.125, alternative="greater", correct=FALSE);

    ##
    ## 1-sample proportions test without continuity correction
    ##
    ## data:  52 out of 400, null probability 0.125
    ## X-squared = 0.091429, df = 1, p-value = 0.3812
    ## alternative hypothesis: true p is greater than 0.125
    ## 95 percent confidence interval:
    ##  0.1048085 1.0000000
    ## sample estimates:
    ##        p 
    ##    0.13 
```
::::















### On a Population Variance

In many practical situations, we are interested in testing whether the
variability in a population (i.e., its variance) has changed. This is
especially important in quality control, finance, and experimental
science. When we have data from a single normal population and want to
test a claim about the population variance, we use the chi-squared
($\chi^2$) test for one variance. This method assumes that the
underlying population is normally distributed and the sample
observations are independent.

<!-- **Hypothesis Tests for One Variance** -->

<!-- - Data from a single normal population; independent observations -->

<!-- - Variance unknown -->

<!-- - Large or small sample -->

To test $H_0 : \sigma^2 = \sigma^2_0$, our options for the null and alternative hypotheses are:

- $H_0: \sigma^2 = \sigma^2_0$ vs. $H_a: \sigma^2 > \mu_0$

- $H_0: \sigma^2 = \sigma^2_0$ vs. $H_a: \sigma^2 < \mu_0$

- $H_0: \sigma^2 = \sigma^2_0$ vs. $H_a: \sigma^2 \neq \mu_0$

We calculate the $\chi^2$-statistic:

$$\chi^2_* = \frac{(n - 1)s^2}{\sigma_0^2} \sim \chi^2_{n - 1}$$

In terms of a variable $\chi^2_{n-1}$ which follows a $t$-distribution at $n-1$ degrees of freedom, the
p-value for a test of $H_0$ against:

- $H_a: \sigma^2 > \sigma_0$ is $P(\chi^2 > \chi^2_*)$

- $H_a: \sigma^2 < \sigma_0$ is $P(\chi^2 < \chi^2_*)$

- $H_a: \sigma^2 \ne \sigma_0$ is $2P(\chi^2 > |\chi^2_*|)$




<!-- **Hypothesis Test**\ -->
<!-- $$\begin{aligned} -->
<!-- H_0 &: \sigma^2 = \sigma_0^2 \\ -->
<!-- H_a &: \sigma^2 \ne \sigma_0^2 \quad \text{(or } \sigma^2 > \sigma_0^2 \text{ or } \sigma^2 < \sigma_0^2\text{)} -->
<!-- \end{aligned}$$ Assume $H_0$ is true, then: -->

<!-- $$\text{Test statistic:} \quad \chi^2_* = \frac{(n - 1)s^2}{\sigma_0^2} \sim \chi^2_{n - 1}$$ -->

<!-- ::: tcolorbox -->
<!-- **Decision rules:** -->

<!-- $H_a : \sigma^2 \ne \sigma_0^2$.\ -->
<!-- Reject $H_0$ if $\chi^2_* > \chi^2_{n-1;\alpha/2}$ or if -->
<!-- $\chi^2_* < \chi^2_{n-1;1-\alpha/2}$. -->

<!-- $H_a : \sigma^2 > \sigma_0^2$.\ -->
<!-- Reject $H_0$ if $\chi^2_* > \chi^2_{n-1;\alpha}$ or if -->
<!-- $P[\chi^2_{n-1} > \chi^2_*]$ is too small. -->

<!-- $H_a : \sigma^2 < \sigma_0^2$.\ -->
<!-- Reject $H_0$ if $\chi^2_* < \chi^2_{n-1;1-\alpha}$ or if -->
<!-- $P[\chi^2_{n-1} < \chi^2_*]$ is too small. -->


::: {.remark}
This test is not robust to departures from Normality.
:::

:::: {.example}
A company produces metal pipes of a standard length, and claims that the
standard deviation of the length is at most 1.2 cm. One of its clients
decides to test this claim by taking a sample of 25 pipes and checking
their lengths. They found that the standard deviation of the sample is
1.5 cm. Does this undermine the company's claim? Use $\alpha = 0.05$.\
*Note: Assume length is Normally distributed.*

**Solution**

$$\begin{aligned}
H_0 &: \sigma^2 \leq 1.2^2 \\
H_a &: \sigma^2 > 1.2^2
\end{aligned}$$

$$\chi^2_* = \frac{(n-1)s^2}{\sigma^2} = \frac{(25-1) \cdot 1.5^2}{1.2^2} = 37.5$$

$$\text{P-value} = P[\chi^2_{24} > 37.5] \approx 0.0389$$

**R Code**

```r
    1 - pchisq(37.5, df = 24);
    ## [1] 0.0389818
```

**Conclusion**\
We reject $H_0 : \sigma^2 \leq 1.2^2$. We have evidence to indicate that
the variance of the length of metal pipes is more than $1.2^2$.
::::

<!-- ::: tcolorbox -->
<!-- **Assumptions:** $Y_1, Y_2, \ldots, Y_n$ constitute a random sample from -->
<!-- a Normal distribution with $E(Y_i) = \mu$ and $V(Y_i) = \sigma^2$. -->

<!-- **Hypotheses:** -->

<!-- \[ -->
<!-- \begin{aligned} -->
<!-- & H_0 : \sigma^2 = \sigma_0^2 \\ -->
<!-- & H_a : -->
<!-- \begin{cases} -->
<!-- \sigma^2 > \sigma_0^2 & \text{(upper-tailed alternative)} \\ -->
<!-- \sigma^2 < \sigma_0^2 & \text{(lower-tailed alternative)} \\ -->
<!-- \sigma^2 \ne \sigma_0^2 & \text{(two-tailed alternative)} -->
<!-- \end{cases} -->
<!-- \end{aligned} -->
<!-- \] -->

<!-- **Test statistic:** -->
<!-- \[ -->
<!-- \chi^2 = \frac{(n - 1)S^2}{\sigma_0^2} -->
<!-- \] -->

<!-- **Rejection Region:** -->

<!-- \[ -->
<!-- \begin{aligned} -->
<!-- & \chi^2 > \chi^2_\alpha && \text{(upper-tailed RR)} \\ -->
<!-- & \chi^2 < \chi^2_{1-\alpha} && \text{(lower-tailed RR)} \\ -->
<!-- & \chi^2 > \chi^2_{\alpha/2} \quad \text{or} \quad \chi^2 < \chi^2_{1-\alpha/2} && \text{(two-tailed RR)} -->
<!-- \end{aligned} -->
<!-- \] -->


<!-- ::: -->

::: example
A manufacturer of car batteries claims that the life of his batteries is
approximately Normally distributed with a standard deviation equal to
0.9 year. If a random sample of 10 of these batteries has a standard
deviation of 1.2 years, do you think that $\sigma > 0.9$ year? Use a
0.05 level of significance.

**Step 1. State hypotheses.** $$\begin{aligned}
H_0 &: \sigma^2 = 0.81 \\
H_a &: \sigma^2 > 0.81
\end{aligned}$$

**Step 2. Compute test statistic.**\
$S^2 = 1.44$, $n = 10$, and $$\chi^2 = \frac{(9)(1.44)}{0.81} = 16$$

**Step 3. Find Rejection Region.**\
From the chi-squared table, the null hypothesis is rejected when
$\chi^2 > 16.919$, where $\nu = 9$ degrees of freedom.

```{r fig.cap="Right-tailed chi-squared distribution with critical value at 16.919", echo=FALSE, fig.align='center', out.width='70%', warning=FALSE, message=FALSE}
library(ggplot2)

# Parameters
df <- 9
alpha <- 0.05
x_crit <- qchisq(1 - alpha, df = df)

# Data
x <- seq(0, 30, length.out = 500)
y <- dchisq(x, df = df)
curve_df <- data.frame(x = x, y = y)

# Rejection region (shaded)
x_fill <- seq(x_crit, 30, length.out = 200)
y_fill <- dchisq(x_fill, df = df)
fill_df <- data.frame(x = x_fill, y = y_fill)

# Plot
ggplot(curve_df, aes(x = x, y = y)) +
  geom_line(color = "#000000", size = 1.2) +  # chi-squared curve
  geom_area(data = fill_df, aes(x = x, y = y),
            fill = "#619CFF", alpha = 0.5) +  # rejection region
  geom_vline(xintercept = x_crit, color = "#619CFF", linetype = "solid", size = 1) +
  annotate("text", x = x_crit, y = -0.003, label = "16.919",
           size = 4, vjust = 1, color = "#000000") +
  annotate("text", x = x_crit + 4, y = 0.015,
           label = expression(alpha == 0.05), parse = TRUE,
           size = 5, color = "#619CFF") +
  scale_x_continuous(limits = c(0, 30)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank()
  )
```




**Step 4. Conclusion.**\
The $\chi^2$ statistic is not significant at the 0.05 level. We conclude
that there is insufficient evidence to claim that $\sigma > 0.9$ year.
:::












## Two Sample Hypothesis Tests

We now move to examining hypothesis tests on 2 independent samples.
We will use the notation

<table class="param-table">
  <tr>
    <pd class="term">$\Theta_1$</td>
    <pd class="sep"> : </pd>
    <pd class="alt">Parameter of interest from Population 1.</pd>
  </tr>
</table>

<table class="param-table">
  <tr>
    <pd class="term">$\Theta_2$</td>
    <pd class="sep"> : </pd>
    <pd class="alt">Parameter of interest from Population 2.</pd>
  </tr>
</table>

<table class="param-table">
  <tr>
    <pd class="term">$\Theta_0$</td>
    <pd class="sep"> : </pd>
    <pd class="alt">Numerical value hypothesized under the null hypothesis.</pd>
  </tr>
</table>


Hypothesis tests of interest on 2 samples are


<table class="hypothesis-table">
  <tr>
    <td class="term">$H_0: \Theta_1 - \Theta_2 = \Theta_0$</td>
    <td class="sep">vs.</td>
    <td class="alt">$H_a: \Theta_1 - \Theta_2 > \Theta_0$</td>
  </tr>
  <tr>
    <td class="term">$H_0: \Theta_2 - \Theta_2 = \Theta_0$</td>
    <td class="sep">vs.</td>
    <td class="alt">$H_a: \Theta_1 - \Theta_2 < \Theta_0$</td>
  </tr>
  <tr>
    <td class="term">$H_0: \Theta_1 - \Theta_2 = \Theta_0$</td>
    <td class="sep">vs.</td>
    <td class="alt">$H_a: \Theta_1 - \Theta_2 \neq \Theta_0$</td>
  </tr>
</table>


Some other potential hypothesis tests of interest on 2 samples are

<table class="hypothesis-table">
  <tr>
    <td class="term">$H_0: \displaystyle\frac{\Theta_1}{\Theta_2} = \Theta_0$</td>
    <td class="sep">vs.</td>
    <td class="alt">$H_a: \displaystyle\frac{\Theta_1}{\Theta_2} > \Theta_0$</td>
  </tr>
  <tr>
    <td class="term">$H_0: \displaystyle\frac{\Theta_1}{\Theta_2} = \Theta_0$</td>
    <td class="sep">vs.</td>
    <td class="alt">$H_a: \displaystyle\frac{\Theta_1}{\Theta_2} < \Theta_0$</td>
  </tr>
  <tr>
    <td class="term">$H_0: \displaystyle\frac{\Theta_1}{\Theta_2} = \Theta_0$</td>
    <td class="sep">vs.</td>
    <td class="alt">$H_a: \displaystyle\frac{\Theta_1}{\Theta_2} \neq \Theta_0$</td>
  </tr>
</table>


::: {.remark}
A very common test is to determine whether $\Theta_1 = \Theta_2$.
Therefore, testing
<table class="hypothesis-table">
  <tr>
    <td class="sep">$H_0: \Theta_1 - \Theta_2 = \Theta_0$</td>
  </tr>
</table>
is equivalent to setting $\Theta_0 = 0$ and writing the null hypothesis as
<table class="hypothesis-table">
  <tr>
    <td class="sep">$H_0: \Theta_1 = \Theta_2$</td>
  </tr>
</table>

:::


### On a Difference of Means

In this section, our parameter of interest is $\mu_1 - \mu_2$
where $\mu_1$ is the mean from one population and 
$\mu_2$ is the mean from another population.



#### When $\sigma_1$ and $\sigma_2$ are Known

To test $H_0 : \mu_1 - \mu_2 = \mu_0$, when population variances are unknown and assumed equal, 
our options for the null and alternative hypotheses are:

- $H_0: \mu_1 - \mu_2 = \mu_0$ vs. $H_a: \mu_1 - \mu_2 > \mu_0$

- $H_0: \mu_1 - \mu_2 = \mu_0$ vs. $H_a: \mu_1 - \mu_2 < \mu_0$

- $H_0: \mu_1 - \mu_2 = \mu_0$ vs. $H_a: \mu_1 - \mu_2 \neq \mu_0$

We calculate the $z$-statistic:
$$z = \frac{\bar{x}_1 - \bar{x}_2 - \mu_0}{\sqrt{\displaystyle\frac{\sigma_1^2}{n_1} + \frac{\sigma_2^2}{n_2}}}$$




#### When $\sigma_1$ and $\sigma_2$ are Not Known


#### When $\sigma_1 = \sigma_2$

To test $H_0 : \mu_1 - \mu_2 = \mu_0$, when population variances are unknown and assumed equal, 
our options for the null and alternative hypotheses are:

- $H_0: \mu_1 - \mu_2 = \mu_0$ vs. $H_a: \mu_1 - \mu_2 > \mu_0$

- $H_0: \mu_1 - \mu_2 = \mu_0$ vs. $H_a: \mu_1 - \mu_2 < \mu_0$

- $H_0: \mu_1 - \mu_2 = \mu_0$ vs. $H_a: \mu_1 - \mu_2 \neq \mu_0$

We calculate the $t$-statistic:

$$z = \frac{\bar{x}_1 - \bar{x}_2 - \mu_0}{\sqrt{\displaystyle\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}} \sim t_{\min(n_1 - 1, \, n_2 - 1)}$$
where the pooled standard deviation is defined as:
$$s_p^2 = \frac{(n_1 - 1)s_1^2 + (n_2 - 1)s_2^2}{n_1 + n_2 - 2}$$

In terms of a variable $T$ which follows a standard normal distribution, 
the p-value for a test of $H_0$ against:

- $H_a: \mu_1 - \mu_2 > \mu_0$ is $P(T > z^{*})$

- $H_a: \mu_1 - \mu_2 < \mu_0$ is $P(T < z^{*})$

- $H_a: \mu_1 - \mu_2 \neq \mu_0$ is $2P(T > |z^{*}|)$





:::::::::::::::::::: example
Comparing Two Population Means Managerial Success Indexes for Two Groups
(With Equal Variances Assumed)

Behavioural researchers have developed an index designed to measure
managerial success. The index (measured on a 100-point scale) is based
on the manager's length of time in the organization and their level
within the term; the higher the index, the more successful the manager.
Suppose a researcher wants to compare the average index for the two
groups of managers at a large manufacturing plant. Managers in group 1
engage in **high volume of interactions** with people outside the
managers' work unit (such interaction include phone and face-to-face
meetings with customers and suppliers, outside meetings, and public
relation work). Managers in group 2 **rarely interact** with people
outside their work unit. Independent random samples of 12 and 15
managers are selected from groups 1 and 2, respectively, and success
index of each is recorded.

**Response variable:** Managerial Success Indexes (quantitative,
continuous, 0--100 scale)

**Explanatory variable:** Type of group (nominal categorical: *Group 1 =
interaction with outsiders*, *Group 2 = fewer interactions*)

**R Code**
```r
    # Importing data file into R
    success = read.csv(file = "success.csv", header = TRUE);

    # Getting names of variables
    names(success);

    # Seeing first few observations
    head(success);

    # Attaching data file
    attach(success);
```

**R Code**

```r
    ## [1] "Success_Index" "Group"
    ##   Success_Index Group
    ## 1            65     1
    ## 2            66     1
    ## 3            58     1
    ## 4            70     1
    ## 5            78     1
    ## 6            53     1
```

**R code (Descriptive Statistics)**

```r
    # loading library mosaic
    library(mosaic)

    favstats(Success_Index ~ Group)
```

**R Code (Descriptive Statistics)**

```r
    ## .group min   Q1 median   Q3 max     mean       sd  n
    ##       1  53 62.25  65.50 69.25  78 65.33333 6.610368 12
    ##       2  34 42.50  50.00 54.50  68 49.46667 9.334014 15
```

**R Code (Descriptive Statistics)**

```r
    summary(Success_Index[Group == 1]);
    length(Success_Index[Group == 1]);
    sd(Success_Index[Group == 1]);

    summary(Success_Index[Group == 2]);
    length(Success_Index[Group == 2]);
    sd(Success_Index[Group == 2]);
```

Note: Group 1 = "interaction with outsiders" and Group 2 = "fewer
interactions".

**R Output**

```r
    ##  Min.  1st Qu.  Median    Mean  3rd Qu.    Max. 
    ##  53.00   62.25   65.50   65.33   69.25   78.00 
    ## [1] 12
    ## [1] 6.610368

    ##  Min.  1st Qu.  Median    Mean  3rd Qu.    Max. 
    ##  34.00   42.50   50.00   49.47   54.50   68.00 
    ## [1] 15
    ## [1] 9.334014
```

**Nearly Normal Condition (Group 1: "interaction with outsiders"):**

```r
    stem(Success_Index[Group == 1]);
```

**R Output**

```r
    ## 
    ##  The decimal point is 1 digit(s) to the right of the |
    ## 
    ##  5 | 38
    ##  6 | 0335689
    ##  7 | 018
```

**Nearly Normal Condition (Group 2: "fewer interactions"):**

```r
    stem(Success_Index[Group == 2]);
```

**R Output**

```r
    ## 
    ##  The decimal point is 1 digit(s) to the right of the |
    ## 
    ##  3 | 46
    ##  4 | 22368
    ##  5 | 023367
    ##  6 | 28
```

**Nearly Normal Condition (Group 1: "interaction with outsiders"):**

```r
    qqnorm(Success_Index[Group == 1]);
    qqline(Success_Index[Group == 1]);
```
```{r fig.cap="Q-Q Plot for Group 1: Interaction with Outsiders", echo=FALSE, fig.align='center', out.width='60%', warning=FALSE, message=FALSE}
library(ggplot2)

# Group 1 data
success_group1 <- c(65, 66, 58, 70, 78, 53, 62, 64, 68, 69, 60, 67)

# Create data frame for ggplot Q-Q plot
df <- data.frame(success = success_group1)

# Q-Q plot with ggplot2
ggplot(df, aes(sample = success)) +
  stat_qq(color = "#619CFF", size = 2) +
  stat_qq_line(color = "#000000") +
  labs(
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_minimal(base_size = 14)
```

**Nearly Normal Condition (Group 2: "fewer interactions"):**

```r
    qqnorm(Success_Index[Group == 2]);
    qqline(Success_Index[Group == 2]);
```

```{r fig.cap="Q-Q Plot for Group 2: Fewer Interactions", echo=FALSE, fig.align='center', out.width='60%', warning=FALSE, message=FALSE}
library(ggplot2)

# Group 2 data
success_group2 <- c(34, 40, 45, 46, 50, 42, 44, 48, 54, 49, 52, 53, 55, 60, 68)

# Data frame for plotting
df <- data.frame(success = success_group2)

# ggplot Q-Q plot
ggplot(df, aes(sample = success)) +
  stat_qq(color = "#619CFF", size = 2) +
  stat_qq_line(color = "#000000") +
  labs(
    x = "Theoretical Quantiles",
    y = "Sample Quantiles"
  ) +
  theme_minimal(base_size = 14)
```

:::

**Nearly Normal Condition:**

```r
    boxplot(Success_Index ~ Group, col = c("red", "blue"))
```

```{r fig.cap="Boxplot of Success Index by Group", echo=FALSE, fig.align='center', out.width='60%', warning=FALSE, message=FALSE}
library(ggplot2)

# Group 1 and Group 2 data
group1 <- c(65, 66, 58, 70, 78, 53, 62, 64, 68, 69, 60, 67)
group2 <- c(34, 40, 45, 46, 50, 42, 44, 48, 54, 49, 52, 53, 55, 60, 68)

# Create data frame
df <- data.frame(
  Success_Index = c(group1, group2),
  Group = factor(c(rep(1, length(group1)), rep(2, length(group2))),
                 labels = c("Group 1", "Group 2"))
)

# Create boxplot using ggplot2
ggplot(df, aes(x = Group, y = Success_Index, fill = Group)) +
  geom_boxplot() +
  scale_fill_manual(values = c("#F8766D", "#619CFF")) +
  labs(x = "Group", y = "Success Index") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```


[]{#fig:boxplot-success label="fig:boxplot-success"}
:::

**Boxplot with `ggplot2`:**

```r
    # loading library;
    library(ggplot2);

    # converting a numeric variable into factor (categorical data)
    group <- factor(Group);

    # bp: just a name (not code) to store boxplots;
    bp <- ggplot(success,
                 aes(x = group, y = Success_Index, fill = group));

    our.labs <- c("Interaction with Outsiders", "Fewer Interactions");

    bp +
      geom_boxplot() +
      scale_x_discrete(labels = our.labs);
```
      




**Checking the Assumptions and Conditions**

**Independent Group Assumption:** The success index in group 1 is
unrelated to the success index in group 2.

**Randomization Condition:** The 27 managers were randomly and
independently selected (12 for group 1, and 15 for group 2).

**Nearly Normal Condition:** The two boxplots of success indexes do not
show skewness; the two stemplots/histograms of success indexes are
unimodal, fairly symmetric and approximately bell-shaped. Q--Q plots
also suggest the normality assumption is reasonable.

**Equal Variances Assumption:** The two boxplots of success indexes
appear to have the same spread; thus, the samples appear to have come
from populations with approximately the same variance.

Since the conditions are satisfied, it is appropriate to construct a $t$
confidence interval with\
$df = 12 + 15 - 2 = 25$.\
**From the data, the following statistics were calculated:**

$$\begin{aligned}
n_1 &= 12 &\quad n_2 &= 15 \\
\bar{x}_1 &= 65.33 &\quad \bar{x}_2 &= 49.47 \\
s_1^2 &= 6.61^2 &\quad s_2^2 &= 9.33^2
\end{aligned}$$

**The pooled variance estimator is:**

$$s_p^2 = \frac{(n_1 - 1)s_1^2 + (n_2 - 1)s_2^2}{n_1 + n_2 - 2} 
= \frac{(12 - 1)(6.61^2) + (15 - 1)(9.33^2)}{12 + 15 - 2}
= 67.97$$

**The number of degrees of freedom is:**
$$\nu = n_1 + n_2 - 2 = 12 + 15 - 2 = 25$$










#### When $\sigma_1 \neq \sigma_2$

To test $H_0 : \mu_1 - \mu_2 = \mu_0$, when population variances are unknown and assumed different, 
our options for the null and alternative hypotheses are:

- $H_0: \mu_1 - \mu_2 = \mu_0$ vs. $H_a: \mu_1 - \mu_2 > \mu_0$

- $H_0: \mu_1 - \mu_2 = \mu_0$ vs. $H_a: \mu_1 - \mu_2 < \mu_0$

- $H_0: \mu_1 - \mu_2 = \mu_0$ vs. $H_a: \mu_1 - \mu_2 \neq \mu_0$

We calculate the $t$-statistic:

$$t^* = \frac{\bar{x}_1 - \bar{x}_2 - \mu_0}{\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}} \sim t_{\min(n_1 -1, n_2 -1)}$$

In terms of a variable $T$ which follows a standard normal distribution, 
the p-value for a test of $H_0$ against:

- $H_a: \mu_1 - \mu_2 > \mu_0$ is $P(T > z^{*})$

- $H_a: \mu_1 - \mu_2 < \mu_0$ is $P(T < z^{*})$

- $H_a: \mu_1 - \mu_2 \neq \mu_0$ is $2P(T > |z^{*}|)$

::: {.remark}
Note a more accurate approximation of the degrees of freedom is given by
$$df = \frac{\left( \frac{s_1^2}{n_1} + \frac{s_2^2}{n_2} \right)^2}
{
\left( \frac{1}{n_1 - 1} \left( \frac{s_1^2}{n_1} \right)^2 \right)
+
\left( \frac{1}{n_2 - 1} \left( \frac{s_2^2}{n_2} \right)^2 \right)
}$$

This approximation is accurate when both sample sizes $n_1$ and $n_2$
are 5 or larger.
:::



::: example
"Conservationists have despaired over destruction of tropical rain
forest by logging, clearing, and burning". These words begin a report on
a statistical study of the effects of logging in Borneo. Here are data
on the number of tree species in 12 unlogged forest plots and 9 similar
plots logged 8 years earlier:

Unlogged: 22 18 22 20 15 21 13 13 19 13 19 15\
Logged : 17 4 18 14 18 15 15 10 12

Does logging significantly reduce the mean number of species in a plot
after 8 years? State the hypotheses and do a t test. Is the result
significant at the 5% level?\
**Solution:**\
1. State hypotheses. $H_0: \mu_1 = \mu_2$ vs. $H_a: \mu_1 > \mu_2$,
where $\mu_1$ is the mean number of species in unlogged plots and
$\mu_2$ is the mean number of species in plots logged 8 years earlier.

2\. Test statistic.
$$t^* \;=\;\frac{\bar x_1 - \bar x_2}{\sqrt{s_1^2/n_1 + s_2^2/n_2}}
\;=\;2.1140$$
$$(\bar x_1 = 17.5,\;\bar x_2 = 13.6666,\;s_1 = 3.5290,\;s_2 = 4.5,\;n_1 = 12,\;n_2 = 9)$$

3\. P-value. Using Table, we have $df = 8$, and
$0.025 < \text{P-value} < 0.05$.

4\. Conclusion. Since P-value $< 0.05$, we reject $H_0$. There is strong
evidence that the mean number of species in unlogged plots is greater
than that for logged plots 8 years after logging.
:::

:::::::::::::::: example
A company that sells educational materials reports statistical studies
to convince customers that its materials improve learning. One new
product supplies "directed reading activities" for classroom use. These
activities should improve the reading ability of elementary school
pupils.

A consultant arranges for a third-grade class of 21 students to take
part in these activities for an eight-week period. A control classroom
of 23 third-graders follows the same curriculum without the activities.
At the end of the eight weeks, all students are given a Degree of
Reading Power (DRP) test, which measures the aspects of reading ability
that the treatment is designed to improve. The data appear in the
following table.

:::{.smalltbl}
+---------------------------+-------------------------------+
| **Treatment**             | **Control**                   |
+:====:+:====:+:====:+:====:+:=====:+:=====:+:=====:+:=====:+
| 24   | 61   | 59   | 46   | 42    | 33    | 46    | 37    |
+------+------+------+------+-------+-------+-------+-------+
| 43   | 44   | 52   | 43   | 43    | 41    | 10    | 42    |
+------+------+------+------+-------+-------+-------+-------+
| 58   | 67   | 62   | 57   | 55    | 19    | 17    | 55    |
+------+------+------+------+-------+-------+-------+-------+
| 71   | 49   | 54   |      | 26    | 54    | 60    | 28    |
+------+------+------+------+-------+-------+-------+-------+
| 43   | 53   | 57   |      | 62    | 20    | 53    | 48    |
+------+------+------+------+-------+-------+-------+-------+
| 49   | 56   | 33   |      | 37    | 85    | 42    |       |
+------+------+------+------+-------+-------+-------+-------+
:::

Because we hope to show that the treatment (Group 1) is better than the
control (Group 2), the hypotheses are:

$$H_0 : \mu_1 = \mu_2$$ $$H_a : \mu_1 > \mu_2$$

```r
    # Step 1. Entering data
    treatment = c(24, 61, 59, 46, 43, 44, 52, 43, 58, 67, 62, 57, 
                  71, 49, 54, 43, 53, 57, 49, 56, 33);
    control = c(42, 33, 46, 37, 43, 41, 10, 42, 55, 19, 17, 55, 
                26, 54, 60, 28, 62, 20, 53, 48, 37, 85, 42);
```

**Nearly Normal Condition (treatment):**

```r
    # Making stemplot;

    stem(treatment);
```

```r
    ##
    ## The decimal point is 1 digit(s) to the right of the |
    ##
    ## 2 | 4
    ## 3 | 3
    ## 4 | 3334699
    ## 5 | 23467789
    ## 6 | 127
    ## 7 | 1
```

**Nearly Normal Condition (control):**

```r
    # Making stemplot;

    stem(control);
```

```r
    ##
    ## The decimal point is 1 digit(s) to the right of the |
    ##
    ## 0 | 079
    ## 2 | 068377
    ## 4 | 12223683455
    ## 6 | 02
    ## 8 | 5
```

**Nearly Normal Condition (treatment):**

```r
    # Making Q-Q plot;
    qqnorm(treatment, pch=19, col="red", main="Treatment");
    qqline(treatment, lty=2);
```
```{r fig.cap="Q-Q Plot for Treatment Group", echo=FALSE, fig.align='center', out.width='60%', warning=FALSE, message=FALSE}
library(ggplot2)

# Step 1: Enter data
treatment <- c(24, 61, 59, 46, 43, 44, 52, 43, 58, 67, 62, 57,
               71, 49, 54, 43, 53, 57, 49, 56, 33)

# Step 2: Create data frame
df <- data.frame(value = treatment)

# Step 3: Generate ggplot Q-Q plot
ggplot(df, aes(sample = value)) +
  stat_qq(color = "#F8766D", size = 2) +
  stat_qq_line(color = "#619CFF", linetype = "dashed") +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal(base_size = 14)
```


**Nearly Normal Condition (control):**

```r
    # Making Q-Q plot;
    qqnorm(treatment, pch=19, col="red", main="Control");
    qqline(control, lty=2);
```

```{r fig.cap="Q-Q Plot for Control Group", echo=FALSE, fig.align='center', out.width='60%', warning=FALSE, message=FALSE}
library(ggplot2)

# Step 1: Enter data
control <- c(42, 33, 46, 37, 43, 41, 10, 42, 55, 19, 17, 55,
             26, 54, 60, 28, 62, 20, 53, 48, 37, 85, 42)

# Step 2: Create data frame
df <- data.frame(value = control)

# Step 3: Generate Q-Q plot using ggplot2
ggplot(df, aes(sample = value)) +
  stat_qq(color = "#F8766D", size = 2) +
  stat_qq_line(color = "#619CFF", linetype = "dashed") +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal(base_size = 14)
```

Stemplots suggest that there is a mild outlier in the control group but
no deviation from Normality serious enough to prevent us from using t
procedures. Normal Q-Q plots for both groups confirm that both are
roughly Normal. The summary statistics are:

```r
    summary(treatment);

    ## Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
    ## 24.00   44.00   53.00   51.48   58.00   71.00

    summary(control);

    ## Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
    ## 10.00   30.50   42.00   41.52   53.50   85.00
```

```r
    # Step 1. Entering data;

    treatment = c(24, 61, 59, 46, 43, 44, 52, 43, 58, 67, 62, 57, 
                  71, 49, 54, 43, 53, 57, 49, 56, 33);
                  
    control = c(42, 33, 46, 37, 43, 41, 10, 42, 55, 19, 17, 55, 
                26, 54, 60, 28, 62, 20, 53, 48, 37, 85, 42);

    # Step 2. Hypothesis Test

    t.test(treatment, control, alternative="greater")
```

**Hypothesis Test (using R)**

```r
    ##
    ##  Welch Two Sample t-test
    ## 
    ## data:  treatment and control
    ## t = 2.3106, df = 37.855, p-value = 0.01305
    ## alternative hypothesis: true difference in means is greater than 0
    ## 95 percent confidence interval:
    ##  2.333784      Inf
    ## sample estimates:
    ## mean of x mean of y 
    ##    51.47619 41.52174 
```

**Hypothesis Test (using table)**

```r
    round (mean(treatment) ,2);

    ##[1] 51.48

    round (sd(treatment) ,2);

    ##[1] 11.01

    round (mean(control) ,2);

    ##[1] 41.52

    round (sd(control) ,2);

    ##[1] 17.15
```

Test statistic.\
$$t^* = \frac{\bar{x}_1 - \bar{x}_2}{\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}} = 2.31$$
$$\quad
(\bar{x}_1 = 51.48, \, \bar{x}_2 = 41.52, \, s_1 = 11.01, \, s_2 = 17.15,
\quad n_1 = 21 \text{ and } n_2 = 23)$$

The conservative approach uses the *t*(20) distribution. The P-value for
the one-sided test is $$\text{P-value} = P\bigl(T \geq 2.31 \bigr)$$
Comparing $t = 2.31$ with the entries in Table 5 for 20 degrees of
freedom, we see that $$0.01 < \text{P-value} < 0.025.$$

Since our P-value is "small", we reject the null hypothesis (Note that
we would reject $H_0$ at the 2.5% significance level). The data strongly
suggest that directed reading activity improves the DRP score.

The design of the DRP study is not ideal. Random assignment of students
was not possible in a school environment, so existing third-grade
classes were used. The effect of the reading programs is therefore
confounded with any other differences between the two classes. The
classes were chosen to be as similar as possible in variables such as
the social and economic status of the students. Pretesting showed that
the two classes were on the average quite similar in reading ability at
the beginning of the experiment. To avoid the effect of two different
teachers, the same teacher taught reading in both classes during the
eight-week period of the experiment. We can therefore be somewhat
confident that our two-sample procedure is detecting the effect of the
treatment and not some other difference between the classes.
:::

<!-- ::: nt -->
<!-- **Why do we use hypothesis tests?** -->

<!-- To conduct tests on parameters and to quantify the degree of certainty -->
<!-- with probability. -->

<!-- The smaller the p-value, the stronger the evidence against $H_0$ -->
<!-- (provided assumptions are met). -->
<!-- ::: -->

<!-- **Understanding Significance Levels** -->

<!-- Suppose you have the following limited information: -->

<!-- - A test was statistically significant at the 5% level. -->

<!-- We **cannot** make this conclusion for certain. That is, -->
<!-- $$p\text{-value} < 0.05 \quad \not\Rightarrow \quad p\text{-value} < 0.01.$$ -->

<!-- - A test was statistically significant at the 1% level. -->

<!-- Yes! Because $$p\text{-value} < 0.01 < 0.05.$$ -->


::: {.remark}
**The Fold Rule**

A rule which can be used to quickly determine whether population
variances are equal or unequal using sample variances.

If $$\frac{\max(s_1, s_2)}{\min(s_1, s_2)} < \sqrt{2} 
\quad \text{then we can consider } \sigma_1^2 = \sigma_2^2$$

and

$$\frac{\max(s_1^2, s_2^2)}{\min(s_1^2, s_2^2)} < 2 
\quad \text{then we can consider } \sigma_1^2 = \sigma_2^2$$

This is quick and simple technique. 
It is only a rule and not as strong as conducting the hypothesis tests for equality of varianges

$$H_0: \sigma_1^2 = \sigma_2^2 \quad \text{vs} \quad H_a: \sigma_1^2 \neq \sigma_2^2$$
which was covered in Section \@ref(sec:RatioVars)
:::

#### Assumptions












### On a Difference of Proportions

We can use hypothesis test to compare proportions from two
independent groups as well. 

To test $H_0 : p_1 - p_2 = p_0$, our options for the null and alternative hypotheses are:

- $H_0: p_1 - p_2 = p_0$ vs. $H_a: p_1 - p_2 > p_0$

- $H_0: p_1 - p_2 = p_0$ vs. $H_a: p_1 - p_2 < p_0$

- $H_0: p_1 - p_2 = p_0$ vs. $H_a: p_1 - p_2 \neq p_0$


We calculate the $z$-statistic:

$$z_* = \frac{\hat{p_1} - \hat{p_2} - p_0}{\sqrt{\hat{p}(1-\hat{p})\left(\displaystyle\frac{1}{n_1}+ \frac{1}{n_2}\right)}} \sim N(0,1).$$

where

$$\hat{p} = \frac{x_1 + x_2}{n_1+n_2}$$ 

and $n_1$ and $n_2$ represent sample sizes of each group.

In terms of a variable $Z$ which follows a standard normal distribution, 
the p-value for a test of $H_0$ against:

- $H_a: p_1 - p_2 > p_0$ is $P(Z > z^{*})$

- $H_a: p_1 - p_2 < p_0$ is $P(Z < z^{*})$

- $H_a: p_1 - p_2 \neq p_0$ is $2P(Z > |z^{*}|)$


::::: example
Nicotine patches are often used to help smokers quit. Does giving
medicine to fight depression help? A randomized double-blind experiment
assigned 244 smokers who wanted to stop to receive nicotine patches and
another 245 to receive both a patch and the anti-depression drug
bupropion. Results: After a y ear, 40 subjects in the nicotine patch
group had abstained from smoking, as had 87 in the patch-plus-drug
group. How significant is the evidence that the medicine increases the
success rate? State hypotheses, calculate a test statistic, use Table 6
to give its P-value, and state y our conclusion. (Use $\alpha = 0.01$)\
Solution:

**Step 1: State Hypothesis**

$H_0: p_1 = p_2$ and $H_a: p_1 < p_2$

**Step 2: Find test statistics**

$\hat{p_1} = \frac{40}{244} = 0.1639$ and
$\hat{p_2} = \frac{87}{245} = 0.3551$. Then
$\hat{p} = \frac{40+87}{244+245} = 0.2597$.

Now,
$z_* = \frac{\hat{p_1} - \hat{p_2}}{\sqrt{\hat{p}(1-\hat{p})(\frac{1}{n_1} + \frac{1}{n_2})}} = -4.82$

**Step 3: Compute p-value**

P-value $= P(Z < z_*) = P(Z < -4.82) < 0.0003$

**Step 4: Conclusion**

Since p-value $< 0.0003 < \alpha = 0.01$, we reject the null hypothesis
that $p_1 = p_2$. The data provide very strong evidence that bupropion
increases success rate.

**R-code:**

Input

::: tcolorbox
    successes=c(87, 40);

    totals=c(245, 244);

    prop.test(successes, totals, alternative="greater",correct=FALSE);
:::

Output

::: tcolorbox

    ## 
    ## 2-sample test for equality of proportions without 
    ## continuity correction 
    ## 
    ## data: x and n
    ## X-squared = 23.237, df = 1, p-value = 7.161e-07
    ## alternative hypothesis: greater 
    ## 95 percent confidence interval: 
    ## 0.1275385 1.0000000 
    ## sample estimates:
    ##    prop 1    prop 2 
    ## 0.3551020 0.1639344
:::
:::::

#### Assumptions

- Independent Response Assumption: Within each group , we need
independent responses from the cases. We cannot check that for certain,
but randomization provides evidence of independence. So, we need to
check the following:

  - Randomization Condition: The data in each group should be drawn
  independently and at random from a population or generated by a
  completely randomized designed experiment.

  - The 10 % Condition: If the data are sampled without replacement, the
  sample should not exceed 10 % of the population. If samples are bigger
  than 10 % of the target population, random draws are no longer
  approximately independent.

  - Independent Groups Assumption: The two groups we are comparing must be
  independent from each other.

- Sample Size Condition Each of the groups must be big enough. As
with individual proportions, we need larger group s to estimate
proportions that are near 0% and 100%. We check the success / failure
condition for each group.

  - Success / Failure Condition: Both groups are big enough that at least
  10 successes and at least 10 failures have been observed in each group
  or will be expected in each (when testing hypothesis).


<!-- Two-sided significance tests (later we will discuss this concept) -->
<!-- are robust against violations of this condition. In this case, we can -->
<!-- conduct significance tests with smaller sample sizes. In practice, the -->
<!-- two-sided significance test works well if there are at least five -->
<!-- successes and five failures in each sample. -->










### On a Ratio of Variances {#sec:RatioVars}

Let's begin this type of hypothesis test with a case. The question is:
how do you know whether the homogeneity of variance assumption is
satisﬁed? One simple method involves just looking at two sample
variances. Logically, if two population variances are equal, then the
two sample variances should be very similar. When the two sample
variances are reasonably close, you can be reasonably conﬁdent that the
homogeneity assumption is satisﬁed and proceed with, for example,
Student t-interval. However, when one sample variance is three or four
times larger than the other, then there is reason for a concern. The
common statistical procedure for comparing population variances
$\sigma_1^2$ and $\sigma_2^2$ makes an inference about the ratio of
$\sigma_1^2$/$\sigma_2^2$.


To test $H_0 : \mu_d = \mu_0$, our options for the null and alternative hypotheses are:

- $H_0: \sigma_1^2 = \sigma_2^2$ vs. $H_a: \sigma_1^2 > \sigma_2^2$

- $H_0: \sigma_1^2 = \sigma_2^2$ vs. $H_a: \sigma_1^2 < \sigma_2^2$

- $H_0: \sigma_1^2 = \sigma_2^2$ vs. $H_a: \sigma_1^2 \neq \sigma_2^2$


We calculate the $F$-statistic:

$$F^{*} = \frac{s_1^2}{s_2^2} \sim F_{n_1-1,n_2-1}.$$

In terms of a variable $F$ which follows an $F$-distribution with $n_1-1$ and $n_2-1$ degrees of freedom, the
p-value for a test of $H_0$ against:

- $H_a: \sigma_1^2 > \sigma_2^2$ is $P(F > F^{*})$

- $H_a: \sigma_1^2 < \sigma_2^2$ is $P(F < F^{*})$

- $H_a: \sigma_1^2 \neq \sigma_2^2$ is $2P(F > |F^{*}|)$



<!-- **Step 1: Stating the Structure of Testing Hypothesis** -->


<!-- **Step 2: Computing Test Statistics** -->

<!-- ::: definition -->
<!-- The test statistics of two sample hypothesis test on variances is given -->
<!-- by: $$F_* = \frac{s_1^2}{s_2^2} \sim F_{n_1-1,n_2-1}.$$ -->
<!-- ::: -->

<!-- **Decision Rules** -->

<!-- - $H_0: \sigma_1^2 = \sigma_2^2$ and $H_a:\sigma_1^2 \neq \sigma_2^2$. -->
<!--   If $F^* > F_{n_1-1,n_2-1,\alpha/2}$ or -->
<!--   $F^* < F_{n_1-1,n_2-1,1 - \alpha/2}$, then we reject $H_0$. Otherwise, -->
<!--   we do not reject it. -->

<!-- - $H_0: \sigma_1^2 = \sigma_2^2$ and $H_a: \sigma_1^2 > \sigma_2^2$. If -->
<!--   $F^*> F_{n_1-1,n_2-1,\alpha}$ or $P(F_{n_1-1,n_2-1} > F^*)$ is too -->
<!--   small, then we reject $H_0$. Otherwise, we do not reject it. -->

<!-- - $H_0: \sigma_1^2 = \sigma_2^2$ and $H_a: \sigma_1^2 < \sigma_2^2$. if -->
<!--   $F^* < F_{n_1-1,n_2-1,1-\alpha}$ or $P(F_{n_1-1,n_2-1} < F^*)$ is too -->
<!--   small, then we reject $H_0$. Otherwise, we do not reject it. -->


#### Assumptions

- The two samples are independent.
- The populations from which the samples are drawn are normally distributed.
- The data are measured on a continuous scale.












## Hypothesis Tests on Paired Data

When observations in sample 1 matches with an observation in sample 2.
Observations in sample 1 are, usually, highly, correlated with
observations in sample 2, these data are often called matched pairs. For
each pair (the same cases), we form: Difference = observation in sample
2 - observation in sample 1. Thus, we have one single sample of
differences scores. For example, in longitudinal studies: Pre- and
post-survey of attitudes towards statistics (Same student is measured
twice: Time 1 (pre) and Time 2 (post). We measure change in the
attitudes: Post - Pre (for each student). Often these types of studies
are called, repeated measures.\
Paired Data Condition: the data must be quantitative and paired.\
Independence Assumption:

- If the data are paired, recall the measurements are not independent since 2 measurements are taken on each unit. 
  However, since each unit is sampled independently, the differences must be independent of each other.

- The pairs should be a random sample.

- In experimental design, the order of the two treatments may be
  randomly assigned, or the treatments ma y be randomly assigned to one
  member of each pair.

- In a before-and-after study, we may believe that the observed
  differences are representative sample of a population of interest. If
  there is any doubt, we need to include a control group to be able to
  draw conclusions.

- If samples are bigger than 10 % of the target population, we need to
  acknowledge this and note in our report. When we sample from a ﬁnite
  population, we should be careful not to sample more than 10 % of that
  population. Sampling too large a fraction of the population calls the
  independence assumption into question.

Recall Section \@ref(sec:CIsOnPaired) where we covered confidence intervals on paired data, 
we introduced the table below:

::: {#smalltbl .smalltbl #direct}
| Sample Units | Measurement 1 (\(M_{1}\)) | Measurement 2 (\(M_{2}\)) | Difference (\(M_{2}-M_{1}\))    |
|:------------:|:-------------------------:|:-------------------------:|:-------------------------------:|
|            1 | \(x_{11}\)               | \(x_{12}\)               | \(x_{d1} = x_{12} - x_{11}\)      |
|            2 | \(x_{21}\)               | \(x_{22}\)               | \(x_{d2} = x_{22} - x_{21}\)      |
|            3 | \(x_{31}\)               | \(x_{32}\)               | \(x_{d3} = x_{32} - x_{31}\)      |
|   \(\vdots\) | \(\vdots\)               | \(\vdots\)               | \(\vdots\)                        |
|            n | \(x_{n1}\)               | \(x_{n2}\)               | \(x_{dn} = x_{n2} - x_{n1}\)      |
:::



From that table, we can get $\bar{x}_{d}$, which is the mean, variance and
standard deviation of the difference. We need these values to continue
our analysis.

To test $H_0 : \mu_d = \mu_0$, our options for the null and alternative hypotheses are:

- $H_0: \mu_d = \mu_0$ vs. $H_a: \mu_d > \mu_0$

- $H_0: \mu_d = \mu_0$ vs. $H_a: \mu_d < \mu_0$

- $H_0: \mu_d = \mu_0$ vs. $H_a: \mu_d \neq \mu_0$


We calculate the $t$-statistic:

$$t^* = \dfrac{\bar{x}_{d} - \mu_0}{s_{d} / \sqrt{n}} \sim t_{n-1}$$
In terms of a variable $T$ which follows a $t$-distribution at $n-1$ degrees of freedom, the
p-value for a test of $H_0$ against:

- $H_a: \mu > \mu_0$ is $P(T > t^{*})$

- $H_a: \mu < \mu_0$ is $P(T < t^{*})$

- $H_a: \mu \ne \mu_0$ is $2P(T > |t^{*}|)$



:::: example
In an effort to determine whether a new type of fertilizer is more
effective than the type currently in use, researchers took 12 two-acre
plots of land scattered throughout the county. Each plot was divided
into two equal-size sub plots, one of which was treated with the new
fertilizer. Wheat was planted, and the crop yields were measured.

::: center
| Plot    | 1  | 2  | 3  | 4  | 5  | 6  | 7  | 8  | 9  | 10 | 11 | 12 |
|:--------|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|
| Current | 56 | 45 | 68 | 72 | 61 | 69 | 57 | 55 | 60 | 72 | 75 | 66 |
| New     | 60 | 49 | 66 | 73 | 59 | 67 | 61 | 60 | 58 | 75 | 72 | 68 |
:::

Can we conclude at the 5% signiﬁcance level that the new fertilizer is
more eﬀective than the current one?

**Solution:**

You can verify that the mean and standard deviation of the twelve
difference measurements are $\bar{x}_{d} = new - current = 1$ and
$s_d = 3.0151$.

**Step 1: State Hypothesis**

$H_0: \mu_d = 0$ and $H_a: \mu_d>0$

**Step 2: Find test statistics**

$t^{*} = \displaystyle\frac{\bar{x}_{d} - 0}{s_d/\sqrt{n}} = \frac{1}{3.0151/\sqrt{12}} = 1.1489$

**Step 3: Compute p-value**

Using t-distribution table with df $=11$, then $0.10 < p-value < 0.15$.

**Step 4: Conclusion**

Since p-value $> \alpha = 0.05$, we can't reject $H_0$. There is not
enough evidence to infer that the new fertilizer is better.
::::





#### Assumptions

- The differences between paired observations are approximately normally distributed.
- The samples are randomly and independently drawn from the population.

## Exercises {#sec:ch6exercises}

---

<div class="exercise-box">
<div class="exercise-label">Question 1</div>
A cola manufacturer calibrates its filling line to a target mean fill volume of $\mu_0 = 355$ mL. The population standard deviation is known to be $\sigma = 4$ mL. A quality engineer randomly selects $n = 64$ cans and records a sample mean of $\bar{x} = 353.2$ mL.

Test $H_0: \mu = 355$ vs. $H_a: \mu < 355$ at $\alpha = 0.05$.

(a) Compute the $z$ test statistic using the formula $z^* = \dfrac{\bar{x} - \mu_0}{\sigma / \sqrt{n}}$.
(b) Compute the p-value using `pnorm()` and fill in the correct argument.
(c) Based on your p-value, state a conclusion at $\alpha = 0.05$.

<div class="webr-exercise">
<div class="webr-exercise-header">Interactive R — One-Sample z Test (Known σ)</div>
<textarea class="webr-editor" rows="13">n     <- 64
xbar  <- 353.2
mu0   <- 355
sigma <- 4

# (a) Compute the z test statistic
z_star <- (xbar - mu0) / (sigma / sqrt(n))
z_star

# (b) Compute the p-value for Ha: mu < mu0  (lower-tailed)
p_value <- pnorm(___)
p_value
</textarea>
<div class="webr-controls">
  <button class="webr-run-btn" disabled>▶ Run</button>
  <button class="webr-reset-btn">↺ Reset</button>
</div>
<div class="webr-output-area"></div>
<div class="webr-canvas-area"><canvas></canvas></div>
</div>
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- The $z$ statistic for a known-$\sigma$ test is $z^* = (\bar{x} - \mu_0) / (\sigma / \sqrt{n})$.
- For a lower-tailed alternative $H_a: \mu < \mu_0$, the p-value is $P(Z < z^*) = $ `pnorm(z_star)`.
- `pnorm(q)` returns $P(Z \leq q)$ for $Z \sim N(0,1)$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">

```r
n     <- 64
xbar  <- 353.2
mu0   <- 355
sigma <- 4

# (a) z test statistic
z_star <- (xbar - mu0) / (sigma / sqrt(n))
z_star
# [1] -3.6

# (b) p-value (lower-tailed)
p_value <- pnorm(z_star)
p_value
# [1] 0.0001591086
```

**(a)** $z^* = \dfrac{353.2 - 355}{4 / \sqrt{64}} = \dfrac{-1.8}{0.5} = -3.6$

**(b)** p-value $= P(Z < -3.6) \approx 0.00016$

**(c)** Since p-value $\approx 0.00016 < \alpha = 0.05$, we **reject $H_0$**. There is sufficient evidence at the 5% level of significance to conclude that the mean fill volume has decreased below 355 mL.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 2</div>
For a one-sample $z$ test, the test statistic is $z^* = 1.75$. Fill in the correct `pnorm()` expression for the p-value under each alternative hypothesis.

(a) $H_a: \mu > \mu_0$ (upper-tailed)
(b) $H_a: \mu < \mu_0$ (lower-tailed)
(c) $H_a: \mu \neq \mu_0$ (two-tailed)

<div class="webr-exercise">
<div class="webr-exercise-header">Interactive R — P-Value Calculations for Three Alternatives</div>
<textarea class="webr-editor" rows="10">z_star <- 1.75

# (a) Ha: mu > mu0   (upper-tailed)
p_value_a <- ___

# (b) Ha: mu < mu0   (lower-tailed)
p_value_b <- ___

# (c) Ha: mu != mu0  (two-tailed)
p_value_c <- ___
</textarea>
<div class="webr-controls">
  <button class="webr-run-btn" disabled>▶ Run</button>
  <button class="webr-reset-btn">↺ Reset</button>
</div>
<div class="webr-output-area"></div>
<div class="webr-canvas-area"><canvas></canvas></div>
</div>
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Upper-tailed: p-value $= P(Z > z^*) = $ `1 - pnorm(z_star)`
- Lower-tailed: p-value $= P(Z < z^*) = $ `pnorm(z_star)`
- Two-tailed: p-value $= 2 \times P(Z > |z^*|) = $ `2 * (1 - pnorm(abs(z_star)))`
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">

```r
z_star <- 1.75

# (a) upper-tailed
p_value_a <- 1 - pnorm(z_star)
p_value_a
# [1] 0.04005916

# (b) lower-tailed
p_value_b <- pnorm(z_star)
p_value_b
# [1] 0.9599408

# (c) two-tailed
p_value_c <- 2 * (1 - pnorm(abs(z_star)))
p_value_c
# [1] 0.08011832
```

**(a)** $P(Z > 1.75) \approx 0.0401$

**(b)** $P(Z < 1.75) \approx 0.9599$

**(c)** $2 \times P(Z > 1.75) \approx 0.0801$

Note that for a fixed $z^*$, the two-tailed p-value is always twice the upper-tailed p-value, and the upper-tailed and lower-tailed p-values always sum to 1.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 3</div>
A nutritionist believes that a new diet plan reduces daily calorie intake below 2000 kcal. She records the daily calorie intake of $n = 10$ participants:

$$1850,\ 1920,\ 1780,\ 1960,\ 1840,\ 1990,\ 1870,\ 1930,\ 1800,\ 1910$$

Assume calorie intake is approximately normally distributed. Use `t.test()` to test $H_0: \mu = 2000$ vs. $H_a: \mu < 2000$ at $\alpha = 0.05$. Fill in the correct values for `mu` and `alternative`.

<div class="webr-exercise">
<div class="webr-exercise-header">Interactive R — One-Sample t Test (Unknown σ)</div>
<textarea class="webr-editor" rows="6">calories <- c(1850, 1920, 1780, 1960, 1840, 1990, 1870, 1930, 1800, 1910)

# Run a one-sample t-test
# H0: mu = 2000  vs  Ha: mu < 2000
result <- t.test(calories, mu = ___, alternative = "___")
result</textarea>
<div class="webr-controls">
  <button class="webr-run-btn" disabled>▶ Run</button>
  <button class="webr-reset-btn">↺ Reset</button>
</div>
<div class="webr-output-area"></div>
<div class="webr-canvas-area"><canvas></canvas></div>
</div>
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- `t.test(x, mu = mu0, alternative = "less")` tests $H_a: \mu < \mu_0$.
- The `alternative` argument accepts `"less"`, `"greater"`, or `"two.sided"`.
- When $\sigma$ is unknown, `t.test()` uses the $t$ distribution with $df = n - 1$ degrees of freedom.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">

```r
calories <- c(1850, 1920, 1780, 1960, 1840, 1990, 1870, 1930, 1800, 1910)

result <- t.test(calories, mu = 2000, alternative = "less")
result
#
#   One Sample t-test
#
# data:  calories
# t = -5.3083, df = 9, p-value = 0.0002386
# alternative hypothesis: true mean is less than 2000
# 95 percent confidence interval:
#      -Inf 1921.699
# sample estimates:
# mean of x
#      1885
```

The sample mean is $\bar{x} = 1885$ kcal. With $t^* \approx -5.31$ and $df = 9$, the p-value $\approx 0.00024 < \alpha = 0.05$. We **reject $H_0$**. There is sufficient evidence that the mean daily calorie intake under the new diet plan is less than 2000 kcal.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 4</div>
A student newspaper reports that only 25% of UTM students hold part-time jobs. A researcher surveys $n = 120$ randomly selected students and finds $x = 42$ who hold part-time jobs. Test whether the true proportion exceeds 25% ($H_0: p = 0.25$ vs. $H_a: p > 0.25$) at $\alpha = 0.05$.

(a) Compute the sample proportion $\hat{p} = x / n$.
(b) Compute the $z$ test statistic: $z^* = \dfrac{\hat{p} - p_0}{\sqrt{p_0(1-p_0)/n}}$.
(c) Compute the p-value using `pnorm()`.

<div class="webr-exercise">
<div class="webr-exercise-header">Interactive R — One-Sample z Test for a Proportion</div>
<textarea class="webr-editor" rows="14">x  <- 42
n  <- 120
p0 <- 0.25

# (a) Compute sample proportion
phat <- ___ / ___

# (b) Compute z test statistic
z_star <- (phat - p0) / sqrt(p0 * (1 - p0) / n)
z_star

# (c) Compute p-value for Ha: p > p0  (upper-tailed)
p_value <- 1 - pnorm(___)
p_value
</textarea>
<div class="webr-controls">
  <button class="webr-run-btn" disabled>▶ Run</button>
  <button class="webr-reset-btn">↺ Reset</button>
</div>
<div class="webr-output-area"></div>
<div class="webr-canvas-area"><canvas></canvas></div>
</div>
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- $\hat{p} = x / n$ where $x$ is the number of successes and $n$ is the sample size.
- The standard error under $H_0$ is $\sqrt{p_0(1-p_0)/n}$ — use $p_0$, not $\hat{p}$, here.
- For an upper-tailed test ($H_a: p > p_0$), the p-value is $P(Z > z^*) = $ `1 - pnorm(z_star)`.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">

```r
x  <- 42
n  <- 120
p0 <- 0.25

# (a) Sample proportion
phat <- x / n
phat
# [1] 0.35

# (b) z test statistic
z_star <- (phat - p0) / sqrt(p0 * (1 - p0) / n)
z_star
# [1] 2.529822

# (c) p-value (upper-tailed)
p_value <- 1 - pnorm(z_star)
p_value
# [1] 0.005713023
```

**(a)** $\hat{p} = 42/120 = 0.35$

**(b)** $z^* = \dfrac{0.35 - 0.25}{\sqrt{0.25 \times 0.75 / 120}} = \dfrac{0.10}{0.03953} \approx 2.53$

**(c)** p-value $= P(Z > 2.53) \approx 0.0057 < \alpha = 0.05$

We **reject $H_0$**. There is sufficient evidence at the 5% level to conclude that more than 25% of UTM students hold part-time jobs.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 5</div>
A national health study reports that 20% of adults have high blood pressure. A local clinic randomly selects $n = 300$ patients and finds $x = 72$ with high blood pressure. Test whether the local rate differs from the national figure ($H_0: p = 0.20$ vs. $H_a: p \neq 0.20$) at $\alpha = 0.05$.

Use `prop.test()` with `correct = FALSE`. Fill in the arguments, then state a conclusion by comparing the p-value to $\alpha = 0.05$.

<div class="webr-exercise">
<div class="webr-exercise-header">Interactive R — One-Sample Proportion Test with prop.test()</div>
<textarea class="webr-editor" rows="6"># One-sample proportion test
# H0: p = 0.20  vs  Ha: p != 0.20

result <- prop.test(x = ___, n = ___, p = ___,
                    alternative = "___", correct = ___)
result</textarea>
<div class="webr-controls">
  <button class="webr-run-btn" disabled>▶ Run</button>
  <button class="webr-reset-btn">↺ Reset</button>
</div>
<div class="webr-output-area"></div>
<div class="webr-canvas-area"><canvas></canvas></div>
</div>
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- `prop.test(x, n, p, alternative, correct)` tests a single proportion.
- For a two-sided test use `alternative = "two.sided"`.
- Set `correct = FALSE` to skip the continuity correction (matching the $z$ formula from Section 6.2).
- Compare the p-value in the output to $\alpha = 0.05$ to decide whether to reject $H_0$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">

```r
result <- prop.test(x = 72, n = 300, p = 0.20,
                    alternative = "two.sided", correct = FALSE)
result
#
#   1-sample proportions test without continuity correction
#
# data:  72 out of 300, null probability 0.2
# X-squared = 3, df = 1, p-value = 0.08327
# alternative hypothesis: two.sided
# 95 percent confidence interval:
#  0.1912025 0.2940115
# sample estimates:
#    p
# 0.24
```

$\hat{p} = 72/300 = 0.24$. The p-value $\approx 0.083 > \alpha = 0.05$, so we **fail to reject $H_0$**. There is not sufficient evidence at the 5% level to conclude that the local rate of high blood pressure differs from the national figure of 20%.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 6</div>
A bottled water plant fills bottles to a target mean weight of $\mu_0 = 500$ g. The population standard deviation is known to be $\sigma = 8$ g. A quality control inspector randomly selects $n = 25$ bottles and records a sample mean of $\bar{x} = 496.5$ g. The plant manager wants to know whether the mean fill weight has dropped below the target.

You may use the Distribution Calculator in Section 4.1.5 to find the required tail probability.

(a) State $H_0$ and $H_a$.
(b) Compute the test statistic.
(c) Find the p-value.
(d) State a conclusion at $\alpha = 0.05$.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Since $\sigma$ is known and the question asks whether the mean has *dropped*, use a lower-tailed $z$ test.
- The test statistic is $z^* = (\bar{x} - \mu_0) / (\sigma / \sqrt{n})$.
- The p-value for $H_a: \mu < \mu_0$ is $P(Z < z^*)$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** $H_0: \mu = 500$ g $\quad$ vs. $\quad H_a: \mu < 500$ g

**(b)** Since $\sigma$ is known, we use the $z$ statistic:
$$z^* = \frac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} = \frac{496.5 - 500}{8 / \sqrt{25}} = \frac{-3.5}{1.6} = -2.1875$$

**(c)** For $H_a: \mu < 500$ (lower-tailed), using $Z \sim N(0,1)$:
$$\text{p-value} = P(Z < -2.1875) \approx 0.0143$$

**(d)** Since p-value $\approx 0.0143 < \alpha = 0.05$, we **reject $H_0$**. There is sufficient evidence at the 5% level to conclude that the mean fill weight has dropped below the target of 500 g.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 7</div>
An athletics coach believes the mean sprint time of her athletes has improved (i.e., decreased) from the historical mean of $\mu_0 = 12.5$ seconds over a 100-metre course. She records sprint times for a random sample of $n = 36$ athletes and obtains a sample mean of $\bar{x} = 12.1$ s. From many years of records, the population standard deviation is known to be $\sigma = 1.2$ s.

You may use the Distribution Calculator in Section 4.1.5 to find the required tail probability.

(a) State $H_0$ and $H_a$.
(b) Compute the test statistic.
(c) Find the p-value.
(d) State a conclusion at $\alpha = 0.05$.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- "Improved" in sprint time means a lower time, so the alternative is lower-tailed.
- $\sigma$ is known, so use the $z$ test statistic.
- The p-value for $H_a: \mu < \mu_0$ is $P(Z < z^*)$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** $H_0: \mu = 12.5$ s $\quad$ vs. $\quad H_a: \mu < 12.5$ s

**(b)**
$$z^* = \frac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} = \frac{12.1 - 12.5}{1.2 / \sqrt{36}} = \frac{-0.4}{0.2} = -2.000$$

**(c)** For $H_a: \mu < 12.5$ (lower-tailed):
$$\text{p-value} = P(Z < -2.000) \approx 0.0228$$

**(d)** Since p-value $\approx 0.0228 < \alpha = 0.05$, we **reject $H_0$**. There is sufficient evidence at the 5% level to conclude that the mean sprint time has decreased from 12.5 seconds.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 8</div>
A professor claims the average midterm score in her large statistics course is $\mu_0 = 70$ points. The population standard deviation is known from past semesters to be $\sigma = 9.1$ points. After a curriculum redesign, a random sample of $n = 49$ students yields a sample mean of $\bar{x} = 72.6$ points. Is there evidence that the mean score has increased?

You may use the Distribution Calculator in Section 4.1.5 to find the required tail probability.

(a) State $H_0$ and $H_a$.
(b) Compute the test statistic.
(c) Find the p-value.
(d) State a conclusion at $\alpha = 0.05$.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- "Has the mean score *increased*?" calls for an upper-tailed alternative: $H_a: \mu > \mu_0$.
- The p-value for $H_a: \mu > \mu_0$ is $P(Z > z^*) = 1 - P(Z \le z^*)$.
- $\sigma$ is known, so use the $z$ test.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** $H_0: \mu = 70 \quad$ vs. $\quad H_a: \mu > 70$

**(b)**
$$z^* = \frac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} = \frac{72.6 - 70}{9.1 / \sqrt{49}} = \frac{2.6}{1.3} = 2.000$$

**(c)** For $H_a: \mu > 70$ (upper-tailed):
$$\text{p-value} = P(Z > 2.000) = 1 - P(Z \le 2.000) \approx 0.0228$$

**(d)** Since p-value $\approx 0.0228 < \alpha = 0.05$, we **reject $H_0$**. There is sufficient evidence at the 5% level to conclude that the mean midterm score increased after the curriculum redesign.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 9</div>
A road-safety researcher claims the mean reaction time for licensed drivers is $\mu_0 = 0.65$ seconds. The population standard deviation is known to be $\sigma = 0.10$ s. A random sample of $n = 100$ drivers is tested and the sample mean reaction time is $\bar{x} = 0.672$ s. The researcher wants to test whether the mean reaction time differs from the claimed value.

You may use the Distribution Calculator in Section 4.1.5 to find the required tail probability.

(a) State $H_0$ and $H_a$.
(b) Compute the test statistic.
(c) Find the p-value.
(d) State a conclusion at $\alpha = 0.05$.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- "Differs from" means a two-tailed test: $H_a: \mu \ne \mu_0$.
- For a two-tailed test, p-value $= 2 \times P(Z > |z^*|) = 2(1 - P(Z \le |z^*|))$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** $H_0: \mu = 0.65$ s $\quad$ vs. $\quad H_a: \mu \ne 0.65$ s

**(b)**
$$z^* = \frac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} = \frac{0.672 - 0.65}{0.10 / \sqrt{100}} = \frac{0.022}{0.01} = 2.200$$

**(c)** For $H_a: \mu \ne 0.65$ (two-tailed):
$$\text{p-value} = 2 \times P(Z > 2.200) = 2(1 - P(Z \le 2.200)) \approx 2(0.0139) = 0.0278$$

**(d)** Since p-value $\approx 0.0278 < \alpha = 0.05$, we **reject $H_0$**. There is sufficient evidence at the 5% level to conclude that the mean reaction time differs from 0.65 seconds.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 10</div>
A cosmetics company claims its face cream reduces mean skin roughness to $\mu_0 = 3.0$ on a 0–10 scale (higher = rougher). Dermatologists apply the cream to a random sample of $n = 64$ volunteers and measure a sample mean roughness of $\bar{x} = 3.4$. The population standard deviation is known to be $\sigma = 1.6$. Test whether the true mean roughness exceeds the company's claim.

You may use the Distribution Calculator in Section 4.1.5 to find the required tail probability.

(a) State $H_0$ and $H_a$.
(b) Compute the test statistic.
(c) Find the p-value.
(d) State a conclusion at $\alpha = 0.01$.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- The question asks whether the mean is *higher* than claimed, so use an upper-tailed test.
- Note that $\alpha = 0.01$ here — compare the p-value to 0.01, not 0.05.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** $H_0: \mu = 3.0 \quad$ vs. $\quad H_a: \mu > 3.0$

**(b)**
$$z^* = \frac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} = \frac{3.4 - 3.0}{1.6 / \sqrt{64}} = \frac{0.4}{0.2} = 2.000$$

**(c)** For $H_a: \mu > 3.0$ (upper-tailed):
$$\text{p-value} = P(Z > 2.000) \approx 0.0228$$

**(d)** Since p-value $\approx 0.0228 > \alpha = 0.01$, we **fail to reject $H_0$**. There is not sufficient evidence at the 1% level to conclude that the true mean skin roughness exceeds 3.0. *(Note: the result would be significant at $\alpha = 0.05$, but not at the stricter $\alpha = 0.01$.)*
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 11</div>
A health study reports that adults sleep a mean of $\mu_0 = 7.0$ hours per night. A nutritionist administers a dietary supplement to a random sample of $n = 16$ volunteers for four weeks. At the end of the study the sample mean sleep time is $\bar{x} = 7.8$ hours with a sample standard deviation of $s = 1.2$ hours. The nutritionist believes the supplement increases sleep time. Assume sleep times are approximately normally distributed.

You may use the Distribution Calculator in Section 4.1.5 to find the required tail probability.

(a) State $H_0$ and $H_a$.
(b) Compute the test statistic.
(c) Find the p-value.
(d) State a conclusion at $\alpha = 0.05$.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- $\sigma$ is unknown, so use the $t$ statistic with $df = n - 1 = 15$.
- For $H_a: \mu > \mu_0$ (upper-tailed), p-value $= P(T_{15} > t^*)$.
- You can bound the p-value from the $t$-table by locating where $t^*$ falls relative to tabulated critical values at $df = 15$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** $H_0: \mu = 7.0$ hr $\quad$ vs. $\quad H_a: \mu > 7.0$ hr

**(b)** Since $\sigma$ is unknown, we use the $t$ statistic with $df = n - 1 = 15$:
$$t^* = \frac{\bar{x} - \mu_0}{s / \sqrt{n}} = \frac{7.8 - 7.0}{1.2 / \sqrt{16}} = \frac{0.8}{0.3} \approx 2.667$$

**(c)** For $H_a: \mu > 7.0$ (upper-tailed), using $T \sim t_{15}$:
$$\text{p-value} = P(T_{15} > 2.667) \approx 0.0088$$
From the $t$-table: $t_{15,\,0.01} = 2.602 < 2.667 < 2.947 = t_{15,\,0.005}$, so $0.005 < \text{p-value} < 0.01$.

**(d)** Since p-value $< \alpha = 0.05$, we **reject $H_0$**. There is sufficient evidence at the 5% level to conclude that the mean sleep time for volunteers on the supplement exceeds 7.0 hours per night.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 12</div>
A fast-food chain states its drive-through service takes a mean of $\mu_0 = 3.5$ minutes. A regional manager suspects service has slowed at a particular outlet. She times $n = 20$ randomly selected customers and records $\bar{x} = 3.9$ minutes and $s = 0.8$ minutes. Assume service times are approximately normally distributed.

You may use the Distribution Calculator in Section 4.1.5 to find the required tail probability.

(a) State $H_0$ and $H_a$.
(b) Compute the test statistic.
(c) Find the p-value.
(d) State a conclusion at $\alpha = 0.05$.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- The manager suspects the mean has *increased*, so use an upper-tailed test: $H_a: \mu > \mu_0$.
- $\sigma$ is unknown, so use the $t$ statistic with $df = n - 1 = 19$.
- Upper-tailed p-value $= P(T_{19} > t^*)$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** $H_0: \mu = 3.5$ min $\quad$ vs. $\quad H_a: \mu > 3.5$ min

**(b)** With $df = n - 1 = 19$:
$$t^* = \frac{\bar{x} - \mu_0}{s / \sqrt{n}} = \frac{3.9 - 3.5}{0.8 / \sqrt{20}} = \frac{0.4}{0.1789} \approx 2.236$$

**(c)** For $H_a: \mu > 3.5$ (upper-tailed), using $T \sim t_{19}$:
$$\text{p-value} = P(T_{19} > 2.236) \approx 0.0188$$
From the $t$-table: $t_{19,\,0.025} = 2.093 < 2.236 < 2.539 = t_{19,\,0.01}$, so $0.01 < \text{p-value} < 0.025$.

**(d)** Since p-value $\approx 0.019 < \alpha = 0.05$, we **reject $H_0$**. There is sufficient evidence at the 5% level to conclude that the mean drive-through service time at this outlet has increased above 3.5 minutes.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 13</div>
A psychologist claims the population mean anxiety score on a standardised scale is $\mu_0 = 15$. She collects a random sample of $n = 10$ patients and records $\bar{x} = 17.2$ with $s = 3.6$. She wants to test whether the mean anxiety score differs from 15. Assume scores are approximately normally distributed.

You may use the Distribution Calculator in Section 4.1.5 to find the required tail probability.

(a) State $H_0$ and $H_a$.
(b) Compute the test statistic.
(c) Find the p-value.
(d) State a conclusion at $\alpha = 0.05$.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- "Differs from" calls for a two-tailed test: $H_a: \mu \ne \mu_0$.
- $\sigma$ is unknown, so use the $t$ statistic with $df = n - 1 = 9$.
- Two-tailed p-value $= 2 \times P(T_9 > |t^*|)$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** $H_0: \mu = 15 \quad$ vs. $\quad H_a: \mu \ne 15$

**(b)** With $df = n - 1 = 9$:
$$t^* = \frac{\bar{x} - \mu_0}{s / \sqrt{n}} = \frac{17.2 - 15}{3.6 / \sqrt{10}} = \frac{2.2}{1.1384} \approx 1.933$$

**(c)** For $H_a: \mu \ne 15$ (two-tailed), using $T \sim t_9$:
$$\text{p-value} = 2 \times P(T_9 > 1.933) \approx 2(0.0427) = 0.0853$$
From the $t$-table: $t_{9,\,0.05} = 1.833 < 1.933 < 2.262 = t_{9,\,0.025}$, so $0.05 < \text{p-value} < 0.10$.

**(d)** Since p-value $\approx 0.085 > \alpha = 0.05$, we **fail to reject $H_0$**. There is not sufficient evidence at the 5% level to conclude that the mean anxiety score differs from 15.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 14</div>
A botanist claims that a new plant food reduces the mean germination time of seeds below the standard $\mu_0 = 8.0$ days. She plants $n = 12$ seeds treated with the food and records $\bar{x} = 7.0$ days and $s = 1.5$ days. Assume germination times are approximately normally distributed.

You may use the Distribution Calculator in Section 4.1.5 to find the required tail probability.

(a) State $H_0$ and $H_a$.
(b) Compute the test statistic.
(c) Find the p-value.
(d) State a conclusion at $\alpha = 0.05$.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- The claim is that germination time has *decreased*, so use $H_a: \mu < \mu_0$.
- $\sigma$ is unknown, so use the $t$ statistic with $df = n - 1 = 11$.
- Lower-tailed p-value $= P(T_{11} < t^*)$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** $H_0: \mu = 8.0$ days $\quad$ vs. $\quad H_a: \mu < 8.0$ days

**(b)** With $df = n - 1 = 11$:
$$t^* = \frac{\bar{x} - \mu_0}{s / \sqrt{n}} = \frac{7.0 - 8.0}{1.5 / \sqrt{12}} = \frac{-1.0}{0.4330} \approx -2.309$$

**(c)** For $H_a: \mu < 8.0$ (lower-tailed), using $T \sim t_{11}$:
$$\text{p-value} = P(T_{11} < -2.309) \approx 0.0207$$
From the $t$-table: $t_{11,\,0.025} = 2.201 < 2.309 < 2.718 = t_{11,\,0.01}$, so $0.01 < \text{p-value} < 0.025$.

**(d)** Since p-value $\approx 0.021 < \alpha = 0.05$, we **reject $H_0$**. There is sufficient evidence at the 5% level to conclude that the new plant food reduces mean germination time below 8.0 days.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 15</div>
A university registrar claims that 40% of students live off campus. A housing researcher believes the true proportion is lower. In a random sample of $n = 200$ students, $x = 65$ reported living off campus.

You may use the Distribution Calculator in Section 4.1.5 to find the required tail probability.

(a) State $H_0$ and $H_a$.
(b) Compute the sample proportion $\hat{p}$ and the test statistic.
(c) Find the p-value.
(d) State a conclusion at $\alpha = 0.05$.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- The researcher believes the proportion is *lower*, so use $H_a: p < p_0$.
- The test statistic is $z^* = (\hat{p} - p_0) / \sqrt{p_0(1-p_0)/n}$, using $p_0$ in the denominator.
- Lower-tailed p-value $= P(Z < z^*)$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** $H_0: p = 0.40 \quad$ vs. $\quad H_a: p < 0.40$

**(b)** Sample proportion:
$$\hat{p} = \frac{x}{n} = \frac{65}{200} = 0.325$$
Test statistic:
$$z^* = \frac{\hat{p} - p_0}{\sqrt{p_0(1-p_0)/n}} = \frac{0.325 - 0.40}{\sqrt{0.40 \times 0.60 / 200}} = \frac{-0.075}{0.03464} \approx -2.165$$

**(c)** For $H_a: p < 0.40$ (lower-tailed):
$$\text{p-value} = P(Z < -2.165) \approx 0.0152$$

**(d)** Since p-value $\approx 0.0152 < \alpha = 0.05$, we **reject $H_0$**. There is sufficient evidence at the 5% level to conclude that fewer than 40% of students live off campus.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 16</div>
A candy manufacturer claims that 30% of candies in each bag are red. A consumer randomly selects $n = 100$ candies from several bags and finds $x = 26$ that are red. Test whether the true proportion of red candies differs from 30%.

You may use the Distribution Calculator in Section 4.1.5 to find the required tail probability.

(a) State $H_0$ and $H_a$.
(b) Compute $\hat{p}$ and the test statistic.
(c) Find the p-value.
(d) State a conclusion at $\alpha = 0.05$.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- "Differs from" calls for a two-tailed test: $H_a: p \ne p_0$.
- Two-tailed p-value $= 2 \times P(Z > |z^*|) = 2(1 - P(Z \le |z^*|))$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** $H_0: p = 0.30 \quad$ vs. $\quad H_a: p \ne 0.30$

**(b)** Sample proportion:
$$\hat{p} = \frac{26}{100} = 0.26$$
Test statistic:
$$z^* = \frac{\hat{p} - p_0}{\sqrt{p_0(1-p_0)/n}} = \frac{0.26 - 0.30}{\sqrt{0.30 \times 0.70 / 100}} = \frac{-0.04}{0.04583} \approx -0.873$$

**(c)** For $H_a: p \ne 0.30$ (two-tailed):
$$\text{p-value} = 2 \times P(Z > 0.873) = 2(1 - P(Z \le 0.873)) \approx 2(0.1913) = 0.3827$$

**(d)** Since p-value $\approx 0.383 > \alpha = 0.05$, we **fail to reject $H_0$**. There is not sufficient evidence to conclude that the true proportion of red candies differs from 30%.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 17</div>
A manufacturer claims fewer than 5% of its electronic components are defective. A quality inspector samples $n = 500$ components and finds $x = 15$ that are defective. Test whether the defective rate is below 5%.

You may use the Distribution Calculator in Section 4.1.5 to find the required tail probability.

(a) State $H_0$ and $H_a$.
(b) Compute $\hat{p}$ and the test statistic.
(c) Find the p-value.
(d) State a conclusion at $\alpha = 0.05$.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- The claim is that the defect rate is *below* 5%, so use $H_a: p < 0.05$.
- Lower-tailed p-value $= P(Z < z^*)$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** $H_0: p = 0.05 \quad$ vs. $\quad H_a: p < 0.05$

**(b)** Sample proportion:
$$\hat{p} = \frac{15}{500} = 0.030$$
Test statistic:
$$z^* = \frac{\hat{p} - p_0}{\sqrt{p_0(1-p_0)/n}} = \frac{0.030 - 0.05}{\sqrt{0.05 \times 0.95 / 500}} = \frac{-0.020}{0.009747} \approx -2.052$$

**(c)** For $H_a: p < 0.05$ (lower-tailed):
$$\text{p-value} = P(Z < -2.052) \approx 0.0201$$

**(d)** Since p-value $\approx 0.020 < \alpha = 0.05$, we **reject $H_0$**. There is sufficient evidence at the 5% level to conclude that the true defective rate is below 5%.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 18</div>
A national survey reports that 60% of households own at least one pet. A researcher randomly selects $n = 150$ households in one city. Of these, $x = 102$ report owning at least one pet. Is the local pet-ownership rate higher than the national rate?

You may use the Distribution Calculator in Section 4.1.5 to find the required tail probability.

(a) State $H_0$ and $H_a$.
(b) Compute $\hat{p}$ and the test statistic.
(c) Find the p-value.
(d) State a conclusion at $\alpha = 0.05$.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- "Is the local rate *higher*?" calls for an upper-tailed test: $H_a: p > p_0$.
- Upper-tailed p-value $= P(Z > z^*) = 1 - P(Z \le z^*)$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** $H_0: p = 0.60 \quad$ vs. $\quad H_a: p > 0.60$

**(b)** Sample proportion:
$$\hat{p} = \frac{102}{150} = 0.680$$
Test statistic:
$$z^* = \frac{\hat{p} - p_0}{\sqrt{p_0(1-p_0)/n}} = \frac{0.680 - 0.60}{\sqrt{0.60 \times 0.40 / 150}} = \frac{0.080}{0.04000} = 2.000$$

**(c)** For $H_a: p > 0.60$ (upper-tailed):
$$\text{p-value} = P(Z > 2.000) = 1 - P(Z \le 2.000) \approx 0.0228$$

**(d)** Since p-value $\approx 0.0228 < \alpha = 0.05$, we **reject $H_0$**. There is sufficient evidence at the 5% level to conclude that the local pet-ownership rate exceeds the national figure of 60%.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 19</div>
A statistics professor conjectures that more than 50% of students pass a particularly challenging midterm. Out of $n = 80$ students, $x = 46$ passed. Test the professor's conjecture.

You may use the Distribution Calculator in Section 4.1.5 to find the required tail probability.

(a) State $H_0$ and $H_a$.
(b) Compute $\hat{p}$ and the test statistic.
(c) Find the p-value.
(d) State a conclusion at $\alpha = 0.05$.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- "More than 50%" means $H_a: p > 0.50$.
- Upper-tailed p-value $= P(Z > z^*)$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** $H_0: p = 0.50 \quad$ vs. $\quad H_a: p > 0.50$

**(b)** Sample proportion:
$$\hat{p} = \frac{46}{80} = 0.575$$
Test statistic:
$$z^* = \frac{\hat{p} - p_0}{\sqrt{p_0(1-p_0)/n}} = \frac{0.575 - 0.50}{\sqrt{0.50 \times 0.50 / 80}} = \frac{0.075}{0.05590} \approx 1.342$$

**(c)** For $H_a: p > 0.50$ (upper-tailed):
$$\text{p-value} = P(Z > 1.342) \approx 0.0899$$

**(d)** Since p-value $\approx 0.090 > \alpha = 0.05$, we **fail to reject $H_0$**. There is not sufficient evidence at the 5% level to conclude that more than 50% of students pass the midterm.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 20</div>
Three independent research scenarios are given below. For **each scenario**: (i) identify the appropriate test (one-sample $z$ test for a mean, one-sample $t$ test for a mean, or one-sample $z$ test for a proportion); (ii) state $H_0$ and $H_a$; (iii) compute the test statistic; (iv) find the p-value; and (v) state a conclusion at the given $\alpha$.

You may use the Distribution Calculator in Section 4.1.5 to find required tail probabilities.

**(a)** A physiotherapist records grip strength for $n = 25$ patients and reports $\bar{x} = 18.8$ kg and $s = 3.5$ kg. The population standard deviation is **unknown**. Historical data show the mean for healthy adults is $\mu_0 = 17.0$ kg. Test whether the sample mean differs from the historical norm at $\alpha = 0.05$.

**(b)** A pollster claims 25% of voters support Candidate X. In a random sample of $n = 200$ voters, $x = 56$ indicate support for Candidate X. Test whether the true support level differs from 25% at $\alpha = 0.05$.

**(c)** A fitness app reports its users average $\mu_0 = 46.0$ minutes of exercise per day. The population standard deviation is **known** to be $\sigma = 6$ minutes. A researcher samples $n = 36$ users and records $\bar{x} = 43.8$ minutes. Test whether the mean is lower than reported at $\alpha = 0.05$.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Use the **$z$ test for a mean** when $\sigma$ is known.
- Use the **$t$ test for a mean** when $\sigma$ is unknown and data are approximately normal.
- Use the **$z$ test for a proportion** when the parameter of interest is a population proportion $p$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a) One-sample $t$ test for a mean** ($\sigma$ unknown)

$H_0: \mu = 17.0$ kg $\quad$ vs. $\quad H_a: \mu \ne 17.0$ kg, with $df = 24$.

$$t^* = \frac{\bar{x} - \mu_0}{s / \sqrt{n}} = \frac{18.8 - 17.0}{3.5 / \sqrt{25}} = \frac{1.8}{0.7} \approx 2.571$$

$$\text{p-value} = 2 \times P(T_{24} > 2.571) \approx 0.0167$$

From the $t$-table: $t_{24,\,0.01} = 2.492 < 2.571 < 2.797 = t_{24,\,0.005}$, so $0.01 < \text{p-value} < 0.02$.

Since p-value $< \alpha = 0.05$, **reject $H_0$**. There is sufficient evidence that mean grip strength differs from 17.0 kg.

---

**(b) One-sample $z$ test for a proportion**

$H_0: p = 0.25 \quad$ vs. $\quad H_a: p \ne 0.25$.

$$\hat{p} = \frac{56}{200} = 0.280, \qquad z^* = \frac{0.280 - 0.25}{\sqrt{0.25 \times 0.75 / 200}} = \frac{0.030}{0.03062} \approx 0.980$$

$$\text{p-value} = 2 \times P(Z > 0.980) \approx 2(0.1635) = 0.3271$$

Since p-value $\approx 0.327 > \alpha = 0.05$, **fail to reject $H_0$**. There is not sufficient evidence that support for Candidate X differs from 25%.

---

**(c) One-sample $z$ test for a mean** ($\sigma$ known)

$H_0: \mu = 46.0$ min $\quad$ vs. $\quad H_a: \mu < 46.0$ min.

$$z^* = \frac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} = \frac{43.8 - 46.0}{6 / \sqrt{36}} = \frac{-2.2}{1.0} = -2.200$$

$$\text{p-value} = P(Z < -2.200) \approx 0.0139$$

Since p-value $\approx 0.014 < \alpha = 0.05$, **reject $H_0$**. There is sufficient evidence that mean daily exercise time for app users is less than 46.0 minutes.
</div>
</details>
