# Hypothesis Tests

## Introduction 

A hypothesis test is a statistical method used to make inferences about a population based on sample data. It involves formulating a null hypothesis and an alternative hypothesis, calculating a test statistic, and determining the likelihood of observing the data under the null hypothesis. Each of these terms will be explained in more detail below.

::: {.definition}
An inferential procedure to determine whether there is sufficient
evidence to suggest a condition for a population parameter using
statistics from a sample.
:::

The main goal of hypothesis testing is to assess the strength of evidence against a null hypothesis, which is a statement of no effect or no difference. The alternative hypothesis represents the research hypothesis or the condition we are interested in testing.

One of the steps in hypothesis testing is to calculate a p-value, which quantifies the evidence against the null hypothesis. The p-value is the probability of observing a test statistic at least as extreme as the one calculated from the sample data, assuming that the null hypothesis is true.

::: {.definition}
The p-value of a hypothesis test is the probability of observing a test statistic at least as extreme as the one calculated in step 3, assuming that the null hypothesis is true by pure chance alone.
:::


The procedure to conduct any hypothesis often follows a standard sequence of steps.

##### Steps {.unnumbered .unlisted}

**[ [Reorder and start with checking assumptions first?] ]{style="color: red"}**


1.  Decide on a level of significance ($\alpha$).

2.  State the null hypothesis ($H_0$) and the alternative hypothesis ($H_a$).
    
3.  Calculate the appropriate test statistic.

4.  Use the test statistic and a reference distribution to calculate a
    p-value.

5.  Compare p-value to $\alpha$ to make a conclusion.

We will now examine each of the steps above in more detail.

**Step 1: Decide on a Level of Significance ($\alpha$)**

This is a threshold for decision making which the study or researcher is willing to implement.
Depends on tolerance for consequences of errors, sample size, nature of
the study, and variability.

Some common Common values are $\alpha = 0.10$, $\alpha = 0.05$, $\alpha = 0.01$.


**Step 2: State the Null Hypothesis and the Alternative Hypothesis**

We will use the following notation for describing the null and alternative hypotheses:

<table class="param-table">
  <tr>
    <pd class="term">$\Theta$</td>
    <pd class="sep"> : </pd>
    <pd class="alt">Parameter of interest.</pd>
  </tr>
</table>

<table class="param-table">
  <tr>
    <pd class="term">$\Theta_0$</td>
    <pd class="sep"> : </pd>
    <pd class="alt">Numerical value of the parameter of interest hypothesized under the null hypothesis.</pd>
  </tr>
</table>

<!-- <table class="param-table"> -->
<!--   <tr> -->
<!--     <pd class="term">$\Theta_0$</pd> -->
<!--     <pd class="sep"> : </pd> -->
<!--     <pd class="alt">Numerical value of the parameter of interest hypothesized under the null hypothesis.</pd> -->
<!--   </tr> -->
<!-- </table> -->

<!-- $$\begin{array}{lcl} -->
<!-- \Theta    & : & \text{parameter of interest}\\ -->
<!-- \Theta_0  & : & \text{numerical value of the parameter of interest hypothesized under the}\\ -->
<!--           &   & \text{null hypothesis.} -->
<!-- \end{array}$$ -->

A hypothesis test is a comparison between two mutually exclusive hypotheses which are:



<table class="definition-table">
  <tr>
    <td>Null hypothesis $(H_0)$</td>
    <td>: </td>
    <td>Represents the current belief or the conservative belief or the status quo. It is a statement of no effect or no difference.</td>
  </tr>
  <tr>
    <td>Alternative hypothesis $(H_a)$</td>
    <td>: </td>
    <td>Represents the research hypothesis, or what you are requested to test or trying to prove.</td>
  </tr>
</table>




<!-- ::: {.smalltblfit} -->
<!-- | Null hypothesis        | Represents the current belief or the conservative Represents the current belief or the conservative Represents the current belief or the conservative Represents the current belief or the conservative Represents the current belief or the conservative |  -->
<!-- | :--------------------- | :----------   |  -->
<!-- | Alternative hypothesis | $x_1$ | -->
<!-- ::: -->



<!-- $$\begin{array}{lcl} -->
<!-- \text{Null hypothesis} \, (H_0)         & : & \text{Represents the current belief or the conservative}\\ -->
<!--                                         &   & \text{belief or the status quo. It is a statement of no effect}\\ -->
<!--                                         &   & \text{or no difference.}\\ -->
<!-- \text{Alternative hypothesis} \, (H_a)  & : & \text{Represents the research hypothesis, or what you are}\\ -->
<!--                                         &   & \text{requested to test or trying to prove).} -->
<!-- \end{array}$$ -->

Our choices for the null and alternative hypothesis are:

<!-- $$\begin{array}{lccl} -->
<!-- H_0: \Theta = \Theta_0 & (\Theta \leq \Theta_0) & \quad & H_a: \Theta > \Theta_0 \quad {\text{one-sided (one-tailed)}} \\ -->
<!-- H_0: \Theta = \Theta_0 & (\Theta \geq \Theta_0) &       & H_a: \Theta < \Theta_0 \quad {\text{one-sided (one-tailed)}} \\ -->
<!-- H_0: \Theta = \Theta_0 &                        &       & H_a: \Theta \neq \Theta_0 \quad {\text{two-sided (two-tailed)}} -->
<!-- \end{array}$$ -->

<!-- $$\begin{array}{l lcl} -->
<!-- 1. &  H_0: \Theta = \Theta_0 & \quad  & H_a: \Theta > \Theta_0\\ -->
<!-- 2. &  H_0: \Theta = \Theta_0 &        & H_a: \Theta < \Theta_0\\ -->
<!-- 3. &  H_0: \Theta = \Theta_0 &        & H_a: \Theta \neq \Theta_0 -->
<!-- \end{array}$$ -->

<!-- - $H_0: \Theta = \Theta_0$ vs. $H_a: \Theta > \Theta_0$ -->
<!-- - $H_0: \Theta = \Theta_0$ vs. $H_a: \Theta < \Theta_0$ -->
<!-- - $H_0: \Theta = \Theta_0$ vs. $H_a: \Theta \neq \Theta$ -->


<table class="hypothesis-table">
  <tr>
    <td class="term">$H_0: \Theta = \Theta_0$</td>
    <td class="sep">vs.</td>
    <td class="alt">$H_a: \Theta > \Theta_0$</td>
  </tr>
  <tr>
    <td class="term">$H_0: \Theta = \Theta_0$</td>
    <td class="sep">vs.</td>
    <td class="alt">$H_a: \Theta < \Theta_0$</td>
  </tr>
  <tr>
    <td class="term">$H_0: \Theta = \Theta_0$</td>
    <td class="sep">vs.</td>
    <td class="alt">$H_a: \Theta \neq \Theta_0$</td>
  </tr>
</table>





:::{.remark}
The first two hypothesis tests above are referred to as one-sided or one-tailed tests.
and the third hypothesis test is referred to as a two-sided or two-tailed test.
:::

:::{.remark}
The one-sided hypothesis test 
$$\begin{array}{lcl}
H_0: \Theta = \Theta_0 & \quad  & H_a: \Theta > \Theta_0\\
\end{array}$$
is equivalent to
$$\begin{array}{lcl}
H_0: \Theta \leq \Theta_0 & \quad  & H_a: \Theta > \Theta_0\\
\end{array}$$

and the other one-sided hypothesis test 

$$\begin{array}{lcl}
H_0: \Theta = \Theta_0 & \quad  & H_a: \Theta < \Theta_0\\
\end{array}$$
is equivalent to
$$\begin{array}{lcl}
H_0: \Theta \geq \Theta_0 & \quad  & H_a: \Theta < \Theta_0\\
\end{array}$$
:::


**Step 3: Calculate an appropriate test statistic**

Depends on the hypothesis test conducted and the information available.

::: definition
$$\text{test statistic} = \frac{\text{(a statistic)} - \text{(hypothesized value of parameters under } H_0 \text{)}}{\text{standard error of statistic}}$$
:::

The test statistic follows a reference distribution. 
Common distributions which thetest statistic follows are the standard normal distrubution, $t$-distribution, 
$F$-distribution and $\chi^2$-distribution.
<!-- $Z$, $t$, $F$, $\chi^2$ -->

**Step 4: Calculate the p-value**

<!-- It quantifies the evidence against the null hypothesis. -->

We use the test statistic, reference distribution, and refer back to $H_a$.

```{r fig.cap="Right-tailed test: p-value is the area to the right of the test statistic", fig.align='center', fig.show='hold', out.width='48%', echo=FALSE}

library(ggplot2)

# --- First Plot: Standard blue bracket lower ---
x <- seq(-4, 4, length = 1000)
y <- dnorm(x)
test_stat <- 1.5
p_x <- seq(test_stat, 4, length = 500)
p_y <- dnorm(p_x)
x_left <- test_stat + 0.1
x_right <- 3.4
y_top <- 0.2
y_mid <- 0.185
y_bot <- 0.17

p1 <- ggplot() +
  geom_line(aes(x, y), color = "black", size = 1.2) +
  geom_area(aes(x = p_x, y = p_y), fill = "skyblue", alpha = 0.6) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  geom_segment(aes(x = test_stat, xend = test_stat, y = 0, yend = dnorm(test_stat)),
               color = "black", linewidth = 1.3) +
  geom_segment(aes(x = x_left, xend = x_left, y = y_bot, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_left, xend = x_right, y = y_mid, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_right, xend = x_right, y = y_mid, yend = y_bot), color = "black") +
  annotate("text", x = (x_left + x_right)/2, y = y_top, label = "p-value", size = 5) +
  annotate("text", x = -2.7, y = 0.35, label = "H[0]:~theta==theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = -2.7, y = 0.29, label = "H[a]:~theta > theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = test_stat, y = 0.005, label = "test stat", size = 4.5, vjust = 1.5) +
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank())

print(p1)

# --- Second Plot: Blue with widest shading and highest bracket ---
y <- dnorm(x)
test_stat <- -1.75
p_x <- seq(test_stat, 4, length = 500)
p_y <- dnorm(p_x)
x_left <- test_stat + 0.3
x_right <- 3.4
y_top <- 0.45
y_mid <- 0.435
y_bot <- 0.42

p2 <- ggplot() +
  geom_line(aes(x, y), color = "black", size = 1.2) +
  geom_area(aes(x = p_x, y = p_y), fill = "skyblue", alpha = 0.6) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  geom_segment(aes(x = test_stat, xend = test_stat, y = 0, yend = dnorm(test_stat)),
               color = "black", linewidth = 1.3) +
  geom_segment(aes(x = x_left, xend = x_left, y = y_bot, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_left, xend = x_right, y = y_mid, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_right, xend = x_right, y = y_mid, yend = y_bot), color = "black") +
  annotate("text", x = (x_left + x_right)/2, y = y_top, label = "p-value", size = 5) +
  annotate("text", x = -2.7, y = 0.35, label = "H[0]:~theta==theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = -2.7, y = 0.29, label = "H[a]:~theta > theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = test_stat, y = 0.005, label = "test stat", size = 4.5, vjust = 1.5) +
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank())

print(p2)
```

```{r fig.cap="Left-tailed test: p-value is the area to the left of the test statistic", fig.align='center', fig.show='hold', out.width='48%', echo=FALSE}

# First plot (lower pink bracket)
library(ggplot2)
x <- seq(-4, 4, length = 1000)
y <- dnorm(x)
test_stat <- -1.5
p_x <- seq(-4, test_stat, length = 500)
p_y <- dnorm(p_x)
x_left <- -3.4
x_right <- test_stat - 0.1
y_top <- 0.17
y_mid <- 0.155
y_bot <- 0.14
p1 <- ggplot() +
  geom_line(aes(x, y), color = "black", linewidth = 1.2) +
  geom_area(aes(x = p_x, y = p_y), fill = "pink", alpha = 0.6) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  geom_segment(aes(x = test_stat, xend = test_stat, y = 0, yend = dnorm(test_stat)), color = "black", linewidth = 1.3) +
  geom_segment(aes(x = x_left, xend = x_left, y = y_bot, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_left, xend = x_right, y = y_mid, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_right, xend = x_right, y = y_mid, yend = y_bot), color = "black") +
  annotate("text", x = (x_left + x_right) / 2, y = y_top, label = "p-value", size = 5) +
  annotate("text", x = 2.5, y = 0.35, label = "H[0]:~theta==theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = 2.5, y = 0.29, label = "H[a]:~theta < theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = test_stat, y = 0.005, label = "test stat", size = 4.5, vjust = 1.5) +
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), panel.grid = element_blank())

print(p1)

# Second plot (curve slightly higher, bracket above)
y <- dnorm(x) * 0.95
test_stat <- 1.5
p_x <- seq(-4, test_stat, length = 500)
p_y <- dnorm(p_x) * 0.95
x_left <- -3.4
x_right <- test_stat - 0.1
y_top <- 0.43
y_mid <- 0.3989
y_bot <- 0.39
p2 <- ggplot() +
  geom_line(aes(x, y), color = "black", linewidth = 1.2) +
  geom_area(aes(x = p_x, y = p_y), fill = "pink", alpha = 0.6) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  geom_segment(aes(x = test_stat, xend = test_stat, y = 0, yend = dnorm(test_stat) * 0.95), color = "black", linewidth = 1.3) +
  geom_segment(aes(x = x_left, xend = x_left, y = y_bot, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_left, xend = x_right, y = y_mid, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_right, xend = x_right, y = y_mid, yend = y_bot), color = "black") +
  annotate("text", x = (x_left + x_right) / 2, y = y_top, label = "p-value", size = 5) +
  annotate("text", x = 2.5, y = 0.34, label = "H[0]:~theta==theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = 2.5, y = 0.28, label = "H[a]:~theta < theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = test_stat, y = 0.005, label = "test stat", size = 4.5, vjust = 1.5) +
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), panel.grid = element_blank())

print(p2)
```


```{r fig.cap="Two-tailed test: p-value is the total area in both tails beyond ±test statistic", echo=FALSE, fig.align='center', out.width='70%'}
library(ggplot2)

# Generate bell curve
x <- seq(-4, 4, length = 1000)
y <- dnorm(x) * 0.95

# Test statistics for two tails
left_stat <- -1.75
right_stat <- 1.75

# Shaded green tail areas
p_x_left <- seq(-4, left_stat, length = 300)
p_y_left <- dnorm(p_x_left) * 0.95
p_x_right <- seq(right_stat, 4, length = 300)
p_y_right <- dnorm(p_x_right) * 0.95

# Bracket height - slightly higher than previous "blue" reference
y_top <- 0.11
y_mid <- 0.095
y_bot <- 0.08

# Bracket horizontal positions
x_l1 <- -3.4
x_l2 <- left_stat - 0.1
x_r1 <- right_stat + 0.1
x_r2 <- 3.4

# Create plot
p <- ggplot() +
  geom_line(aes(x, y), color = "black", linewidth = 1.2) +
  geom_area(aes(x = p_x_left, y = p_y_left), fill = "lightgreen", alpha = 0.6) +
  geom_area(aes(x = p_x_right, y = p_y_right), fill = "lightgreen", alpha = 0.6) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  geom_segment(aes(x = left_stat, xend = left_stat, y = 0, yend = dnorm(left_stat) * 0.95),
               color = "black", linewidth = 1.3) +
  geom_segment(aes(x = right_stat, xend = right_stat, y = 0, yend = dnorm(right_stat) * 0.95),
               color = "black", linewidth = 1.3) +

  # Left bracket
  geom_segment(aes(x = x_l1, xend = x_l1, y = y_bot, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_l1, xend = x_l2, y = y_mid, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_l2, xend = x_l2, y = y_mid, yend = y_bot), color = "black") +
  annotate("text", x = (x_l1 + x_l2)/2, y = y_top, label = "1/2 p-value", size = 5) +

  # Right bracket
  geom_segment(aes(x = x_r1, xend = x_r1, y = y_bot, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_r1, xend = x_r2, y = y_mid, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_r2, xend = x_r2, y = y_mid, yend = y_bot), color = "black") +
  annotate("text", x = (x_r1 + x_r2)/2, y = y_top, label = "1/2 p-value", size = 5) +

  # Hypotheses on the right
  annotate("text", x = 2.5, y = 0.25, label = "H[0]:~theta==theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = 2.5, y = 0.195, label = "H[a]:~theta != theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = left_stat, y = 0.005, label = "test stat", size = 4.5, vjust = 1.5) +
  annotate("text", x = right_stat, y = 0.005, label = "test stat", size = 4.5, vjust = 1.5) +

  # Style
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank()
  )
# PRINT the plot
p
```



**Step 5: Compare *p*-value to level of significance $\alpha$ and make a conclusion**

- If the *p*-value $< \alpha$, we say there is sufficient evidence against $H_0$ and the hypothesis test rejects $H_0$
  in favor of $H_a$.

- If the *p*-value $> \alpha$, thenwe say there is insufficient evidence against $H_0$ and we do not reject $H_0$ 
  (in this situation we can also say fail to reject $H_0$).

<!-- **Note:** It is not good practice to give conclusions in the context of -->
<!-- stating we *accept $H_0$* or *accept $H_a$*. -->


**About the p-value of the test statistics**

- The p-value is a conditional probability.

- It is not the probability that $H_0$ (null hypothesis: current belief)
  is true.

- It is: P(observed statistic value --- $H_0$).
  Given $H_0$ (the null hypothesis), because $H_0$ gives the parameter
  values that we need to find required probability.

- The p-value serves as a measure of the strength of the evidence against
  the null hypothesis (but it should not serve as a hard and fast rule
  for decision).

- If the p-value = 0.03 (for example) all we can say is that there is 3%
  chance of observing the statistic value we actually observed (or one
  even more inconsistent with the null value).

- The p-value is the chance (the proportion) of getting a, for instance,
  $\hat{p}$ as far as or further from $H_0$ than the value observed.

- The p-value is the probability of getting at least something (e.g., sample
  proportion $\hat{p}$) more extreme (e.g., unusual, unlikely, or rare)
  than what we have already found (our observed value of $\hat{p}$) that
  provide even stronger evidence against $H_0$.

- The more extreme the z-score (large in absolute values) are the ones
  that denote farther departure of the observed value (e.g., our
  $\hat{p}$) from the parameter value ($p_0$) in $H_0$.

- In the one-sided test, e.g., $H_a : p > p_0$, p-value is one-tailed
  probability. This is the probability that sample proportion $\hat{p}$
  falls at least as far from $p_0$ in one direction as the observed
  value of $\hat{p}$.

- In the two-sided test, e.g., $H_a : p \ne p_0$, p-value is two-tailed
  probability. This is the probability that sample proportion $\hat{p}$
  falls at least as far from $p_0$ in either direction as the observed
  value of $\hat{p}$.

The probability, computed assuming that $H_0$ is true, that the test
statistic would take a value as extreme or more extreme than that
actually observed is called the **P-value** of the test. The smaller the
P-value, the stronger the evidence against $H_0$ provided by the data.

Small P-values are evidence against $H_0$, because they say that the
observed result is unlikely to occur when $H_0$ is true. Large P-values
fail to give evidence against $H_0$.

**The P-value Scale**

- If P-value $<$ 0.001, we have very strong evidence against $H_0$.

- If 0.001 $\leq$ P-value $<$ 0.01, we have strong evidence against
  $H_0$.

- If 0.01 $\leq$ P-value $<$ 0.05, we have evidence against $H_0$.

- If 0.05 $\leq$ P-value $<$ 0.075, we have some evidence against $H_0$.

- If 0.075 $\leq$ P-value $<$ 0.10, we have slight evidence against
  $H_0$.

**Use p-value Method to Make a Decision (Reject or Fail to Reject
$H_0$)**

But how small is small p-value?\
We would need to choose an $\alpha$-level (significance-level): a number
such that if:

- $P\text{–value} \leq \alpha$-level, we reject $H_0$; We can conclude
  $H_a$ (we have evidence to support our claim). Often we phrase as a
  statistically significant result at that specified $\alpha$-level.

- $P\text{–value} > \alpha$-level, we fail to reject $H_0$; We cannot
  conclude $H_a$ (we have not enough evidence to support our claim;
  thus, $H_0$ is plausible - We do not accept $H_0$). Often we phrase as
  the result is not statistically significant at that specified
  $\alpha$-level.

- The default $\alpha$-level (significance-level) is typically
  $\alpha = 0.05$ (but it can be different based on the context of the
  study - it is usually not higher than 0.10).

The p-value in the previous example was extremely small (less than
0.0001). That is a strong evidence to suggest that more than 5% of
children have genetic abnormalities. However, it does not say that the
percentage of sampled children with genetic abnormalities was "a lot
more than 5%". That is, the p-value by itself says nothing about how
much greater the percentage might be. The confidence interval provides
that information.\
To assess the difference in practical terms, we should also construct a
confidence interval:

$$0.1198 \pm (1.96 \times 0.0166)$$ $$0.1198 \pm 0.0324$$
$$(0.0874,\ 0.1522)$$

Interpretation: We are 95% Confident that the true percentage of
children with genetic abnormalities is between 8.74% and 15.22%.\
95% CI for $p$: (9.1%, 15.6%) -- We are 95% confident that the true
percentage of all children that have genetic abnormalities is between
approximately 9.1% and 15.6%. Since both values of this CI are more than
the hypothesized value of $p = 0.05$ (5%), we can further infer that
this true percentage is more than 5%.\
**Do environmental chemicals cause congenital abnormalities?**

We do not know that environmental chemicals cause genetic abnormalities.
We merely have evidence that suggests that a greater percentage of
children are diagnosed with genetic abnormalities now, compared to the
1980s.

**More About P-values**

- Big p-values just mean that what we have observed is not surprising.
  It means that the results are in line with our assumption that the
  null hypothesis models the world, so we have no reason to reject it.

- A big p-value does not prove that the null hypothesis is true.

- When we see a big p-value, all we can say is: we cannot reject $H_0$
  (we fail to reject $H_0$) -- we cannot conclude $H_a$ (We have no
  evidence to support $H_a$).
