# Hypothesis Tests

## Introduction 

A hypothesis test is a statistical method used to make inferences about a population based on sample data. It involves formulating a null hypothesis and an alternative hypothesis, calculating a test statistic, and determining the likelihood of observing the data under the null hypothesis. Each of these terms will be explained in more detail below.

::: {.definition}
An inferential procedure to determine whether there is sufficient
evidence to suggest a condition for a population parameter using
statistics from a sample.
:::

The main goal of hypothesis testing is to assess the strength of evidence against a null hypothesis, which is a statement of no effect or no difference. The alternative hypothesis represents the research hypothesis or the condition we are interested in testing.

One of the steps in hypothesis testing is to calculate a p-value, which quantifies the evidence against the null hypothesis. The p-value is the probability of observing a test statistic at least as extreme as the one calculated from the sample data, assuming that the null hypothesis is true.

::: {.definition}
The p-value of a hypothesis test is the probability of observing a test statistic at least as extreme as the one calculated in step 3, assuming that the null hypothesis is true by pure chance alone.
:::


The procedure to conduct any hypothesis often follows a standard sequence of steps.

##### Steps {.unnumbered .unlisted}

**[ [Reorder and start with checking assumptions first?] ]{style="color: red"}**


1.  Decide on a level of significance ($\alpha$).

2.  State the null hypothesis ($H_0$) and the alternative hypothesis ($H_a$).
    
3.  Calculate the appropriate test statistic.

4.  Use the test statistic and a reference distribution to calculate a
    p-value.

5.  Compare p-value to $\alpha$ to make a conclusion.

We will now examine each of the steps above in more detail.

**Step 1: Decide on a Level of Significance ($\alpha$)**

This is a threshold for decision making which the study or researcher is willing to implement.
Depends on tolerance for consequences of errors, sample size, nature of
the study, and variability.

Some common Common values are $\alpha = 0.10$, $\alpha = 0.05$, $\alpha = 0.01$.


**Step 2: State the Null Hypothesis and the Alternative Hypothesis**

We will use the following notation for describing the null and alternative hypotheses:

<table class="param-table">
  <tr>
    <pd class="term">$\Theta$</td>
    <pd class="sep"> : </pd>
    <pd class="alt">Parameter of interest.</pd>
  </tr>
</table>

<table class="param-table">
  <tr>
    <pd class="term">$\Theta_0$</td>
    <pd class="sep"> : </pd>
    <pd class="alt">Numerical value of the parameter of interest hypothesized under the null hypothesis.</pd>
  </tr>
</table>

<!-- <table class="param-table"> -->
<!--   <tr> -->
<!--     <pd class="term">$\Theta_0$</pd> -->
<!--     <pd class="sep"> : </pd> -->
<!--     <pd class="alt">Numerical value of the parameter of interest hypothesized under the null hypothesis.</pd> -->
<!--   </tr> -->
<!-- </table> -->

<!-- $$\begin{array}{lcl} -->
<!-- \Theta    & : & \text{parameter of interest}\\ -->
<!-- \Theta_0  & : & \text{numerical value of the parameter of interest hypothesized under the}\\ -->
<!--           &   & \text{null hypothesis.} -->
<!-- \end{array}$$ -->

A hypothesis test is a comparison between two mutually exclusive hypotheses which are:



<table class="definition-table">
  <tr>
    <td>Null hypothesis $(H_0)$</td>
    <td>: </td>
    <td>Represents the current belief or the conservative belief or the status quo. It is a statement of no effect or no difference.</td>
  </tr>
  <tr>
    <td>Alternative hypothesis $(H_a)$</td>
    <td>: </td>
    <td>Represents the research hypothesis, or what you are requested to test or trying to prove.</td>
  </tr>
</table>




<!-- ::: {.smalltblfit} -->
<!-- | Null hypothesis        | Represents the current belief or the conservative Represents the current belief or the conservative Represents the current belief or the conservative Represents the current belief or the conservative Represents the current belief or the conservative |  -->
<!-- | :--------------------- | :----------   |  -->
<!-- | Alternative hypothesis | $x_1$ | -->
<!-- ::: -->



<!-- $$\begin{array}{lcl} -->
<!-- \text{Null hypothesis} \, (H_0)         & : & \text{Represents the current belief or the conservative}\\ -->
<!--                                         &   & \text{belief or the status quo. It is a statement of no effect}\\ -->
<!--                                         &   & \text{or no difference.}\\ -->
<!-- \text{Alternative hypothesis} \, (H_a)  & : & \text{Represents the research hypothesis, or what you are}\\ -->
<!--                                         &   & \text{requested to test or trying to prove).} -->
<!-- \end{array}$$ -->

Our choices for the null and alternative hypothesis are:

<!-- $$\begin{array}{lccl} -->
<!-- H_0: \Theta = \Theta_0 & (\Theta \leq \Theta_0) & \quad & H_a: \Theta > \Theta_0 \quad {\text{one-sided (one-tailed)}} \\ -->
<!-- H_0: \Theta = \Theta_0 & (\Theta \geq \Theta_0) &       & H_a: \Theta < \Theta_0 \quad {\text{one-sided (one-tailed)}} \\ -->
<!-- H_0: \Theta = \Theta_0 &                        &       & H_a: \Theta \neq \Theta_0 \quad {\text{two-sided (two-tailed)}} -->
<!-- \end{array}$$ -->

<!-- $$\begin{array}{l lcl} -->
<!-- 1. &  H_0: \Theta = \Theta_0 & \quad  & H_a: \Theta > \Theta_0\\ -->
<!-- 2. &  H_0: \Theta = \Theta_0 &        & H_a: \Theta < \Theta_0\\ -->
<!-- 3. &  H_0: \Theta = \Theta_0 &        & H_a: \Theta \neq \Theta_0 -->
<!-- \end{array}$$ -->

<!-- - $H_0: \Theta = \Theta_0$ vs. $H_a: \Theta > \Theta_0$ -->
<!-- - $H_0: \Theta = \Theta_0$ vs. $H_a: \Theta < \Theta_0$ -->
<!-- - $H_0: \Theta = \Theta_0$ vs. $H_a: \Theta \neq \Theta$ -->


<table class="hypothesis-table">
  <tr>
    <td class="term">$H_0: \Theta = \Theta_0$</td>
    <td class="sep">vs.</td>
    <td class="alt">$H_a: \Theta > \Theta_0$</td>
  </tr>
  <tr>
    <td class="term">$H_0: \Theta = \Theta_0$</td>
    <td class="sep">vs.</td>
    <td class="alt">$H_a: \Theta < \Theta_0$</td>
  </tr>
  <tr>
    <td class="term">$H_0: \Theta = \Theta_0$</td>
    <td class="sep">vs.</td>
    <td class="alt">$H_a: \Theta \neq \Theta_0$</td>
  </tr>
</table>





:::{.remark}
The first two hypothesis tests above are referred to as one-sided or one-tailed tests.
and the third hypothesis test is referred to as a two-sided or two-tailed test.
:::

:::{.remark}
The one-sided hypothesis test 
$$\begin{array}{lcl}
H_0: \Theta = \Theta_0 & \quad  & H_a: \Theta > \Theta_0\\
\end{array}$$
is equivalent to
$$\begin{array}{lcl}
H_0: \Theta \leq \Theta_0 & \quad  & H_a: \Theta > \Theta_0\\
\end{array}$$

and the other one-sided hypothesis test 

$$\begin{array}{lcl}
H_0: \Theta = \Theta_0 & \quad  & H_a: \Theta < \Theta_0\\
\end{array}$$
is equivalent to
$$\begin{array}{lcl}
H_0: \Theta \geq \Theta_0 & \quad  & H_a: \Theta < \Theta_0\\
\end{array}$$
:::


**Step 3: Calculate an appropriate test statistic**

Depends on the hypothesis test conducted and the information available.

::: definition
$$\text{test statistic} = \frac{\text{(a statistic)} - \text{(hypothesized value of parameters under } H_0 \text{)}}{\text{standard error of statistic}}$$
:::

The test statistic follows a reference distribution. 
Common distributions which thetest statistic follows are the standard normal distrubution, $t$-distribution, 
$F$-distribution and $\chi^2$-distribution.
<!-- $Z$, $t$, $F$, $\chi^2$ -->

**Step 4: Calculate the p-value**

<!-- It quantifies the evidence against the null hypothesis. -->

We use the test statistic, reference distribution, and refer back to $H_a$.

```{r fig.cap="Right-tailed test: p-value is the area to the right of the test statistic", fig.align='center', fig.show='hold', out.width='48%', echo=FALSE}

library(ggplot2)

# --- First Plot: Standard blue bracket lower ---
x <- seq(-4, 4, length = 1000)
y <- dnorm(x)
test_stat <- 1.5
p_x <- seq(test_stat, 4, length = 500)
p_y <- dnorm(p_x)
x_left <- test_stat + 0.1
x_right <- 3.4
y_top <- 0.2
y_mid <- 0.185
y_bot <- 0.17

p1 <- ggplot() +
  geom_line(aes(x, y), color = "black", size = 1.2) +
  geom_area(aes(x = p_x, y = p_y), fill = "skyblue", alpha = 0.6) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  geom_segment(aes(x = test_stat, xend = test_stat, y = 0, yend = dnorm(test_stat)),
               color = "black", linewidth = 1.3) +
  geom_segment(aes(x = x_left, xend = x_left, y = y_bot, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_left, xend = x_right, y = y_mid, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_right, xend = x_right, y = y_mid, yend = y_bot), color = "black") +
  annotate("text", x = (x_left + x_right)/2, y = y_top, label = "p-value", size = 5) +
  annotate("text", x = -2.7, y = 0.35, label = "H[0]:~theta==theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = -2.7, y = 0.29, label = "H[a]:~theta > theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = test_stat, y = 0.005, label = "test stat", size = 4.5, vjust = 1.5) +
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank())

print(p1)

# --- Second Plot: Blue with widest shading and highest bracket ---
y <- dnorm(x)
test_stat <- -1.75
p_x <- seq(test_stat, 4, length = 500)
p_y <- dnorm(p_x)
x_left <- test_stat + 0.3
x_right <- 3.4
y_top <- 0.45
y_mid <- 0.435
y_bot <- 0.42

p2 <- ggplot() +
  geom_line(aes(x, y), color = "black", size = 1.2) +
  geom_area(aes(x = p_x, y = p_y), fill = "skyblue", alpha = 0.6) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  geom_segment(aes(x = test_stat, xend = test_stat, y = 0, yend = dnorm(test_stat)),
               color = "black", linewidth = 1.3) +
  geom_segment(aes(x = x_left, xend = x_left, y = y_bot, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_left, xend = x_right, y = y_mid, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_right, xend = x_right, y = y_mid, yend = y_bot), color = "black") +
  annotate("text", x = (x_left + x_right)/2, y = y_top, label = "p-value", size = 5) +
  annotate("text", x = -2.7, y = 0.35, label = "H[0]:~theta==theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = -2.7, y = 0.29, label = "H[a]:~theta > theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = test_stat, y = 0.005, label = "test stat", size = 4.5, vjust = 1.5) +
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank())

print(p2)
```

```{r fig.cap="Left-tailed test: p-value is the area to the left of the test statistic", fig.align='center', fig.show='hold', out.width='48%', echo=FALSE}

# First plot (lower pink bracket)
library(ggplot2)
x <- seq(-4, 4, length = 1000)
y <- dnorm(x)
test_stat <- -1.5
p_x <- seq(-4, test_stat, length = 500)
p_y <- dnorm(p_x)
x_left <- -3.4
x_right <- test_stat - 0.1
y_top <- 0.17
y_mid <- 0.155
y_bot <- 0.14
p1 <- ggplot() +
  geom_line(aes(x, y), color = "black", linewidth = 1.2) +
  geom_area(aes(x = p_x, y = p_y), fill = "pink", alpha = 0.6) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  geom_segment(aes(x = test_stat, xend = test_stat, y = 0, yend = dnorm(test_stat)), color = "black", linewidth = 1.3) +
  geom_segment(aes(x = x_left, xend = x_left, y = y_bot, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_left, xend = x_right, y = y_mid, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_right, xend = x_right, y = y_mid, yend = y_bot), color = "black") +
  annotate("text", x = (x_left + x_right) / 2, y = y_top, label = "p-value", size = 5) +
  annotate("text", x = 2.5, y = 0.35, label = "H[0]:~theta==theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = 2.5, y = 0.29, label = "H[a]:~theta < theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = test_stat, y = 0.005, label = "test stat", size = 4.5, vjust = 1.5) +
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), panel.grid = element_blank())

print(p1)

# Second plot (curve slightly higher, bracket above)
y <- dnorm(x) * 0.95
test_stat <- 1.5
p_x <- seq(-4, test_stat, length = 500)
p_y <- dnorm(p_x) * 0.95
x_left <- -3.4
x_right <- test_stat - 0.1
y_top <- 0.43
y_mid <- 0.3989
y_bot <- 0.39
p2 <- ggplot() +
  geom_line(aes(x, y), color = "black", linewidth = 1.2) +
  geom_area(aes(x = p_x, y = p_y), fill = "pink", alpha = 0.6) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  geom_segment(aes(x = test_stat, xend = test_stat, y = 0, yend = dnorm(test_stat) * 0.95), color = "black", linewidth = 1.3) +
  geom_segment(aes(x = x_left, xend = x_left, y = y_bot, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_left, xend = x_right, y = y_mid, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_right, xend = x_right, y = y_mid, yend = y_bot), color = "black") +
  annotate("text", x = (x_left + x_right) / 2, y = y_top, label = "p-value", size = 5) +
  annotate("text", x = 2.5, y = 0.34, label = "H[0]:~theta==theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = 2.5, y = 0.28, label = "H[a]:~theta < theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = test_stat, y = 0.005, label = "test stat", size = 4.5, vjust = 1.5) +
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), panel.grid = element_blank())

print(p2)
```


```{r fig.cap="Two-tailed test: p-value is the total area in both tails beyond ±test statistic", echo=FALSE, fig.align='center', out.width='70%'}
library(ggplot2)

# Generate bell curve
x <- seq(-4, 4, length = 1000)
y <- dnorm(x) * 0.95

# Test statistics for two tails
left_stat <- -1.75
right_stat <- 1.75

# Shaded green tail areas
p_x_left <- seq(-4, left_stat, length = 300)
p_y_left <- dnorm(p_x_left) * 0.95
p_x_right <- seq(right_stat, 4, length = 300)
p_y_right <- dnorm(p_x_right) * 0.95

# Bracket height - slightly higher than previous "blue" reference
y_top <- 0.11
y_mid <- 0.095
y_bot <- 0.08

# Bracket horizontal positions
x_l1 <- -3.4
x_l2 <- left_stat - 0.1
x_r1 <- right_stat + 0.1
x_r2 <- 3.4

# Create plot
p <- ggplot() +
  geom_line(aes(x, y), color = "black", linewidth = 1.2) +
  geom_area(aes(x = p_x_left, y = p_y_left), fill = "lightgreen", alpha = 0.6) +
  geom_area(aes(x = p_x_right, y = p_y_right), fill = "lightgreen", alpha = 0.6) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1) +
  geom_segment(aes(x = left_stat, xend = left_stat, y = 0, yend = dnorm(left_stat) * 0.95),
               color = "black", linewidth = 1.3) +
  geom_segment(aes(x = right_stat, xend = right_stat, y = 0, yend = dnorm(right_stat) * 0.95),
               color = "black", linewidth = 1.3) +

  # Left bracket
  geom_segment(aes(x = x_l1, xend = x_l1, y = y_bot, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_l1, xend = x_l2, y = y_mid, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_l2, xend = x_l2, y = y_mid, yend = y_bot), color = "black") +
  annotate("text", x = (x_l1 + x_l2)/2, y = y_top, label = "1/2 p-value", size = 5) +

  # Right bracket
  geom_segment(aes(x = x_r1, xend = x_r1, y = y_bot, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_r1, xend = x_r2, y = y_mid, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_r2, xend = x_r2, y = y_mid, yend = y_bot), color = "black") +
  annotate("text", x = (x_r1 + x_r2)/2, y = y_top, label = "1/2 p-value", size = 5) +

  # Hypotheses on the right
  annotate("text", x = 2.5, y = 0.25, label = "H[0]:~theta==theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = 2.5, y = 0.195, label = "H[a]:~theta != theta[0]", parse = TRUE, size = 6) +
  annotate("text", x = left_stat, y = 0.005, label = "test stat", size = 4.5, vjust = 1.5) +
  annotate("text", x = right_stat, y = 0.005, label = "test stat", size = 4.5, vjust = 1.5) +

  # Style
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank()
  )
# PRINT the plot
p
```



**Step 5: Compare *p*-value to level of significance $\alpha$ and make a conclusion**

- If the *p*-value $< \alpha$, we say there is sufficient evidence against $H_0$ and the hypothesis test rejects $H_0$
  in favor of $H_a$.

- If the *p*-value $> \alpha$, thenwe say there is insufficient evidence against $H_0$ and we do not reject $H_0$ 
  (in this situation we can also say fail to reject $H_0$).

<!-- **Note:** It is not good practice to give conclusions in the context of -->
<!-- stating we *accept $H_0$* or *accept $H_a$*. -->


**About the p-value of the test statistics**

- The p-value is a conditional probability.

- It is not the probability that $H_0$ (null hypothesis: current belief)
  is true.

- It is: P(observed statistic value --- $H_0$).
  Given $H_0$ (the null hypothesis), because $H_0$ gives the parameter
  values that we need to find required probability.

- The p-value serves as a measure of the strength of the evidence against
  the null hypothesis (but it should not serve as a hard and fast rule
  for decision).

- If the p-value = 0.03 (for example) all we can say is that there is 3%
  chance of observing the statistic value we actually observed (or one
  even more inconsistent with the null value).

- The p-value is the chance (the proportion) of getting a, for instance,
  $\hat{p}$ as far as or further from $H_0$ than the value observed.

- The p-value is the probability of getting at least something (e.g., sample
  proportion $\hat{p}$) more extreme (e.g., unusual, unlikely, or rare)
  than what we have already found (our observed value of $\hat{p}$) that
  provide even stronger evidence against $H_0$.

- The more extreme the z-score (large in absolute values) are the ones
  that denote farther departure of the observed value (e.g., our
  $\hat{p}$) from the parameter value ($p_0$) in $H_0$.

- In the one-sided test, e.g., $H_a : p > p_0$, p-value is one-tailed
  probability. This is the probability that sample proportion $\hat{p}$
  falls at least as far from $p_0$ in one direction as the observed
  value of $\hat{p}$.

- In the two-sided test, e.g., $H_a : p \ne p_0$, p-value is two-tailed
  probability. This is the probability that sample proportion $\hat{p}$
  falls at least as far from $p_0$ in either direction as the observed
  value of $\hat{p}$.

The probability, computed assuming that $H_0$ is true, that the test
statistic would take a value as extreme or more extreme than that
actually observed is called the **P-value** of the test. The smaller the
P-value, the stronger the evidence against $H_0$ provided by the data.

Small P-values are evidence against $H_0$, because they say that the
observed result is unlikely to occur when $H_0$ is true. Large P-values
fail to give evidence against $H_0$.

**The P-value Scale**

- If P-value $<$ 0.001, we have very strong evidence against $H_0$.

- If 0.001 $\leq$ P-value $<$ 0.01, we have strong evidence against
  $H_0$.

- If 0.01 $\leq$ P-value $<$ 0.05, we have evidence against $H_0$.

- If 0.05 $\leq$ P-value $<$ 0.075, we have some evidence against $H_0$.

- If 0.075 $\leq$ P-value $<$ 0.10, we have slight evidence against
  $H_0$.

**Use p-value Method to Make a Decision (Reject or Fail to Reject
$H_0$)**

But how small is small p-value?\
We would need to choose an $\alpha$-level (significance-level): a number
such that if:

- $P\text{–value} \leq \alpha$-level, we reject $H_0$; We can conclude
  $H_a$ (we have evidence to support our claim). Often we phrase as a
  statistically significant result at that specified $\alpha$-level.

- $P\text{–value} > \alpha$-level, we fail to reject $H_0$; We cannot
  conclude $H_a$ (we have not enough evidence to support our claim;
  thus, $H_0$ is plausible - We do not accept $H_0$). Often we phrase as
  the result is not statistically significant at that specified
  $\alpha$-level.

- The default $\alpha$-level (significance-level) is typically
  $\alpha = 0.05$ (but it can be different based on the context of the
  study - it is usually not higher than 0.10).

The p-value in the previous example was extremely small (less than
0.0001). That is a strong evidence to suggest that more than 5% of
children have genetic abnormalities. However, it does not say that the
percentage of sampled children with genetic abnormalities was "a lot
more than 5%". That is, the p-value by itself says nothing about how
much greater the percentage might be. The confidence interval provides
that information.\
To assess the difference in practical terms, we should also construct a
confidence interval:

$$0.1198 \pm (1.96 \times 0.0166)$$ $$0.1198 \pm 0.0324$$
$$(0.0874,\ 0.1522)$$

Interpretation: We are 95% Confident that the true percentage of
children with genetic abnormalities is between 8.74% and 15.22%.\
95% CI for $p$: (9.1%, 15.6%) -- We are 95% confident that the true
percentage of all children that have genetic abnormalities is between
approximately 9.1% and 15.6%. Since both values of this CI are more than
the hypothesized value of $p = 0.05$ (5%), we can further infer that
this true percentage is more than 5%.\
**Do environmental chemicals cause congenital abnormalities?**

We do not know that environmental chemicals cause genetic abnormalities.
We merely have evidence that suggests that a greater percentage of
children are diagnosed with genetic abnormalities now, compared to the
1980s.

**More About P-values**

- Big p-values just mean that what we have observed is not surprising.
  It means that the results are in line with our assumption that the
  null hypothesis models the world, so we have no reason to reject it.

- A big p-value does not prove that the null hypothesis is true.

- When we see a big p-value, all we can say is: we cannot reject $H_0$
  (we fail to reject $H_0$) -- we cannot conclude $H_a$ (We have no
  evidence to support $H_a$).














## One Sample Hypothesis Tests

### On a Population Mean

#### When $\sigma$ is Known


<!-- Draw a simple random sample of size $n$ from a Normal population that -->
<!-- has unknown mean $\mu$ and known standard deviation $\sigma$.  -->
To test $H_0 : \mu = \mu_0$, in the situation where the population standard deviation $\sigma$ is known,
our options for the null and alternative hypotheses are:

- $H_0: \mu = \mu_0$ vs. $H_a: \mu > \mu_0$

- $H_0: \mu = \mu_0$ vs. $H_a: \mu < \mu_0$

- $H_0: \mu = \mu_0$ vs. $H_a: \mu \neq \mu_0$

<!-- <table class="hypothesis-table"> -->
<!--   <tr> -->
<!--     <td class="term" style="text-align: left !important;">$H_0: \mu = \mu_0$</td> -->
<!--     <td class="sep" style="text-align: left !important;">vs.</td> -->
<!--     <td class="alt" style="text-align: left !important;">$H_a: \mu > \mu_0$</td> -->
<!--   </tr> -->
<!--   <tr> -->
<!--     <td class="term" style="text-align: left !important;">$H_0: \mu = \mu_0$</td> -->
<!--     <td class="sep" style="text-align: left !important;">vs.</td> -->
<!--     <td class="alt" style="text-align: left !important;">$H_a: \mu < \mu_0$</td> -->
<!--   </tr> -->
<!--   <tr> -->
<!--     <td class="term" style="text-align: left !important;">$H_0: \mu = \mu_0$</td> -->
<!--     <td class="sep" style="text-align: left !important;">vs.</td> -->
<!--     <td class="alt" style="text-align: left !important;">$H_a: \mu \neq \mu_0$</td> -->
<!--   </tr> -->
<!-- </table> -->


We calculate the one-sample $z$-statistic:

$$z^{*} = \frac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} \sim N(0,1)$$

The reference distribution is the standard normal distribution.

In terms of a variable $Z$ having the standard normal distribution, the
p-value for a test of $H_0$ against:

- $H_a: \mu > \mu_0$ is $P(Z > z^{*})$

- $H_a: \mu < \mu_0$ is $P(Z < z^{*})$

- $H_a: \mu \ne \mu_0$ is $2P(Z > |z^{*}|)$


::::: example
Deer are a common sight on the UTM campus. Suppose an ecologist is
interested in the average mass of adult white-tailed does (female deer)
around the Mississauga campus to determine whether they are healthy for
the upcoming winter. The ecologist captures a sample of 36 adult females
around the UTM and measures the average mass of this sample to be 42.53
kg.

From previous studies conducted in the area, the average mass of healthy
does was reported to be 45 kg. Conduct a hypothesis test at the 5%
significance level to determine whether the mass of does around UTM has
decreased. Assume the standard deviation is known to be 5.25 kg.

**1. Level of significance.** $\alpha = 0.05$

**2. State the null and alternative hypotheses.**
$$H_0: \mu = 45 \qquad H_a: \mu < 45$$

**3. Calculate appropriate test statistic.**

Given: $$n = 36, \quad \bar{x} = 42.53, \quad \sigma = 5.25$$

Since $\sigma$ is known, the test statistic is:
$$z^* = \frac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} = \frac{42.53 - 45}{5.25/\sqrt{36}} = -2.82$$

Reference distribution: standard normal

**4. Calculate p-value**

$$\text{p-value} = P(Z < -2.82) \approx 0.0024$$
```{r fig.cap="Left-tailed p-value for the test statistic \\( z^* = -2.82 \\)", fig.align='center', echo=FALSE, out.width='70%'}
library(ggplot2)

# Normal curve data
x <- seq(-4, 4, length = 1000)
y <- dnorm(x)

# Shaded left-tail region
test_stat <- -2.82
p_x <- seq(-4, test_stat, length = 500)
p_y <- dnorm(p_x)

# Bracket near pink area
x_left <- -3.6
x_right <- test_stat - 0.1
y_bot <- 0.005
y_mid <- 0.028
y_top <- 0.04

# Build plot
p <- ggplot() +
  geom_line(aes(x, y), color = "black", linewidth = 1.4) +
  geom_area(aes(x = p_x, y = p_y), fill = "orchid", alpha = 0.65) +
  geom_hline(yintercept = 0, color = "black", linewidth = 1.1) +

  # Test statistic line and label
  geom_segment(aes(x = test_stat, xend = test_stat, y = 0, yend = dnorm(test_stat)), 
               color = "black", linewidth = 1.2) +
  annotate("text", x = test_stat, y = -0.02, label = "z* = -2.82", size = 6, fontface = "italic") +

  # Bracket and p-value label
  geom_segment(aes(x = x_left, xend = x_left, y = y_bot, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_left, xend = x_right, y = y_mid, yend = y_mid), color = "black") +
  geom_segment(aes(x = x_right, xend = x_right, y = y_mid, yend = y_bot), color = "black") +
  annotate("text", x = (x_left + x_right)/2, y = y_top, label = "p-value", size = 5.2) +

  # Hypotheses
  annotate("text", x = 2.4, y = 0.32, label = "H[0]:~mu==45", parse = TRUE, size = 6.2) +
  annotate("text", x = 2.4, y = 0.26, label = "H[a]:~mu < 45", parse = TRUE, size = 6.2) +

  # Axes
  scale_x_continuous(breaks = c(-4, -2, 0, 2, 4), limits = c(-4, 4), expand = c(0, 0)) +
  scale_y_continuous(limits = c(-0.03, 0.42)) +

  theme_minimal() +
  labs(x = NULL, y = NULL) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  )

p
```


**5. Compare p-value with level of significance $\alpha$ and make a
conclusion:**

$$0.0024 < 0.05 \Rightarrow \text{p-value} < \alpha$$

There is sufficient evidence at the 5% level of significance to reject
the null that does this winter weigh the same as in the past and to
conclude the alternative that does this winter weigh less than 45 kg.

**R code:**

::: tcolorbox
    # Find test stat
    z_test_stat = (42.53 - 45) / (5.25 / sqrt(36))
    z_test_stat
    [1] -2.822857

    # Find the p-value
    # Since the alternative is Ha : mu < 45
    p-value = pnorm(z_test_stat)
    [1] 0.00237989
:::

*Note:* The `pnorm()` function in R, by default, returns the cumulative
probability (area) to the left of the given value.\

**R code: Using BSDA package**

::: tcolorbox
    # Using the BSDA library. install BSDA if it is not already installed.
    # install.packages("BSDA")
    > library(BSDA)
    > # Conduct the z-test with the zsum.test function
    > zsum.test(mean.x = 42.53, sigma.x = 5.24, n.x = 36, mu = 45, alternative = "less")

            One-sample z-Test

    data:  Summarized x
    z = -2.8282, p-value = 0.00234
    alternative hypothesis: true mean is less than 45
    95 percent confidence interval:
     NA 43.96651
    sample estimates:
    mean of x 
        42.53 
:::

**Interpretation:**

There is sufficient evidence at the 5% level of significance to reject
the null hypothesis. We conclude that the average mass of does this
winter is significantly less than 45 kg.
:::::


:::::: example
Diet colas use artificial sweeteners to avoid sugar. These sweeteners
gradually lose their sweetness over time. Manufacturers therefore test
new colas for loss of sweetness before marketing them. Trained tasters
sip the cola along with drinks of standard sweetness and score the cola
on a "sweetness score" of 1 to 10. The cola is then stored for a month
at high temperature to imitate the effect of four months' storage at
room temperature. Each taster scores the cola again after storage. This
is a matched pairs experiment. Our data are the differences (score
before storage minus score after storage) in the tasters' scores. The
bigger these differences, the bigger the loss of sweetness.

Suppose we know that for any cola, the sweetness loss scores vary from
taster to taster according to a Normal distribution with standard
deviation $\sigma = 1$. The mean $\mu$ for all tasters measures loss of
sweetness.

The following are the sweetness losses for a new cola as measured by 10
trained tasters:

$$2.0, 0.4, 0.7, 2.0, -0.4, 2.2, -1.3, 1.2, 1.1, 2.3$$

Are these data good evidence that the cola lost sweetness in storage?\
**Solution**\
$\mu$ = mean sweetness loss for the population of **all** tasters.\
**Step 1:** State hypotheses. $$\begin{aligned}
H_0 &: \mu = 0 \\
H_a &: \mu > 0
\end{aligned}$$ **Step 2:** Test statistic:
$z_\star = \dfrac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} = \dfrac{1.02 - 0}{1 / \sqrt{10}} = 3.23$\
**Step 3:** P-value. $P(Z > z_\star) = P(Z > 3.23) = 0.0006$\
**Step 4:** Conclusion. We would rarely observe a mean as large as 1.02
if $H_0$ were true. The small p-value provides strong evidence against
$H_0$, supporting $H_a: \mu > 0$. That is, the mean sweetness loss is
likely positive.

**R code (Simulation)**

::: tcolorbox
    # n = sample size;
    n<-10;
    mu.zero<-0;
    sigma<-1;
    sigma.xbar<-sigma/sqrt(n);

    # x bar = sample mean with 10 obs;
    x.bar<-rnorm(1,mean=mu.zero,sd=sigma.xbar);
    x.bar;

    ## [1] 0.3265859

    # z.star = test statistic;
    z.star<-(x.bar-mu.zero)/sigma.xbar;
    z.star;

    ## [1] 1.032755
:::

**R code (10,000 Simulations)**

::: tcolorbox
    n <- 10;
    mu.zero <- 0;
    sigma <- 1;
    sigma.xbar <- sigma / sqrt(n);
    # x bar = sample mean with 10 obs;
    # m = number of simulations;
    m <- 10000;
    x.bar <- rnorm(m, mean = mu.zero, sd = sigma.xbar);

    # z.star = test statistic;
    z.star <- (x.bar - mu.zero) / sigma.xbar;
    hist(z.star, xlab = "differences", col = "blue");
:::


```{r fig.cap="Histogram of \\( z^\\star \\) values from 10,000 simulations under \\( H_0 \\)", fig.align='center', echo=FALSE}
n <- 10
mu.zero <- 0
sigma <- 1
sigma.xbar <- sigma / sqrt(n)

# m = number of simulations;
m <- 10000
x.bar <- rnorm(m, mean = mu.zero, sd = sigma.xbar)

# z.star = test statistic;
z.star <- (x.bar - mu.zero) / sigma.xbar

# Create the histogram
hist(z.star, col = "blue", xlab = "differences", main = "Histogram of z.star")
```


**R code (Empirical p-value)**

::: tcolorbox
    ## P-value

    p_value <- length(z.star[z.star > 3.23]) / m;

    p_value

    ## [1] 8e-04
:::
::::::


::: example
The National Center for Health Statistics reports that the systolic
blood pressure for males 35 to 44 years of age has mean 128 and standard
deviation 15.

The medical director of a large company looks at the medical records of
72 executives in this age group and finds that the mean systolic blood
pressure in this sample is $\bar{x} = 126.07$. Is this evidence that the
company's executives have a different mean blood pressure from the
general population?

Suppose we know that executives' blood pressures follow a Normal
distribution with standard deviation $\sigma = 15$.

**Solution:** Let $\mu$ be the mean systolic blood pressure of the
executive population.

1.  **State hypotheses:**\
    $$\begin{aligned}
    H_0 &: \mu = 128 \\
    H_a &: \mu \ne 128
    \end{aligned}$$

2.  **Test statistic:**\
    $$z^{*} = \frac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} = \frac{126.07 - 128}{15 / \sqrt{72}} = -1.09$$

3.  **P-value:**\
    $$2P(Z > |z^{*}|) = 2P(Z > 1.09) = 2(1 - 0.8621) = 0.2758$$

4.  **Conclusion:**\
    More than 27% of the time, a simple random sample of size 72 from
    the general male population would have a mean blood pressure at
    least as far from 128 as that of the executive sample. The observed
    $\bar{x} = 126.07$ is therefore not good evidence that executives
    differ from other men.
:::

<!-- ::: tcolorbox -->
<!-- There are four steps in carrying out a significance test: -->

<!-- 1.  State the hypotheses. -->

<!-- 2.  Calculate the test statistic. -->

<!-- 3.  Find the p-value. -->

<!-- 4.  State your conclusion in the context of your specific setting. -->

<!-- Once you have stated your hypotheses and identified the proper test, you -->
<!-- or your computer can do Steps 2 and 3 by following a recipe. -->
<!-- ::: -->


::: example
Consider the following hypothesis test:

$$\begin{aligned}
H_0 &: \mu = 20 \\
H_a &: \mu < 20
\end{aligned}$$

A sample of 50 provided a sample mean of 19.4. The population standard
deviation is 2.

1.  Compute the value of the test statistic.

2.  What is the p-value?

3.  Using $\alpha = 0.05$, what is your conclusion?

**Solution:**

1.  **Test statistic:**
    $$z^{*} = \frac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} = \frac{19.4 - 20}{2 / \sqrt{50}} = -2.1213$$

2.  **P-value:** $$P(Z < z^{*}) = P(Z < -2.1213) = 0.0169$$

3.  **Conclusion:**\
    Since the p-value $= 0.0169 < \alpha = 0.05$, we reject
    $H_0 : \mu = 20$. We conclude that $\mu < 20$.
:::

::: example
Consider the following hypothesis test:

$$\begin{aligned}
H_0 &: \mu = 25 \\
H_a &: \mu > 25
\end{aligned}$$

A sample of 40 provided a sample mean of 26.4. The population standard
deviation is 6.

1.  Compute the value of the test statistic.

2.  What is the p-value?

3.  Using $\alpha = 0.01$, what is your conclusion?

**Solution:**

1.  **Test statistic:**
    $$z^{*} = \frac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} = \frac{26.4 - 25}{6 / \sqrt{40}} = 1.4757$$

2.  **P-value:** $$P(Z > z^{*}) = P(Z > 1.4757) = 0.0700$$

3.  **Conclusion:**\
    Since the p-value $= 0.0700 > \alpha = 0.01$, we **cannot reject**
    $H_0 : \mu = 25$.\
    We conclude that we don't have enough evidence to claim that
    $\mu > 25$.
:::

::: example
Consider the following hypothesis test:

$$\begin{aligned}
H_0 &: \mu = 15 \\ 
H_a &: \mu \ne 15
\end{aligned}$$

A sample of 50 provided a sample mean of 14.15. The population standard
deviation is 3.

1.  Compute the value of the test statistic.

2.  What is the p-value?

3.  Using $\alpha = 0.05$, what is your conclusion?

**Solution:**

1.  **Test statistic:**
    $$z^{*} = \frac{\bar{x} - \mu_0}{\sigma / \sqrt{n}} = \frac{14.15 - 15}{3 / \sqrt{50}} = -2.0034$$

2.  **P-value:**
    $$2P(Z > |z^{*}|) = 2P(Z > |{-2.0034}|) = 2P(Z > 2.0034) = 0.0451$$

3.  **Conclusion:**\
    Since P-value $= 0.0451 < \alpha = 0.05$, we reject
    $H_0 : \mu = 15$.\
    We conclude that $\mu \ne 15$.

**Confidence Interval Interpretation:**

The 95% confidence interval for $\mu$ is:
$$\bar{x} \pm z^{*} \left( \frac{\sigma}{\sqrt{n}} \right)$$
$$14.15 \pm 1.96 \left( \frac{3}{\sqrt{50}} \right)
= (13.3184,\ 14.9815)$$

Since the hypothesized value $\mu_0 = 15$ falls **outside** this
interval, we again reject $H_0 : \mu = 15$.
:::

<!-- ::: {. remark} -->
<!-- *A level $\alpha$ two-sided significance test rejects a hypothesis -->
<!-- $H_0 : \mu = \mu_0$ exactly when the value $\mu_0$ falls outside a level -->
<!-- $1 - \alpha$ confidence interval for $\mu$.* -->
<!-- ::: -->









#### When $\sigma$ is Not Known

To test $H_0 : \mu = \mu_0$, in the situation where the population standard deviation $\sigma$ is not known,
our options for the null and alternative hypotheses are:

- $H_0: \mu = \mu_0$ vs. $H_a: \mu > \mu_0$

- $H_0: \mu = \mu_0$ vs. $H_a: \mu < \mu_0$

- $H_0: \mu = \mu_0$ vs. $H_a: \mu \neq \mu_0$

We calculate the one-sample $t$-statistic:
$$t^{*} = \frac{\bar{x} - \mu_0}{s / \sqrt{n}} \sim t_{n-1}$$
In terms of a variable $T$ which follows the $t$-distribution at $n-1$ degrees of freedom, the
p-value for a test of $H_0$ against:

- $H_a: \mu > \mu_0$ is $P(T > t^{*})$

- $H_a: \mu < \mu_0$ is $P(T < t^{*})$

- $H_a: \mu \ne \mu_0$ is $2P(T > |t^{*}|)$


::: example
Researchers studied the physiological effects of laughter. They measured
heart rates (in beats per minute) of **$n = 25$** subjects (ages 18--34)
while they laughed. They obtained:
$$\bar{x} = 73.5, \quad s = 6, \quad \alpha = 0.05$$ It is well known
that the resting heart rate is 71 bpm. Is there evidence that the mean
heart rate during laughter exceeds 71 bpm?

**Step 1: State the hypotheses.** $$\begin{aligned}
H_0 &: \mu = 71 \\
H_a &: \mu > 71
\end{aligned}$$

**Step 2: Check assumptions.**

- The sample is an independent random sample of individuals aged 18--34.

- The population of heart rates during laughter is normally distributed.

**Step 3: Compute the test statistic.**

Since $\sigma$ is unknown, we use the $t$ statistic:
$$t^\ast = \frac{\bar{x} - \mu_0}{s / \sqrt{n}} = \frac{73.5 - 71}{6 / \sqrt{25}} = 2.083$$

Reference distribution: $t$ distribution with $n - 1 = 25 - 1 = 24$
degrees of freedom.

**Step 4: Determine the p-value.**

Using the $t$ distribution with 24 df: $$0.01 < \text{p-value} < 0.025$$

**Step 5: Make a conclusion.**

Since $\text{p-value} < \alpha = 0.05$, we reject $H_0$.

*There is sufficient evidence at the 5% level of significance to reject
the null that the mean is 71 bpm in favor of the alternative that the
mean is greater than 71 bpm for people who are laughing.*
:::

::::: example
A researcher is asked to test the hypothesis that the average price of a
2-star (CAA rating) motel room has decreased since last year. Last year,
a study showed that the prices were Normally distributed with a mean of
\$89.50.

A random sample of twelve 2-star motels produced the following room
prices:

$$\text{\$85.00, 92.50, 87.50, 89.90, 90.00, 82.50, 87.50, 90.00, 85.00, 89.00, 91.50, 87.50}$$

At the 5% level of significance, can we conclude that the mean price has
decreased?

**Solution:**

Let $\mu$ be the true average price of a 2-star motel room.

1.  **State hypotheses.** $$\begin{aligned}
    H_0 &: \mu = 89.5 \\
    H_a &: \mu < 89.5
    \end{aligned}$$

2.  **Compute test statistic.**
    $$t^\ast = \frac{\bar{x} - \mu_0}{s / \sqrt{n}} = \frac{88.1583 - 89.5}{2.9203 / \sqrt{12}} = -1.5915$$

3.  **Find the p-value.**\
    With $df = 11$, the p-value (from the $t$-distribution table) is
    between 0.05 and 0.10.

4.  **Conclusion.**\
    Since the p-value $> 0.05$, we **fail to reject** $H_0$.\
    There is not sufficient evidence to conclude that the average price
    of 2-star motels has decreased this year.

**R code (One Sample t-test)**

::: tcolorbox
    # Step 1. Entering data;
    prices=c(85.00, 92.50, 87.50, 89.90, 90.00, 82.50,
             87.50, 90.00, 85.00, 89.00, 91.50, 87.50);

    # Step 2. Hypothesis test;
    t.test(prices, alternative="less", mu=89.5);
:::

**R output**

::: tcolorbox
    ## 
    ##  One Sample t-test
    ## 
    ## data:  prices
    ## t = -1.5915, df = 11, p-value = 0.0699
    ## alternative hypothesis: true mean is less than 89.5
    ## 95 percent confidence interval:
    ##  -Inf 89.67229
    ## sample estimates:
    ## mean of x 
    ##  88.15833 
:::
:::::

::: tcolorbox
Draw an SRS of size $n$ from a large population having unknown mean
$\mu$.

To *test the hypothesis* $H_0 : \mu = \mu_0$, compute the *one-sample
$t$ statistic* $$t^\ast = \frac{\bar{x} - \mu_0}{s / \sqrt{n}}$$

In terms of a variable $T$ having the $t_{n - 1}$ distribution, the
p-value for a test of $H_0$ against

$$\begin{aligned}
H_a : \mu > \mu_0 &\quad \text{is} \quad P(T \ge t^\ast) \\
H_a : \mu < \mu_0 &\quad \text{is} \quad P(T \le t^\ast) \\
H_a : \mu \ne \mu_0 &\quad \text{is} \quad 2P(T \ge |t^\ast|)
\end{aligned}$$

These p-values are exact if the population distribution is Normal and
are approximately correct for large $n$ in other cases.
:::

:::: example
We are conducting a two-sided one-sample $t$-test for the hypotheses:
$$\begin{aligned}
H_0 &: \mu = 64 \\
H_a &: \mu \ne 64
\end{aligned}$$

based on a sample of $n = 15$ observations, with test statistic
$t^* = 2.12$.

**a) Degrees of freedom:**

$$df = n - 1 = 15 - 1 = 14$$

**b) Critical values and p-value bounds:**

From the $t$-distribution table for $df = 14$:

- $t = 1.761$ corresponds to a two-tailed probability of 0.10

- $t = 2.145$ corresponds to a two-tailed probability of 0.05

Since $t^* = 2.12$ falls between these values, the two-sided p-value
satisfies: $$0.05 < \text{p-value} < 0.10$$

**c) Significance:**

- At the 10% level: **Yes**, since the p-value $< 0.10$

- At the 5% level: **No**, since the p-value $> 0.05$

**d) Exact two-sided p-value using R:**

::: tcolorbox
    # Compute exact two-sided p-value for t* = 2.12 with df = 14
    2 * (1 - pt(2.12, df = 14))

    ## [1] 0.05235683
:::

Thus, the exact two-sided p-value is approximately **0.0524**,
confirming the bracketing result.
::::















