# Statistical Power

## Introduction

::: definition
The probability that a fixed level $\alpha$ significance test will
reject $H_0$ when a particular alternative value of the parameter is
true is called the **power** of the test against that alternative.
:::

The statistical power of a test is its ability to detect an effect if it
exists in reality.

It is the probability of correctly rejecting $H_0$ when $H_0$ is false
in reality.

$$\text{Power} = P(\text{reject } H_0 \mid H_0 \text{ false})$$

$$0 < \text{power} < 1$$

Power close to 1 (high power):\
Test is good at detecting effects.\
Power close to 0 (low power):\
Test is not reliable (i.e., we expect the test will not reject $H_0$
when $H_0$ is false).

Power is affected by:

- **The effect**\
  ([larger differences between reality and the null are easier to
  detect]{style="color: blue"})

- **Sample size**\
  ([larger samples increase power]{style="color: blue"})

- **Significance level** ($\alpha$)\
  ([as $\alpha$ increases, easier to reject $H_0$]{style="color: blue"})

- **Variability in data**\
  ([lower variability, higher power]{style="color: blue"})




## Type I and Type II Errors 

It is possible to make an incorrect conclusion on a hypothesis test.\
**Type I:** Incorrectly reject $H_0$ when $H_0$ is true in reality.\
**Type II:** Incorrectly fail to reject $H_0$ when $H_0$ is false in
reality.\
**Reality vs Conclusion Table:**

::: center
+:-:+:----------------------------------------------------------------------------------------------------------------------:+
|   |                                               $H_0$ True                                  $H_0$ False                  |
|   |   -------------------------- -------------------------------------------- -------------------------------------------- |
|   |   **Reject $H_0$**             [Type I ($\alpha$)]{style="color: red"}     [No error ]{style="color: green!50!black"}  |
|   |   **Fail to reject $H_0$**    [No error ]{style="color: green!50!black"}    [Type II ($\beta$)]{style="color: red"}    |
+---+------------------------------------------------------------------------------------------------------------------------+
:::

***Note:*** Type I errors are generally considered worse.

Let $\beta$ be the probability of a Type II error. Then:

$$\text{Power} = 1 - \beta = 1 - P(\text{Type II})$$

::::: example
The cola maker determines that a sweetness loss is too large to accept
if the mean response for all tasters is $\mu = 1.1$. Will a 5%
significance test detect this?

**Hypotheses:** $$\begin{aligned}
H_0\!:&\ \mu = 0 \\
H_A\!:&\ \mu > 0
\end{aligned}$$

Assume: $$n = 10, \quad \sigma = 1, \quad \alpha = 0.05$$

**Step 1: Determine the rejection region.**

Since the test is one-sided with $\alpha = 0.05$, we find:
$$z_{\text{crit}} = 1.645 \quad \text{(from Z-table)}$$

We reject $H_0$ if: $$Z^* > 1.645$$

::: center

```{r fig.cap="Rejection region for $Z$ with $\\alpha = 0.05$", echo=FALSE, fig.align='center', out.width='60%', warning=FALSE}
library(ggplot2)

# Data
x <- seq(-3, 3, length = 1000)
y <- dnorm(x)
x_reject <- seq(1.645, 3, length = 500)
y_reject <- dnorm(x_reject)

# Plot
ggplot(data.frame(x, y), aes(x, y)) +
  geom_line(color = "#619CFF", linewidth = 1.2) +
  geom_area(data = data.frame(x = x_reject, y = y_reject), 
            aes(x = x, y = y), fill = "#00BA38", alpha = 0.4) +
  geom_vline(xintercept = 1.645, linetype = "dashed") +
  annotate("text", x = 1.645, y = 0.02, 
           label = expression(z[crit] == 1.645), hjust = -0.1, size = 3.5) +
  annotate("text", x = 0, y = 0.12, 
           label = expression("Do not reject " * H[0]), size = 4) +
  annotate("text", x = 2.3, y = 0.25, 
           label = expression("Reject " * H[0]), size = 4) +
  labs(x = "", y = "") +
  theme_minimal(base_size = 13)
```

:::

**Step 2: Find the equivalent critical value of** $\bar{x}$.

Since $\sigma$ is known,
$$Z = \frac{\bar{x} - \mu_0}{\sigma/\sqrt{n}} \quad \Rightarrow \quad
1.645 = \frac{\bar{x}_{\text{crit}} - 0}{1/\sqrt{10}} \Rightarrow \bar{x}_{\text{crit}} \approx 0.520$$

So we reject $H_0$ if $\bar{x} > 0.520$.

**Step 3: Calculate the power when $\mu = 1.1$ is true.**

$$P\left(Z > \frac{0.520 - 1.1}{1/\sqrt{10}}\right)
\approx P(Z > -1.83) = 1 - 0.0336 = 0.9664$$

**Interpretation:** There is a 96.6% chance the test correctly detects
$\mu = 1.1$.

::: center
```{r fig.cap="Power curve showing shaded rejection area under \\( H_A \\)", echo=FALSE, fig.align='center', out.width='60%'}
library(ggplot2)

# Parameters
mu_0 <- 0
mu_a <- 1.1
sigma <- 1 / sqrt(10)
x_vals <- seq(-1, 3, length.out = 1000)

# Distributions
df <- data.frame(
  x = x_vals,
  y0 = dnorm(x_vals, mean = mu_0, sd = sigma),
  yA = dnorm(x_vals, mean = mu_a, sd = sigma)
)

# Rejection region under H_A
x_reject <- seq(0.52, 3, length.out = 500)
y_reject <- dnorm(x_reject, mean = mu_a, sd = sigma)

# Critical value height (for vertical dashed line)
y_crit <- dnorm(0.52, mean = mu_a, sd = sigma)

ggplot(df, aes(x)) +
  geom_line(aes(y = y0), color = "#619CFF", linewidth = 1.2) +
  geom_line(aes(y = yA), color = "#F8766D", linewidth = 1.2) +
  geom_area(data = data.frame(x = x_reject, y = y_reject),
            aes(x = x, y = y), fill = "#00BA38", alpha = 0.4) +
  geom_vline(xintercept = 0.52, linetype = "dashed") +
  annotate("text", x = 0.52, y = -0.01, label = expression(bar(x)[crit] == 0.520), vjust = 1.2, size = 3.5) +
  labs(x = "", y = "") +
  theme_minimal(base_size = 17) +
  annotate("text", x = 0.6, y = 0.8 * max(df$y0), label = expression(H[0] * ":  " * mu == 0), color = "#619CFF", size = 4) +
  annotate("text", x = 1.6, y = 0.8 * max(df$yA), label = expression(H[A] * ":  " * mu == 1.1), color = "#F8766D", size = 4)
```

:::

This result can also be visualized using a power curve, which shows how
the probability of correctly rejecting $H_0$ increases with the true
mean $\mu$.

```{r fig.cap="Power curve for a one-sided test with points at \\( \\mu = 0.52 \\) and \\( \\mu = 1.1 \\)", echo=FALSE, fig.align='center'}
library(ggplot2)
library(dplyr)

# Step 1: Define parameters
n <- 10
alpha <- 0.05
mu0 <- 0
sigma <- 1
z_crit <- qnorm(1 - alpha)
xbar_crit <- mu0 + z_crit * sigma / sqrt(n)

# Step 2: Sequence of true means
mu_vals <- seq(0, 3, length.out = 100)

# Step 3: Calculate power
power <- 1 - pnorm((xbar_crit - mu_vals) / (sigma / sqrt(n)))
df <- data.frame(mu = mu_vals, power = power)

# Step 4: Highlight points at mu = 0.52 and 1.1
highlight_points <- data.frame(
  mu = c(0.52, 1.1),
  power = 1 - pnorm((xbar_crit - c(0.52, 1.1)) / (sigma / sqrt(n)))
)

# ggplot version
ggplot(df, aes(x = mu, y = power)) +
  geom_line(color = "#619CFF", linewidth = 1.2, linetype = "dashed") +
  geom_point(data = highlight_points, aes(x = mu, y = power),
             color = "#F8766D", size = 3) +
  labs(
    x = expression(mu),
    y = "power"
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  theme_minimal(base_size = 14)
```
:::::






If we reject $H_0$ when in fact $H_0$ is true, this is a **Type I
error.**\
If we fail to reject $H_0$ when in fact $H_a$ is true, this is a **Type
II error.**\
The **significance level** $\alpha$ of any fixed level test is the
probability of a Type I error.\
The **power** of a test against any alternative is 1 minus the
probability of a Type II error for that alternative.\
The probability of making a Type II error is denoted by $\beta$.

::: tcolorbox
**Type I Error**\
$H_0$ is true, but sampling variation in the data leads you to reject
$H_0$, you've made a Type I error.\
When $H_0$ is true, a Type I error occurs if $H_0$ is rejected.\
**Type II Error**\
$H_0$ is false, but sampling variation in the data does not lead you to
reject $H_0$, you've made a Type II error.\
When $H_0$ is false, a Type II error occurs if $H_0$ is **NOT**
rejected.
:::

:::: example
According to Access and Support to Education and Training Survey (2008),
of 4,756 adult Canadians, 1,581 indicated that they worked at a job or
business at anytime (between July 2007 and June 2008), regardless of the
number of hours per week.

Is there evidence to suggest that the true proportion $p$ is greater
than 0.50?

$$\begin{aligned}
H_0\!:&\ p = 0.50 \\
H_a\!:&\ p > 0.50
\end{aligned}$$

**R Output**

```r
    prop.test(x = 1581, n = 4756, p = 0.50,
              alternative = "greater", correct = FALSE)

    ## 
    ##  1-sample proportions test without continuity
    ##  correction
    ## 
    ## data:  1581 out of 4756, null probability 0.5
    ## X-squared = 534.24, df = 1, p-value = 1
    ## alternative hypothesis: true p is greater than 0.5
    ## 95 percent confidence interval:
    ##  0.3212845 1.0000000
    ## sample estimates:
    ##         p 
    ## 0.3324222 
```

**$P\text{-value} > \alpha = 0.05$**; we Fail to Reject $H_0$.

**This means we could be making a Type II error.**We indicated that
there is no evidence to conclude that the true proportion of adult
Canadians who worked at a job or business at anytime (between July 2007
and June 2008), regardless of the number of hours per week, was more
than 0.50 --- this conclusion implies that $H_0: p = 0.50$ is plausible,
but we could be wrong.
::::

::: example
Bottles of a popular cola are supposed to contain 300 milliliters (ml)
of cola. There is some variation from bottle to bottle because the
filling machinery is not perfectly precise. The distribution of contents
is Normal with standard deviation $\sigma = 3$ ml. Will inspecting 6
bottles discover underfilling?

The hypotheses are: $$\begin{aligned}
H_0\!:&\ \mu = 300 \\
H_a\!:&\ \mu < 300
\end{aligned}$$

A 5% significance test rejects $H_0$ if $z_* \leq -1.645$, where the
test statistic $z_*$ is: $$z_* = \frac{\bar{x} - 300}{3 / \sqrt{6}}$$

Power calculations help us see how large a shortfall in the bottle
contents the test can be expected to detect. Find the power of this test
against the alternative $\mu = 299$.

**Step 1. Write the rule for rejecting $H_0$ in terms of $\bar{x}$.**

We know that $\sigma = 3$, so the $z$ test rejects $H_0$ at the
$\alpha = 0.05$ level when:
$$z = \frac{\bar{x} - 300}{3 / \sqrt{6}} < -1.645$$

This is the same as:
$$\bar{x} < 300 - 1.645 \cdot \frac{3}{\sqrt{6}} \quad \Rightarrow \quad \bar{x} < 297.985$$

**Step 2. The power is the probability of this event under the condition
that the alternative $\mu = 299$ is true.**

To calculate this probability, standardize $\bar{x}$ using $\mu = 299$:
$$\begin{aligned}
\text{power} &= P(\bar{x} < 297.985 \mid \mu = 299) \\
&= P\left( Z < \frac{297.985 - 299}{3 / \sqrt{6}} \right) \\
&= P(Z < -0.83) = 0.2033
\end{aligned}$$
:::

## Using Power to Determine Sample Size

When designing a study, one of the most important decisions is how large
a sample to collect. If the sample size is too small, even meaningful
effects may go undetected due to low statistical power. On the other
hand, collecting an unnecessarily large sample can be inefficient and
costly. By using power calculations, researchers can determine the
minimum sample size needed to detect an effect of a given size with a
specified probability (power), while controlling for Type I error. This
section introduces how statistical power is used in planning and
justifying sample sizes before conducting a hypothesis test.

::: example
Suppose an experimenter wishes to test $$\begin{aligned}
H_0\!:&\ \mu = 100 \\
H_a\!:&\ \mu > 100
\end{aligned}$$ at the $\alpha = 0.05$ level of significance and wants
$1 - \beta$ to equal 0.60 when $\mu = 103$. What is the smallest (i.e.,
cheapest) sample size that will achieve that objective? Assume that the
variable being measured is Normally distributed with $\sigma = 14$.\
**Step 1. Write the rule for rejecting $H_0$ in terms of $\bar{x}_*$.**

By definition, $$\alpha = P(\text{we reject } H_0 \mid \mu = 100)
= P\left( \bar{X} > \bar{x}_* \mid \mu = 100 \right)
= P\left( Z > \frac{\bar{x}_* - 100}{14 / \sqrt{n}} \right) = 0.05$$

From the standard normal table, $P(Z > 1.645) = 0.05$, so:
$$\bar{x}_* = 100 + 1.645 \cdot \frac{14}{\sqrt{n}}$$

**Step 2. The power is the probability of this event under the condition
that the alternative $\mu = 103$ is true.**

To calculate this probability, standardize $\bar{x}$ using $\mu = 103$:

$$\begin{aligned}
\text{power} &= 1 - \beta = P(\bar{X} > \bar{x}_* \mid \mu = 103) \\
&= P\left( Z > \frac{\bar{x}_* - 103}{14 / \sqrt{n}} \right) = 0.60
\end{aligned}$$

From the standard normal table,
$P(Z > -0.25) \approx 0.5987 \approx 0.60$, so:
$$\frac{\bar{x}_* - 103}{14 / \sqrt{n}} = -0.25
\Rightarrow
\bar{x}_* = 103 - 0.25 \cdot \frac{14}{\sqrt{n}}$$

**Step 3. Solving for $n$**

From Steps 1 and 2:
$$100 + 1.645 \cdot \frac{14}{\sqrt{n}} = 103 - 0.25 \cdot \frac{14}{\sqrt{n}}$$

Solving for $n$:
$$\left( \frac{(1.645 + 0.25) \cdot 14}{103 - 100} \right)^2 \approx 78.2045$$

Therefore, a minimum of 79 observations must be taken to guarantee that
the hypothesis test will have the desired power of at least 0.60.
:::

::: example
A vending machine advertises that it dispenses 225 ml cups of coffee
($\sigma = 7$ ml). You believe the mean volume of coffee per cup is
something less than 225 ml. You plan to sample 40 cups of coffee from
this machine to test your hypothesis.

1.  If the true mean volume of coffee per cup is 223 ml, what is the
    power of your test at $\alpha = 0.05$? *(Homework)*

2.  How many coffee cups should you sample if you want to raise the
    power in part (a) to 0.80?

**Solution (b):**

**Step 1. Write the rule for rejecting $H_0$ in terms of $\bar{x}_*$.**

By definition: $$\begin{aligned}
\alpha &= P(\text{we reject } H_0 \mid \mu = 225) \\
&= P(\bar{X} < \bar{x}_* \mid \mu = 225) \\
&= P\left( Z < \frac{\bar{x}_* - 225}{7 / \sqrt{n}} \right) = 0.05
\end{aligned}$$

From the standard normal table, $P(Z < -1.645) = 0.05$, so:
$$\bar{x}_* = 225 - 1.645 \cdot \frac{7}{\sqrt{n}}$$

**Step 2. The power is the probability of this event under the condition
that the alternative $\mu = 223$ is true.**

To calculate this probability, standardize $\bar{x}$ using $\mu = 223$:

$$\begin{aligned}
\text{power} &= 1 - \beta = P(\bar{X} < \bar{x}_* \mid \mu = 223) \\
&= P\left( Z < \frac{\bar{x}_* - 223}{7 / \sqrt{n}} \right) = 0.80
\end{aligned}$$

From the table, $P(Z < 0.84) = 0.7995 \approx 0.80$, so:
$$\frac{\bar{x}_* - 223}{7 / \sqrt{n}} = 0.84
\quad \Rightarrow \quad
\bar{x}_* = 223 + 0.84 \cdot \frac{7}{\sqrt{n}}$$

**Step 3. Solving for $n$**

From Steps 1 and 2:
$$225 - 1.645 \cdot \frac{7}{\sqrt{n}} = 223 + 0.84 \cdot \frac{7}{\sqrt{n}}$$

Solving for $n$:
$$n = \left( \frac{(1.645 + 0.84) \cdot 7}{225 - 223} \right)^2 \approx 75.6465$$

Therefore, a minimum of 76 observations must be taken to guarantee that
the hypothesis test will have the desired precision.
:::

::: example
A newsletter reports that 90% of adults drink milk. The researchers are
interested in investigating if fewer than 90% of adults drink milk (at
$\alpha = 0.05$). They collect a random sample of 200 adults in a
certain region.

1.  Calculate power of the test if the percentage of adults who drink
    milk is really 85%.

    $$\begin{aligned}
    \text{Under } H_0 &: \hat{p}^* = 0.90 - 1.645 \sqrt{\frac{0.90(1 - 0.90)}{200}} = 0.8651 \\
    \text{Power} &= P(\text{Reject } H_0 \mid p = 0.85) \\
    &= P(\hat{p} < 0.8651 \mid p = 0.85) \\
    &= P\left(Z < \frac{0.8651 - 0.85}{\sqrt{\frac{0.85(1 - 0.85)}{200}}} \right) \\
    &= P(Z < 0.5980) \approx 0.7250
    \end{aligned}$$

    **R Output**

    ```r
        pnorm(0.5980, mean = 0, sd = 1)
        # [1] 0.72508
    ```

2.  Calculate beta if the percentage of adults who drink milk is really
    85%.

    $$\begin{aligned}
    \beta &= 1 - \text{Power} = 1 - 0.725 = 0.275
    \end{aligned}$$

3.  How many adults should you sample if you want to raise the power in
    part (a) to 0.80?

    **Step 1: Determine rejection cutoff under** $H_0 : p = 0.90$
    $$\begin{aligned}
    \hat{p}^* = 0.90 - 1.645 \sqrt{\frac{0.90(1 - 0.90)}{n}}
    \end{aligned}$$

    **Step 2: Set power to 0.80 under** $p = 0.85$ $$\begin{aligned}
    P\left( \frac{\hat{p} - 0.85}{\sqrt{\frac{0.85(1 - 0.85)}{n}}} < \frac{\hat{p}^* - 0.85}{\sqrt{\frac{0.85(1 - 0.85)}{n}}} \right) &= 0.80 \\
    \frac{\hat{p}^* - 0.85}{\sqrt{\frac{0.85(1 - 0.85)}{n}}} &= 0.8416 \\
    \hat{p}^* &= 0.85 + 0.8416 \sqrt{\frac{0.85(1 - 0.85)}{n}}
    \end{aligned}$$

    **R Output**

    ```r
        qnorm(0.80, mean = 0, sd = 1)
        # [1] 0.8416212
    ```

    **Step 3: Equating expressions for** $\hat{p}^*$ $$\begin{aligned}
    0.90 - 1.645 \sqrt{\frac{0.90(1 - 0.90)}{n}} &= 0.85 + 0.8416 \sqrt{\frac{0.85(1 - 0.85)}{n}} \\
    0.05 &= \left(1.645 \sqrt{0.90(0.10)} + 0.8416 \sqrt{0.85(0.15)}\right) \cdot \frac{1}{\sqrt{n}} \\
    \sqrt{n} &= \frac{0.4935 + 0.3005}{0.05} = 15.87 \Rightarrow n \approx 253
    \end{aligned}$$
:::


:::: example
A newsletter reports that 90% of adults drink milk. The researchers are
interested in investigating if less than 90% of adults drink milk (at
$\alpha = 0.05$). They collect a **random sample of 100 adults** in a
certain region.\
Calculate power of the test if the percentage of adults who drink milk
is really 85%.

We test: $$\begin{aligned}
H_0 &: p = 0.90 \\
H_a &: p < 0.90
\end{aligned}$$

The rejection region is determined by: $$\begin{aligned}
\alpha &= 0.05 = P(\text{reject } H_0 \mid H_0 \text{ is true}) \\
P(Z < -1.645) &= 0.05 \quad \text{(Z critical value is -1.645)}
\end{aligned}$$

Now calculate power under the alternative $p = 0.85$:

$$\begin{aligned}
\text{Power} &= P(\text{reject } H_0 \mid H_0 \text{ is false}) \\
&= P\left( \frac{\hat{p} - 0.90}{\sqrt{\frac{0.90(1 - 0.90)}{100}}} < -1.645 \,\middle|\, p = 0.85 \right) \\
&= P(\hat{p} < 0.85065 \mid p = 0.85) \\
&= P\left( Z < \frac{0.85065 - 0.85}{\sqrt{\frac{0.85(1 - 0.85)}{100}}} \right) \\
&= P(Z < 0.0182) \approx 0.5072
\end{aligned}$$

**R Output**

```r
    pnorm(0.0182, mean = 0, sd = 1)
    # [1] 0.5072603
```
::::

:::: example
A newsletter reports that 90% of adults drink milk. The researchers are
interested in investigating if less than 90% of adults drink milk (at
$\alpha = 0.05$). They collect a **random sample of 50 adults** in a
certain region.\
Calculate power of the test if the percentage of adults who drink milk
is really 85%.

$$\begin{aligned}
H_0 &: p = 0.90 \\
H_a &: p < 0.90
\end{aligned}$$

The rejection region is determined by: $$\begin{aligned}
\alpha &= 0.05 = P(\text{reject } H_0 \mid H_0 \text{ is true}) \\
P(Z < -1.645) &= 0.05 \quad \text{(Z critical value is -1.645)}
\end{aligned}$$

**Power:** $$\begin{aligned}
\text{Power} &= P(\text{reject } H_0 \mid H_0 \text{ is false}) \\
&= P\left( \frac{\hat{p} - 0.90}{\sqrt{\frac{0.90(1-0.90)}{50}}} < -1.645 \;\middle|\; p = 0.85 \right) \\
&= P(\hat{p} < 0.85065 \mid p = 0.85) \\
&= P\left( Z < \frac{0.85065 - 0.85}{\sqrt{\frac{0.85(1 - 0.85)}{50}}} \right) \\
&= P(Z < -0.3921) \\
&\approx 0.3475
\end{aligned}$$

**R Output**

```r
    pnorm(-0.3921, mean = 0, sd = 1)
    # [1] 0.3474922
```
::::

If we keep $\alpha$ at the same size, larger sample sizes increase the
power of the test because sampling variability (sampling distributions)
are much narrower.

::: tcolorbox
**If $H_a : \mu > \mu_0$, then**
$$\text{Power}(\mu_a) = \pi(\mu_a) = P \left[ Z \geq Z_{\alpha} + \frac{\sqrt{n}(\mu_0 - \mu_a)}{\sigma} \right]$$

**where** $\mu_a > \mu_0$.

**If $H_a : \mu < \mu_0$, then**
$$\text{Power}(\mu_a) = \pi(\mu_a) = P \left[ Z \leq -Z_{\alpha} + \frac{\sqrt{n}(\mu_0 - \mu_a)}{\sigma} \right]$$

**where** $\mu_a < \mu_0$.

**If $H_a : \mu \ne \mu_0$, then** $$\begin{aligned}
\text{Power}(\mu_a) &= \pi(\mu_a) \\
&= 1 - P \left[ -Z_{\alpha/2} + \frac{\sqrt{n}(\mu_0 - \mu_a)}{\sigma} \leq Z \leq Z_{\alpha/2} + \frac{\sqrt{n}(\mu_0 - \mu_a)}{\sigma} \right]
\end{aligned}$$

**where** $\mu_a \ne \mu_0$.

**You need to know:**

- Standard deviation, $\sigma$

- Significance level, $\alpha$

- Effect size to detect, $\mu_0 - \mu_a$
:::

Calculations of power (or of error probabilities) are useful for
planning studies because we can make these calculations before we have
any data. Once we actually have data, it is more common to report a
P-value rather than a reject-or-not decision at a fixed significance
level $\alpha$. The P-value measures the strength of the evidence
provided by the data against $H_0$. It leaves any action or decision
based on that evidence up to each individual. Different people may
require different strengths of evidence.

## Exercises {#sec:ch7exercises}

---

<div class="exercise-box">
<div class="exercise-label">Question 1</div>
A battery manufacturer's current cell design has a known mean lifetime of $\mu_0 = 500$ hours, with population standard deviation $\sigma = 20$ hours. Engineers develop a new electrode chemistry that they believe *increases* mean battery life, and plan to test
$$H_0: \mu = 500 \qquad \text{vs.} \qquad H_a: \mu > 500$$
using a random sample of $n = 36$ batteries at $\alpha = 0.05$.

Complete the R code below to (a) find the critical value $z_{\text{crit}}$ for this upper-tailed test, (b) convert it into a rejection cutoff for $\bar{x}$, (c) find the power of the test against the alternative $\mu_a = 510$ hours, and (d) find $\beta$ for that same alternative.

<div class="webr-exercise">
<div class="webr-exercise-header">Exercise</div>
<textarea class="webr-editor" rows="19">mu0   <- 500
sigma <- 20
n     <- 36
alpha <- 0.05
mua   <- 510

# (a) Critical z value for an upper-tailed test
z_crit <- qnorm(___)
z_crit

# (b) Convert z_crit to a cutoff for xbar
xbar_crit <- mu0 + z_crit * sigma / sqrt(n)
xbar_crit

# (c) Power against the alternative mu = 510
power <- 1 - pnorm((xbar_crit - mua) / (sigma / sqrt(n)))
power

# (d) Type II error probability
beta <- ___
beta</textarea>
<div class="webr-controls">
  <button class="webr-run-btn" disabled>▶ Run</button>
  <button class="webr-reset-btn">↺ Reset</button>
</div>
<div class="webr-output-area"></div>
<div class="webr-canvas-area"><canvas></canvas></div>
</div>
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- For an upper-tailed test, the rejection region is in the right tail, so $z_{\text{crit}} = z_{1-\alpha}$; in R this is `qnorm(1 - alpha)`.
- Once you have $z_{\text{crit}}$, convert to $\bar{x}_{\text{crit}} = \mu_0 + z_{\text{crit}}(\sigma/\sqrt{n})$.
- Power is computed by standardizing $\bar{x}_{\text{crit}}$ under the *alternative* mean $\mu_a$, not under $\mu_0$.
- Recall the relationship between power and $\beta$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">

```r
mu0   <- 500
sigma <- 20
n     <- 36
alpha <- 0.05
mua   <- 510

# (a) Critical z value for an upper-tailed test
z_crit <- qnorm(1 - alpha)
z_crit
# [1] 1.644854

# (b) Convert z_crit to a cutoff for xbar
xbar_crit <- mu0 + z_crit * sigma / sqrt(n)
xbar_crit
# [1] 505.4828

# (c) Power against the alternative mu = 510
power <- 1 - pnorm((xbar_crit - mua) / (sigma / sqrt(n)))
power
# [1] 0.9123145

# (d) Type II error probability
beta <- 1 - power
beta
# [1] 0.08768546
```

**(a)** $z_{\text{crit}} = z_{0.95} \approx 1.6449$.

**(b)** $\bar{x}_{\text{crit}} = 500 + 1.6449(20/\sqrt{36}) \approx 505.483$ hours. The test rejects $H_0$ whenever $\bar{x} > 505.483$.

**(c)** Standardizing $505.483$ using $\mu_a = 510$: $\text{Power} = P(Z > -1.3551) \approx 0.9123$.

**(d)** $\beta = 1 - 0.9123 = 0.0877$.

**Interpretation:** There is approximately a 91.2% probability that this test will correctly detect a true increase in mean battery life to 510 hours. Equivalently, there is about an 8.8% chance the study fails to detect this real improvement — a Type II error.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 2</div>
A pharmaceutical company markets a cholesterol drug and, based on past data, believes the mean LDL level of treated patients is $\mu_0 = 130$ mg/dL, with $\sigma = 15$ mg/dL known from a large historical database. A researcher tests a reformulated version, believing it *lowers* LDL further, using
$$H_0: \mu = 130 \qquad \text{vs.} \qquad H_a: \mu < 130$$
with $n = 25$ patients at $\alpha = 0.01$.

Complete the R code to (a) find $z_{\text{crit}}$ for this lower-tailed test, (b) find the rejection cutoff for $\bar{x}$, (c) find the power against $\mu_a = 122$ mg/dL, and (d) find $\beta$.

<div class="webr-exercise">
<div class="webr-exercise-header">Exercise</div>
<textarea class="webr-editor" rows="19">mu0   <- 130
sigma <- 15
n     <- 25
alpha <- 0.01
mua   <- 122

# (a) Critical z value for a lower-tailed test
z_crit <- qnorm(___)
z_crit

# (b) Convert z_crit to a cutoff for xbar
xbar_crit <- mu0 + z_crit * sigma / sqrt(n)
xbar_crit

# (c) Power against the alternative mu = 122
power <- pnorm((xbar_crit - mua) / (sigma / sqrt(n)))
power

# (d) Type II error probability
beta <- ___
beta</textarea>
<div class="webr-controls">
  <button class="webr-run-btn" disabled>▶ Run</button>
  <button class="webr-reset-btn">↺ Reset</button>
</div>
<div class="webr-output-area"></div>
<div class="webr-canvas-area"><canvas></canvas></div>
</div>
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- For a lower-tailed test, the rejection region is in the left tail, so $z_{\text{crit}} = z_{\alpha}$, a *negative* number: `qnorm(alpha)`.
- The cutoff formula is still $\bar{x}_{\text{crit}} = \mu_0 + z_{\text{crit}}(\sigma/\sqrt{n})$; because $z_{\text{crit}} < 0$, this cutoff falls below $\mu_0$.
- Power is $P(\bar{x} < \bar{x}_{\text{crit}})$ standardized using $\mu_a$, computed directly with `pnorm()` (no need for `1 -`).
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">

```r
mu0   <- 130
sigma <- 15
n     <- 25
alpha <- 0.01
mua   <- 122

# (a) Critical z value for a lower-tailed test
z_crit <- qnorm(alpha)
z_crit
# [1] -2.326348

# (b) Convert z_crit to a cutoff for xbar
xbar_crit <- mu0 + z_crit * sigma / sqrt(n)
xbar_crit
# [1] 123.021

# (c) Power against the alternative mu = 122
power <- pnorm((xbar_crit - mua) / (sigma / sqrt(n)))
power
# [1] 0.6331918

# (d) Type II error probability
beta <- 1 - power
beta
# [1] 0.3668082
```

**(a)** $z_{\text{crit}} = z_{0.01} \approx -2.3263$.

**(b)** $\bar{x}_{\text{crit}} = 130 + (-2.3263)(15/\sqrt{25}) \approx 123.021$ mg/dL. The test rejects $H_0$ whenever $\bar{x} < 123.021$.

**(c)** Standardizing under $\mu_a = 122$: $\text{Power} = P(Z < 0.3403) \approx 0.6332$.

**(d)** $\beta = 1 - 0.6332 = 0.3668$.

**Interpretation:** There is approximately a 63.3% probability that this test will correctly detect a true LDL reduction to 122 mg/dL. There is a 36.7% chance ($\beta$) that the study fails to detect this real reduction.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 3</div>
A city transportation department installs a new traffic-signal timing system that it believes *reduces* average commute time below the historical mean $\mu_0 = 30$ minutes ($\sigma = 8$ minutes, known from years of sensor data). The department tests
$$H_0: \mu = 30 \qquad \text{vs.} \qquad H_a: \mu < 30$$
at $\alpha = 0.05$, and wants to see how the power against the alternative $\mu_a = 27$ minutes changes as the sample size $n$ increases.

Complete the R code to compute the rejection cutoff and the power for each sample size in `n_vec`, then describe the pattern you observe.

<div class="webr-exercise">
<div class="webr-exercise-header">Exercise</div>
<textarea class="webr-editor" rows="14">mu0   <- 30
sigma <- 8
alpha <- 0.05
mua   <- 27
n_vec <- c(15, 30, 45, 60, 90)

z_crit <- qnorm(alpha)

# (a) Vector of rejection cutoffs, one for each sample size
xbar_crit <- mu0 + z_crit * sigma / sqrt(___)
xbar_crit

# (b) Vector of power values, one for each sample size
power_vec <- pnorm((xbar_crit - mua) / (sigma / sqrt(___)))
power_vec</textarea>
<div class="webr-controls">
  <button class="webr-run-btn" disabled>▶ Run</button>
  <button class="webr-reset-btn">↺ Reset</button>
</div>
<div class="webr-output-area"></div>
<div class="webr-canvas-area"><canvas></canvas></div>
</div>
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- R will happily compute `sigma / sqrt(n_vec)` element-by-element, returning a vector of standard errors, one for each entry of `n_vec`.
- Both blanks should be filled with the same vector — the sample sizes you're varying.
- As $n$ increases (with $\alpha$, $\sigma$, $\mu_0$, and $\mu_a$ all held fixed), the standard error shrinks, which pulls the cutoff closer to $\mu_0$ and pushes power in a predictable direction.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">

```r
mu0   <- 30
sigma <- 8
alpha <- 0.05
mua   <- 27
n_vec <- c(15, 30, 45, 60, 90)

z_crit <- qnorm(alpha)

# (a) Vector of rejection cutoffs
xbar_crit <- mu0 + z_crit * sigma / sqrt(n_vec)
xbar_crit
# [1] 26.60240 27.59754 28.03840 28.30120 28.61294

# (b) Vector of power values
power_vec <- pnorm((xbar_crit - mua) / (sigma / sqrt(n_vec)))
power_vec
# [1] 0.4236812 0.6587690 0.8080472 0.8961444 0.9721073
```

| $n$ | 15 | 30 | 45 | 60 | 90 |
|---|---|---|---|---|---|
| $\bar{x}_{\text{crit}}$ | 26.602 | 27.598 | 28.038 | 28.301 | 28.613 |
| Power | 0.4237 | 0.6588 | 0.8080 | 0.8961 | 0.9721 |

**Interpretation:** Holding $\alpha$, $\sigma$, $\mu_0$, and $\mu_a$ fixed, power increases steadily as $n$ increases — from about 42.4% at $n=15$ to about 97.2% at $n=90$. This happens because a larger sample shrinks the standard error $\sigma/\sqrt{n}$, which narrows the sampling distribution of $\bar{x}$ under both $H_0$ and $H_a$ and pulls the rejection cutoff closer to $\mu_0$, making it easier to correctly detect the true reduction to $\mu_a = 27$ minutes.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 4</div>
An agricultural researcher is testing a new fertilizer that is believed to *increase* mean corn yield above the regional average of $\mu_0 = 150$ bushels per acre ($\sigma = 25$ bushels/acre, known from historical regional data). The test is
$$H_0: \mu = 150 \qquad \text{vs.} \qquad H_a: \mu > 150$$
based on $n = 50$ test plots at $\alpha = 0.05$.

Complete the R code below to construct the power curve — the power of the test plotted against a range of possible true means $\mu_a$ — and to highlight the power at $\mu_a = 157$ and $\mu_a = 161$ bushels/acre.

<div class="webr-exercise">
<div class="webr-exercise-header">Exercise</div>
<textarea class="webr-editor" rows="20">mu0   <- 150
sigma <- 25
n     <- 50
alpha <- 0.05

# (a) Critical value for xbar under an upper-tailed test
z_crit <- qnorm(1 - alpha)
xbar_crit <- mu0 + z_crit * sigma / sqrt(n)
xbar_crit

# (b) Sequence of possible true means to evaluate power over
mu_vals <- seq(___, ___, length.out = 100)

# (c) Power at each value in mu_vals
power_curve <- 1 - pnorm((xbar_crit - mu_vals) / (sigma / sqrt(n)))

# (d) Plot the power curve
plot(mu_vals, power_curve, type = "l", lty = 2, col = "blue",
     xlab = expression(mu), ylab = "power")

# (e) Highlight the power at mu = 157 and mu = 161
points(157, 1 - pnorm((xbar_crit - 157) / (sigma / sqrt(n))), pch = 19, col = "red")
points(161, 1 - pnorm((xbar_crit - ___) / (sigma / sqrt(n))), pch = 19, col = "red")</textarea>
<div class="webr-controls">
  <button class="webr-run-btn" disabled>▶ Run</button>
  <button class="webr-reset-btn">↺ Reset</button>
</div>
<div class="webr-output-area"></div>
<div class="webr-canvas-area"><canvas></canvas></div>
</div>
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- `seq()` needs a starting value and an ending value that together span a sensible range of true means around $\mu_0 = 150$ (150 to 180 is reasonable here, since $\mu_a > \mu_0$ for an upper-tailed test).
- The final `points()` call should highlight the same true mean referenced earlier in that line, $\mu_a = 161$.
- Points further to the right of $\mu_0$ on the power curve should show higher power, since they represent alternatives further from $H_0$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">

```r
mu0   <- 150
sigma <- 25
n     <- 50
alpha <- 0.05

# (a) Critical value for xbar under an upper-tailed test
z_crit <- qnorm(1 - alpha)
xbar_crit <- mu0 + z_crit * sigma / sqrt(n)
xbar_crit
# [1] 155.8154

# (b) Sequence of possible true means to evaluate power over
mu_vals <- seq(150, 180, length.out = 100)

# (c) Power at each value in mu_vals
power_curve <- 1 - pnorm((xbar_crit - mu_vals) / (sigma / sqrt(n)))

# (d) Plot the power curve
plot(mu_vals, power_curve, type = "l", lty = 2, col = "blue",
     xlab = expression(mu), ylab = "power")

# (e) Highlight the power at mu = 157 and mu = 161
points(157, 1 - pnorm((xbar_crit - 157) / (sigma / sqrt(n))), pch = 19, col = "red")
points(161, 1 - pnorm((xbar_crit - 161) / (sigma / sqrt(n))), pch = 19, col = "red")
# Power at mu = 157: 0.6312
# Power at mu = 161: 0.9287
```

**Interpretation:** The rejection cutoff is $\bar{x}_{\text{crit}} \approx 155.815$ bushels/acre. At $\mu_a = 157$, only slightly above the cutoff, the power is about 0.6312 — the test correctly detects this modest true increase about 63% of the time. At $\mu_a = 161$, further from $\mu_0$, power rises to about 0.9287. The power curve is increasing in $\mu_a$ because larger true increases push more of the alternative sampling distribution of $\bar{x}$ past the fixed cutoff $\bar{x}_{\text{crit}}$, leaving less area in the "fail to reject" region.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 5</div>
A manufacturing plant's current process produces defective circuit boards at a rate of $p_0 = 0.08$. Engineers introduce a new process that they believe *reduces* the defect rate, and plan to test
$$H_0: p = 0.08 \qquad \text{vs.} \qquad H_a: p < 0.08$$
using $n = 250$ boards at $\alpha = 0.05$, with the Normal approximation to the sampling distribution of $\hat{p}$.

Complete the R code to (a) find $z_{\text{crit}}$, (b) find the rejection cutoff $\hat{p}^{\,*}$ under $H_0$, (c) find the power against the true alternative $p_a = 0.05$, and (d) find $\beta$.

<div class="webr-exercise">
<div class="webr-exercise-header">Exercise</div>
<textarea class="webr-editor" rows="19">p0    <- 0.08
n     <- 250
alpha <- 0.05
pa    <- 0.05

# (a) Critical z value for a lower-tailed test
z_crit <- qnorm(___)
z_crit

# (b) Rejection cutoff for phat under H0
phat_crit <- p0 + z_crit * sqrt(p0 * (1 - p0) / n)
phat_crit

# (c) Power against the true proportion pa = 0.05
power <- pnorm((phat_crit - pa) / sqrt(pa * (1 - pa) / n))
power

# (d) Type II error probability
beta <- ___
beta</textarea>
<div class="webr-controls">
  <button class="webr-run-btn" disabled>▶ Run</button>
  <button class="webr-reset-btn">↺ Reset</button>
</div>
<div class="webr-output-area"></div>
<div class="webr-canvas-area"><canvas></canvas></div>
</div>
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- The rejection cutoff $\hat{p}^{\,*}$ must be computed using the standard error under $H_0$, based on $p_0 = 0.08$ — not $p_a$.
- Power, on the other hand, is computed by standardizing $\hat{p}^{\,*}$ using the standard error based on the *true* alternative $p_a = 0.05$. Using the wrong standard error at the wrong stage is a common mistake.
- This is a lower-tailed test, so use `qnorm(alpha)` for the negative critical value.
- Recall $\beta = 1 - \text{Power}$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">

```r
p0    <- 0.08
n     <- 250
alpha <- 0.05
pa    <- 0.05

# (a) Critical z value for a lower-tailed test
z_crit <- qnorm(alpha)
z_crit
# [1] -1.644854

# (b) Rejection cutoff for phat under H0
phat_crit <- p0 + z_crit * sqrt(p0 * (1 - p0) / n)
phat_crit
# [1] 0.05177746

# (c) Power against the true proportion pa = 0.05
power <- pnorm((phat_crit - pa) / sqrt(pa * (1 - pa) / n))
power
# [1] 0.5513015

# (d) Type II error probability
beta <- 1 - power
beta
# [1] 0.4486985
```

**(a)** $z_{\text{crit}} = z_{0.05} \approx -1.6449$ (using the Normal approximation).

**(b)** $\hat{p}^{\,*} = 0.08 + (-1.6449)\sqrt{0.08(0.92)/250} \approx 0.0518$. The test rejects $H_0$ whenever $\hat{p} < 0.0518$.

**(c)** Standardizing under $p_a = 0.05$: $\text{Power} = P(Z < 0.1290) \approx 0.5513$.

**(d)** $\beta = 1 - 0.5513 = 0.4487$.

**Interpretation:** There is approximately a 55.1% probability that this test will correctly detect a true reduction in the defect rate to 5%. There is a 44.9% chance ($\beta$) that the study fails to detect this real improvement, even though the new process is genuinely better — this reflects the relatively small gap between $p_0 = 0.08$ and $p_a = 0.05$.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 6</div>
A quality-control engineer is explaining statistical power to a new colleague at a manufacturing plant.

(a) In your own words, define the statistical power of a test.
(b) Explain what it means for a test to have high power. What does it mean for a test to have low power?
(c) Give one concrete consequence, in a quality-control setting, of running a test that has low power.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Power is a conditional probability — it is defined given that a specific condition about $H_0$ holds. Start from that definition.
- "High" and "low" power should be described in terms of how reliably the test detects a real effect, not in terms of the p-value of any single sample.
- For part (c), think about what happens when a real problem exists but the test is unlikely to flag it.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** The power of a test is the probability that the test correctly rejects $H_0$ when a particular alternative value of the parameter is actually true:
$$\text{Power} = P(\text{reject } H_0 \mid H_0 \text{ is false})$$

**(b)** A test with power close to 1 (high power) is very likely to detect a real effect when one exists — it reliably rejects $H_0$ when the alternative is true. A test with power close to 0 (low power) is unreliable: even when the alternative is true, the test will often fail to reject $H_0$, so a real effect can easily go undetected.

**(c)** If a quality-control test for detecting a shift in a manufacturing process has low power, a genuine problem — such as a machine that has drifted out of calibration and started producing defective parts — could go undetected for a long time, because the test is unlikely to signal that anything is wrong even when something truly is.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 7</div>
An environmental agency tests soil samples from a former industrial site to determine whether the site is contaminated above a legal safety threshold. They test
$$H_0: \text{site is not contaminated} \qquad \text{vs.} \qquad H_a: \text{site is contaminated}$$

Fill in the four blank cells of the truth-versus-decision table below, and then answer the questions that follow.

| | $H_0$ True (site is safe) | $H_0$ False (site is contaminated) |
|---|---|---|
| **Reject $H_0$** (declare contaminated) | ______ (i) | ______ (ii) |
| **Fail to reject $H_0$** (declare safe) | ______ (iii) | ______ (iv) |

(a) Fill in each of (i)–(iv) with one of: Type I error, Type II error, correct decision.
(b) Which single cell corresponds to the probability $\alpha$?
(c) Which single cell corresponds to the probability $\beta$?
(d) In this particular context, briefly explain which type of error (I or II) has more serious real-world consequences, and why.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- A Type I error can only occur in the column where $H_0$ is actually true.
- A Type II error can only occur when the decision is "fail to reject $H_0$."
- $\alpha$ is defined as $P(\text{reject } H_0 \mid H_0 \text{ true})$; $\beta$ is defined analogously for the opposite decision and the opposite truth.
- For part (d), think about the cost of concluding the site is safe when it is not, versus concluding it is contaminated when it is not.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)**

| | $H_0$ True (site is safe) | $H_0$ False (site is contaminated) |
|---|---|---|
| **Reject $H_0$** (declare contaminated) | (i) Type I error | (ii) Correct decision |
| **Fail to reject $H_0$** (declare safe) | (iii) Correct decision | (iv) Type II error |

**(b)** Cell (i) corresponds to $\alpha = P(\text{reject } H_0 \mid H_0 \text{ true})$ — declaring a genuinely safe site contaminated.

**(c)** Cell (iv) corresponds to $\beta = P(\text{fail to reject } H_0 \mid H_0 \text{ false})$ — declaring a genuinely contaminated site safe.

**(d)** In this context, the Type II error (cell (iv)) is arguably more serious: declaring a truly contaminated site "safe" could allow continued exposure to a health hazard, with consequences that are difficult to reverse once people have been affected. The Type I error (cell (i)) is costly — unnecessary remediation spending on a site that was actually safe — but it is generally less harmful to public health than leaving real contamination undetected.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 8</div>
An automobile parts supplier tests every airbag deployment sensor before shipment. The test is set up as
$$H_0: \text{sensor functions correctly} \qquad \text{vs.} \qquad H_a: \text{sensor is defective}$$

(a) Describe, in the context of this scenario, what a Type I error would mean and what its practical consequence would be.
(b) Describe, in the context of this scenario, what a Type II error would mean and what its practical consequence would be.
(c) Briefly explain why a manufacturer in this industry might be willing to accept a higher Type I error rate in exchange for a lower Type II error rate.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- A Type I error occurs when $H_0$ is rejected but $H_0$ is actually true — think about what "rejecting $H_0$" means here.
- A Type II error occurs when the test fails to reject $H_0$ but $H_0$ is actually false.
- For part (c), compare the cost of discarding a good sensor to the cost of shipping a defective one.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** A Type I error occurs if the test concludes a sensor is defective (rejects $H_0$) when the sensor actually functions correctly. The practical consequence is a working sensor being scrapped or sent back for unnecessary rework — a cost in materials and production time, but not a safety risk.

**(b)** A Type II error occurs if the test fails to reject $H_0$ — concluding the sensor functions correctly — when the sensor is actually defective. The practical consequence is a faulty sensor being shipped and installed in a vehicle, where it might fail to deploy the airbag in a crash — a serious safety risk.

**(c)** Because a Type II error here carries a severe safety consequence while a Type I error carries only a financial and operational cost, the manufacturer would reasonably prefer a testing procedure with higher power (lower $\beta$) against defective sensors, even if that means tolerating a somewhat higher false-positive (Type I error) rate that discards a few functioning sensors.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 9</div>
A company redesigns its website checkout page and tests whether the new design *increases* the proportion of visitors who complete a purchase, using
$$H_0: p = p_0 \qquad \text{vs.} \qquad H_a: p > p_0$$
The analysis rejects $H_0$ and the company concludes that the new design increased the conversion rate.

(a) What type of error, if any, could the company be making by reaching this conclusion?
(b) Explain, in the context of this scenario, what it would mean for that error to have actually occurred.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Only one type of error is possible whenever $H_0$ has been rejected — think about which one occurs when $H_0$ is true but is rejected anyway.
- The other type of error (associated with failing to reject $H_0$) is not possible here, because that is not the decision that was made.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** Because the company rejected $H_0$, the only possible error is a **Type I error**. A Type II error cannot occur here, since that error is only possible when the decision is to fail to reject $H_0$.

**(b)** A Type I error would mean that the true conversion rate did not actually increase (the redesign had no real effect, so $H_0: p = p_0$ was actually true), but sampling variability in the observed data happened to produce a sample proportion large enough to lead the company to reject $H_0$ anyway. In that case, the company would be crediting the redesign with an improvement that does not really exist.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 10</div>
A school district pilots a new teaching method that it hopes will *increase* mean standardized test scores above the historical average, using
$$H_0: \mu = \mu_0 \qquad \text{vs.} \qquad H_a: \mu > \mu_0$$
The analysis fails to reject $H_0$, and the district concludes there is not enough evidence that the new method improved scores.

(a) What type of error, if any, could the district be making by reaching this conclusion?
(b) Explain, in the context of this scenario, what it would mean for that error to have actually occurred.
(c) Explain why the district's conclusion should not be phrased as "the new method has no effect."
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Only one type of error is possible whenever a test fails to reject $H_0$ — the same principle as the previous question, applied to the opposite decision.
- Part (c) is about the difference between "failing to reject $H_0$" and "accepting $H_0$."
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** Because the district failed to reject $H_0$, the only possible error is a **Type II error**. A Type I error cannot occur here, since that error requires $H_0$ to have been rejected.

**(b)** A Type II error would mean that the new teaching method truly does raise mean test scores (so $H_a: \mu > \mu_0$ is actually true), but sampling variability in this particular study's data was not strong enough to produce a test statistic in the rejection region, so the real improvement went undetected.

**(c)** Failing to reject $H_0$ only means the data did not provide sufficiently strong evidence against $H_0$ at the chosen significance level — it does not prove that $H_0$ is true. The method may have a real, positive effect that this particular study simply lacked the power to detect (for example, due to a small sample size or high variability in scores). We can say the study failed to reject $H_0$; we should never say $H_0$ was "accepted."
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 11</div>
A clinical research team is evaluating a new therapy in two separate sub-studies.

(a) In sub-study 1, against a particular alternative value of the parameter, the probability of a Type II error is $\beta = 0.35$. What is the power of the test against that alternative? Interpret the result in context.
(b) In sub-study 2, against a different alternative value, the power of the test is $0.72$. What is $\beta$ for that alternative? Interpret the result in context.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Use the relationship $\text{Power} = 1 - \beta$ in both directions: it can be solved for power given $\beta$, or for $\beta$ given power.
- Each part refers to power/$\beta$ against a *specific* alternative — the numbers in (a) and (b) are not both describing the same alternative.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** $\text{Power} = 1 - \beta = 1 - 0.35 = 0.65$. There is a 65% probability that sub-study 1's test correctly rejects $H_0$ against this particular alternative — that is, correctly detects that the new therapy has a real effect of that magnitude.

**(b)** $\beta = 1 - \text{Power} = 1 - 0.72 = 0.28$. There is a 28% probability that sub-study 2's test fails to reject $H_0$ against its alternative, even though the therapy truly has an effect of that magnitude — a Type II error.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 12</div>
Two independent clinical trials test whether a new medication *lowers* mean systolic blood pressure below the baseline level $\mu_0 = 140$ mmHg, both using $\sigma = 12$ mmHg (known), $\alpha = 0.05$, and the same alternative $\mu_a = 134$ mmHg. Trial A enrolls $n = 20$ patients; Trial B enrolls $n = 50$ patients.

(a) Without doing any calculation, state which trial you expect to have greater power, and explain why.
(b) Compute the power of each trial against $\mu_a = 134$ and confirm your answer to part (a).
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Both trials use a lower-tailed test: $H_a: \mu < \mu_0$.
- Increasing $n$ shrinks the standard error $\sigma/\sqrt{n}$, which narrows the sampling distributions of $\bar{x}$ under both $H_0$ and $H_a$.
- Compute the rejection cutoff $\bar{x}_{\text{crit}} = \mu_0 + z_{\alpha}(\sigma/\sqrt{n})$ separately for each $n$, then find the power under $\mu_a = 134$ for each.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** Trial B ($n = 50$) should have greater power than Trial A ($n = 20$), because a larger sample size shrinks the standard error of $\bar{x}$, narrowing the sampling distributions under $H_0$ and $H_a$ and making it easier to distinguish them — with everything else (effect size, $\sigma$, $\alpha$) held fixed.

**(b)** Using $z_{\text{crit}} = z_{0.05} = -1.6449$:

Trial A ($n=20$): $\bar{x}_{\text{crit}} = 140 - 1.6449(12/\sqrt{20}) \approx 135.586$ mmHg.
$$\text{Power}_A = P\left(Z < \frac{135.586 - 134}{12/\sqrt{20}}\right) \approx P(Z < 0.591) \approx 0.7228$$

Trial B ($n=50$): $\bar{x}_{\text{crit}} = 140 - 1.6449(12/\sqrt{50}) \approx 137.209$ mmHg.
$$\text{Power}_B = P\left(Z < \frac{137.209 - 134}{12/\sqrt{50}}\right) \approx P(Z < 1.892) \approx 0.9707$$

This confirms part (a): Trial B's power (about 97.1%) is substantially greater than Trial A's power (about 72.3%), driven entirely by the larger sample size.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 13</div>
Two agricultural research stations test whether a new irrigation technique *increases* mean crop yield above $\mu_0 = 100$ bushels/acre, both using $n = 36$ plots, $\alpha = 0.05$, and the same alternative $\mu_a = 108$ bushels/acre. Station 1 operates in a region with more consistent soil and weather conditions, giving $\sigma = 10$; Station 2 operates in a more variable region, giving $\sigma = 25$.

(a) Without doing any calculation, state which station you expect to have greater power, and explain why.
(b) Compute the power for each station against $\mu_a = 108$ and confirm your answer to part (a).
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Both stations use an upper-tailed test: $H_a: \mu > \mu_0$.
- A larger $\sigma$ increases the standard error $\sigma/\sqrt{n}$ for a fixed $n$, which widens the sampling distributions of $\bar{x}$ under both $H_0$ and $H_a$.
- Compute $\bar{x}_{\text{crit}} = \mu_0 + z_{1-\alpha}(\sigma/\sqrt{n})$ separately for each $\sigma$, then find the power under $\mu_a = 108$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** Station 1 ($\sigma = 10$) should have greater power than Station 2 ($\sigma = 25$), because lower variability in the data narrows the sampling distribution of $\bar{x}$ under both $H_0$ and $H_a$ (for the same $n$), making the two distributions easier to distinguish and the true effect easier to detect.

**(b)** Using $z_{\text{crit}} = z_{0.95} = 1.6449$:

Station 1 ($\sigma=10$): $\bar{x}_{\text{crit}} = 100 + 1.6449(10/\sqrt{36}) \approx 102.741$ bushels/acre.
$$\text{Power}_1 = P\left(Z > \frac{102.741 - 108}{10/\sqrt{36}}\right) \approx P(Z > -3.156) \approx 0.9992$$

Station 2 ($\sigma=25$): $\bar{x}_{\text{crit}} = 100 + 1.6449(25/\sqrt{36}) \approx 106.854$ bushels/acre.
$$\text{Power}_2 = P\left(Z > \frac{106.854 - 108}{25/\sqrt{36}}\right) \approx P(Z > -0.275) \approx 0.6084$$

This confirms part (a): Station 1's power (about 99.9%) is much greater than Station 2's power (about 60.8%), driven entirely by the difference in population variability.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 14</div>
A retail chain trains its staff using a new customer-service program and tests whether it *increases* mean customer satisfaction scores (on a 100-point scale) above $\mu_0 = 72$, using $\sigma = 9$, $n = 40$, and $\alpha = 0.05$ in both cases. Study A evaluates the program against a modest alternative of $\mu_a = 74$; Study B evaluates it against a larger alternative of $\mu_a = 78$.

(a) Without doing any calculation, state which study you expect to have greater power, and explain why.
(b) Compute the power for each alternative and confirm your answer to part (a).
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Both studies use the same rejection cutoff $\bar{x}_{\text{crit}}$, since $\mu_0$, $\sigma$, $n$, and $\alpha$ are identical — only the *true* alternative mean differs between (a) and (b).
- A larger gap between $\mu_0$ and $\mu_a$ (a larger effect size) shifts the alternative sampling distribution further away from the cutoff.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** Study B ($\mu_a = 78$) should have greater power than Study A ($\mu_a = 74$), because the true effect size ($\mu_a - \mu_0$) is larger in Study B, shifting more of the alternative sampling distribution of $\bar{x}$ past the (shared) rejection cutoff.

**(b)** The shared rejection cutoff is $\bar{x}_{\text{crit}} = 72 + 1.6449(9/\sqrt{40}) \approx 74.341$.

Study A ($\mu_a=74$):
$$\text{Power}_A = P\left(Z > \frac{74.341 - 74}{9/\sqrt{40}}\right) \approx P(Z > 0.240) \approx 0.4054$$

Study B ($\mu_a=78$):
$$\text{Power}_B = P\left(Z > \frac{74.341 - 78}{9/\sqrt{40}}\right) \approx P(Z > -2.572) \approx 0.9949$$

This confirms part (a): Study B's power (about 99.5%) is much greater than Study A's power (about 40.5%), because detecting a larger true improvement (6 points vs. 2 points) is easier with the same sample size and variability.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 15</div>
A food safety laboratory tests whether the mean concentration of a contaminant in a food product has *decreased* below the regulatory threshold $\mu_0 = 50$ ppm, using $\sigma = 8$ ppm, $n = 30$, and a true alternative of $\mu_a = 45$ ppm in both cases. Protocol A uses $\alpha = 0.01$; Protocol B uses $\alpha = 0.10$.

(a) Without doing any calculation, state which protocol you expect to have greater power, and explain the trade-off involved.
(b) Compute the power under each protocol and confirm your answer to part (a).
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Both protocols use a lower-tailed test: $H_a: \mu < \mu_0$.
- A larger $\alpha$ moves the rejection cutoff $\bar{x}_{\text{crit}} = \mu_0 + z_\alpha(\sigma/\sqrt{n})$ closer to $\mu_0$, enlarging the rejection region.
- Consider what a larger $\alpha$ does to the probability of a Type I error at the same time it changes power.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** Protocol B ($\alpha = 0.10$) should have greater power than Protocol A ($\alpha = 0.01$), because a larger $\alpha$ makes the rejection region wider (the cutoff moves closer to $\mu_0$), so it is easier to reject $H_0$ under any given alternative. The trade-off is that Protocol B also has a higher probability of a Type I error — falsely concluding the contaminant level has decreased when it has not — since $\alpha$ *is* that Type I error probability.

**(b)**

Protocol A ($\alpha=0.01$, $z_{\text{crit}} = -2.3263$): $\bar{x}_{\text{crit}} = 50 - 2.3263(8/\sqrt{30}) \approx 46.602$ ppm.
$$\text{Power}_A = P\left(Z < \frac{46.602 - 45}{8/\sqrt{30}}\right) \approx P(Z < 1.097) \approx 0.8637$$

Protocol B ($\alpha=0.10$, $z_{\text{crit}} = -1.2816$): $\bar{x}_{\text{crit}} = 50 - 1.2816(8/\sqrt{30}) \approx 48.128$ ppm.
$$\text{Power}_B = P\left(Z < \frac{48.128 - 45}{8/\sqrt{30}}\right) \approx P(Z < 2.142) \approx 0.9839$$

This confirms part (a): Protocol B's power (about 98.4%) exceeds Protocol A's power (about 86.4%), but at the cost of a ten-times-larger Type I error rate (0.10 vs. 0.01).
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 16</div>
A factory manager introduces a new assembly-line procedure and wants to know whether it *increases* the mean number of units produced per shift above the current level of $\mu_0 = 120$ units. The population standard deviation is known to be $\sigma = 18$ units. A random sample of $n = 49$ shifts under the new procedure is planned, and the test will be conducted at $\alpha = 0.05$.

(a) Find the critical value $z_{\text{crit}}$ and express the rejection rule in terms of $z$.
(b) Convert the rejection rule into a rule stated in terms of $\bar{x}$.
(c) Calculate the power of this test against the alternative $\mu_a = 126$ units.
(d) Interpret the power value you found in part (c) in the context of this scenario.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- This is an upper-tailed test, so the rejection rule is $z^* > z_{\text{crit}}$ with $z_{\text{crit}} = z_{1-\alpha}$.
- Convert using $\bar{x}_{\text{crit}} = \mu_0 + z_{\text{crit}}(\sigma/\sqrt{n})$.
- For part (c), standardize $\bar{x}_{\text{crit}}$ using $\mu_a = 126$, not $\mu_0 = 120$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** The hypotheses are $H_0: \mu = 120$ vs. $H_a: \mu > 120$. With $\alpha = 0.05$, $z_{\text{crit}} = z_{0.95} \approx 1.6449$. The rejection rule is: reject $H_0$ if $z^* > 1.6449$.

**(b)** Since $\sigma = 18$ is known,
$$z^* = \frac{\bar{x} - 120}{18/\sqrt{49}} > 1.6449 \quad \Longleftrightarrow \quad \bar{x} > 120 + 1.6449\left(\frac{18}{\sqrt{49}}\right) \approx 124.230$$
So the test rejects $H_0$ whenever $\bar{x} > 124.230$ units.

**(c)** Standardizing the cutoff $124.230$ using the alternative $\mu_a = 126$:
$$\text{Power} = P\left(Z > \frac{124.230 - 126}{18/\sqrt{49}}\right) = P(Z > -0.6885) \approx 0.7544$$

**(d)** There is approximately a 75.4% probability that this test will correctly detect a true increase in mean units produced per shift to 126, if the new assembly-line procedure really does raise output by that amount. Equivalently, there is roughly a 24.6% chance ($\beta$) that the study fails to detect this real improvement.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 17</div>
A call center's average customer hold time has historically been $\mu_0 = 8$ minutes, with $\sigma = 2.4$ minutes known from years of call-log data. After introducing a new call-routing algorithm believed to *reduce* hold times, management plans to sample $n = 64$ calls and test at $\alpha = 0.05$.

(a) Find the critical value $z_{\text{crit}}$ and express the rejection rule in terms of $z$.
(b) Convert the rejection rule into a rule stated in terms of $\bar{x}$.
(c) Calculate the power of this test against the alternative $\mu_a = 7.2$ minutes.
(d) Interpret the power value you found in part (c) in the context of this scenario.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- "Reduce hold times" calls for a lower-tailed alternative: $H_a: \mu < \mu_0$, so $z_{\text{crit}} = z_{\alpha}$, a negative value.
- The cutoff formula is $\bar{x}_{\text{crit}} = \mu_0 + z_{\text{crit}}(\sigma/\sqrt{n})$; because $z_{\text{crit}} < 0$, expect a cutoff below $\mu_0$.
- For part (c), standardize $\bar{x}_{\text{crit}}$ under $\mu_a = 7.2$, and compute the probability with the correct (lower-tailed) direction.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** The hypotheses are $H_0: \mu = 8$ vs. $H_a: \mu < 8$. With $\alpha = 0.05$, $z_{\text{crit}} = z_{0.05} \approx -1.6449$. The rejection rule is: reject $H_0$ if $z^* < -1.6449$.

**(b)** Since $\sigma = 2.4$ is known,
$$z^* = \frac{\bar{x} - 8}{2.4/\sqrt{64}} < -1.6449 \quad \Longleftrightarrow \quad \bar{x} < 8 - 1.6449\left(\frac{2.4}{\sqrt{64}}\right) \approx 7.5065$$
So the test rejects $H_0$ whenever $\bar{x} < 7.5065$ minutes.

**(c)** Standardizing the cutoff $7.5065$ using the alternative $\mu_a = 7.2$:
$$\text{Power} = P\left(Z < \frac{7.5065 - 7.2}{2.4/\sqrt{64}}\right) = P(Z < 1.0218) \approx 0.8466$$

**(d)** There is approximately an 84.7% probability that this test will correctly detect a true reduction in mean hold time to 7.2 minutes, if the new routing algorithm really does reduce hold times by that amount. There is roughly a 15.3% chance ($\beta$) that the study fails to detect this real improvement.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 18</div>
A cereal manufacturer's boxes are filled to a target mean weight of $\mu_0 = 350$ g, with $\sigma = 5$ g known from the filling equipment's specifications. Quality control wants to detect *either* overfilling or underfilling, so they test
$$H_0: \mu = 350 \qquad \text{vs.} \qquad H_a: \mu \neq 350$$
using $n = 40$ boxes at $\alpha = 0.05$.

(a) Find the two critical z-values and the corresponding lower and upper rejection cutoffs for $\bar{x}$.
(b) Calculate the power of this test against the alternative $\mu_a = 352$ g, making sure to account for both tails.
(c) Interpret the power value you found in part (b) in context.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- For a two-sided test, split $\alpha$ evenly between the two tails: use $z_{\alpha/2}$ for both cutoffs.
- The two cutoffs are $\bar{x}_{\text{crit,low}} = \mu_0 - z_{\alpha/2}(\sigma/\sqrt{n})$ and $\bar{x}_{\text{crit,high}} = \mu_0 + z_{\alpha/2}(\sigma/\sqrt{n})$.
- Power is the probability that $\bar{x}$ falls below the *lower* cutoff **or** above the *upper* cutoff, standardized under $\mu_a = 352$ — do not simply double a one-sided power value without checking that both tails' contributions are actually meaningful.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** With $\alpha = 0.05$ split into two tails, $z_{\alpha/2} = z_{0.025} \approx 1.9600$. The standard error is $\sigma/\sqrt{n} = 5/\sqrt{40} \approx 0.7906$.
$$\bar{x}_{\text{crit,low}} = 350 - 1.9600(0.7906) \approx 348.451 \qquad \bar{x}_{\text{crit,high}} = 350 + 1.9600(0.7906) \approx 351.549$$
The test rejects $H_0$ whenever $\bar{x} < 348.451$ or $\bar{x} > 351.549$.

**(b)** Standardizing both cutoffs under $\mu_a = 352$:
$$z_{\text{low}} = \frac{348.451 - 352}{0.7906} \approx -4.490 \qquad z_{\text{high}} = \frac{351.549 - 352}{0.7906} \approx -0.570$$
$$\text{Power} = P(Z < -4.490) + P(Z > -0.570) \approx 0.0000 + 0.7156 = 0.7156$$
Here, the lower-tail contribution is essentially zero (since $\mu_a = 352$ is above $\mu_0$, it is extremely unlikely for $\bar{x}$ to fall below the *lower* cutoff), so almost all the power comes from the upper tail. This shows why doubling a one-sided power value is not generally valid — the two tails do not contribute equally when $\mu_a$ is much closer to one cutoff than the other.

**(c)** There is approximately a 71.6% probability that this two-sided test will correctly detect that the mean fill weight has shifted to 352 g. There is roughly a 28.4% chance ($\beta$) that this real shift goes undetected.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 19</div>
A university testing office administers a standardized diagnostic exam with a historical mean score of $\mu_0 = 78$ points and known standard deviation $\sigma = 12$ points. A tutoring center wants to know whether its students score higher on average, and uses a rejection rule (already derived from a chosen significance level) that rejects $H_0$ whenever $\bar{x} > 82$, based on a sample of $n = 36$ tutored students.

Suppose the tutoring center's students truly have a mean score of $\mu_a = 84$ points.

(a) Calculate $\beta$, the probability that this test fails to reject $H_0$ even though $\mu_a = 84$ is the true mean.
(b) Calculate the power of the test against this same alternative.
(c) Interpret both values in context.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- You are given the cutoff directly ($\bar{x}_{\text{crit}} = 82$), so there is no need to find $z_{\text{crit}}$ from $\alpha$ first.
- $\beta$ is the probability of the *complementary* event to rejection, standardized under the true alternative mean $\mu_a = 84$: $\beta = P(\bar{x} < 82 \mid \mu = 84)$.
- Once you have $\beta$, use the relationship between power and $\beta$ for part (b) — no separate standardization is needed.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** Standardizing the cutoff $\bar{x}_{\text{crit}} = 82$ under $\mu_a = 84$:
$$\beta = P(\bar{x} < 82 \mid \mu = 84) = P\left(Z < \frac{82 - 84}{12/\sqrt{36}}\right) = P(Z < -1.00) \approx 0.1587$$

**(b)** $\text{Power} = 1 - \beta = 1 - 0.1587 = 0.8413$.

**(c)** There is approximately an 84.1% probability that this test correctly detects the tutoring center's true mean score of 84. A Type II error would occur if the study failed to detect this real 6-point improvement even though the tutored students' true mean score had increased — this happens with probability $\beta \approx 0.1587$.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 20</div>
A regional environmental agency is testing whether a new emissions regulation has *decreased* the mean concentration of a pollutant below the pre-regulation level of $\mu_0 = 42 \ \mu\text{g/m}^3$ ($\sigma = 9$, $n = 40$, $\alpha = 0.05$, lower-tailed test). The rejection cutoff works out to $\bar{x}_{\text{crit}} \approx 39.659$. The agency computed the power of this test against three different possible true post-regulation means:

| True mean $\mu_a$ ($\mu\text{g/m}^3$) | 41 | 39 | 37 |
|---|---|---|---|
| Power | 0.1731 | 0.6784 | 0.9692 |

(a) Describe the pattern shown in this table as $\mu_a$ moves further from $\mu_0 = 42$.
(b) Explain, in terms of the sampling distribution of $\bar{x}$, why power behaves this way.
(c) If the true post-regulation mean turned out to be $\mu_a = 41$, would you be confident that a study using this design would detect the improvement? Explain.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Compare how far each value of $\mu_a$ is from $\mu_0 = 42$, and see how that distance lines up with the power values given.
- Think about where the fixed cutoff $\bar{x}_{\text{crit}} = 39.659$ sits relative to each alternative sampling distribution (each one centered at a different $\mu_a$).
- For part (c), a power value well below 0.5 says something specific about how often the test would succeed if repeated.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** As $\mu_a$ moves further below $\mu_0 = 42$ (from 41, to 39, to 37), power increases — from about 17.3%, to about 67.8%, to about 96.9%.

**(b)** The rejection cutoff $\bar{x}_{\text{crit}} = 39.659$ stays fixed regardless of the true mean. As $\mu_a$ moves further below this fixed cutoff, more of the sampling distribution of $\bar{x}$ under that alternative (centered at $\mu_a$, with standard error $\sigma/\sqrt{n}$) falls below $\bar{x}_{\text{crit}}$, which is exactly the rejection region for this lower-tailed test. A small true decrease (like $\mu_a = 41$) leaves the alternative distribution mostly straddling the cutoff, so only a small fraction falls in the rejection region; a large true decrease (like $\mu_a = 37$) shifts nearly the entire distribution below the cutoff.

**(c)** No — with power only about 0.1731 against $\mu_a = 41$, this design would correctly detect a true decrease to 41 only about 17% of the time; roughly 83% of the time it would fail to detect this real (but small) improvement. A much larger sample size would be needed to reliably detect an effect this modest.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 21</div>
A demographer believes the mean age of residents in a rapidly growing suburb has *increased* above the historical value of $\mu_0 = 37$ years, with $\sigma = 6$ years known from census records. She wants to test $H_0: \mu = 37$ vs. $H_a: \mu > 37$ at $\alpha = 0.05$, and wants the power against the alternative $\mu_a = 39$ years to be at least $0.85$.

(a) Write the rejection rule for $\bar{x}_*$ in terms of the (unknown) sample size $n$.
(b) Write the expression for power in terms of $n$, and set it equal to $0.85$.
(c) Solve for the minimum required sample size.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Set up two expressions for $\bar{x}_*$: one from $\alpha = P(\bar{X} > \bar{x}_* \mid \mu = 37)$, and one from $0.85 = P(\bar{X} > \bar{x}_* \mid \mu = 39)$.
- Both expressions equal $\bar{x}_*$, so set them equal to each other and solve for $n$.
- Remember that $n$ must be a whole number, and that rounding down would leave the achieved power slightly below the target.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** By definition,
$$\alpha = P(\bar{X} > \bar{x}_* \mid \mu = 37) = P\left(Z > \frac{\bar{x}_* - 37}{6/\sqrt{n}}\right) = 0.05$$
Since $P(Z > 1.6449) = 0.05$,
$$\bar{x}_* = 37 + 1.6449\left(\frac{6}{\sqrt{n}}\right)$$

**(b)** The power condition is
$$\text{Power} = P(\bar{X} > \bar{x}_* \mid \mu = 39) = P\left(Z > \frac{\bar{x}_* - 39}{6/\sqrt{n}}\right) = 0.85$$
Since $P(Z > -1.0364) \approx 0.85$,
$$\frac{\bar{x}_* - 39}{6/\sqrt{n}} = -1.0364 \quad \Longrightarrow \quad \bar{x}_* = 39 - 1.0364\left(\frac{6}{\sqrt{n}}\right)$$

**(c)** Setting the two expressions for $\bar{x}_*$ equal:
$$37 + 1.6449\left(\frac{6}{\sqrt{n}}\right) = 39 - 1.0364\left(\frac{6}{\sqrt{n}}\right)$$
$$n = \left[\frac{(1.6449 + 1.0364)(6)}{39 - 37}\right]^2 \approx 64.70$$
The calculation gives $n = 64.70$, so the minimum required sample size is $\boxed{65}$ residents, because sample size must be a whole number and rounding down to 64 would leave the achieved power slightly below the target of 0.85.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 22</div>
A pharmaceutical company believes a new medication *decreases* mean recovery time below the current standard of $\mu_0 = 14$ days, with $\sigma = 3$ days known from prior trials. It plans to test $H_0: \mu = 14$ vs. $H_a: \mu < 14$ at $\alpha = 0.05$, and wants the power against the alternative $\mu_a = 12.5$ days to be at least $0.90$.

(a) Write the rejection rule for $\bar{x}_*$ in terms of the (unknown) sample size $n$.
(b) Write the expression for power in terms of $n$, and set it equal to $0.90$.
(c) Solve for the minimum required sample size.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- This is a lower-tailed test, so start from $\alpha = P(\bar{X} < \bar{x}_* \mid \mu = 14)$.
- Set up a second expression for $\bar{x}_*$ from the power condition $0.90 = P(\bar{X} < \bar{x}_* \mid \mu = 12.5)$, then equate the two expressions for $\bar{x}_*$.
- Round the final answer for $n$ up, not down.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** By definition,
$$\alpha = P(\bar{X} < \bar{x}_* \mid \mu = 14) = P\left(Z < \frac{\bar{x}_* - 14}{3/\sqrt{n}}\right) = 0.05$$
Since $P(Z < -1.6449) = 0.05$,
$$\bar{x}_* = 14 - 1.6449\left(\frac{3}{\sqrt{n}}\right)$$

**(b)** The power condition is
$$\text{Power} = P(\bar{X} < \bar{x}_* \mid \mu = 12.5) = P\left(Z < \frac{\bar{x}_* - 12.5}{3/\sqrt{n}}\right) = 0.90$$
Since $P(Z < 1.2816) \approx 0.90$,
$$\frac{\bar{x}_* - 12.5}{3/\sqrt{n}} = 1.2816 \quad \Longrightarrow \quad \bar{x}_* = 12.5 + 1.2816\left(\frac{3}{\sqrt{n}}\right)$$

**(c)** Setting the two expressions for $\bar{x}_*$ equal:
$$14 - 1.6449\left(\frac{3}{\sqrt{n}}\right) = 12.5 + 1.2816\left(\frac{3}{\sqrt{n}}\right)$$
$$n = \left[\frac{(1.6449 + 1.2816)(3)}{14 - 12.5}\right]^2 \approx 34.26$$
The calculation gives $n = 34.26$, so the minimum required sample size is $\boxed{35}$ patients, because sample size must be a whole number and rounding down to 34 would leave the achieved power slightly below the target of 0.90.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 23</div>
A public health survey reports that $p_0 = 0.30$ of adults in a region meet recommended weekly exercise guidelines. After a city-wide fitness campaign, researchers want to test whether the true proportion has *increased*, using
$$H_0: p = 0.30 \qquad \text{vs.} \qquad H_a: p > 0.30$$
with $n = 400$ adults surveyed at $\alpha = 0.05$, using the Normal approximation to the sampling distribution of $\hat{p}$.

(a) Find the rejection cutoff $\hat{p}^{\,*}$ under $H_0$.
(b) Calculate the power of this test against the true alternative $p_a = 0.35$.
(c) Interpret the power value in context.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- The rejection cutoff must use the standard error under $H_0$, based on $p_0 = 0.30$: $\hat{p}^{\,*} = p_0 + z_{1-\alpha}\sqrt{p_0(1-p_0)/n}$.
- Power requires standardizing $\hat{p}^{\,*}$ using the standard error based on $p_a = 0.35$ instead — the two stages use different standard errors.
- This is an upper-tailed test, so use $P(Z > \cdot)$ for the power calculation.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** With $z_{\text{crit}} = z_{0.95} \approx 1.6449$:
$$\hat{p}^{\,*} = 0.30 + 1.6449\sqrt{\frac{0.30(0.70)}{400}} \approx 0.3377$$
The test rejects $H_0$ whenever $\hat{p} > 0.3377$.

**(b)** Standardizing $\hat{p}^{\,*} = 0.3377$ using the true alternative $p_a = 0.35$:
$$\text{Power} = P\left(Z > \frac{0.3377 - 0.35}{\sqrt{0.35(0.65)/400}}\right) = P(Z > -0.5162) \approx 0.6972$$
(This uses the Normal approximation to the sampling distribution of $\hat{p}$, which is reasonable here since $np_a$ and $n(1-p_a)$ are both well above 10.)

**(c)** There is approximately a 69.7% probability that this test correctly detects a true increase in the exercise-guideline compliance rate to 35%. There is roughly a 30.3% chance ($\beta$) that this real increase goes undetected by the survey.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 24</div>
A subscription service currently retains $p_0 = 0.75$ of customers who report being "satisfied" after one year. Management is testing a new loyalty program, believing it will *increase* the satisfaction proportion, and wants to test $H_0: p = 0.75$ vs. $H_a: p > 0.75$ at $\alpha = 0.05$. They want the power against the alternative $p_a = 0.82$ to be at least $0.85$.

(a) Write the rejection rule for $\hat{p}^{\,*}$ in terms of the (unknown) sample size $n$.
(b) Write the expression for power in terms of $n$, and set it equal to $0.85$.
(c) Solve for the minimum required sample size.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Follow the same two-equation strategy used for the mean-test sample-size problems, but with $\hat{p}^{\,*} = p_0 + z_{1-\alpha}\sqrt{p_0(1-p_0)/n}$ under $H_0$ and a second expression for $\hat{p}^{\,*}$ under $p_a = 0.82$ with power $= 0.85$.
- The two standard-error terms (one using $p_0$, one using $p_a$) do **not** simplify to the same expression — keep them separate until the final equation.
- Round the final $n$ up, not down.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** By definition,
$$\alpha = P(\hat{p} > \hat{p}^{\,*} \mid p = 0.75) = P\left(Z > \frac{\hat{p}^{\,*} - 0.75}{\sqrt{0.75(0.25)/n}}\right) = 0.05$$
Since $P(Z > 1.6449) = 0.05$,
$$\hat{p}^{\,*} = 0.75 + 1.6449\sqrt{\frac{0.75(0.25)}{n}}$$

**(b)** The power condition is
$$\text{Power} = P(\hat{p} > \hat{p}^{\,*} \mid p = 0.82) = P\left(Z > \frac{\hat{p}^{\,*} - 0.82}{\sqrt{0.82(0.18)/n}}\right) = 0.85$$
Since $P(Z > -1.0364) \approx 0.85$,
$$\hat{p}^{\,*} = 0.82 - 1.0364\sqrt{\frac{0.82(0.18)}{n}}$$

**(c)** Setting the two expressions for $\hat{p}^{\,*}$ equal:
$$0.75 + 1.6449\sqrt{\frac{0.75(0.25)}{n}} = 0.82 - 1.0364\sqrt{\frac{0.82(0.18)}{n}}$$
$$0.07 = \frac{1}{\sqrt{n}}\left(1.6449\sqrt{0.1875} + 1.0364\sqrt{0.1476}\right) = \frac{1.1105}{\sqrt{n}}$$
$$\sqrt{n} = \frac{1.1105}{0.07} \approx 15.865 \quad \Longrightarrow \quad n \approx 251.64$$
The calculation gives $n = 251.64$, so the minimum required sample size is $\boxed{252}$ customers, because sample size must be a whole number and rounding down to 251 would leave the achieved power slightly below the target of 0.85.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 25</div>
A hospital's 30-day readmission rate has historically been $p_0 = 0.22$. A new discharge-planning program is designed to *reduce* readmissions, and the hospital is deciding among three proposed pilot study sizes before committing resources: $n = 100$, $n = 200$, and $n = 300$ patients. The test in each case is
$$H_0: p = 0.22 \qquad \text{vs.} \qquad H_a: p < 0.22 \qquad \text{at } \alpha = 0.05$$
and the hospital wants to know the power against a true post-program readmission rate of $p_a = 0.16$ for each design.

(a) Compute the rejection cutoff $\hat{p}^{\,*}$ and the power against $p_a = 0.16$ for each of the three sample sizes.
(b) Based on your results, which design would you recommend if the hospital can only afford to enroll $n = 200$ patients but wants power of at least $0.80$? Explain.
(c) Suppose the pilot study is run with $n = 300$ patients, the test is conducted, and a p-value of $0.032$ is reported. Explain the conceptual difference between this p-value and the power calculated in part (a) for $n = 300$ — in particular, why the p-value alone (without the power calculations above) would not tell the hospital anything about how reliably the *design* could have detected a true reduction to $p_a = 0.16$ before the study was ever run.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Compute $\hat{p}^{\,*} = p_0 + z_{\alpha}\sqrt{p_0(1-p_0)/n}$ for each $n$ (this is a lower-tailed test), then standardize under $p_a = 0.16$ to get power for each design.
- For part (b), compare the computed power values directly to the target of $0.80$.
- For part (c), think about *when* each quantity can be calculated (before vs. after data collection) and what each one conditions on.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** Using $z_{\text{crit}} = z_{0.05} \approx -1.6449$:

| $n$ | $\hat{p}^{\,*}$ | Power against $p_a = 0.16$ |
|---|---|---|
| 100 | 0.1519 | 0.4122 |
| 200 | 0.1718 | 0.6758 |
| 300 | 0.1807 | 0.8355 |

For example, at $n = 300$: $\hat{p}^{\,*} = 0.22 - 1.6449\sqrt{0.22(0.78)/300} \approx 0.1807$, and
$$\text{Power} = P\left(Z < \frac{0.1807 - 0.16}{\sqrt{0.16(0.84)/300}}\right) \approx P(Z < 0.976) \approx 0.8355$$

**(b)** At $n = 200$, the power against $p_a = 0.16$ is only about $0.6758$, which falls short of the target of $0.80$. None of the three proposed sizes below 300 would meet the target — only the $n = 300$ design achieves power above $0.80$ (about $0.8355$). If the hospital is limited to $n = 200$, it should recognize that this design has roughly a 32% chance of failing to detect a real reduction to 16%, and should either seek to enroll more patients or accept this comparatively higher risk of a Type II error.

**(c)** The p-value of $0.032$ is calculated *after* the data are collected, and it measures the strength of evidence in this *particular sample* against $H_0: p = 0.22$ — it says nothing about $p_a = 0.16$ specifically, nor about any other possible alternative. Power, by contrast, is calculated *before* the data are collected (during study design) and is a property of the testing procedure itself against a *specific* stated alternative ($p_a = 0.16$): it answers "if the true rate really were 0.16, how often would a study of this design correctly detect it?" A design can have low power against $p_a = 0.16$ and still occasionally produce a small p-value in an actual sample (as happened here), or have high power and still fail to reject $H_0$ in a particular unlucky sample. The p-value from a single completed study cannot retroactively tell the hospital how reliable its chosen sample size was — that question can only be answered by the power calculations performed in part (a).
</div>
</details>
