# Analysis of Variance

ANOVA (Analysis of Variance) is used to compare multiple means.<br>
Suppose we have $k$ groups $(k \geqslant 3)$ and each group is given a treatment.

The ANOVA model is:

\[
Y_{ij} = \mu + \tau_i + \varepsilon_{ij}, \quad \varepsilon_{ij} \sim N(0, \sigma^2)
\]

Where:  
- \( Y_{ij} \): outcome on unit \( j \) in group \( i \)  
- \( \mu \): overall mean  
- \( \tau_i \): treatment-specific effect for group \( i \)  
- \( \varepsilon_{ij} \): error term, assumed normally distributed  
- \( \mu_i = \mu + \tau_i \): group mean

The equivalent form of the ANOVA model is:

\[
Y_{ij} = \mu_i + \varepsilon_{ij}, \quad \varepsilon_{ij} \sim \mathcal{N}(0, \sigma^2)
\]


\[
\begin{array}{c|c|c|c|c}
\text{Group 1} & \text{Group 2} & \cdots & \text{Group } K \\
\hline
X_{11} & X_{21} & \cdots & X_{K1} \\
X_{12} & X_{22} & \cdots & X_{K2} \\
\vdots & \vdots & \ddots & \vdots \\
X_{1n_1} & X_{2n_2} & \cdots & X_{Kn_K} \\
\hline
\bar{X}_1 & \bar{X}_2 & \cdots & \bar{X}_K \\
S_1^2 & S_2^2 & \cdots & S_K^2 \\
n_1 & n_2 & \cdots & n_K \\
\end{array}
\]

Where:  
- \( \bar{X}_i \): sample mean of group \( i \)  
- \( S_i^2 \): sample variance of group \( i \)  
- \( n_i \): sample size of group \( i \)


**Total Sample Size and Overall Mean**

Total sample size:

\[
n = n_1 + n_2 + \cdots + n_k
\]

Overall sample mean:

\[
\bar{X} = \frac{1}{n} \sum_{i=1}^k \sum_{j=1}^{n_i} X_{ij} = \frac{1}{n} \sum_{i=1}^k n_i \bar{X}_i
\]

---

**Sources of Variation**

Variation **within** groups (SSW / SSE):

\[
\text{SSE} = \sum_{i=1}^k (n_i - 1) \cdot S_i^2
\]

This represents **error** or **residual** variation.

---

Variation **between** groups (SSB / SSTr):

\[
\text{SSTr} = \sum_{i=1}^k n_i \cdot (\bar{X}_i - \bar{X})^2
\]

This captures the **treatment effect** (how group means differ from overall mean).

---

**Total Variation**

\[
\text{SSTotal} = \sum_{i=1}^k \sum_{j=1}^{n_i} (X_{ij} - \bar{X})^2
\]

---

**Decomposition Identity**

\[
\text{SSTotal} = \text{SSTr} + \text{SSE}
\]


**ANOVA Table**

| Source     | df     | SS       | MS                          | F-Statistic         |
|------------|--------|----------|-----------------------------|---------------------|
| Treatment  | k - 1  | SSTr     | MSTr = SSTr / (k - 1)       | F* = MSTr / MSE     |
| Error      | n - k  | SSE      | MSE = SSE / (n - k)         |                     |
| Total      | n - 1  | SSTotal  |                             |                     |
                          


Where \( MS = \dfrac{SS}{df} \)

---

**Hypothesis Test**
\[
\begin{aligned}
H_0\!: &\quad \mu_1 = \mu_2 = \cdots = \mu_k \\
&\quad \text{(all means are equal → all treatments equally effective)} \\
\\
H_a\!: &\quad \text{At least one } \mu_j \text{ (for } j = 1, \dots, k\text{) is different} \\
&\quad \text{(→ at least one treatment has a different effect)}
\end{aligned}
\]



**Test Statistic**

\[
F^* = \frac{\text{MSTr}}{\text{MSE}} = \frac{\text{SSTr} / (k - 1)}{\text{SSE} / (n - k)} \sim F_{(k - 1,\ n - k)}
\]

---

**Reference Distribution**

The test statistic follows an \( F \)-distribution with:

- numerator degrees of freedom = \( k - 1 \)  
- denominator degrees of freedom = \( n - k \)

---

**P-value**

```{r fig.cap="F-distribution with shaded p-value area", echo=FALSE, fig.align='center', out.width='70%'}
# Load library
library(ggplot2)

# Parameters
df1 <- 3   # numerator degrees of freedom (k - 1)
df2 <- 20  # denominator degrees of freedom (n - k)
f_star <- 3.2

# Generate F-distribution data
x_vals <- seq(0, 6, length.out = 500)
y_vals <- df(x_vals, df1, df2)
f_data <- data.frame(x = x_vals, y = y_vals)
```

```{r fig.cap="F-distribution with shaded p-value area", echo=FALSE, fig.align='center', out.width='70%'}
# Plot
ggplot(f_data, aes(x, y)) +
  geom_line() +
  geom_area(data = subset(f_data, x > f_star), aes(x, y), fill = "#FF68A1", alpha = 0.5) +
  geom_vline(xintercept = f_star, linetype = "dashed") +
  annotate("text", x = f_star + 1.2, y = max(y_vals) * 0.5, label = "p-value", size = 4) +
  annotate("text", x = f_star, y = -0.01, label = "F*", vjust = 1.5) +
  labs(x = "F", y = "Density") +
  theme_minimal(base_size = 12)
```

**The Data Matrix**

The following table shows last year's sales data for a small business. The sample is put into a matrix format in which each of the three columns corresponds to one of the three countries in which the company does business. The numbers in the cell represent the sales (in units of $\$ 1000$ ) made in that country last year. These data will be used to develop the theory underlying **Analysis of Variance**, or, for short, **ANOVA**.


|  | Country A | Country B | Country C |
| :---: | :---: | :---: | :---: |
|  | 6 | 10 | 14 |
|  | 10 | 8 | 13 |
|  | 7 | 12 | 11 |
|  | 9 | 10 | 10 |
| Average | 8 | 10 | 12 |

Altogether, there were 12 sales last year that totaled \$120 – so the average sale was \$10.  

The column (country) averages are:  
Column 1 (Country A): \$8  
Column 2 (Country B): \$10  
Column 3 (Country C): \$12


Now we will begin our study of how to make a statistically valid prediction of the next sales figure. In that regard, there are two possible situations that can occur:

1. The country of the next sale (observation) is **not** known.  
2. The country of the next sale is known.


Situation 1.

Without any additional information, the best prediction is the sample mean $\$ 10$. This prediction is best in the least squares sense - that is, if $\$ 10$ had been used to predict each of the 12 observations in the sample, then the total of the squared errors $S S_{\text {Total }}$ would be as small as possible. In our data set, $S S_{\text {Total }}$ equals 60 . That figure can be verified by calculating $\sum\left(x_{i}-10\right)^{2}$ for each observation $x_{i}$ of the sample.

Situation 2. One-factor ANOVA Model.

If the country of the next sale is known, then two different predictions are possible for the next sales figure:

- The sample mean $\$ 10$.
- The mean of the sales of the country in which the next sale will occur. (In this case, $\$ 8$ if the next sale will occur in Country A, $\$ 10$ in Country B, or $\$ 12$ in Country C.) This prediction **ignores** the information present in the sales figures from the other two countries.


**The Null Hypothesis for One-Factor ANOVA**

We have discussed the prediction possibilities for one-factor ANOVA models. Now, we will learn how to test the statistical significance of a one-factor ANOVA model.
Let's suppose that we want to predict the next sales figure, and that we know the country in which this sale will occur. Without any statistical testing, we can always by default use the sample mean $\$ 10$ to predict the next sale. The default prediction, the sample mean, doesn't use any information about the country (column) in which the sale will occur.

However, if instead we use the mean of the observations in only one column (the column that corresponds to the particular country in which we know the next sale will occur), then we have to test the null hypothesis

$$
H_{0}: \mu_{C O L 1}, \mu_{C O L 2}, \mu_{C O L 3}, \text { are equal }
$$

and reject it in favor of the alternative hypothesis\

$$
H_{a}: \mu_{C O L 1}, \mu_{C O L 2}, \mu_{C O L 3}, \text { are NOT all equal }
$$

If the null hypothesis is rejected, then we can be statistically confident that the column means are not all equal, and therefore that the individual column means (i. e., $\$ 8, \$ 10, \$ 12$ ) can be used to predict the amount of the next sale. If the next sale was going to occur in Country A , then the prediction would be $\$ 8$. If the next sale was going to occur in Country B, then the prediction would be $\$ 10$. If the next sale was going to occur in Country C , then the prediction would be $\$ 12$.

**The One-Factor ANOVA F Test**

To test the null hypothesis stated above, we have to calculate an F-statistic. If $F_{*}>F_{(c-1, n-c), \alpha}$, then reject $H_{0}$, and use the sample column means to predict future observations. Otherwise, do not reject $H_{0}$ and use the overall sample mean to predict future observations.

**ANOVA Table**

To see how this $F_{*}$ is calculated, see the ANOVA Table below.

| Source of <br> Variation | Degrees of <br> Freedom <br> $(\mathrm{df})$ | Sum of <br> Square <br> $(\mathrm{SS})$ | Mean Sum of <br> Squares <br> $($ MSS $)$ | F Ratio |
| :---: | :---: | :---: | :---: | :---: |
| Treatments | $\mathrm{c}-1$ | SST | $\frac{\text { SST }}{c-1}$ | $F=\frac{\text { MST }}{\text { MSE }}$ |
| Error | $\mathrm{n}-\mathrm{c}$ | SSE | $\frac{\text { SSE }}{n-c}$ |  |
| Total | $\mathrm{n}-1$ | $\mathrm{SS}_{\mathrm{Total}}$
 |  |  |

**Calculation of $\mathrm{SS}_{\mathrm{Total}}$**

If **no** model is used, then the predictions for each of the 12 observations will be 10 . If these predictions are used, the squared error of these 12 predictions is given in the table below.

| Country A | Country B | Country C |
| :---: | :---: | :---: |
| 16 | 0 | 16 |
| 0 | 4 | 9 |
| 9 | 4 | 1 |
| 1 | 0 | 0 |

Prediction Errors Squared when NO Factor is used $($ Total $)=60$.

**Calculation of SSE**

If the column model is used, then the 12 observations would have the following 12 predictions, where $\$ 8$ is the average for the first column, $\$ 10$ is the average for the second column, and $\$ 12$ is the average for the third column.

| Country A | Country B | Country C |
| :---: | :---: | :---: |
| 8 | 10 | 12 |
| 8 | 10 | 12 |
| 8 | 10 | 12 |
| 8 | 10 | 12 |

**Calculation of SSE**

Using the above 12 predictions, the errors squared are shown in the table below.

| Country A | Country B | Country C |
| :---: | :---: | :---: |
| 4 | 0 | 4 |
| 4 | 4 | 1 |
| 1 | 4 | 1 |
| 1 | 0 | 4 |

Errors Squared when the Column Factor is used $($ Total $)=28$.

**Calculation of SST**

The units explained by the column model are calculated by finding the square of each prediction change when moving from NO model to the column model. The following table presents the square of each prediction change:

| Country A | Country B | Country C |
| :---: | :---: | :---: |
| 4 | 0 | 4 |
| 4 | 0 | 4 |
| 4 | 0 | 4 |
| 4 | 0 | 4 |

Table of the Square of the Prediction Change when Moving from NO Model to the Column Model $($ Total $)=32$.

**ANOVA Table**

The ANOVA Table for the column factor can now be filled in as shown below:

| Source of <br> Variation | Degrees of <br> Freedom <br> $(\mathrm{df})$ | Sum of <br> Square <br> $(\mathrm{SS})$ | Mean Sum of <br> Squares <br> $($ MSS $)$ | F Ratio |
| :---: | :---: | :---: | :---: | :---: |
| Treatments | 2 | 32 | $\frac{32}{2}=16$ | $\frac{16}{3.1111}=5.1428$ |
| Error | 9 | 28 | $\frac{28}{9}=3.1111$ |  |
| Total | 11 | 60 |  |  |

So for this one-factor ANOVA model, $F_{*}=5.1428$.

**Conclusion**

If the null hypothesis is true, then the F-statistic should be a value from the $F_{2,9}$ distribution. Referring to the table that contains the upper 0.05 cut-off points of F distributions, we see that $F_{(2,9), 0.05}=4.256$. Since 5.1428 is greater than 4.256 , this tells us that the F -statistic is in the upper 0.05 of the $F_{2,9}$ distribution. Therefore we reject the null hypothesis at the 0.05 significance level, and we conclude that the country means are not all the same. Thus, the prediction for the next sale in a known country is the mean of all the previous sales in that country.

```r
# R Code;
salesA=c(6,10,7,9);
salesB=c (10, 8,12,10);
salesC=c(14,13,11,10);
sales=c(salesA,salesB,salesC);
country=c(rep(1,4),rep(2,4),rep(3,4));
oneway.test(sales~country,var.equal=TRUE);
```

```r
##
## One-way analysis of means
##
## data: sales and country
## F = 5.1429, num df = 2, denom df = 9, p-value =
## 0.0324
```

```r
# R Code;
# Another way: using lm;
salesA=c(6,10,7,9);
salesB=c (10, 8,12,10);
salesC=c(14,13,11,10);
sales=c(salesA,salesB,salesC);
country=c(rep(1,4),rep(2,4),rep(3,4));
anova(lm(sales~factor(country)));
```

```r
## Analysis of Variance Table
##
## Response: sales
## Df Sum Sq Mean Sq F value Pr(>F)
## factor(country) 2 32 16.0000 5.1429 0.0324 *
## Residuals 9 28 3.1111
## ---
## Signif. codes:
## 0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
```

Officially, to use the predictions from an ANOVA model, three assumptions about the populations from which the sample was taken must be satisfied:

1. Each population has a Normal distribution.
2. Each population has the **same** standard deviation $\sigma$.
3. The observations are mutually independent of one another.

**Formulas**

Sum of Squares for Treatments (a.k.a. between-treatments variation or Explained)

$$
S S T=\sum_{j=1}^{k} n_{j}\left(\bar{x}_{j}-\overline{\bar{x}}\right)^{2}
$$

Sum of Squares for Error (a.k.a. within-treatments variation or Unexplained)

$$
S S E=\sum_{j=1}^{k} \sum_{i=1}^{n_{j}}\left(x_{i j}-\bar{x}_{j}\right)^{2}=\left(n_{1}-1\right) s_{1}^{2}+\ldots+\left(n_{k}-1\right) s_{k}^{2}
$$

**Mean Square for Treatments**

$$
M S T=\frac{S S T}{k-1}
$$

**Mean Square for Error**

$$
M S E=\frac{S S E}{n-k}
$$


**Test Statistic**

$$
F=\frac{M S T}{M S E}
$$

Exercise

A statistics practitioner calculated the following statistics:

\[
\begin{array}{|c|c|c|c|}
\hline
\textbf{Statistic} & \textbf{1} & \textbf{2} & \textbf{3} \\
\hline
n & 5 & 5 & 5 \\
\hline
\bar{x} & 10 & 15 & 20 \\
\hline
s^2 & 50 & 50 & 50 \\
\hline
\end{array}
\]

Complete the ANOVA table.

Solution

$$
\begin{aligned}
& \overline{\bar{x}}=\frac{5(10)+5(15)+5(20)}{5+5+5}=15 \\
& S S T=5(10-15)^{2}+5(15-15)^{2}+5(20-15)^{2}=250 \\
& S S E=(5-1)(50)+(5-1)(50)+(5-1)(50)=600
\end{aligned}
$$

ANOVA Table

| Source of Variation | Degrees of Freedom (df) | Sum of Square (SS) | Mean Sum of Squares (MSS) | F Ratio |
| :--- | :--- | :--- | :--- | :--- |
| Treatments | 2 | 250 | $\frac{250}{2}=125$ | $\frac{125}{50}=2.50$ |
| Error | 12 | 600 | $\frac{600}{12}=50$ |  |
| Total | 14 | 850 |  |  |

Exercise

A statistics practitioner calculated the following statistics:

\[
\begin{array}{|c|c|c|c|}
\hline
\textbf{Statistic} & \textbf{1} & \textbf{2} & \textbf{3} \\
\hline
n & 4 & 4 & 4 \\
\hline
\bar{x} & 20 & 22 & 25 \\
\hline
s^2 & 10 & 10 & 10 \\
\hline
\end{array}
\]


Complete the ANOVA table.

Solution

$$
\begin{aligned}
& \overline{\bar{x}}=\frac{4(20)+4(22)+4(25)}{4+4+4}=22.33 \\
& S S T=4(20-22.33)^{2}+4(22-22.33)^{2}+4(25-22.33)^{2}=50.67 \\
& S S E=(4-1)(10)+(4-1)(10)+(4-1)(10)=90
\end{aligned}
$$

ANOVA Table

| Source of Variation | Degrees of Freedom (df) | Sum of Square (SS) | Mean Sum of Squares (MSS) | F Ratio |
| :--- | :--- | :--- | :--- | :--- |
| Treatments | 2 | 50.67 | $\frac{50.67}{2}=25.33$ | $\frac{25.33}{10}=2.53$ |
| Error | 9 | 90 | $\frac{90}{9}=10$ |  |
| Total | 11 | 140.67 |  |  |

Exercise

A consumer organization was concerned about the differences between the advertised sizes of containers and the actual amount of product. In a preliminary study, six packages of three different brands of margarine that are supposed to contain 500 ml were measured. The differences from 500 ml are listed here. Do these data provide sufficient evidence to conclude that differences exist between the three brands? Use $\alpha=0.05$.

| Brand 1 | Brand 2 | Brand 3 |
| :---: | :---: | :---: |
| 1 | 2 | 1 |
| 3 | 2 | 2 |
| 3 | 4 | 4 |
| 0 | 3 | 2 |
| 1 | 0 | 3 |
| 0 | 4 | 4 |

Exercise on 3 bounds

Consider  
$\alpha = 0.05$

| Brand 1       | Brand 2       | Brand 3       |
|:-------------:|:-------------:|:-------------:|
| $x_{11}=1$    | $x_{21}=2$    | $x_{31}=1$    |
| $x_{12}=3$    | $x_{22}=2$    | $x_{32}=2$    |
| 3             | 4             | 4             |
| 0             | 3             | 2             |
| 1             | 0             | 3             |
| $x_{16}=0$    | $x_{26}=4$    | $x_{36}=4$    |
| $n_1=6$       | $n_2=6$       | $n_3=6$       |



Overall Sample Size

\[
n = n_1 + n_2 + n_3 = 18
\]

Sample Means

**Group 1:**
\[
\bar{X}_1 = \frac{1}{n_1} \sum_{j=1}^{n_1} x_{1j} = \frac{1}{6}(1 + 3 + \dots + 0) = 1.33
\]

**Group 2:**
\[
\bar{X}_2 = \frac{1}{n_2} \sum_{j=1}^{n_2} x_{2j} = \frac{1}{6}(2 + 2 + \dots + 4) = 2.50
\]

**Group 3:**
\[
\bar{X}_3 = \frac{1}{n_3} \sum_{j=1}^{n_3} x_{3j} = \frac{1}{6}(1 + 2 + \dots + 4) = 2.67
\]

Overall Mean

\[
\bar{X} = \frac{1}{n} \sum_{i=1}^{k} \sum_{j=1}^{n_i} x_{ij} = \frac{1 + 3 + \dots + 0 + 2 + 2 + \dots + 4 + 1 + 2 + \dots + 4}{18}
\]

\[
= \frac{1}{n} \sum_{i=1}^{k} n_i \bar{X}_i = \frac{1}{18} \left[ (6)(1.33) + (6)(2.5) + (6)(2.67) \right]
\]

\[
= 2.17
\]


| Brand 1         | Brand 2         | Brand 3         |
|:---------------:|:---------------:|:---------------:|
| $x_{11} = 1$    | $x_{21} = 2$    | $x_{31} = 1$    |
| $x_{12} = 3$    | $x_{22} = 2$    | $x_{32} = 2$    |
| 3               | 4               | 4               |
| 0               | 3               | 2               |
| 1               | 0               | 3               |
| $x_{16} = 0$    | $x_{26} = 4$    | $x_{36} = 4$    |
| $n_1 = 6$       | $n_2 = 6$       | $n_3 = 6$       |
| $\bar{x}_1 = 1.33$ | $\bar{x}_2 = 2.50$ | $\bar{x}_3 = 2.67$ |



Sample Variances

**Group 1:**

\[
S_1^2 = \frac{1}{n_1 - 1} \sum_{j=1}^{n_1} (x_{1j} - \bar{x}_1)^2
\]

\[
= \frac{1}{6 - 1} \left[ (1 - 1.33)^2 + \dots + (0 - 1.33)^2 \right]
\]

\[
= 1.87
\]

---

**Group 2:**

\[
S_2^2 = \dots = 2.30
\]

---

**Group 3:**

\[
S_3^2 = \dots = 1.47
\]


| Brand 1          | Brand 2          | Brand 3          |
|:----------------:|:----------------:|:----------------:|
| $x_{11} = 1$     | $x_{21} = 2$     | $x_{31} = 1$     |
| $x_{12} = 3$     | $x_{22} = 2$     | $x_{32} = 2$     |
| 3                | 4                | 4                |
| 0                | 3                | 2                |
| 1                | 0                | 3                |
| $x_{16} = 0$     | $x_{26} = 4$     | $x_{36} = 4$     |
| $n_1 = 6$        | $n_2 = 6$        | $n_3 = 6$        |
| $\bar{x}_1 = 1.33$ | $\bar{x}_2 = 2.50$ | $\bar{x}_3 = 2.67$ |
| $S_1^2 = 1.87$   | $S_2^2 = 2.30$   | $S_3^2 = 1.47$   |

Overall mean:  
\[
\bar{x} = 2.17, \quad n = 18
\]

Within (Error)

\[
SSE = \sum_{i=1}^{3} (n_i - 1) S_i^2 = (6 - 1)(1.87) + (6 - 1)(2.30) + (6 - 1)(1.47)
\]

\[
= \boxed{28.20}
\]

---

Between (Treatment effect)

\[
SS_{\text{Tot}} = \sum_{i=1}^{3} n_i (\bar{x}_i - \bar{x})^2 = 6(1.33 - 2.17)^2 + 6(2.50 - 2.17)^2 + 6(2.67 - 2.17)^2
\]

\[
= \boxed{6.39}
\]

ANOVA Table

\[
MS = \frac{SS}{df}
\]


| Source of Variation | Degrees of Freedom (df) | Sum of Squares (SS) | Mean Sum of Squares (MSS)     | F Ratio                         |
| :------------------ | :---------------------- | :------------------ | :----------------------------- | :------------------------------ |
| Treatment           | $3 - 1 = 2$             | 6.39                | $\frac{6.39}{2} = 3.195$       | $\frac{3.195}{1.88} = 1.70$     |
| Error               | $18 - 3 = 15$           | 28.20               | $\frac{28.20}{15} = 1.88$      |                                |
| Total               | $18 - 1 = 17$           | 34.59               | $\times$                       |                                |



Hypothesis Test

\[
\begin{aligned}
H_0\!: &\quad \mu_1 = \mu_2 = \mu_3 \\
&\quad \text{(all means are equal} \rightarrow \text{ all treatments equally effective)} \\
\\
H_a\!: &\quad \text{At least one } \mu_j,\ j = 1, \dots, 3\ \text{ is different} \\
&\quad \text{(} \rightarrow \text{ at least one treatment has a different effect)}
\end{aligned}
\]


Test Statistic

\[
F^* = \frac{MS_{\text{Tr}}}{MSE} = \frac{SS_{\text{Tr}}/(k-1)}{SSE/(n-k)} = 1.70 \sim F_{(2,15)}
\]

F distribution with numerator df = 2 and denominator df = 15.

\[
\text{p-value} > 0.100 > 0.05 \quad (\alpha)
\]

Conclusion of ANOVA

There is **insufficient evidence to reject**  
\[
H_0: \mu_1 = \mu_2 = \mu_3
\]

The analysis from ANOVA suggests that all 3 groups have the same mean.



What if $H_0$ had been rejected?

Suppose we rejected $H_0$.  
Which mean (or means) are different?

\[
\mu_1,\ \mu_2,\ \mu_3
\]

Pairwise Comparisons

We would compare the means in **pairs**:

- $\mu_1$ vs. $\mu_2$  
- $\mu_1$ vs. $\mu_3$  
- $\mu_2$ vs. $\mu_3$

Number of comparisons:  
\[
\binom{k}{2}
\]

Solution

Step 1. State Hypotheses.

$\mu_{i}=$ population mean for differences from 500 ml (brand $i$, where
$i=1,2,3$ ).

$H_{0}: \mu_{1}=\mu_{2}=\mu_{3}$

$H_{a}$ : At least two means differ.

Step 2. Compute test statistic.

|  | Brand 1 | Brand 2 | Brand 3 |
| :---: | :---: | :---: | :---: |
| Mean | 1.33 | 2.50 | 2.67 |
| Variance | 1.87 | 2.30 | 1.47 |

Grand mean = \( \bar{x} = 2.17 \).

\[
SST = 6(1.33 - 2.17)^2 + 6(2.50 - 2.17)^2 + 6(2.67 - 2.17)^2 = 6.387 \approx 6.39
\]

\[
SSE = (6 - 1)(1.87) + (6 - 1)(2.30) + (6 - 1)(1.47) = 28.20
\]


 ANOVA Table

| Source of Variation | Degrees of Freedom (df) | Sum of Square (SS) | Mean Sum of Squares (MSS) | F Ratio |
| :--- | :--- | :--- | :--- | :--- |
| Treatments | 2 | 6.39 | $\frac{6.39}{2}=3.195$ | $\frac{3.195}{1.88}=1.70$ |
| Error | 15 | 28.20 | $\frac{28.20}{15}=1.88$ |  |
| Total | 17 | 34.59 |  |  |

Step 3. Find Rejection Region.
We reject the null hypothesis only if

$$
F>F_{\alpha, k-1, n-k}
$$

If we let $\alpha=0.05$, the rejection region for this exercise is

$$
F>F_{0.05,2,15}=3.682
$$

Step 4. Conclusion.

We found the value of the test statistic to be $F=1.70$. Since $F=1.70<F_{0.05,2,15}=3.682$, we can't reject $H_{0}$. Thus, there is **not** evidence to infer that the average differences differ between the three brands.

```r
# R Code;
brand1=c(1,3,3,0,1,0);
brand2=c (2,2,4,3,0,4);
brand3=c(1,2,4,2,3,4);
differences=c(brand1,brand2,brand3);
brand=c(rep(1,6),rep(2,6),rep(3,6));
oneway.test(differences~brand,var.equal=TRUE);
```

```r
##
## One-way analysis of means
##
## data: differences and brand
## F = 1.6864, num df = 2, denom df = 15, p-value
## = 0.2185
##
```

Margarine example

```r
> brand = as.factor(brand)
> anova_model = aov(differences ~ brand)
> anova(anova_model)
```


|  | Df | Sum Sq | Mean Sq | $F$ value |
| :--- | ---: | ---: | ---: | :---: |
| brand | 2 | 6.3333 | 3.1667 | 1.6864 |
| Residuals | 15 | 28.1667 | 1.8778 | 0.2185 |




Exercise

The friendly folks a the Internal Revenue Service (IRS) in the United States and Canada Revenue Agency (CRA) are always looking for ways to improve the wording and format of its tax return forms. Three new forms have been developed recently. To determine which, if any, are superior to the current form, 120 individuals were asked to participate in an experiment. Each of the three new forms and the currently used form were filled out by 30 different people. The amount of time (in minutes) taken by each person to complete the task was recorded.
What conclusions can be drawn from these data?

R Code

```r
#Step 1. Entering data;
# importing data;
# url of tax return forms;
forms_url =
"https://mcs.utm.utoronto.ca/"nosedal/data/tax-forms.txt"
forms_data= read.table(forms_url,header=TRUE);
names(forms_data);
forms_data[1:4, ];
```


R Code

```r
## [1] "Form1" "Form2" "Form3" "Form4"
##   Form1 Form2 Form3 Form4
## 1    23    88   116   103
## 2    59   114   123   122
## 3    68    81    64   105
## 4   122    41   136    73
```

R Code

```r
#Step 2. ANOVA;
time1=forms_data$Form1;
time2=forms_data$Form2;
time3=forms_data$Form3;
time4=forms_data$Form4;
length(forms_data$Form1);
times=c(time1,time2,time3,time4);
forms=c(rep(1,30),rep(2,30),rep(3,30),rep(4,30));
oneway.test(times~forms,var.equal=TRUE)
```

R Code

```r
## [1] 30
##
## One-way analysis of means
##
## data: times and forms
## F = 2.9358, num df = 3, denom df = 116, p-value
## = 0.03632
```

Assumptions of ANOVA

Model

\[
Y_{ij} = \mu + \tau_i + \epsilon_{ij} \quad \text{where} \quad \epsilon_{ij} \sim \mathcal{N}(0, \sigma^2)
\]

or equivalently,

\[
Y_{ij} = \mu_i + \epsilon_{ij} \quad \text{with} \quad \epsilon_{ij} \sim \mathcal{N}(0, \sigma^2)
\]

*Observations of a group are assigned the same group mean.*

**Assumptions**

1.Observations are independent

2.Error terms (residuals) are Normal

3.All groups have the same population variance

\[
\sigma_1^2 = \sigma_2^2 = \cdots = \sigma_k^2 = \sigma^2
\]

This common variance is estimated using the **Mean Squared Error (MSE)** from ANOVA:

\[
\hat{\sigma}^2 = \text{MSE}
\]




**Multiple Comparisons**

Example

Because of foreign competition, North American automobile manufacturers have become more concerned with quality. One aspect of quality is the cost of repairing damage caused by accidents. A manufacturer is considering several new types of bumpers. To test how well they react to low-speed collisions, 10 bumpers of each of four different types were installed on mid-size cars, which were then driven into a wall at 5 miles per hour. The cost of repairing the damage in each case was assessed. The data are shown below.

a. Is there sufficient evidence at the $5 \%$ significance level to infer that the bumpers differ in their reactions to low-speed collisions?

b. If differences exist, which bumpers differ? Apply Fisher's LSD method with the Bonferroni adjustment.


| Bumper 1 | Bumper 2 | Bumper 3 | Bumper 4 |
| :--- | :--- | :--- | :--- |
| 610 | 404 | 599 | 272 |
| 354 | 663 | 426 | 405 |
| 234 | 521 | 429 | 197 |
| 399 | 518 | 621 | 363 |
| 278 | 499 | 426 | 297 |
| 358 | 374 | 414 | 538 |
| 379 | 562 | 332 | 181 |
| 548 | 505 | 460 | 318 |
| 196 | 375 | 494 | 412 |
| 444 | 438 | 637 | 499 |

```r
cost_bumper1 = c(610, 354, 234, 399, 278, 358, 379, 548, 196, 444)
cost_bumper2 = c(404, 663, 521, 518, 499, 374, 562, 505, 375, 438)
cost_bumper3 = c(599, 426, 429, 621, 426, 414, 332, 460, 494, 637)
cost_bumper4 = c(272, 405, 197, 363, 297, 538, 181, 318, 412, 499)
bumper_data = rbind(
    data.frame(cost = cost_bumper1, type = "Bumper 1"),
    data.frame(cost = cost_bumper2, type = "Bumper 2"),
    data.frame(cost = cost_bumper3, type = "Bumper 3"),
    data.frame(cost = cost_bumper4, type = "Bumper 4")
)
# data in long form at
bumper_data
```

```r
library(mosaic)
result = do.call(rbind, lapply(list(cost_bumper1, cost_bumper2,
    cost_bumper3, cost_bumper4), fav_stats))
rownames(result) = paste("Bumper", 1:4)
result
```

|  | min | Q1 | median | Q3 | max | mean | sd | n | missing |
| :--- | :--- | :--- | :--- | :--- | :--- | :--- | :--- | :--- | :--- |
| Bumper 1 | 196 | 297.00 | 368.5 | 432.75 | 610 | 380.0 | 130.09313 | 10 | 0 |
| Bumper 2 | 374 | 412.50 | 502.0 | 520.25 | 663 | 485.9 | 90.53968 | 10 | 0 |
| Bumper 3 | 332 | 426.00 | 444.5 | 572.75 | 637 | 483.8 | 102.10866 | 10 | 0 |
| Bumper 4 | 181 | 278.25 | 340.5 | 410.25 | 538 | 348.2 | 118.52688 | 10 | 0 |

```r
    # Create anova model
    bumper_model = aov(cost ~ type, data = bumper_data)
    # Get ANOVA table
    anova(bumper_model)
```

Bumper Example

Hypotheses

Let:

- $H_0$: $\mu_1 = \mu_2 = \mu_3 = \mu_4$  
- $H_a$: At least one $\mu_i,\ i = 1, \dots, 4$ is different

Test Statistic

$F = 4.0563 \sim F_{(3,36)}$

$p$-value = 0.01395 $<\ 0.05$ (α)

Sufficient evidence to reject $H_0$ and conclude that at least one group of bumpers has a different mean repair cost.


Which group/groups are different?

Perform **pairwise comparisons**  
(i.e., create pairwise confidence intervals)


For this example:

We perform the following comparisons:

1. $\mu_1 - \mu_2$  
2. $\mu_1 - \mu_3$  
3. $\mu_1 - \mu_4$  
4. $\mu_2 - \mu_3$  
5. $\mu_2 - \mu_4$  
6. $\mu_3 - \mu_4$  

We compute $\binom{K}{2}$ comparisons, where $K = 4$ (the number of groups):

$$
\binom{4}{2} = 6
$$


**Pairwise confidence Intervals**

Fisher's LSD

\[
(\bar{x}_i - \bar{x}_j) \pm t_{(n - k,\ \alpha/2)} \cdot \sqrt{MSE \left( \frac{1}{n_1} + \frac{1}{n_2} \right)}
\]


Bonferroni

\[
(\bar{x}_i - \bar{x}_j) \pm t_{(n - k,\ \alpha / (2m))} \cdot \sqrt{MSE \left( \frac{1}{n_1} + \frac{1}{n_2} \right)}
\]

<span style="color:red">\( m = \binom{k}{2} \) &nbsp;&nbsp;&nbsp;&nbsp; # pairwise comparisons</span>



**Note:** _not_ conducting pairwise hypothesis tests



Fisher's is easier to calculate, has more statistical power; however, Bonferroni maintains experiment-wise error rate.

- **Fisher:** \( \alpha_E = 1 - (1 - \alpha)^m \)
- **Bonferroni:** \( \alpha_E = 1 - \left( 1 - \frac{\alpha}{m} \right)^m \)



Fisher

- \( m = 1, \quad \alpha = 0.05 \quad \Rightarrow \quad 1 - (1 - \alpha)^1 = 0.05 \)
- \( m = 2, \quad \alpha = 0.05 \quad \Rightarrow \quad 1 - (1 - \alpha)^2 \geq 0.05 \)
- \( m = 3, \quad \alpha = 0.05 \quad \Rightarrow \quad 1 - (1 - \alpha)^3 \geq 1 - (1 - \alpha)^2 \)
- \( \cdots \)

As \( m \uparrow \), error rate \( \uparrow \)


Bonferroni

As \( m \uparrow \),  
\[
\alpha_E = \left( 1 - \frac{\alpha}{m} \right)^m
\]  
remains _almost constant_


Solution a)

The test statistic is $F_{*}=4.06$ and the $P-$ value $=0.0139$. There is enough statistical evidence to infer that there are differences between some of the bumpers. The question is now, Which bumpers differ?

Fisher's Least Significant Difference Method

The confidence interval estimator is

$$
\left(\bar{x}_{i}-\bar{x}_{j}\right) \pm t_{\alpha / 2} \sqrt{M S E\left(\frac{1}{n_{i}}+\frac{1}{n_{j}}\right)}
$$

Least Significant Difference (definition)

We define the least significant difference LSD as

$$
L S D=t_{\alpha / 2} \sqrt{M S E\left(\frac{1}{n_{i}}+\frac{1}{n_{j}}\right)}
$$

A simple way of determining whether differences exist between each pair of population means is to compare the absolute value of the difference between their two sample means and LSD. In other words, we will conclude that $\mu_{i}$ and $\mu_{j}$ differ if

$$
\left|\bar{x}_{i}-\bar{x}_{j}\right|>L S D
$$

LSD will be the same for all pairs of means if all $k$ sample sizes are equal. If some sample sizes differ, LSD must be calculated for each combination.
It can be argued that this method is flawed because it will increase the probability of committing a Type I error. That is, it is more likely that the analysis of variance to conclude that a difference exists in some of the population means when in fact none differ.

The true probability of making at least one Type I error is called the experimentwise Type I error rate, denoted $\alpha_{E}$. The experimentwise Type I error rate can be calculated as

$$
\alpha_{E}=1-(1-\alpha)^{C}
$$

Here $C$ is the number of pairwise comparisons, which can be calculated by $C=\frac{k(k-1)}{2}$. It can be shown that

$$
\alpha_{E} \leq C \alpha
$$

which means that if we want the probability of making at least one Type I error to be no more than $\alpha_{E}$, we simply specify $\alpha=\frac{\alpha_{E}}{C}$. The resulting procedure is called the **Bonferroni adjustment**.

Solution b)

Let's use our example to illustrate Fisher's LSD method and the Bonferroni adjustment. The four sample means and standard deviations are
$\bar{y}_{1}=380$ and $s_{1}=130.0931$

$\bar{y}_{2}=485.9$ and $s_{2}=90.5396$

$\bar{y}_{3}=483.8$ and $s_{3}=102.1086$

$\bar{y}_{4}=348.2$ and $s_{4}=118.5268$


The pairwise absolute differences are

$$
\begin{aligned}
& \left|\bar{y}_{1}-\bar{y}_{2}\right|=|380-485.9|=105.9 \\
& \left|\bar{y}_{1}-\bar{y}_{3}\right|=|380-483.8|=103.8 \\
& \left|\bar{y}_{1}-\bar{y}_{4}\right|=|380-348.2|=31.8 \\
& \left|\bar{y}_{2}-\bar{y}_{3}\right|=|485.9-483.8|=2.1 \\
& \left|\bar{y}_{2}-\bar{y}_{4}\right|=|485.9-348.2|=137.7 \\
& \left|\bar{y}_{3}-\bar{y}_{4}\right|=|483.8-348.2|=135.6
\end{aligned}
$$


We have that \( MSE = 12,\!399 \) and \( \nu = n - k = 40 - 4 = 36 \).  
If we perform the LSD procedure with the Bonferroni adjustment, the number of pairwise comparisons is 6. We set \( \alpha = 0.05 / 6 = 0.0083 \). Thus
\[
t_{\alpha/2,\, n-k} = t_{0.00415, 36} = 2.7935 \quad \text{(using R)}
\]

```r
qt(0.00415, 36)
## [1] -2.793555
```

\[
LSD = t_{\alpha/2} \sqrt{ MSE \left( \frac{1}{n_i} + \frac{1}{n_j} \right) } 
\approx 2.7935 \cdot \sqrt{ 12399 \left( \frac{1}{10} + \frac{1}{10} \right) } 
= 139.109
\]


Now no pair of means differ because all the absolute values of the differences between sample means are less than 139.1095.
The drawback to the LSD procedure is that we increase the probability of at least one Type I error. The Bonferroni adjustment corrects this problem.

Bumper example

```r
library(mosaic)
result = do.call(rbind, lapply(list(cost_bumper1, cost_bumper2,
    cost_bumper3, cost_bumper4), fav_stats))
rownames(result) = paste("Bumper", 1:4)
result
```

|   |         | min | Q1     | median | Q3     | max | mean  | sd     | n  |
|---|---------|-----|--------|--------|--------|-----|-------|--------|----|
| $\rightarrow$ | Bumper 1 | 196 | 297.00 | 368.5  | 432.75 | 610 | 380.0 | 130.09 | 10 |
| $\rightarrow$ | Bumper 2 | 374 | 412.50 | 502.0  | 520.25 | 663 | 485.9 | 90.54  | 10 |
|   | Bumper 3 | 332 | 426.00 | 444.5  | 572.75 | 637 | 483.8 | 102.11 | 10 |
|   | Bumper 4 | 181 | 278.25 | 340.5  | 410.25 | 538 | 348.2 | 118.53 | 10 |


<div style="text-align: center;">
```{r fig.cap="95% Confidence Intervals for Mean Cost by Bumper Group", echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)

# Create bumper summary manually (or from fav_stats if already available)
bumper_summary <- data.frame(
  Bumper = paste("Bumper", 1:4),
  mean = c(380.0, 485.9, 483.8, 348.2),
  sd = c(130.09313, 90.53968, 102.10866, 118.52688),
  n = c(10, 10, 10, 10)
)

bumper_summary <- bumper_summary %>%
  mutate(
    se = sd / sqrt(n),
    lower = mean - qt(0.975, df = n - 1) * se,
    upper = mean + qt(0.975, df = n - 1) * se
  )

ggplot(bumper_summary, aes(x = Bumper, y = mean)) +
  geom_point(size = 3, color = "#619CFF") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, color = "#619CFF") +
  labs(
    title = "",
    y = "Mean Repair Cost",
    x = "Bumper Group"
  ) +
  theme_minimal()
```
</div>

We want to calculate a 95% pairwise confidence interval for the difference between **Bumper 1** and **Bumper 2** using Fisher's LSD and Bonferroni.

We test: \( \mu_1 - \mu_2 \)

Fisher's LSD Formula:

\[
(\bar{X}_1 - \bar{X}_2) \pm t_{(n - k,\ \alpha/2)} \cdot \sqrt{ MSE \left( \frac{1}{n_1} + \frac{1}{n_2} \right) }
\]

Plug in:

\[
(380 - 485.9) \pm t_{(36,\ 0.025)} \cdot \sqrt{12399 \left( \frac{1}{10} + \frac{1}{10} \right)}
\]

From R:

```r
qt(1 - 0.025, 36)
## [1] 2.028
```

\[
= -105.9 \pm 2.028 \cdot \sqrt{12399 \cdot \left( \frac{1}{10} + \frac{1}{10} \right)}
\]

\[
= -105.9 \pm 100.9942
\]

\[
= (-206.894,\ -4.9058)
\]



Since the entire interval is **negative**, that is:  
\[
\mu_1 - \mu_2 \leq 0
\]

Conclusion: 
Fisher’s LSD suggests \( \mu_2 > \mu_1 \)

Bonferroni Formula:

\[
(\bar{X}_i - \bar{X}_j) \pm t_{(n - k,\ \alpha/(2m))} \cdot \sqrt{ MSE \left( \frac{1}{n_1} + \frac{1}{n_2} \right) }
\]

Plug in:

\[
(380 - 485.9) \pm t_{(36,\ \alpha/2m)} \cdot \sqrt{12399 \left( \frac{1}{10} + \frac{1}{10} \right)}
\]

Where:

\[
\alpha/2m = \frac{0.05}{2 \times 6} = 0.004167
\]

From R:

```r
qt(1 - 0.004167, 36)
## [1] 2.79197
```

\[
= -105.9 \pm 2.79197 \cdot \sqrt{12399 \left( \frac{1}{10} + \frac{1}{10} \right)}
= -105.9 \pm 139.0338
\]

\[
= (-244.9335,\ 33.13347)
\]


Since the interval contains **zero**, that is:

\[
\mu_1 - \mu_2 \in (-244,\ 33)
\]

Conclusion: 
Bonferroni informs us it is **plausible** that \( \mu_1 = \mu_2 \)

<div style="text-align: center;">

```{r fig.cap="Comparison of LSD and Bonferroni Confidence Intervals for \\(\\mu_1 - \\mu_2\\). LSD is narrower; Bonferroni adjusts for multiple comparisons.", echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)

# Values for Bumper 1 vs Bumper 2
diff <- 380.0 - 485.9
MSE <- 12399
n <- 10
SE <- sqrt(MSE * (1/n + 1/n))
df <- 36
m <- 6

t_lsd <- qt(1 - 0.025, df)
t_bonf <- qt(1 - 0.05 / (2 * m), df)

CI_lsd <- c(diff - t_lsd * SE, diff + t_lsd * SE)
CI_bonf <- c(diff - t_bonf * SE, diff + t_bonf * SE)

ci_data <- data.frame(
  Method = c("LSD", "Bonferroni"),
  lower = c(CI_lsd[1], CI_bonf[1]),
  upper = c(CI_lsd[2], CI_bonf[2]),
  center = diff
)

ggplot(ci_data, aes(x = Method, y = center)) +
  geom_point(size = 3, color = "#F8766D") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, color = "#F8766D") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#000000") +
labs(
  title = expression("Confidence Interval for " * mu[1] - mu[2]),
  y = expression("Difference in Means (" * mu[1] - mu[2] * ")"),
  x = "Method"
) +
  theme_minimal(base_size = 12)
```

</div>


Using `DescTools::PostHocTest()` for Fisher's LSD

We perform Fisher’s LSD post-hoc test using the `DescTools` package:

This returns the following multiple comparisons of means:

95% Family-wise Confidence Level – Fisher LSD

| Comparison        | diff   | lwr.ci   | upr.ci   | p-value | 
|------------------|--------|----------|----------|---------|
| Bumper 2 - Bumper 1 | 105.9  | 4.9053   | 206.8947 | 0.0404  | 
| Bumper 3 - Bumper 1 | 103.8  | 2.8053   | 204.7947 | 0.0443  | 
| Bumper 4 - Bumper 1 | -31.8  | -132.7947| 69.1946  | 0.5271  |        
| Bumper 3 - Bumper 2 | -2.1   | -103.0947| 98.8946  | 0.9666  |        
| Bumper 4 - Bumper 2 | -137.7 | -238.6947| -36.7054 | 0.0089  | 
| Bumper 4 - Bumper 3 | -135.6 | -236.5947| -34.6053 | 0.0099  | 

Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1


Using `DescTools::PostHocTest()` for Bonferroni Correction

We perform a Bonferroni-adjusted post-hoc test using the `DescTools` package:

```r
library(DescTools)
PostHocTest(bumper_model, method = "bonferroni")
```

This returns the following multiple comparisons of means:

95% Family-wise Confidence Level – Bonferroni

| Comparison        | diff   | lwr.ci     | upr.ci     | p-value |
|------------------|--------|------------|------------|---------|
| Bumper 2 - Bumper 1 | 105.9  | -33.1341   | 244.9341   | 0.2423  |        
| Bumper 3 - Bumper 1 | 103.8  | -35.2341   | 242.8341   | 0.2657  |        
| Bumper 4 - Bumper 1 | -31.8  | -170.8341  | 107.2341   | 1.0000  |        
| Bumper 3 - Bumper 2 | -2.1   | -141.1341  | 136.9341   | 1.0000  |        
| Bumper 4 - Bumper 2 | -137.7 | -276.7341  | 1.3341     | 0.0535  |        
| Bumper 4 - Bumper 3 | -135.6 | -274.6341  | 3.4341     | 0.0595  | 

Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

**Bonferroni vs. LSD Comparison**

| **Bonferroni** | **LSD** |
|----------------|---------|
| Controls for experiment-wise error | Does **not** control for experiment-wise error |
| \( \alpha_E = 1 - (1 - \frac{\alpha}{m})^m \) | \( \alpha_E = 1 - (1 - \alpha)^m \) |
| Suitable when control of Type I error is required | Suitable when control of Type I error is **not** a strict concern |
| Higher risk of Type II error | Lower risk of Type II error |
| Lower power | Higher power |

Example

An apple juice manufacturer has developed a new product - a liquid concentrate that, when mixed with water, produces 1 liter of apple juice. The product has several attractive features. First, it is more convenient that canned apple juice. Second, because the apple juice that is sold in cans is actually made from concentrate, the quality of the new product is at least as high as canned apple juice. Third, the cost of the new product is slightly lower than canned apple juice. The marketing manager has to decide how to market the new product. She can create advertising that emphasizes convenience, quality, or price.

To facilitate a decision, she conducts an experiment. In three different cities that are similar in size and demographic makeup, she launches the product with advertising stressing the convenience of the liquid concentrate in one city. In the second city, the advertisements emphasize the quality of the product. Advertising that highlights the relatively low cost of the liquid concentrate is used in the third city. The number of packages sold weekly is recorded for the 20 weeks following the beginning of the campaign.

These data are available at:

```r
ad_url =
"https://mcs.utm.utoronto.ca/ ~nosedal/data/ad-strategies.t
```

The marketing manager wants to know if differences in sales exist between the three advertising strategies.

To illustrate Fisher's LSD method and the Bonferroni adjustment, consider the dataset described above and assume we tested to determine whether population means differ using a $5 \%$ significance level. The three sample means are: $577.55,653.0$, and 608.65 .
The pair-wise absolute differences are

$$
\begin{aligned}
& \left|\bar{x}_{1}-\bar{x}_{2}\right|=|577.55-653.0|=|-75.45|=75.45 \\
& \left|\bar{x}_{1}-\bar{x}_{3}\right|=|577.55-608.65|=|-31.10|=31.10 \\
& \left|\bar{x}_{2}-\bar{x}_{3}\right|=|653.0-608.65|=|44.35|=44.35
\end{aligned}
$$

If we conduct the LSD procedure with $\alpha=0.05$ we find $\left|t_{\alpha / 2, n-k}\right|=\left|t_{0.025,57}\right|=|-2.0024655|=2.0024655$

```r
qt(0.025, 57)
## [1] -2.002465
```
$$
\text{LSD} = t_{\alpha / 2} \cdot \sqrt{MSE \left( \frac{1}{n_i} + \frac{1}{n_j} \right)} 
\approx 2.002 \cdot \sqrt{8894 \left( \frac{1}{20} + \frac{1}{20} \right)} = 59.71
$$
We can see that only one pair of sample means differ by more than 59.71.  
That is, \( |\bar{x}_1 - \bar{x}_2| = 75.45 \), and the other two differences are less than LSD.  
We conclude that only \( \mu_1 \) and \( \mu_2 \) differ.

If we perform the LSD procedure with the Bonferroni adjustment,  
the number of pairwise comparisons is 3 (calculated as  \( C = \frac{k(k - 1)}{2} = \frac{3(2)}{2} \)).  
We set \( \alpha = 0.05 / 3 = 0.0167 \). Thus,  
\[
t_{\alpha/2,\, n-k} = t_{0.0083,\, 57} = -2.4682794
\]
and


$$
t_{\alpha / 2} \sqrt{M S E\left(\frac{1}{n_{i}}+\frac{1}{n_{j}}\right)} \approx 2.467 \sqrt{8894\left(\frac{1}{20}+\frac{1}{20}\right)}=73.54
$$

Again we conclude that only $\mu_{1}$ and $\mu_{2}$ differ. Notice, however, in the second calculation LSD is larger.
The drawback to LSD is that we increase the probability of at least one Type I error. The Bonferroni adjustment corrects this problem.

## Exercises {#sec:ch8exercises}

### Part A: Regular Practice Questions

---

<div class="exercise-box">
<div class="exercise-label">Question 1</div>
A professor wants to compare the average final exam scores of students taught by three different teaching methods: online, in-person, and hybrid. Each student is taught using only one method.

Which statistical method should be used?

A. One-sample $t$-test

B. Two-sample $t$-test

C. One-way ANOVA

D. Chi-square test
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- A one-sample $t$-test compares one group mean to a fixed value.
- A two-sample $t$-test compares exactly two group means.
- ANOVA is designed to compare means across **three or more** independent groups.
- A chi-square test is used for categorical data (counts/proportions), not means.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**C. One-way ANOVA.**

There are $k = 3$ groups (online, in-person, hybrid) and the response variable (exam score) is continuous. When comparing means across three or more independent groups, one-way ANOVA is the appropriate method. A two-sample $t$-test can only handle $k = 2$ groups.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 2</div>
A researcher wants to know whether four brands of batteries have the same average lifetime. Let $\mu_1, \mu_2, \mu_3, \mu_4$ be the population mean lifetimes (in hours) for the four brands.

Write the null and alternative hypotheses for a one-way ANOVA test.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- The null hypothesis in ANOVA states that **all** group means are equal.
- The alternative hypothesis states that **at least two** means differ — not that all means differ.
- Do not write $H_a$ as "$\mu_1 \ne \mu_2 \ne \mu_3 \ne \mu_4$"; that would mean all pairs differ.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
$$H_0: \mu_1 = \mu_2 = \mu_3 = \mu_4$$
$$H_a: \text{At least two battery brands have different mean lifetimes.}$$

$H_0$ asserts no difference among all four brands. $H_a$ is non-specific — it only requires one pair to differ, not all pairs.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 3</div>
A company compares the average repair cost for 4 bumper types. There are 10 cars in each group. The cars were **randomly assigned** to bumper types, and each car was tested once.

Which ANOVA assumption is mainly supported by the random assignment?

A. Normality

B. Equal variances

C. Independence

D. Large sample size
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- **Normality**: addressed by checking the distribution of residuals, not by random assignment.
- **Equal variances**: assessed by comparing within-group spreads, not by design.
- **Independence**: observations are independent when each unit is assigned to exactly one group at random, with no repeated measurements.
- **Large sample size**: this is not a formal ANOVA assumption.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**C. Independence.**

Random assignment ensures that each car is independently assigned to one bumper type, with no car influencing another car's result. This directly supports the independence assumption: $\epsilon_{ij} \overset{iid}{\sim} N(0, \sigma^2)$. Normality and equal variances must be assessed separately (e.g., residual plots, Levene's test).
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 4</div>
A study compares average weekly sales under three advertising strategies: convenience, quality, and price. There are 8 weeks of data for each strategy ($n = 24$ total, $k = 3$ groups).

The partial ANOVA table is:

| Source    | df | SS  | MS | F |
|:----------|:--:|:---:|:--:|:-:|
| Treatment |    | 240 |    |   |
| Error     |    | 420 |    |   |
| Total     |    |     |    |   |

Complete the ANOVA table.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- $df_{\text{Treatment}} = k - 1$, $\quad df_{\text{Error}} = n - k$, $\quad df_{\text{Total}} = n - 1$.
- $MS = SS / df$ for each row.
- $F = MS_{\text{Treatment}} / MS_{\text{Error}}$.
- $SS_{\text{Total}} = SS_{\text{Treatment}} + SS_{\text{Error}}$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
With $k = 3$ groups and $n = 24$ total observations:

$$df_{\text{Treatment}} = k - 1 = 2, \quad df_{\text{Error}} = n - k = 21, \quad df_{\text{Total}} = n - 1 = 23$$

$$SS_{\text{Total}} = 240 + 420 = 660$$

$$MS_{\text{Treatment}} = \frac{240}{2} = 120, \quad MS_{\text{Error}} = \frac{420}{21} = 20, \quad F = \frac{120}{20} = 6.00$$

| Source    | df | SS  | MS  | F    |
|:----------|:--:|:---:|:---:|:----:|
| Treatment |  2 | 240 | 120 | 6.00 |
| Error     | 21 | 420 |  20 |      |
| Total     | 23 | 660 |     |      |
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 5</div>
Using the ANOVA table from Question 4, the critical value is $F_{0.05,\,2,\,21} = 3.47$.

At $\alpha = 0.05$, state the conclusion.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- The rejection region for ANOVA is always in the **right tail**: reject $H_0$ if $F^* > F_{\alpha,\,df_T,\,df_E}$.
- Compare the computed $F = 6.00$ to the critical value $F_{0.05,2,21} = 3.47$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
Since $F^* = 6.00 > F_{0.05,\,2,\,21} = 3.47$, we **reject $H_0$**.

There is sufficient evidence at $\alpha = 0.05$ to conclude that not all advertising strategies produce the same mean weekly sales. At least two strategies differ in their average sales.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 6</div>
A one-way ANOVA comparing four tax return forms gives:

$$F^* = 2.94, \quad \text{p-value} = 0.036$$

At $\alpha = 0.05$, what should we conclude?
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Compare the p-value to $\alpha = 0.05$.
- If p-value $< \alpha$: reject $H_0$.
- The p-value is the probability of observing a test statistic at least as extreme as $F^*$, **assuming $H_0$ is true**.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
Since $\text{p-value} = 0.036 < \alpha = 0.05$, we **reject $H_0$**.

There is sufficient evidence at the 5% significance level to conclude that at least two forms have different average completion times.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 7</div>
In Question 6, the ANOVA test rejects $H_0$. Can we immediately conclude which two forms have different mean completion times?
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- ANOVA tests a single global null hypothesis: $\mu_1 = \mu_2 = \cdots = \mu_k$.
- Rejecting $H_0$ tells us *that* differences exist, but not *where* they are.
- Post-hoc methods (pairwise comparisons) are needed to identify specific pairs.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**No.** ANOVA only tells us that **at least two** group means differ — it does not identify which pair(s). To determine which specific groups differ, we need **pairwise comparison** methods such as:

- **Fisher's LSD**: higher power, but does not control the experiment-wise Type I error rate.
- **Bonferroni adjustment**: more conservative; controls the experiment-wise error rate by using $\alpha' = \alpha / m$, where $m = \binom{k}{2}$ is the number of pairwise comparisons.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 8</div>
A small business records sales in three countries over four weeks each:

| Country A | Country B | Country C |
|:---------:|:---------:|:---------:|
|     6     |    10     |    14     |
|    10     |     8     |    13     |
|     7     |    12     |    11     |
|     9     |    10     |    10     |

Find the mean sales for each country and the overall grand mean.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Group mean: $\bar{x}_i = \dfrac{1}{n_i} \sum_{j=1}^{n_i} x_{ij}$.
- Grand mean: $\bar{x} = \dfrac{\text{sum of all observations}}{\text{total number of observations}} = \dfrac{\sum_i \sum_j x_{ij}}{n}$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
$$\bar{x}_A = \frac{6+10+7+9}{4} = \frac{32}{4} = 8$$

$$\bar{x}_B = \frac{10+8+12+10}{4} = \frac{40}{4} = 10$$

$$\bar{x}_C = \frac{14+13+11+10}{4} = \frac{48}{4} = 12$$

Total of all observations: $32 + 40 + 48 = 120$. Total sample size: $n = 12$.

$$\bar{x} = \frac{120}{12} = 10$$
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 9</div>
Using the sales data from Question 8:

**(a)** If the next observation's country is **unknown**, what single value gives the best prediction of sales?

**(b)** If the next observation is **known to be from Country C** and ANOVA has already rejected $H_0$, what value should be used as the prediction?
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Without group information, the best single predictor is the **grand mean** $\bar{x}$.
- Once ANOVA rejects $H_0$ (groups differ) and the group is known, the **group mean** $\bar{x}_i$ is a better predictor than $\bar{x}$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**(a)** Use the grand mean $\bar{x} = \mathbf{10}$.

Without knowing which country, the best predictor is the overall average. This is the prediction under $H_0$ (no group differences).

**(b)** Use Country C's sample mean $\bar{x}_C = \mathbf{12}$.

When ANOVA rejects $H_0$, the group means are significantly different. Knowing the observation comes from Country C, the group-specific mean gives a more accurate prediction than the grand mean.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 10</div>
Using the group means from Question 8 ($\bar{x}_A = 8$, $\bar{x}_B = 10$, $\bar{x}_C = 12$, $\bar{x} = 10$, $n_i = 4$ each), compute the **treatment sum of squares**:

$$SS_{\text{Treatment}} = \sum_{i=1}^{k} n_i\,(\bar{x}_i - \bar{x})^2$$
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- $SS_{\text{Treatment}}$ measures how much the group means vary **around the grand mean**.
- Compute the squared deviation of each group mean from $\bar{x}$, multiply by $n_i$, and sum.
- A larger $SS_{\text{Treatment}}$ means more between-group variation.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
Each group has $n_i = 4$ and $\bar{x} = 10$:

$$SS_{\text{Treatment}} = 4(8-10)^2 + 4(10-10)^2 + 4(12-10)^2$$
$$= 4(4) + 4(0) + 4(4) = 16 + 0 + 16 = \mathbf{32}$$
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 11</div>
For the country sales example, $SS_{\text{Treatment}} = 32$, $SS_{\text{Error}} = 28$, $n = 12$, $k = 3$.

Complete the ANOVA table.

| Source    | df | SS | MS | F |
|:----------|:--:|:--:|:--:|:-:|
| Treatment |    | 32 |    |   |
| Error     |    | 28 |    |   |
| Total     |    |    |    |   |
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- $df_{\text{Treatment}} = k - 1 = 2$, $df_{\text{Error}} = n - k = 9$, $df_{\text{Total}} = n - 1 = 11$.
- $MS_{\text{Treatment}} = 32/2$, $MS_{\text{Error}} = 28/9$.
- $F = MS_{\text{Treatment}} / MS_{\text{Error}}$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
$$df_T = 2,\quad df_E = 9,\quad df_{\text{Total}} = 11$$
$$MS_{\text{Treatment}} = \frac{32}{2} = 16,\quad MS_{\text{Error}} = \frac{28}{9} \approx 3.1111$$
$$F = \frac{16}{3.1111} \approx 5.1429$$

| Source    | df | SS | MS     | F      |
|:----------|:--:|:--:|:------:|:------:|
| Treatment |  2 | 32 | 16     | 5.1429 |
| Error     |  9 | 28 | 3.1111 |        |
| Total     | 11 | 60 |        |        |

From R: `oneway.test(sales ~ country, var.equal = TRUE)` gives $F = 5.1429$, p-value $= 0.0324 < 0.05$.
We **reject $H_0$**: not all countries have the same mean sales.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 12</div>
A consumer organisation compares the difference between advertised and actual margarine amounts (in grams) for three brands. Six packages were sampled per brand:

| Brand 1 | Brand 2 | Brand 3 |
|:-------:|:-------:|:-------:|
|    1    |    2    |    1    |
|    3    |    2    |    2    |
|    3    |    4    |    4    |
|    0    |    3    |    2    |
|    1    |    0    |    3    |
|    0    |    4    |    4    |

The ANOVA output gives $F^* = 1.6864$, p-value $= 0.2185$.

At $\alpha = 0.05$, is there sufficient evidence that the brands differ in their mean shortfall?
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Compare p-value = 0.2185 to $\alpha = 0.05$.
- If p-value $\ge \alpha$: fail to reject $H_0$.
- A large p-value means the observed $F$ is not unusual under $H_0$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
Since p-value $= 0.2185 > \alpha = 0.05$, we **fail to reject $H_0$**.

There is **not** sufficient evidence to conclude that the mean shortfall differs among the three margarine brands. The observed difference in sample means could easily be due to random variation.

```r
b1 <- c(1, 3, 3, 0, 1, 0)
b2 <- c(2, 2, 4, 3, 0, 4)
b3 <- c(1, 2, 4, 2, 3, 4)
differences <- c(b1, b2, b3)
brand <- factor(rep(c("Brand1","Brand2","Brand3"), each = 6))
oneway.test(differences ~ brand, var.equal = TRUE)
# F = 1.6864, p-value = 0.2185
```
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 13</div>
In ANOVA, a **large** $F$ statistic usually means:

A. Within-group variation is much larger than between-group variation

B. Between-group variation is large relative to within-group variation

C. All observations are exactly equal

D. The sample size must be too small
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Recall $F = MS_{\text{Treatment}} / MS_{\text{Error}}$.
- $MS_{\text{Treatment}}$ measures **between-group** variation.
- $MS_{\text{Error}}$ measures **within-group** variation.
- A large $F$ means the numerator (between) is much larger than the denominator (within).
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**B. Between-group variation is large relative to within-group variation.**

$$F = \frac{MS_{\text{Treatment}}}{MS_{\text{Error}}} = \frac{\text{between-group variation}}{\text{within-group variation}}$$

When group means are very different (large $MS_{\text{Treatment}}$) relative to the spread within each group (small $MS_{\text{Error}}$), $F$ is large, providing strong evidence against $H_0$.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 14</div>
The boxplot below compares completion times (minutes) for four tax return forms. Observe the medians and spreads of the four groups.

```{r ch8-q14-box, echo=FALSE, fig.height=3.5, fig.width=6, fig.align='center', message=FALSE, warning=FALSE}
library(ggplot2)
set.seed(258)
times <- c(
  c(45, 58, 72, 38, 61, 53, 79, 46, 65, 55),
  c(82, 99, 74, 110, 88, 102, 79, 95, 107, 84),
  c(98, 148, 75, 162, 109, 143, 62, 155, 105, 131),
  c(78, 93, 106, 70, 117, 86, 99, 91, 113, 85)
)
forms <- rep(c("Form 1","Form 2","Form 3","Form 4"), each = 10)
df_forms <- data.frame(time = times, form = factor(forms))
ggplot(df_forms, aes(x = form, y = time, fill = form)) +
  geom_boxplot(width = 0.45, color = "grey30",
               outlier.shape = 16, outlier.size = 2) +
  scale_fill_manual(values = c("#74b9ff","#55efc4","#fd79a8","#ffeaa7")) +
  labs(x = "Tax Return Form", y = "Completion time (minutes)",
       caption = "n = 10 per form  |  ANOVA: F = 2.94, p-value = 0.036") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.caption = element_text(color = "#555", size = 9))
```

A one-way ANOVA gives $F^* = 2.94$, p-value $= 0.036$. Which conclusion is best at $\alpha = 0.05$?

A. All four forms have exactly the same mean completion time.

B. There is evidence that at least two form means differ.

C. Form 3 must be different from every other form.

D. ANOVA cannot be used with four groups.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Compare p-value $= 0.036$ to $\alpha = 0.05$.
- ANOVA's $H_a$ states "at least two means differ" — it does not pinpoint which pair.
- Option C overclaims: ANOVA cannot identify which specific pairs differ without post-hoc tests.
- Option D is false: ANOVA is designed for $k \ge 2$ groups.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**B. There is evidence that at least two form means differ.**

Since p-value $= 0.036 < \alpha = 0.05$, we reject $H_0$. The boxplot shows Form 3 has a visibly higher median and wider spread, which is consistent with at least one group differing. However, ANOVA alone cannot confirm that Form 3 specifically differs from every other form — post-hoc pairwise comparisons (e.g., Fisher's LSD, Bonferroni) are needed to identify which pairs differ.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 15</div>
In an ANOVA $F$-test, the p-value is calculated as the **right-tail** area of the $F$-distribution. Explain why the right tail is used rather than the left tail or both tails.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- The $F$ statistic is $F = MS_{\text{Treatment}} / MS_{\text{Error}}$.
- Under $H_0$, both numerator and denominator estimate the same $\sigma^2$, so $F \approx 1$.
- Evidence against $H_0$ (group means differ) makes $MS_{\text{Treatment}}$ larger than $MS_{\text{Error}}$, pushing $F$ into the right tail.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
The $F$ statistic measures the ratio of between-group to within-group variation:

$$F = \frac{MS_{\text{Treatment}}}{MS_{\text{Error}}}$$

Under $H_0$ (all group means equal), both $MS_{\text{Treatment}}$ and $MS_{\text{Error}}$ estimate the same population variance $\sigma^2$, so $F$ is close to 1. When $H_0$ is false (group means differ), $MS_{\text{Treatment}}$ becomes inflated relative to $MS_{\text{Error}}$, producing **large** values of $F$.

Therefore, **large** $F$ values constitute evidence against $H_0$, and the rejection region lies entirely in the **right tail** of the $F$-distribution:

$$\text{p-value} = P(F_{k-1,\,n-k} \ge F^*)$$
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 16</div>
A study compares $k = 5$ training programs. The ANOVA test rejects $H_0$. How many pairwise comparisons are needed to compare all possible pairs of programs?
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- The number of pairwise comparisons among $k$ groups is $\dbinom{k}{2} = \dfrac{k(k-1)}{2}$.
- Each pair $(\mu_i, \mu_j)$ with $i < j$ is counted once.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
$$m = \binom{k}{2} = \frac{k(k-1)}{2} = \frac{5 \times 4}{2} = \mathbf{10}$$

There are **10 pairwise comparisons** in total:
$(1,2),\ (1,3),\ (1,4),\ (1,5),\ (2,3),\ (2,4),\ (2,5),\ (3,4),\ (3,5),\ (4,5)$.

This is why the Bonferroni adjustment uses $\alpha' = \alpha / m = 0.05 / 10 = 0.005$ per comparison — to control the experiment-wise Type I error rate.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 17</div>
A researcher compares 4 bumper types. The ANOVA test rejects $H_0$. They want to compare every pair of bumpers using pairwise confidence intervals.

Which method is more conservative: **Fisher's LSD** or the **Bonferroni adjustment**? Briefly explain.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- "More conservative" means it is harder to declare a significant difference (wider intervals, lower Type I error per comparison).
- Fisher's LSD uses $\alpha$ unchanged for each comparison.
- Bonferroni uses $\alpha' = \alpha / m$ (a smaller significance level per comparison, so larger critical value $t$).
- A larger $t$ critical value → wider confidence interval → harder to reject.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**Bonferroni is more conservative.**

For $k = 4$ groups, there are $m = \binom{4}{2} = 6$ pairwise comparisons. Fisher's LSD uses the full $\alpha = 0.05$ for each comparison, while Bonferroni adjusts to $\alpha' = 0.05 / 6 \approx 0.0083$.

The smaller $\alpha'$ requires a larger $t$ critical value, which produces **wider** confidence intervals. A larger LSD means fewer pairs will be declared significantly different. The trade-off is:

| Method | Power | Experiment-wise error control |
|:---|:---|:---|
| Fisher's LSD | Higher | Does **not** control $\alpha_E$ |
| Bonferroni | Lower | Controls $\alpha_E \le \alpha$ |

Fisher's LSD increases the probability of at least one false positive ($\alpha_E = 1 - (1-\alpha)^m$); Bonferroni corrects this at the cost of lower power.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 18</div>
For the bumper repair cost study, the pairwise absolute difference between Bumper 1 and Bumper 2 is $|\bar{x}_1 - \bar{x}_2| = 105.9$.

Using Fisher's LSD method with $MSE = 12{,}399$, $n_i = 10$, and $t_{0.025,\,36} \approx 2.028$:

$$LSD = t_{\alpha/2} \sqrt{MSE \left(\frac{1}{n_i} + \frac{1}{n_j}\right)} \approx 2.028\sqrt{12399 \times 0.2} \approx 101.0$$

Should we conclude that $\mu_1$ and $\mu_2$ differ under Fisher's LSD?
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- The Fisher's LSD decision rule: conclude $\mu_i \ne \mu_j$ if $|\bar{x}_i - \bar{x}_j| > LSD$.
- Compare 105.9 to $LSD \approx 101.0$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
$$LSD = 2.028\sqrt{12399 \times 0.2} = 2.028 \times 49.797 \approx 101.0$$

Since $|\bar{x}_1 - \bar{x}_2| = 105.9 > LSD \approx 101.0$, we **conclude that $\mu_1$ and $\mu_2$ differ** under Fisher's LSD.

The 95% Fisher confidence interval for $\mu_1 - \mu_2$ is:
$$-105.9 \pm 101.0 = (-206.9,\ -4.9)$$

Since the entire interval is **negative**, there is evidence that $\mu_2 > \mu_1$ (Bumper 2 has higher mean repair cost than Bumper 1).
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 19</div>
For the same Bumper 1 vs. Bumper 2 comparison ($|\bar{x}_1 - \bar{x}_2| = 105.9$), now apply the **Bonferroni adjustment**.

With $k = 4$ groups, $m = 6$ pairwise comparisons, $\alpha' = 0.05/6 \approx 0.0083$, and $t_{0.00415,\,36} \approx 2.792$:

$$LSD_{\text{Bonf}} = 2.792\sqrt{12399 \times 0.2} \approx 139.1$$

Should we conclude that $\mu_1$ and $\mu_2$ differ under the Bonferroni adjustment?
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- The Bonferroni decision rule is the same: conclude $\mu_i \ne \mu_j$ if $|\bar{x}_i - \bar{x}_j| > LSD_{\text{Bonf}}$.
- Compare 105.9 to 139.1.
- Note that $LSD_{\text{Bonf}} > LSD_{\text{Fisher}}$, making Bonferroni harder to reject.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
Since $|\bar{x}_1 - \bar{x}_2| = 105.9 < LSD_{\text{Bonf}} \approx 139.1$, we **fail to conclude** that $\mu_1$ and $\mu_2$ differ under the Bonferroni adjustment.

The Bonferroni 95% confidence interval for $\mu_1 - \mu_2$ is:
$$-105.9 \pm 139.1 = (-245.0,\ 33.2)$$

Since the interval **contains zero**, the difference is not statistically significant after adjustment.

This illustrates the trade-off: Fisher's LSD (LSD $\approx$ 101.0) declares this pair significant, while Bonferroni (LSD $\approx$ 139.1) does not. Bonferroni is more conservative but better controls the experiment-wise error rate.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question 20</div>
A one-way ANOVA comparing four bumper types gives p-value $= 0.0139$. At $\alpha = 0.05$, write a conclusion in context.
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Compare p-value = 0.0139 to $\alpha = 0.05$.
- State whether to reject or fail to reject $H_0$.
- Express the conclusion in the context of the problem (bumper repair costs).
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
Since p-value $= 0.0139 < \alpha = 0.05$, we **reject $H_0$**.

There is sufficient evidence at the 5% significance level to conclude that **at least two bumper types have different mean repair costs**. Further pairwise comparisons (Fisher's LSD or Bonferroni) are needed to identify which specific bumpers differ.
</div>
</details>

---

### Part B: WebR Practice Questions

Use the following dataset for Questions B1–B10. Three countries' weekly sales data:

```
salesA <- c(6, 10, 7, 9)
salesB <- c(10, 8, 12, 10)
salesC <- c(14, 13, 11, 10)
```

---

<div class="exercise-box">
<div class="exercise-label">Question B1</div>
Combine the three group vectors into a single response variable `sales`. Fill in the blanks and click **▶ Run**.
<div class="webr-exercise">
<div class="webr-exercise-header">Exercise</div>
<textarea class="webr-editor" rows="6">salesA <- c(6, 10, 7, 9)
salesB <- c(10, 8, 12, 10)
salesC <- c(14, 13, 11, 10)

sales <- c(____, ____, ____)
sales</textarea>
<div class="webr-controls">
  <button class="webr-run-btn" disabled>▶ Run</button>
  <button class="webr-reset-btn">↺ Reset</button>
</div>
<div class="webr-output-area"></div>
<div class="webr-canvas-area"><canvas></canvas></div>
</div>
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- `c(...)` concatenates vectors end-to-end.
- The result should have $4 + 4 + 4 = 12$ elements in order: A values, then B values, then C values.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
```r
salesA <- c(6, 10, 7, 9)
salesB <- c(10, 8, 12, 10)
salesC <- c(14, 13, 11, 10)

sales <- c(salesA, salesB, salesC)
sales
# [1]  6 10  7  9 10  8 12 10 14 13 11 10
```
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question B2</div>
Create the group label variable `country` as a factor, with 4 repetitions of each label. Fill in the blank and click **▶ Run**.
<div class="webr-exercise">
<div class="webr-exercise-header">Exercise</div>
<textarea class="webr-editor" rows="5">country <- factor(c(rep("A", 4), rep("B", 4), rep("C", ____)))
country</textarea>
<div class="webr-controls">
  <button class="webr-run-btn" disabled>▶ Run</button>
  <button class="webr-reset-btn">↺ Reset</button>
</div>
<div class="webr-output-area"></div>
<div class="webr-canvas-area"><canvas></canvas></div>
</div>
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- `rep("C", n)` repeats the string `"C"` $n$ times.
- Country C has 4 observations, just like A and B.
- `factor()` converts the character vector into a categorical variable with levels.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
```r
country <- factor(c(rep("A", 4), rep("B", 4), rep("C", 4)))
country
# [1] A A A A B B B B C C C C
# Levels: A B C
```
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question B3</div>
Use `tapply()` to compute the mean sales for each country. Fill in the blanks and click **▶ Run**.
<div class="webr-exercise">
<div class="webr-exercise-header">Exercise</div>
<textarea class="webr-editor" rows="8">salesA <- c(6, 10, 7, 9)
salesB <- c(10, 8, 12, 10)
salesC <- c(14, 13, 11, 10)
sales   <- c(salesA, salesB, salesC)
country <- factor(c(rep("A", 4), rep("B", 4), rep("C", 4)))

tapply(___, ___, ___)</textarea>
<div class="webr-controls">
  <button class="webr-run-btn" disabled>▶ Run</button>
  <button class="webr-reset-btn">↺ Reset</button>
</div>
<div class="webr-output-area"></div>
<div class="webr-canvas-area"><canvas></canvas></div>
</div>
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- `tapply(X, INDEX, FUN)` applies `FUN` to subsets of `X` defined by `INDEX`.
- Here: `X = sales`, `INDEX = country`, `FUN = mean`.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
```r
tapply(sales, country, mean)
#  A  B  C
#  8 10 12
```
Country A has the lowest mean (8), Country C the highest (12).
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question B4</div>
Compute the **overall sample mean** of all 12 sales observations. Fill in the blank and click **▶ Run**.
<div class="webr-exercise">
<div class="webr-exercise-header">Exercise</div>
<textarea class="webr-editor" rows="7">salesA <- c(6, 10, 7, 9)
salesB <- c(10, 8, 12, 10)
salesC <- c(14, 13, 11, 10)
sales   <- c(salesA, salesB, salesC)

mean(___)</textarea>
<div class="webr-controls">
  <button class="webr-run-btn" disabled>▶ Run</button>
  <button class="webr-reset-btn">↺ Reset</button>
</div>
<div class="webr-output-area"></div>
<div class="webr-canvas-area"><canvas></canvas></div>
</div>
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Apply `mean()` to the full `sales` vector (all 12 values combined).
- This is the grand mean $\bar{x} = \sum_{i,j} x_{ij} / n$.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
```r
mean(sales)
# [1] 10
```
The grand mean is $\bar{x} = 10$, equal to the average of the three group means $(8 + 10 + 12)/3 = 10$ because the group sizes are equal.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question B5</div>
Use `tapply()` to compute the sample **variance** for each country. Fill in the blank and click **▶ Run**.
<div class="webr-exercise">
<div class="webr-exercise-header">Exercise</div>
<textarea class="webr-editor" rows="8">salesA <- c(6, 10, 7, 9)
salesB <- c(10, 8, 12, 10)
salesC <- c(14, 13, 11, 10)
sales   <- c(salesA, salesB, salesC)
country <- factor(c(rep("A", 4), rep("B", 4), rep("C", 4)))

tapply(sales, country, ___)</textarea>
<div class="webr-controls">
  <button class="webr-run-btn" disabled>▶ Run</button>
  <button class="webr-reset-btn">↺ Reset</button>
</div>
<div class="webr-output-area"></div>
<div class="webr-canvas-area"><canvas></canvas></div>
</div>
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Replace `mean` with `var` to compute within-group sample variances.
- The ANOVA equal-variance assumption requires these group variances to be similar.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
```r
tapply(sales, country, var)
#        A        B        C
# 2.666667 2.666667 2.666667
```
All three groups have the same sample variance ($\approx 2.67$), which supports the equal-variance assumption of ANOVA.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question B6</div>
Fit a one-way ANOVA model and display the ANOVA table. Fill in the blanks and click **▶ Run**.
<div class="webr-exercise">
<div class="webr-exercise-header">Exercise</div>
<textarea class="webr-editor" rows="9">salesA <- c(6, 10, 7, 9)
salesB <- c(10, 8, 12, 10)
salesC <- c(14, 13, 11, 10)
sales   <- c(salesA, salesB, salesC)
country <- factor(c(rep("A", 4), rep("B", 4), rep("C", 4)))

model <- aov(____ ~ ____)
anova(model)</textarea>
<div class="webr-controls">
  <button class="webr-run-btn" disabled>▶ Run</button>
  <button class="webr-reset-btn">↺ Reset</button>
</div>
<div class="webr-output-area"></div>
<div class="webr-canvas-area"><canvas></canvas></div>
</div>
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- `aov(response ~ group)` fits the one-way ANOVA model.
- The response variable goes on the left of `~`; the grouping factor goes on the right.
- `anova(model)` prints the ANOVA table with df, SS, MS, F, and p-value.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
```r
model <- aov(sales ~ country)
anova(model)
# Analysis of Variance Table
# Response: sales
#           Df Sum Sq Mean Sq F value  Pr(>F)
# country    2     32  16.000  5.1429 0.03240 *
# Residuals  9     28   3.111
```
$F^* = 5.1429$, p-value $= 0.0324 < 0.05$ → reject $H_0$. Not all country means are equal.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question B7</div>
Run a one-way ANOVA using `oneway.test()` assuming **equal variances**. Fill in the blank and click **▶ Run**.
<div class="webr-exercise">
<div class="webr-exercise-header">Exercise</div>
<textarea class="webr-editor" rows="8">salesA <- c(6, 10, 7, 9)
salesB <- c(10, 8, 12, 10)
salesC <- c(14, 13, 11, 10)
sales   <- c(salesA, salesB, salesC)
country <- factor(c(rep("A", 4), rep("B", 4), rep("C", 4)))

oneway.test(sales ~ country, var.equal = ___)</textarea>
<div class="webr-controls">
  <button class="webr-run-btn" disabled>▶ Run</button>
  <button class="webr-reset-btn">↺ Reset</button>
</div>
<div class="webr-output-area"></div>
<div class="webr-canvas-area"><canvas></canvas></div>
</div>
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Set `var.equal = TRUE` to assume equal variances (classical one-way ANOVA).
- Setting `var.equal = FALSE` gives Welch's ANOVA, which does not assume equal variances.
- Both approaches give the same result when group variances are identical (as here).
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
```r
oneway.test(sales ~ country, var.equal = TRUE)
# One-way analysis of means
# F = 5.1429, num df = 2, denom df = 9, p-value = 0.0324
```
Same result as `aov()`: $F = 5.1429$, p-value $= 0.0324$. Both functions implement the same classical ANOVA when `var.equal = TRUE`.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question B8</div>
The code below has a bug: the response and group variables are **reversed**. Fix the error and click **▶ Run**.
<div class="webr-exercise">
<div class="webr-exercise-header">Exercise</div>
<textarea class="webr-editor" rows="9">salesA <- c(6, 10, 7, 9)
salesB <- c(10, 8, 12, 10)
salesC <- c(14, 13, 11, 10)
sales   <- c(salesA, salesB, salesC)
country <- factor(c(rep("A", 4), rep("B", 4), rep("C", 4)))

# Fix the bug below
model <- aov(country ~ sales)
anova(model)</textarea>
<div class="webr-controls">
  <button class="webr-run-btn" disabled>▶ Run</button>
  <button class="webr-reset-btn">↺ Reset</button>
</div>
<div class="webr-output-area"></div>
<div class="webr-canvas-area"><canvas></canvas></div>
</div>
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- The `aov()` formula is always `response ~ group`.
- `sales` is the **response** (what we measure); `country` is the **group** (what we compare).
- Reversing them tries to predict a categorical variable from a numeric one, which is wrong for ANOVA.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
```r
model <- aov(sales ~ country)   # response ~ group
anova(model)
# Response: sales
#           Df Sum Sq Mean Sq F value Pr(>F)
# country    2     32      16  5.1429 0.0324 *
# Residuals  9     28       3
```
The response variable (`sales`) must appear on the **left** of `~` and the grouping factor (`country`) on the **right**.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question B9</div>
Create a **boxplot** to compare sales across the three countries. Fill in the blanks and click **▶ Run** to see the plot.
<div class="webr-exercise">
<div class="webr-exercise-header">Exercise</div>
<textarea class="webr-editor" rows="11">salesA <- c(6, 10, 7, 9)
salesB <- c(10, 8, 12, 10)
salesC <- c(14, 13, 11, 10)
sales   <- c(salesA, salesB, salesC)
country <- factor(c(rep("A", 4), rep("B", 4), rep("C", 4)))

boxplot(____ ~ ____,
        main = "Sales by Country",
        xlab = "Country",
        ylab = "Sales")</textarea>
<div class="webr-controls">
  <button class="webr-run-btn" disabled>▶ Run</button>
  <button class="webr-reset-btn">↺ Reset</button>
</div>
<div class="webr-output-area"></div>
<div class="webr-canvas-area"><canvas></canvas></div>
</div>
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- The `boxplot()` formula syntax is `boxplot(response ~ group, ...)`.
- Fill in `sales ~ country` to plot sales (y-axis) grouped by country (x-axis).
- The plot should show Country C with a higher median than A and B.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
```r
boxplot(sales ~ country,
        main = "Sales by Country",
        xlab = "Country",
        ylab = "Sales")
```
The boxplot shows Country C with the highest median ($\tilde{x}_C = 12$) and Country A with the lowest ($\tilde{x}_A = 7.5$). The medians increase from A to C, consistent with the ANOVA result that group means differ.
</div>
</details>

---

<div class="exercise-box">
<div class="exercise-label">Question B10</div>
Run the ANOVA below and read the output. Then select the correct conclusion at $\alpha = 0.05$.

A. Do not reject $H_0$; the country means may all be equal.

B. Reject $H_0$; not all country means are equal.

C. Reject $H_0$; all country means are different from each other.

D. ANOVA cannot be used here because the sample size is too small.
<div class="webr-exercise">
<div class="webr-exercise-header">Exercise</div>
<textarea class="webr-editor" rows="8">salesA <- c(6, 10, 7, 9)
salesB <- c(10, 8, 12, 10)
salesC <- c(14, 13, 11, 10)
sales   <- c(salesA, salesB, salesC)
country <- factor(c(rep("A", 4), rep("B", 4), rep("C", 4)))

oneway.test(sales ~ country, var.equal = TRUE)</textarea>
<div class="webr-controls">
  <button class="webr-run-btn" disabled>▶ Run</button>
  <button class="webr-reset-btn">↺ Reset</button>
</div>
<div class="webr-output-area"></div>
<div class="webr-canvas-area"><canvas></canvas></div>
</div>
</div>

<details data-type="hint">
<summary>💡 Hint</summary>
<div class="content">
- Read the p-value from the output and compare it to $\alpha = 0.05$.
- $H_a$ in ANOVA says "at least two means differ" — **not** "all means are different."
- Option C overclaims: ANOVA cannot tell which specific pairs differ.
- Option D is wrong: ANOVA works with small samples as long as the normality assumption holds approximately.
</div>
</details>

<details data-type="answer">
<summary>✅ Answer</summary>
<div class="content">
**B. Reject $H_0$; not all country means are equal.**

The output gives $F = 5.1429$, p-value $= 0.0324$.

Since p-value $= 0.0324 < \alpha = 0.05$, we **reject $H_0$**. There is sufficient evidence that at least two countries have different mean sales. Option C overclaims — ANOVA only tells us that *some* means differ, not that *all* pairs differ. Post-hoc pairwise comparisons would be needed to identify which specific pairs are different.
</div>
</details>
