# Simple Linear Regression

## Linear Model

In statistics, simple linear regression (SLR) is a linear regression
model with a single explanatory variable. In other words, we use linear
functions to illustrate the relationship of variables (ie. time and
one's height). The goal of simple linear regression is to find the
best-fitting straight line, known as the regression line, that predicts
the dependent variable based on the independent variable. 

### Measures of Linear Relationship

We begin be introducing a quantitaive measurement to
measure the linear relationship between variavles

**Covariance (Sample Covariance)**

In probability theory and statistics, covariance is a measure of the
joint variability of two random variables. The covariance sign shows the
direction of the linear relationship between two variables. If higher
values of one variable tend to occur with higher values of the other
(and lower with lower), the covariance is positive, meaning the
variables move in the same direction. If higher values of one variable
tend to occur with lower values of the other, the covariance is
negative, meaning they move in opposite directions. The size (magnitude)
of the covariance reflects how much the two variables vary together,
based on the variances they share.

::: definition
The sample covariance for 2 variables $X$ and $Y$ is:
$$S_{xy} = \frac{1}{n-1} \cdot \sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y}) = \frac{\sum_{i = 1}^{n}x_i \cdot y_i}{n -1} - \frac{n\bar{x}\bar{y}}{n-1}.$$
These are the two ways to compute covariance. Both will give you the
same answer.
:::

The covariance is a measures of how two variables move together.

- If covariance of two random variables is greater than $0$,
  ($cov(x,y) > 0$), then the two random variables show the same trend.
  That is: if one random variable is increasing, then the other one is
  also increasing; while if one random variable is decreasing, then the
  other one is also decreasing.

- If covariance of two random variables is less than $0$,
  ($cov(x,y) < 0$), then the two random variables show the opposite
  trend. That is: if one random variable is increasing, then the other
  one is decreasing; while if one random variable is decreasing, then
  the other one is increasing.

- If covariance of two random variables is equal to $0$,
  ($cov(x,y) = 0$), then we say that there is no relationship
  (systematically linear) between the two random variables.

Note that covariance is not standardized, so it can be difficult to
interpret directly.

**Coefficient of Correlation**

In statistics, correlation or dependence is any statistical
relationship, whether causal or not, between two random variables or
bivariate data. It helps us understand whether and how changes in one
variable are associated with changes in another. A positive correlation
means that as one variable increases, the other tends to increase as
well, while a negative correlation means that one variable tends to
decrease as the other increases. The degree of correlation is usually
expressed with a correlation coefficient, which ranges from $-1$ to
$+1$.

::: definition
The coefficient of correlation is given by:
$$r_{xy} = \frac{S_{xy}}{S_x \cdot S_y}.$$ Now, $r_{xy} =$ sample
correlation coefficient, $S_{xy} =$ sample covariance, $S_{x} =$ sample
standard deviation of $x$, $S_{y} =$ sample standard deviation of $y$.
Also, remember that the range of coefficient of correlation is between
$-1$ and $+1$: $r_{xy} \in [-1, +1]$.
:::

The correlation r measures the strength and direction of the linear
association between two quantitative variables $x$ and $y$. Although you
calculate a correlation for any scatter plot, $r$ measures only
straight-line relationships. In short, coefficient of correlation is a
measure of the strength of the linear relationship between two random
variables.

- If $r_{xy} \approx +1$, then we say that the two random variables have
  a strong positive correlation (See Figure \@ref(fig:examplePositiveCorr)).

```{r examplePositiveCorr,fig.cap="An illustration of simple linear regression model (with positive correlation between total sleep duration and memory score).", echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
# Load necessary library
library(plotly)

# -----------------------------
# Step 1: Simulate example data
# -----------------------------
set.seed(123)

# Simulate 30 observations of sleep duration (hours)
sleep <- seq(4, 10, length.out = 30)

# Simulate memory scores with noise (linearly dependent on sleep)
memory <- 50 + 5 * sleep + rnorm(30, mean = 0, sd = 5)

# Create a data frame
data <- data.frame(sleep, memory)

# -----------------------------
# Step 2: Fit simple linear model
# -----------------------------
model <- lm(memory ~ sleep, data = data)

# Add predicted (fitted) values to data frame
data$fit <- predict(model)

# -----------------------------
# Step 3: Plot using Plotly
# -----------------------------
fig <- plot_ly(data, x = ~sleep, y = ~memory, type = 'scatter', mode = 'markers',
               marker = list(color = 'rgba(0, 100, 255, 0.7)', size = 8),
               name = 'Data points') %>%
  # add_lines(y = ~fit, mode = 'lines',
  #           line = list(color = '#F8766D', width = 2),
  #           name = 'SLR Line') %>%
  layout(title = "Example of Positive Correlation",
         xaxis = list(title = "X"),
         yaxis = list(title = "Y"),
         legend = list(x = 0.05, y = 0.95))

# Display the plot
fig
```
  


- If $r_{xy} \approx -1$, then we say that the two random variables have
  a strong negative correlation. (See figure \@ref(fig:exampleNegaticeCorr))


```{r exampleNegaticeCorr, fig.cap="An illustration of strong negative correlation ($r_{xy} \\approx -1$).", message=FALSE, warning=FALSE, echo=FALSE, fig.align='center'}
# Load required library
library(plotly)

# -----------------------------
# Step 1: Simulate data
# -----------------------------
set.seed(123)

# Simulate speed (km/h), ranging from 40 to 100
speed <- seq(40, 100, length.out = 30)

# Assume travel time = distance / speed + some noise
distance <- 100  # km
travel_time <- distance / speed + rnorm(30, mean = 0, sd = 0.2)

# Create data frame
data <- data.frame(speed, travel_time)

# -----------------------------
# Step 2: Fit linear regression
# -----------------------------
model <- lm(travel_time ~ speed, data = data)
data$fit <- predict(model)

# -----------------------------
# Step 3: Plot using Plotly
# -----------------------------
fig <- plot_ly(data, x = ~speed, y = ~travel_time, type = 'scatter', mode = 'markers',
               marker = list(color = '#619CFF', size = 8),  # Default Plotly blue
name = 'Data Points') %>%
  # add_trace(y = ~fit, mode = 'lines',
  #           line = list(color = '#F8766D', width = 2),
  #           name = 'SLR Line') %>%
  layout(title = "Example of Negative Correlation",
         xaxis = list(title = "X"),
         yaxis = list(title = "Y"),
         legend = list(x = 0.05, y = 0.95))

# Display plot
fig
```


- If $r_{xy} \approx 0$, then we say that there is essentially no
  correlation between the two random variables. Note that if
  $r \approx 0$, then it suggests a linear relationship doesn't exist
  but other relationship may exist (See Figure \@ref(fig:exampleNoCorr)).
  
 
```{r exampleNoCorr, fig.cap="An illustration of no correlation ($r_{xy} \\approx 0$).", message=FALSE, warning=FALSE, echo=FALSE, fig.align='center'}
# Load library
library(plotly)

# -----------------------------
# Step 1: Simulate no-correlation data
# -----------------------------
set.seed(123)

# Simulate coffee consumption (0 to 8 cups/day)
coffee <- seq(0, 8, length.out = 30)

# Simulate intelligence scores randomly, independent of coffee
intelligence <- rnorm(30, mean = 100, sd = 10)  # No trend with coffee

# Create data frame
data <- data.frame(coffee, intelligence)

# -----------------------------
# Step 2: Plot only data points (no line)
# -----------------------------
fig <- plot_ly(data, x = ~coffee, y = ~intelligence, type = 'scatter', mode = 'markers',
               marker = list(color = 'rgba(0, 100, 255, 0.7)', size = 8),
               name = 'Data points') %>%
  layout(title = "Example of No Correlation",
         xaxis = list(title = "X"),
         yaxis = list(title = "Y"),
         legend = list(x = 0.05, y = 0.95))

# Show plot
fig
```

The interactive plot below allows us to examine the spatial arrangement of data and the resulting correlation.

<center>
```{r CorrShinyApp, echo=FALSE, warning=FALSE, results='asis', out.width='100%'}
knitr::include_app("https://bryan-su.shinyapps.io/desktop/", height = "735")
```
</center>

Note that correlation doesn't imply causation:

- $cor(x,y) \approx +1$ doesn't necessarily imply on increase in $x$
  causes increase in $y$.

- $cor(x,y) \approx -1$ doesn't necessarily imply on increase in $x$
  causes decrease in $y$.

**Properties of Covariance and Correlation**

These two values are symmetric:

- $cov(x,y) = cov(y,x)$;

- $cor(x,y) = cor(y,x)$.

::::::::: example
Five observations taken for two variables follow.

::: {.smalltbl}
| \(x_i\) | \(y_i\) |
|:-------:|:-------:|
|    4    |   50    |
|    6    |   50    |
|   11    |   40    |
|    3    |   60    |
|   16    |   30    |
:::

Compute the sample covariance.

Compute and interpret the sample correlation coefficient.

**Solution:**

Step 1: Compute $\bar{x}$ and $\bar{y}$; $\bar{x} = 8$ and
$\bar{y} = 46$ (check this by yourself).

Step 2: Find $s_x$ and $s_y$.

$$s_x^2 = \frac{1}{5-1} \cdot \sum_{i = 1}^{5}(x_i - \bar{x})^2 = 29.5 \text{ and } s_y^2 = \frac{1}{5-1} \cdot \sum_{i=1}^{5}(y_i - \bar{y})^2 = 130$$
Then: $s_x = 5.4313$ and $s_y = 11.4017$.

Step 3: Find $s_{xy}$ and $r$.

$$\sum_{i=1}^{5}x_i \cdot y_i = 1600, \text{ then } s_{xy} = \frac{1600}{5-1} - \frac{5\cdot8\cdot46}{5-1} = -60 \text{ and } r_{xy} = \frac{s_{xy}}{s_x \cdot s_y} = 0.9688.$$

**R code**

Step 1: Entering data;

::: tcolorbox
    X=c(4,6,11,3,16); 

    Y=c(50,50,40,60,30);
:::

Step 2: Finding means;

::: tcolorbox
    mean(X);

    mean(Y);
:::

Step 3: Finding variances;

::: tcolorbox
    var(X);

    var(Y);
:::

Step 4: Finding standard deviations;

::: tcolorbox
    sd(X);

    sd(Y):
:::

Step 5: Finding covariance and correlation;

::: tcolorbox
    cov(X,Y);

    cor(X,Y);
:::
:::::::::

### Least Squares Method

\noindent The method of least squares is a foundational optimization technique in regression analysis.  Its goal is to determine the parameters of a candidate model that make the model’s predictions as close as possible to the observed data.  Concretely, in the linear setting, we posit that each observation $(x_{i1}, \dots, x_{ip}, y_i)$ satisfies
\[
    y_i \;=\; \beta_0 + \beta_1 x_{i1} + \cdots + \beta_p x_{ip} + \varepsilon_i,
    \quad \varepsilon_i \sim N(0, \sigma^2),
\]
where the predictors $x_{ij}$ are assumed fixed and the error terms $\varepsilon_i$ are independent Gaussian noise.

\bigskip

\noindent Given a sample of $n$ observations, the least‑squares estimates of the coefficients $\beta_0,\dots,\beta_p$ are those values $\hat\beta_0,\dots,\hat\beta_p$ which minimize the sum of square errors (SSE),
\[
    \mathrm{SSE}
    \;=\;
    \sum_{i=1}^n \bigl(y_i - \hat y_i\bigr)^2
    \;=\;
    \sum_{i=1}^n \bigl(y_i - (\beta_0 + \beta_1 x_{i1} + \cdots + \beta_p x_{ip})\bigr)^2.
\]
The resulting fitted values take the form
\[
    \hat y_i
    \;=\;
    \hat\beta_0 + \hat\beta_1\,x_{i1} + \cdots + \hat\beta_p\,x_{ip}.
\]
Here, $\hat\beta_0$ serves as the intercept, while each $\hat\beta_j$ for $j=1,\dots,p$ quantifies the change in the expected response for a one‑unit increase in the corresponding predictor.

\bigskip

\noindent In this course, we will concentrate on the simplest nontrivial case of a single predictor, writing
\[
    \hat y \;=\; b_0 + b_1\,x,
\]
where $b_0$ and $b_1$ denote the intercept and slope of the fitted line, respectively.  By taking partial derivatives of the residual sum of squares with respect to $b_0$ and $b_1$ and setting them to zero, one obtains the classical normal equations.  Their solution yields the familiar closed‑form formulas
\[
    b_1 \;=\; r\,\frac{s_y}{s_x},
    \qquad
    b_0 \;=\; \bar y \;-\; b_1\,\bar x,
\]
where $r$ is the sample correlation between $x$ and $y$, and $s_x$, $s_y$ denote their sample standard deviations.


::: {.remark}
- The distinction between explanatory and response variables is
  essential in Least Squares Method.

- The least-squares line (trendline) always passes through the point
  ($\bar{x}$, $\bar{y}$) on the graph of $y$ against $x$.

- The square of the correlation, $r^2$, is the fraction of the variation
  in the values of $y$ that is explained by the variation in $x$.
:::

:::: example
A tool die maker operates out of a small shop making specialized tools.
He is considering increasing the size of his business and needs to know
more about his costs. One such cost is electricity, which he needs to
operate his machines and lights. He keeps track of his daily electricity
costs and the number of tools that he made that day. These data are
listed next. Determine the fixed and variable electricity costs using
the Least Squares Method.

::: {.smalltl}
| Day | Number of tools (X) | Electricity costs (Y) |
|:---:|:-------------------:|:---------------------:|
|  1  |          7          |         23.80         |
|  2  |          3          |         11.89         |
|  3  |          2          |         15.89         |
|  4  |          5          |         26.11         |
|  5  |          8          |         31.79         |
|  6  |         11          |         39.93         |
|  7  |          5          |         12.27         |
|  8  |         15          |         40.06         |
|  9  |          3          |         21.38         |
| 10  |          6          |         18.65         |
:::

Solution:

Step 1: Entering Data;

::: tcolorbox
    tools=c(7,3,2,5,8,11,5,15,3,6);

    cost=c(23.80,11.89,15.98,26.11,31.79, 39.93,12.27,40.06,21.38,18.65);
:::

Step 2: Finding Slope;

::: tcolorbox
    Sx=sd(tools); 

    Sy=sd(cost); 

    r=cor(tools,cost); 

    b1=r*(Sy/Sx); 

    b1;
     
    ## [1] 2.245882
:::

Step 3: Finding $y$-intercept;

::: tcolorbox
    x.bar=mean(tools); 

    y.bar=mean(cost);

    b0=y.bar - b1*x.bar;

    b0; 

    ## [1] 9.587765
:::

We can also use R-code to draw a graph:

::: tcolorbox
    plot(tools,cost,pch=19);

    abline(least.squares$coeff,col="red");

    # pch=19 tells R to draw solid circles; 

    # abline tells R to add trendline;
:::

Interpretation:

The slope measures the marginal rate of change in the dependent
variable. In this example, the slope is $2.25$, which means that in this
sample, for each one-unit increase in the number of tools, the marginal
increase in the electricity cost is $\$ 2.25$ per tool.

The $y$-intercept is $9.57$; that is, the line strikes the $y$-axis at
$9.57$. However, when $x = 0$, we are producing no tools and hence the
estimated ﬁxed cost of electricity is $\$9.57$ per day .
::::

### Estimating Regression Model Parameters

The regression line which we are going to use is:
$E(Y) = \beta_0 + \beta_1 \cdot x$. This is fitted to the data points
$(x_1, y_1), (x_2, y_2), \ldots, (x_n, y_n)$ by finding the line that is
closest to the data in some sense. There are many ways in which
closeness can be defined, but the method most generally used is to
consider the vertical deviations between the line and the data points:
$y_i - (\beta_0 + \beta_1 \cdot x_i), 1 \leq i \leq n$.

The fitted line is chosen to be the line that minimizes the sum of the
squares of these vertical deviations
$$Q = \sum_{i=1}^{n} [y_i - (\beta_0 + \beta_1 \cdot x_i)]^2$$ and this
is referred to as the least squares fit. (The quantity $Q$ is also
called the **sum of squares for error**, SSE.)

The parameter estimates $\hat{\beta_0}$ and $\hat{\beta_1}$ are
therefore the values that minimize the quantity $Q$. They are found
taking partial derivatives of $Q$ with respect to $\hat{\beta_0}$ and
$\hat{\beta_1}$ and setting the resulting expressions equal to $0$.

```{r fig.cap="An illustration of a simple linear regression model. The red line is the fitted regression line, and the full black vertical lines represent the residuals (SSE). The data points deviate from the line to show errors clearly. Note that the sum of residuals is necessarily $0$.", message=FALSE, warning=FALSE,echo=FALSE, fig.align='center'}
library(ggplot2)

# Data points
x_vals <- c(0.5, 1.5, 1, 2, 2.5, 3, 3.5, 4, 4.5, 5)
y_vals <- c(1.0, 1.4, 1.5, 2.4, 2.1, 2.7, 3.3, 3.4, 4.2, 3.7)
df <- data.frame(x = x_vals, y = y_vals)

# Regression line: y_hat = 0.8x + 0.35
df$y_hat <- 0.8 * df$x + 0.35

# Residual segments
residual_segments <- data.frame(
  x = df$x,
  xend = df$x,
  y = df$y,
  yend = df$y_hat
)

# Generate regression line for plotting
line_df <- data.frame(x = seq(0, 6, length.out = 100))
line_df$y <- 0.8 * line_df$x + 0.35

# Plot
# Plot
ggplot(df, aes(x = x, y = y)) +
  geom_point(color = "#619CFF", size = 2) +  # Blue points

  geom_line(data = line_df, aes(x = x, y = y), color = "#F8766D", size = 1) +  # Red regression line

  geom_segment(data = residual_segments,
               aes(x = x, xend = xend, y = y, yend = yend),
               color = "#000", linetype = "solid") +  # Residual lines

  annotate("text", x = 4.5, y = 5.4,
           label = expression(hat(y) == hat(beta)[0] + hat(beta)[1] * x),
           color = "#F8766D", size = 5, hjust = 0) +  # Equation label

  labs(x = expression(x), y = expression(y)) +
  theme_minimal()
```



The following proof is the derive of $\hat{\beta_0}$ and $\hat{\beta_1}$
for simple linear regression model:
$\hat{y} = \hat{\beta_0} + \hat{\beta_1} \cdot x + \varepsilon$.

Let $x$ be independent variable and $y$ be dependent variable, then the
model of simple linear regression is:
$\hat{y} = \hat{\beta_0} + \hat{\beta_1} \cdot x + \varepsilon$, where
$\hat{\beta_0}$ represents the $y$-intercept, $\hat{\beta_1}$ represents
the slope and $\varepsilon$ is the error term that
$\varepsilon \sim N(0, \sigma^2)$. 

Moreover,
$$\hat{\beta_0} = \bar{y} - \hat{\beta_1}\cdot \bar{x}$$ 
and
$$\hat{\beta_1} = \frac{S_{xy}}{S_{xx}} = \frac{\displaystyle\sum_{i=1}^{n}x_iy_i - n\bar{x}\bar{y}}{\displaystyle\sum_{i=1}^{n}x_i^2 - n\bar{x}^2}$$.


::::: example
Suppose an appliance store conducts a 5-month experiment to determine
the effect of advertising on sales revenue. The results are shown in a
table below. The relationship between sales revenue, $y$, and
advertising expenditure, $x$, is hypothesized to follow a ﬁrst-order
linear model, that is, $y = \beta_0 + \beta_1\cdot x + \varepsilon$, where
$y =$ dependent variable, $x =$ independent variable, $\beta_0$
y-intercept, $\beta_1 =$ slope of the line and $\varepsilon =$ error
variable.

::: {.smalltbl}
| Day | Number of tools (X) | Electricity costs (Y) |
|:---:|:-------------------:|:---------------------:|
|  1  |          7          |         23.80         |
|  2  |          3          |         11.89         |
|  3  |          2          |         15.89         |
|  4  |          5          |         26.11         |
|  5  |          8          |         31.79         |
|  6  |         11          |         39.93         |
|  7  |          5          |         12.27         |
|  8  |         15          |         40.06         |
|  9  |          3          |         21.38         |
| 10  |          6          |         18.65         |
:::

a\) Obtain the least squares estimates of $\beta_0$ and $\beta_1$, and
state the estimated regression function.

b\) Plot the estimated regression function and the data.

**Solution:**

a\) $\bar{x} = 3$, $\bar{y} = 2$, $S_{xx} = 10$, $S_{xy} = 7$

Then, the slope of the least squares line is
$\hat{\beta_1} = \frac{S_{xy}}{S_{xx}} = 0.7$ and
$\hat{\beta_0} = \bar{y} - \hat{\beta_1}\bar{x} = -0.1$. Thus, the least
squares line is $\hat{y} = -0.1 + 0.7x$.

b\) R-code

::: tcolorbox
    plot(x, y, main="Scatterplot: Simple Linear Regression",
    xlab="x", ylab="y", pch=19,col="blue");

    abline(coef(linear.reg), col="red",lty=2);
:::
:::::

### SST, SSE and SSR

Early in Section $15.3$, we introduced a value called the sum of
residual squared (SSE). There are two more values that are important in
simple linear regression, which are total sum of squares (SST) and sum
of squares for regression (SSR). We will introduce all these three
values with figures, so that you may have a better understanding of what
they measure.

**SST (Total Sum of Squares)**

It is defined as the sum over all squared differences between the
observations and their overall mean $\bar{y}$.



```{r fig.cap= "An illustration of Total Sum of Squares (SST). The blue points represent height measurements over time, the dashed line is the mean height $\\bar{y}$, and the solid black vertical lines represent the squared deviations from the mean (SST components).", echo=FALSE, out.width='90%',fig.align='center'}
library(ggplot2)

# Observed data
df <- data.frame(
  x = c(0.5, 1.5, 1, 2, 2.5, 3, 3.5, 4, 4.5, 5),
  y = c(1.3, 2.1, 1.6, 2.7, 2.2, 2.5, 3.7, 3.9, 3.4, 4.2)
)

# Mean line value
y_bar <- mean(df$y)

# Regression line points
reg_line <- data.frame(x = seq(0, 6, length.out = 100))
reg_line$y <- 0.8 * reg_line$x + 0.35

# Vertical segments from each point to mean
df$y_bar <- y_bar

# Plot
ggplot(df, aes(x = x, y = y)) +
  geom_point(color = "#619CFF", size = 2) +  # Blue points
  geom_line(data = reg_line, aes(x = x, y = y),
            color = "#F8766D", size = 1) +  # Red regression line
  geom_hline(yintercept = y_bar, linetype = "dashed", color = "#000000") +  # Mean line
  geom_segment(aes(x = x, xend = x, y = y, yend = y_bar),
               color = "#000000") +  # SST verticals

  # Mean label (bar(y))
  annotate("text", x = 6.3, y = y_bar, label = "bar(y)",
           hjust = 0, vjust = -0.5, parse = TRUE) +

  # Regression equation label: hat(y) == hat(beta)[0] + hat(beta)[1]*x
  annotate("text", x = 4.5, y = 5.4,
           label = "hat(y)==hat(beta)[0]+hat(beta)[1]*x",
           parse = TRUE, color = "#F8766D", size = 5, hjust = 0) +

  labs(x = expression(x), y = expression(y)) +
  theme_minimal()
```


 



::: definition
For any simple linear regression model, SST (Toal sum of squares)
measures the sum over all squared differences between the observations
and their overall mean $\bar{y}$ is given by:
$$SST = \sum_{i=1}^{n}(y_i - \bar{y})^2.$$
:::

**SSE (Sum of Residual Squared)**

It is the sum of the squares of residuals (deviations predicted from
actual empirical values of data). It is a measure of the discrepancy
between the data and an estimation model, such as a linear regression. A
small SSE indicates a tight fit of the model to the data.


```{r fig.cap= "A simple linear regression illustration with residuals shown. Blue points are observed data, red line is the model, dashed line is the average of $y$, and black lines represent residuals $y_i - \\hat{y}_i$.", echo=FALSE, fig.align='center', out.width='90%'}
library(ggplot2)

# Data points from TikZ
df <- data.frame(
  x = c(0.5, 1.5, 1, 2, 2.5, 3, 3.5, 4, 4.5, 5),
  y = c(1.4, 1.1, 2.1, 1.9, 3.1, 2.2, 3.8, 3.4, 4.5, 3.6)
)

# Regression line: y_hat = 0.8x + 0.35
df$y_hat <- 0.8 * df$x + 0.35

# Residual segments
residual_df <- data.frame(
  x = df$x,
  xend = df$x,
  y = df$y,
  yend = df$y_hat
)

# Regression line curve
reg_line_df <- data.frame(x = seq(0, 6, length.out = 100))
reg_line_df$y <- 0.8 * reg_line_df$x + 0.35

# Mean line
y_bar <- 2.7

# Plot
ggplot(df, aes(x = x, y = y)) +
  geom_point(color = "#619CFF", size = 2) +  # Blue points
  geom_line(data = reg_line_df, aes(x = x, y = y), color = "#F8766D", size = 1) +  # Regression line
  geom_segment(data = residual_df,
               aes(x = x, xend = xend, y = y, yend = yend),
               color = "#000000") +  # Residuals
  geom_hline(yintercept = y_bar, linetype = "dashed", color = "#000000") +  # Mean line
  annotate("text", x = 6.7, y = y_bar, label = "bar(y)", parse = TRUE, hjust = 0) +  # Mean label
  annotate("text", x = 4.5, y = 5.0,
           label = "hat(y)==hat(beta)[0]+hat(beta)[1]*x", parse = TRUE,
           color = "#F8766D", size = 5, hjust = 0) +  # Regression equation
  labs(x = expression(x), y = expression(y)) +
  theme_minimal()
```




::: definition
For any simple linear regression model, SSE (Sum of residual squared)
measures the distance between observed data and estimated data, which is
given by: $$SSE = \sum_{i=1}^{n} (y_i - \hat{y_i})^2.$$
:::

Note that SSE (sum of residual squared) is an explained variation.

**SSR (Sum Square Regression)**

It measures the distance between estimated value (estimated dependent
data) and the mean of dependent data ($\bar{y}$).


```{r fig.cap= "An illustration of simple linear regression. The blue points represent monthly height measurements with added variability. The red line is the fitted simple linear regression model. The dashed line shows the mean of the dependent variable, $\\bar{y}$, and the solid black lines illustrate the deviation of the model's predictions from this mean.", echo=FALSE, out.width='90%',fig.align='center', message=FALSE, warning=FALSE, results='hide'}
library(ggplot2)
library(dplyr)

# Create data with both over- and under-predicted points
df <- data.frame(
  x = c(0.5, 1.5, 1, 2, 2.5, 3, 3.5, 4, 4.5, 5),
  y = c(1.2, 2.1, 1.6, 2.5, 2.9, 3.3, 3.7, 4.2, 4.4, 4.7)
)

# Linear model
model <- lm(y ~ x, data = df)
df$y_hat <- predict(model)
# Ensure y_bar is numeric
y_bar <- mean(df$y)

ggplot(df, aes(x = x, y = y)) +
  geom_point(color = "#619CFF", size = 3) +  # Blue points
  geom_abline(intercept = coef(model)[1], slope = coef(model)[2],
              color = "#F8766D", size = 1) +  # Red regression line
  geom_segment(aes(x = x, xend = x, y = y_hat, yend = y_bar),
               color = "#000000") +  # SSR residuals
  geom_hline(yintercept = y_bar, linetype = "dashed", color = "#000000") +  # Mean line
  annotate("text", x = 5.5, y = y_bar,
           label = expression(bar(y)),
           hjust = 0) +  # Mean label
  annotate("text", x = max(df$x) - 0.2, y = max(df$y) - 0.2,
           label = expression(hat(y) == hat(beta)[0] + hat(beta)[1]*x),
           color = "#F8766D", size = 5, hjust = 0) +  # Regression equation
  labs(x = "x", y = "y") +
  theme_minimal(base_size = 14) +
  coord_cartesian(clip = "off")
```




::: definition
For any simple linear regression, the distance between the mean of
dependent value and estimated dependent value is called sum square
regression (SSR), which is given by
$$SSR = \sum_{i=1}^{n}(\hat{y_i} - \bar{y})^2.$$
:::

Note that sum square regression (SSR) is an explained variation.

**Summary**

Now, let's zoom in to see SST, SSE and SSR. Note that the relationship
of these measures is that total deviation is equal to unexplained
deviation (error) plus explained deviation (regression), that is:
$(y_i - \bar{y}) = (y_i - \hat{y_i}) + (\hat{y_i} - \bar{y})$.

We square all three deviations for each one of our data points, and sum
over all $n$ points. Here, cross terms drop out, and we are left with
the following equation:
$$\sum_{i=1}^{n}(y_i - \bar{y})^2 = \sum_{i=1}^{n} (y_i - \hat{y_i})^2 + \sum_{i=1}^{n}(\hat{y_i} - \bar{y})^2,$$
SST = SSE + SSR.

Total sum of squares = Sum of squares for error + Sum of squares for
regression.


```{r fig.cap = "Visual representation of regression variability. The blue point represents the observed value $y_i$, the red point shows the predicted value $\\hat{y}_i$, and the dashed line represents the mean $\\bar{y}$. The green arrow illustrates the total variability (SST), the pink arrow shows the explained variability (SSR), and the orange curly bracket represents the residual (unexplained) variability (SSE).", echo=FALSE, out.width='90%',fig.align='center', message=FALSE, warning=FALSE, results='hide',}
```



**Coefficient of Determination ($r^2$)**

Moreover, we can use SST, SSE and SSR to calculate another value which
is important in simple linear regression, that is coefficient of
determination. It is proportion of variability in $y$ which is explained
by $x$.

::: definition
We define the coefficient of determination as the sum of squares due to
the regression divided by the total sum of squares.
$$r^2 = \frac{SSR}{SST} = 1 - \frac{SSE}{SST}.$$ The coefficient of
determination can be interpreted as the proportion of the variation in
$Y$ that is explained by the regression relationship of $Y$ with $X$ (or
the proportion of the total corrected sum of squares explained by the
regression). Note that: $0 \leq r^2 \leq 1.$
:::




















## Inference for Simple Linear Regression

In previous chapters, we focused on estimating regression parameters and
interpreting the fitted line. In this chapter, we take a step further by
conducting formal inference on the slope and intercept of a simple
linear regression model. We examine the distribution of errors, assess
variability, and introduce the idea of using hypothesis tests and
confidence intervals to evaluate whether the linear relationship
observed in the data is statistically significant.

We begin by introducing the regression model and exploring the
assumptions necessary to perform inference on the coefficients,
particularly the slope.

$$Y = \beta_0 + \beta_1 X + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, \sigma^2)$$

Can perform inference on $\beta_0$ and $\beta_1$, however we are usually
more interested in $\beta_1$

What does the error term $\varepsilon \sim \mathcal{N}(0, \sigma^2)$
mean?\
**At each** value of $X$, the errors are distributed normally with a
mean of zero and a constant variance.



Can verify with residual plots (assumptions).

We estimate $\sigma^2$ with a value we call $S^2$ and use $S^2$ for
inference.

### Estimating Variance in Linear Regression {#estimating-variance-in-linear-regression}

$$Y = \beta_0 + \beta_1 X + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, \sigma^2)$$

**Estimate $\sigma^2$ with $S^2$**:

$$S^2 = \frac{\sum_{i=1}^n e_i^2}{n - 2}
= \frac{\sum_{i=1}^n (y_i - \hat{y}_i)^2}{n - 2}
= \frac{SSE}{n - 2}$$

[*notice similarity*]{style="color: red"}

$$S_x^2 = \frac{\sum_{i=1}^n (x_i - \bar{x})^2}{n - 1}
\quad {\text{(sample variance, }\bar{x} \text{ estimated)}}$$

$$S = +\sqrt{S^2}
\quad {\text{(estimate of standard deviation)}} \\$$

**In calculating $S^2$, why do we divide by $n-2$?**

Since we estimate 2 unknown parameters in the model (both $\beta_0$ and
$\beta_1$), which are used in the calculation of $S^2$.

::: tcolorbox
We have data on an explanatory variable $x$ and a response variable $y$
for $n$ individuals. From the data, calculate the means $\bar{x}$ and
$\bar{y}$ and the standard deviations $S_x$ and $S_y$ of the two
variables, and their correlation $r$. The least-squares regression line
is the line:

$$\hat{y} = b_0 + b_1 x$$

with *slope*

$$b_1 = r \frac{S_y}{S_x}$$

and *intercept*

$$b_0 = \bar{y} - b_1 \bar{x}$$
:::

::: definition
The **least-squares regression line** of $y$ on $x$ is the line that
makes the sum of the squares of the vertical distances of the data
points from the line as small as possible.
:::

::: example
Suppose an appliance store conducts a 5-month experiment to determine
the effect of advertising on sales revenue. The results are shown in a
table below. The relationship between sales revenue, $y$, and
advertising expenditure, $x$, is hypothesized to follow a first-order
linear model, that is,

$$y = \beta_0 + \beta_1 x + \varepsilon$$

where

$$\begin{aligned}
y & = \text{dependent variable} \\
x & = \text{independent variable} \\
\beta_0 & = \text{$y$-intercept} \\
\beta_1 & = \text{slope of the line} \\
\varepsilon & = \text{error variable}
\end{aligned}$$

| Month | Expenditure $x$ (hundreds) | Revenue $y$ (thousands) |
|-------|-----------------------------|--------------------------|
| 1     | 1                           | 1                        |
| 2     | 2                           | 1                        |
| 3     | 3                           | 2                        |
| 4     | 4                           | 2                        |
| 5     | 5                           | 4                        |



The question is this: How can we best use the information in the sample
of five observations in our table to estimate the unknown $y$-intercept
$\beta_0$ and slope $\beta_1$?

We are given:
$$\bar{x} = 3, \quad \bar{y} = 2, \quad S_x = 1.5811, \quad S_y = 1.2247, \quad S_{xy} = 1.75$$

Then, the slope of the least squares line is

$$b_1 = r \frac{S_y}{S_x} = (0.9037) \left( \frac{1.2247}{1.5811} \right) = 0.7$$

and

$$b_0 = \bar{y} - b_1 \bar{x} = 2 - (0.7)(3) = -0.1$$

The least squares line is thus:

$$\hat{y} = -0.1 + 0.7x$$
:::

### Confidence Intervals and Hypothesis Tests 

We have $n$ observations on an explanatory variable $x$ and a response
variable $y$. Our goal is to study or predict the behavior of $y$ for
given values of $x$.

- For any fixed value of $x$, the response $y$ varies according to a
  Normal distribution. Repeated measures $y$ are independent of each
  other.

- The mean response $\mu_y$ has a straight-line relationship with $x$:
  $\mu_y = \beta_0 + \beta_1 x$. The slope $\beta_1$ and intercept
  $\beta_0$ are **unknown** parameters.

- The standard deviation of $y$ (call it $\sigma$) is the same for all
  values of $x$. The value of $\sigma$ is **unknown**. The regression
  model has three parameters, $\beta_0$, $\beta_1$, and $\sigma$.

Thus, if $$\hat{y}_i = \hat{\beta}_0 + \hat{\beta}_1 x_i$$ is the
predicted value of the $i$th $y$ value, then the deviation of the
observed value $y_i$ from $\hat{y}_i$ is the difference
$y_i - \hat{y}_i$ and the sum of squares of deviations to be minimized
is

$$SSE = \sum_{i=1}^{n} (y_i - \hat{y}_i)^2 = \sum_{i=1}^{n} [y_i - (\hat{\beta}_0 + \hat{\beta}_1 x_i)]^2.$$

The quantity *SSE* is also called the **sum of squares for error**.
$$\begin{aligned}
\text{Fitted Value:} \quad & \hat{y}_i = \hat{\beta}_0 + \hat{\beta}_1 x_i \\
\text{Residual:} \quad & \hat{\varepsilon}_i = y_i - \hat{y}_i
\end{aligned}$$

The **regression standard error** is

$$s = \sqrt{\frac{1}{n - 2} \sum \text{residual}^2} 
= \sqrt{\frac{1}{n - 2} \sum_{i=1}^{n} (y_i - \hat{y}_i)^2} 
= \sqrt{\frac{SSE}{n - 2}}$$

Use $s$ to estimate the **unknown** $\sigma$ in the regression model


The standard error of $\hat{\beta}_1$ is the standard deviation of the
sampling distribution of $\hat{\beta}_1$ (estimate of slope $\beta_1$):

$$SE(\hat{\beta}_1) = \frac{s}{\sqrt{\sum_{i=1}^{n} (x_i - \bar{x})^2}} 
= \frac{s}{\sqrt{(n - 1) s_x^2}}$$

**Confidence Interval for the Slope**

$$\hat{\beta}_1 \pm t_{(n-2, \, \alpha/2)} \cdot SE(\hat{\beta}_1)
\quad = \quad 
\hat{\beta}_1 \pm t_{(n-2, \, \alpha/2)} \cdot \frac{s}{\sqrt{\sum_{i=1}^{n} (x_i - \bar{x})^2}}$$

:::: example
Revisit the example on advertising and sales and construct a 95%
confidence interval on the slope. Provide an interpretation of the CI.

From earlier: $$\hat{y} = -0.1 + 0.7x$$

::: center
   **$x$**   **$y$**   **$\hat{y}$**   **$y - \hat{y}$**   **$(y - \hat{y})^2$**   **$(x - \bar{x})^2$**
  --------- --------- --------------- ------------------- ----------------------- -----------------------
      1         1           0.6               0.4                  0.16                      4
      2         1           1.3              -0.3                  0.09                      1
      3         2           2.0               0.0                  0.00                      0
      4         2           2.7              -0.7                  0.49                      1
      5         4           3.4               0.6                  0.36                      4
:::

We are given:
$$\sum x_i = 15, \quad \bar{x} = \frac{15}{5} = 3, \quad SSE = 1.10, \quad \sum (x_i - \bar{x})^2 = 10$$

**Step 1: Estimate variance and standard deviation**
$$s^2 = \frac{SSE}{n-2} = \frac{1.10}{5 - 2} = 0.3667
\quad \Rightarrow \quad
s = \sqrt{0.3667} = 0.6055$$

**Step 2: Compute standard error of $\hat{\beta}_1$**
$$SE(\hat{\beta}_1) = \frac{s}{\sqrt{\sum (x_i - \bar{x})^2}} 
= \frac{0.6055}{\sqrt{10}} = 0.1914$$

**Step 3: Determine critical $t$-value**

$$n - 2 = 3, \quad \alpha = 0.05, \quad \alpha/2 = 0.025
\Rightarrow \quad
t_{(3, 0.025)} = 3.182$$

**Step 4: Construct CI for the slope**
$$\hat{\beta}_1 \pm t_{(n-2, \alpha/2)} \cdot SE(\hat{\beta}_1)
= 0.7 \pm 3.182 \cdot 0.1914 = 0.7 \pm 0.6092$$

$$\Rightarrow \text{CI: } (0.0908, \; 1.3092)$$

**Interpretation:** We are 95% confident the slope ($\beta_1$) for this
model lies between 0.0908 and 1.3092.
::::

<!-- ### Interpreting Confidence Intervals for $\beta_1$ {#interpreting-confidence-intervals-for-beta_1 .unnumbered} -->

::: center
  ----------------- ----------------------------------------------------------
    **Suppose CI:** $(-, -)$
                    Suggests $\beta_1$ has a **negative** sign.
                    Suggests negative correlation, potentially good model.
                    
    **Suppose CI:** $(+, +)$
                    Suggests $\beta_1$ has a **positive** sign.
                    Suggests positive correlation, potentially good model.
                    
    **Suppose CI:** $(-, +)$
                    $\beta_1 = 0$ is plausible.
                    Suggests **no linear relationship** between $x$ and $y$.
  ----------------- ----------------------------------------------------------
:::

In cases where the CI does not contain zero, we can infer the sign of
the slope (just not the steepness)


The **regression standard error** is

$$s = \sqrt{\frac{1}{n - 2} \sum_{i=1}^n (y_i - \hat{y}_i)^2} = \sqrt{\frac{SSE}{n - 2}} = \sqrt{\frac{1.1}{3}} = 0.6055$$

Use $s$ to estimate the **unknown** $\sigma$ in the regression model


A level $C$ confidence interval for the slope $\beta_1$ of the true
regression line is

$$b_1 \pm t^* SE_{b_1}$$

In this formula, the standard error of the least-squares slope $b$ is

$$SE_{b_1} = \frac{s}{\sqrt{\sum (x_i - \bar{x})^2}} = \frac{s}{\sqrt{(n - 1) S_x^2}}$$

and $t^*$ is the critical value for the $t(n - 2)$ density curve with
area $C$ between $-t^*$ and $t^*$.

::: tcolorbox
**Hypotheses:** $$\begin{aligned}
H_0\!: \beta_1 = 0 \quad & \text{vs.} \quad H_a\!: \beta_1 > 0 \\
H_0\!: \beta_1 = 0 \quad & \text{vs.} \quad H_a\!: \beta_1 < 0 \\
H_0\!: \beta_1 = 0 \quad & \text{vs.} \quad H_a\!: \beta_1 \neq 0 \quad \text{\textit{(most common)}}
\end{aligned}$$

**Test Statistic:**
$$t = \frac{\hat{\beta}_1 - 0}{SE(\hat{\beta}_1)} = \frac{\hat{\beta}_1}{\dfrac{s}{\sqrt{\sum_{i=1}^n (x_i - \bar{x})^2}}}$$

**Reference distribution:** $t$ distribution with $n - 2$ degrees of
freedom.

*Note:* A test statistic always follows the form:
$$\text{test stat} = \frac{\text{statistic} - \text{hypothesized value}}{\text{SE(statistic)}}$$
:::

::: example
For the advertising example, perform a two-sided hypothesis test on the
slope.

**Hypotheses:** $$H_0\!: \beta_1 = 0 
\qquad 
H_a\!: \beta_1 \neq 0$$

**Test Statistic:**
$$t^* = \frac{\hat{\beta}_1 - 0}{\dfrac{s}{\sqrt{\sum (x_i - \bar{x})^2}}}
= \frac{0.7 - 0}{0.6055 / \sqrt{10}} = 3.6558$$

**Reference Distribution:** $t$ distribution with $n - 2 = 5 - 2 = 3$
degrees of freedom.

**Decision Rule:**

Using a two-tailed test:
$$\text{p-value} = 2 \cdot P(T_3 > 3.6558) < 0.01 \Rightarrow \text{p-value} < 0.05$$

**Conclusion:** Since $p$-value $< 0.05$, we reject $H_0$ and conclude
$H_a\!: \beta_1 \neq 0$.

*Interpretation:* The slope should be included in the model. There is
significant evidence of a linear relationship between advertising and
sales revenue


For the advertising-sales example, a 95% Confidence Interval for the
slope $\beta_1$ is
$$0.7 \pm 3.182 \left( \frac{0.6055}{\sqrt{10}} \right)$$
$$0.7 \pm 0.6092$$

Thus, we estimate with 95% confidence that the interval from 0.0908 and
1.3092 includes the parameter $\beta_1$.
:::

We can also test hypotheses about the slope $\beta_1$. The most common
hypothesis is

$$H_0 : \beta_1 = 0.$$

A regression line with slope 0 is horizontal. That is, the mean of $y$
does not change at all when $x$ changes. So this $H_0$ says that there
is no true linear relationship between $x$ and $y$.

::: tcolorbox
To test the hypothesis $H_0 : \beta_1 = 0$, compute the $t$ statistic
$$t = \frac{b_1}{SE_{b_1}}.$$

In terms of a random variable $T$ having the $t(n - 2)$ distribution,
the P-value for a test of $H_0$ against: $$\begin{aligned}
H_a : \beta_1 \ne 0 & \quad \text{is } 2P(T > |t|). \\
H_a : \beta_1 > 0 & \quad \text{is } P(T > t). \\
H_a : \beta_1 < 0 & \quad \text{is } P(T < t).
\end{aligned}$$
:::

::::: example
$$\alpha = 0.05$$

1.  $H_0: \beta_1 = 0 \quad \text{vs} \quad H_a: \beta_1 > 0$

2.  $t^* = \frac{b_1}{SE_{b_1}} = \frac{0.7}{0.1914} = 3.6572$

3.  $P\text{-value} = P(T > t) = P(T > 3.6572) \quad \text{d.f.} = n - 2 = 5 - 2 = 3.$\
    Using t-distribution table, $0.01 < P\text{ value} < 0.025$

4.  Since $P\text{-value} < \alpha = 0.05$, we reject $H_0$.

Our example (different $H_a$) $$\alpha = 0.05$$

1.  $H_0: \beta_1 = 0 \quad \text{vs} \quad H_a: \beta_1 \neq 0$

2.  $t^* = \frac{b_1}{SE_{b_1}} = \frac{0.7}{0.1914} = 3.6572$

3.  $P\text{-value} = 2P(T > |t|) = 2P(T > 3.6572) \quad \text{d.f.} = n - 2 = 5 - 2 = 3.$\
    Using Table 3, $0.02 < P\text{ value} < 0.05$

4.  Since $P\text{-value} < \alpha = 0.05$, we reject $H_0$.

**R code**

::: tcolorbox
    x = c(1, 2, 3, 4, 5);
    y = c(1, 1, 2, 2, 4);
    mod = lm(y~x);
    summary(mod);
:::

**R Output**

::: tcolorbox
    ## 
    ## Call:
    ## lm(formula = y~x)
    ## 
    ## Residuals:
    ##       1        2        3        4        5 
    ##  4.000e-01 -3.000e-01 -3.886e-16 -7.000e-01  6.000e-01 
    ## 
    ## Coefficients:
    ##             Estimate Std. Error t value Pr(>|t|)   
    ## (Intercept)  -0.1000     0.6351  -0.157  0.8849    
    ## x             0.7000     0.1915   3.656  0.0354 *  
    ## ---
    ## Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
    ## 
    ## Residual standard error: 0.6055 on 3 degrees of freedom
:::

$$\hat{y} = \hat{\beta}_0 + \hat{\beta}_1 x = -0.10 + 0.70x$$
$$SE(\hat{\beta}_1) = 0.1915$$

By default, R conducts the following test for each coefficient:

$$\begin{aligned}
    H_0&: \beta_j = 0 \\
    H_a&: \beta_j \ne 0 \quad \text{(two sided)}
\end{aligned}$$

Test statistic: $$t^* = \frac{\hat{\beta}_j - 0}{SE(\hat{\beta}_j)}$$

For advertising and sales data: $$\begin{aligned}
    H_0&: \beta_1 = 0 \\
    H_a&: \beta_1 \ne 0
\end{aligned}$$

$$t^* = \frac{\hat{\beta}_1 - 0}{SE(\hat{\beta}_1)} = \frac{0.70}{0.1915} = 3.656$$
:::::

::: nt
**Review**

$$\begin{aligned}
y &= \beta_0 + \beta_1 x + \varepsilon, \quad \varepsilon \sim N(0, \sigma^2) \\
\hat{y} &= \hat{\beta}_0 + \hat{\beta}_1 x \\
\hat{\beta}_1 &= \frac{\sum (x_i - \bar{x})(y_i - \bar{y})}{\sum (x_i - \bar{x})^2} = \frac{s_{xy}}{s_{xx}} \\
\hat{\beta}_0 &= \bar{y} - \hat{\beta}_1 \bar{x} \\
\end{aligned}$$

**r:** coefficient of correlation (strength)\
**r$^2$:** coefficient of determination (% variability)

$$\begin{aligned}
s^2 &= \frac{\sum_{i=1}^n (y_i - \hat{y}_i)^2}{n - 2} = \frac{SSE}{n - 2} \\
s &= \sqrt{s^2} \\
SE(\hat{\beta}_1) &= \frac{s}{\sqrt{s_{xx}}} \\
CI &: \hat{\beta}_1 \pm t_{n - 2, \alpha/2} \cdot SE(\hat{\beta}_1) \\
\end{aligned}$$ Hypothesis test: $$\begin{aligned}
H_0 &: \beta_1 = 0 \\
\text{Test stat:}\quad t &= \frac{\hat{\beta}_1 - 0}{SE(\hat{\beta}_1)}
\end{aligned}$$
:::

We square all three deviations for each one of our data points, and sum
over all $n$ points. Here, cross terms drop out, and we are left with
the following equation:

$$\sum_{i=1}^{n}(y_i - \bar{y})^2 = \sum_{i=1}^{n}(y_i - \hat{y}_i)^2 + \sum_{i=1}^{n}(\hat{y}_i - \bar{y})^2$$

$$\text{SST} = \text{SSE} + \text{SSR}$$

Total sum of squares = Sum of squares for error + Sum of squares for
regression


$$\begin{aligned}
SSE &= \sum_{i=1}^{n} (y_i - \hat{y}_i)^2 \\
    &= \sum_{i=1}^{n} (y_i - \bar{y})^2 - \hat{\beta}_1 \sum_{i=1}^{n} (x_i - \bar{x})(y_i - \bar{y}) \\
    &= S_{YY} - \hat{\beta}_1 S_{XY}
\end{aligned}$$

Notice that this provides an easier computational method of finding
SSE


**R output (Additional example)**

::: tcolorbox
    > summary(model);

    Call:
    lm(formula = camrys$Price ~ Odometer, data = camrys)

    Residuals:
         Min       1Q   Median       3Q      Max 
    -0.68679 -0.27263  0.00521  0.23210  0.70071 

    Coefficients:
                 Estimate Std. Error t value Pr(>|t|)    
    (Intercept)  17.248727   0.182093   94.72   <2e-16 ***
    Odometer     -0.066861   0.004975  -13.44   <2e-16 ***
    ---
    Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

    Residual standard error: 0.3265 on 98 degrees of freedom
    Multiple R-squared:  0.6483,    Adjusted R-squared:  0.6447 
    F-statistic: 180.6 on 1 and 98 DF,  p-value: < 2.2e-16
:::

### ANOVA Table for Simple Linear Regression

Analysis of Variance (ANOVA) is a statistical method used to assess
whether variation in a response variable can be explained by predictor
variables in a regression model. It summarizes sources of variation
using sums of squares, degrees of freedom, and mean squares in a
structured table format.

  **Source of Variation**    **Sum of Squares**   **Degrees of Freedom**        **Mean Square**                **Computed F**
  ------------------------- -------------------- ------------------------ ---------------------------- ------------------------------
  Regression                        SSR                     1                         SSR               $\dfrac{SSR}{SSE / (n - 2)}$
  Error                             SSE                  $n - 2$           $s^2 = \dfrac{SSE}{n - 2}$  
  Total                             SST                  $n - 1$                                       

For the general multivariate regression model: $$\begin{aligned}
    Y &= \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_p X_p + \varepsilon, \\
      &\quad \varepsilon \sim N(0, \sigma^2)
\end{aligned}$$ with $p$ predictors


ANOVA can be used for testing: $$\begin{aligned}
    H_0&: \beta_1 = \beta_2 = \cdots = \beta_p = 0 \\
    H_a&: \text{At least one } \beta_j \ne 0, \quad j = 1, \ldots, p
\end{aligned}$$

**Test Statistic:** $$\begin{aligned}
    F &= \dfrac{MSR}{MSE} = \dfrac{SSR/p}{SSE / (n - p - 1)} \sim F(p, n - p - 1)
\end{aligned}$$

**Reference distribution:** $F$ with numerator df = $p$, denominator df
= $n - p - 1$

::::: example
We fitted a simple linear regression model using:

::: tcolorbox
    x = c(1,2,3,4,5);
    y = c(1,1,2,2,4);
    mod = lm(y~x);
    anova(mod);
:::

The ANOVA output was:

::: tcolorbox
    ## Analysis of Variance Table
    ##
    ## Response: y
    ##             Df  Sum Sq Mean Sq F value   Pr(>F)
    ## x            1     4.9    4.9000  13.364  0.03535 *
    ## Residuals    3     1.1    0.3667
    ## ---
    ## Signif. codes:
    ## 0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
:::

**Interpretation:**

- The regression model includes one predictor $x$, so the degrees of
  freedom for regression is 1.

- The sum of squares for regression is $SSR = 4.9$, and for residuals
  $SSE = 1.1$.

- Mean squares are calculated as:
  $$MSR = \frac{SSR}{1} = 4.9, \quad MSE = \frac{SSE}{n - 2} = \frac{1.1}{3} = 0.3667$$

- The F-statistic is:
  $$F = \frac{MSR}{MSE} = \frac{4.9}{0.3667} \approx 13.364$$

- The p-value is $\approx 0.03535$, indicating that the predictor is
  significant at the 5% level.

**Conclusion:** Since the p-value is less than 0.05, we reject $H_0$ and
conclude that $x$ has a statistically significant linear relationship
with $y$.
:::::

:::::::::::: example
We consider data on apartments near UTM, with price (in thousands of
dollars), area (in 100 square feet), and number of beds and baths.

::: center
  ----------------- ---------------------- ---------- -----------
      **Price**            **Area**         **Beds**   **Baths**
   ($\times 1000$)   ($\times 100$ sq ft)             
         620                 11.0              2           2
         590                 6.5               2           1
         620                 10.0              2           2
         700                 8.4               2           2
         680                 8.0               2           2
         500                 5.7               1           1
         760                 12.0              2           2
         800                 14.0              3           1
         660                 7.3               2           1
  ----------------- ---------------------- ---------- -----------
:::
```{r fig.cap="Plot of Price vs Area for Apartments near UTM",echo=FALSE, fig.align='center', out.width='80%'}
library(ggplot2)

# Data
area <- c(5.5, 6.5, 7.5, 8.0, 8.3, 9.8, 11.0, 12.5, 13.8, 10.2)
price <- c(490, 590, 660, 700, 710, 630, 630, 760, 810, 680)
df <- data.frame(area, price)

# Plot
ggplot(df, aes(x = area, y = price)) +
  geom_point(size = 3, color = "#000000") +
  labs(
    x = "Area (sq ft)",
    y = "Price (x$1000)",
    title = "Plot of Price vs Area for Apartments near UTM"
  ) +
  theme_minimal(base_size = 14)

```


   **Price**   **Area**   $(x - \bar{x})$   $(y - \bar{y})$          $(x - \bar{x})(y - \bar{y})$                 $(x - \bar{x})^2$
  ----------- ---------- ----------------- ----------------- --------------------------------------------- --------------------------------
      620        11.0           1.8              -38.9                          -70.02                                   3.24
      590        6.5           -2.7              -68.9                          186.03                                   7.29
      620        10.0           0.8              -38.9                          -31.12                                   0.64
      700        8.4           -0.8              41.1                           -32.88                                   0.64
      680        8.0           -1.2              21.1                           -25.32                                   1.44
      500        5.7           -3.5             -158.9                          556.15                                  12.25
      760        12.0           2.8              101.1                          283.08                                   7.84
      800        14.0           4.8              141.1                          677.28                                  23.04
      660        7.3           -1.9               1.1                            -2.09                                   3.61
    **Sum**      82.9                                         $\sum (x - \bar{x})(y - \bar{y}) = 1541.11$   $\sum (x - \bar{x})^2 = 60.00$

  : Deviation table for computing $\hat{\beta}_1$ and $\hat{\beta}_0$

The sample means are:
$$\bar{y} = \frac{\sum y}{n} = \frac{5930}{9} = 658.89, \qquad
\bar{x} = \frac{\sum x}{n} = \frac{82.9}{9} = 9.21$$

#### Finding the Regression Coefficients {#finding-the-regression-coefficients .unnumbered}

To compute the least squares regression line, we calculate the slope and
intercept using the formulas:

$$\hat{\beta}_1 = \frac{S_{xy}}{S_{xx}} = \frac{1541.11}{60} = 25.69$$
$$\hat{\beta}_0 = \bar{y} - \hat{\beta}_1 \bar{x} = 658.89 - (25.69)(9.21) = 422.28$$

#### Equation of the Regression Line {#equation-of-the-regression-line .unnumbered}

Using the values above, we write the estimated regression equation as:

$$\hat{y} = \hat{\beta}_0 + \hat{\beta}_1 x = 422.28 + 25.69x$$

This equation gives the predicted apartment price (in \$1000s) based on
area (in 100 sq ft).

#### Interpretation of Coefficients {#interpretation-of-coefficients .unnumbered}

The slope $\hat{\beta}_1 = 25.69$ means that for every additional 100 sq
ft in area, we expect the apartment price to increase by approximately
\$25,690 on average


The intercept $\hat{\beta}_0 = 422.28$ suggests the predicted price when
the area is zero. While this has no practical interpretation in this
context, it is a necessary component of the regression model.

#### Interpolation and Extrapolation {#interpolation-and-extrapolation .unnumbered}

To estimate the price of an apartment with an area of 800 sq ft (i.e.,
$x = 8$), we compute:

$$\hat{y} = 422.28 + 25.69(8) = 627.8 \quad (\$1000)$$

Since 8 is within the range of observed values, this is an example of
**interpolation.**

For an apartment with 2,500 sq ft ($x = 25$):

$$\hat{y} = 422.28 + 25.69(25) = 1064.53 \quad (\$1000)$$

This is an example of **extrapolation**, and such predictions should be
treated with caution since they lie outside the data range


We can create a simple linear regression model in R using the `lm`
command:

**R code**

::: tcolorbox
    lm(y ~ x, data = data_source)
:::

The data is available in the `apt_around_utm.csv` file.

**R code**

::: tcolorbox
    apt = read.csv(file.choose())
    # apt = read.csv("~/PATH_TO_FILE/apt_around_utm.csv")

    apt_model = lm(price ~ area, data = apt)
:::

**R output**

::: tcolorbox
    > apt_model

    Call:
    lm(formula = price ~ area, data = apt)

    Coefficients:
    (Intercept)       area  
         422.26       25.69  
:::

We can compute the residuals and the sum of squared errors (SSE) using
the table below:

::: center
   $y$   $x$    $\hat{y}$   $y - \hat{y}$   $(y - \hat{y})^2$  
  ----- ------ ----------- --------------- ------------------- --
   620   11.0    704.85        -84.85            7198.73       
   590   6.5     589.24         0.76              0.58         
   620   10.0    679.16        -59.16            3499.36       
   700   8.4     638.05         61.95            3837.62       
   680   8.0     627.78         52.22            2727.4        
   500   5.7     568.69        -68.69            4718.13       
   760   12.0    730.54         29.46            868.17        
   800   14.0    781.92         18.08            327.06        
   660   7.3     609.79         50.21            2520.79       
:::

Recall our fitted regression model: $$\hat{y} = 25.69x + 422.26$$

$$\text{SSE} = \sum (y_i - \hat{y}_i)^2 = 25,\!697.83$$

We now conduct a hypothesis test on the slope $\beta_1$ at the 5%
significance level.

**Step 1: Hypotheses**
$$H_0: \beta_1 = 0 \quad \text{vs.} \quad H_a: \beta_1 \ne 0$$

**Step 2: Test statistic**
$$s^2 = \frac{SSE}{n - 2} = \frac{25,\!697.83}{7} = 3670.26$$
$$s = \sqrt{3670.26} = 60.58$$
$$SE(\hat{\beta}_1) = \frac{s}{\sqrt{S_{xx}}} = \frac{60.58}{\sqrt{60}} = 7.82$$
$$t = \frac{\hat{\beta}_1 - 0}{SE(\hat{\beta}_1)} = \frac{25.69}{7.82} = 3.284$$

**Step 3: Conclusion** Using $t$-distribution with 7 degrees of freedom:

$$0.005 < p\text{-value} < 0.01$$

Since $p$-value $< 0.05$, we reject $H_0$ and conclude that there is
sufficient evidence that $\beta_1 \ne 0$.This suggests there is a
statistically significant relationship between price and area for
apartments near UTM.

**Final Check: Total Sum of Squares**

::: center
   $y$   $x$    $\hat{y}$   $(y - \hat{y})^2$   $(y - \bar{y})^2$   $(\hat{y} - \bar{y})^2$  
  ----- ------ ----------- ------------------- ------------------- ------------------------- --
   620   11.0    704.85          7198.73             2112.00                1512.35          
   590   6.5     589.24           0.58               4850.88                4745.68          
   620   10.0    679.16          3499.36             410.73                 1512.35          
   700   8.4     638.05          3837.62             434.20                 1690.12          
   680   8.0     627.78          2727.4              968.04                 445.68           
   500   5.7     568.69          4718.13             8136.08                2524.68          
   760   12.0    730.54          868.17              5133.21               10223.46          
   800   14.0    781.92          327.06              1513.47               19912.35          
   660   7.3     609.79          2520.79             2410.45                 1.23            
:::

$$SSE + SSR = SST = 25,\!697.83 + 39,\!591.06 = 65,\!288.90$$ This
confirms the ANOVA identity: Total = Explained + Residual

We previously estimated the model:

$$\hat{y} = 25.69x + 422.26$$

**Coefficient of Determination and Correlation**

$$r^2 = \frac{SSR}{SST} = \frac{39591.06}{65288.90} = 0.6063 = 60.63\%$$

Interpretation: Approximately 60.63% of the variability in price is
explained by the regression model.

$$r = \pm \sqrt{r^2} = \pm \sqrt{0.6063} = \pm 0.779$$

Since $\hat{\beta}_1 > 0$, we choose the positive root:

$$r = 0.779$$

Interpretation: There is a strong positive correlation between apartment
area and price.

**R code**

::: tcolorbox
    model = lm(y~x, data = data\_source) 
    summary(model)
:::

**R code**

::: tcolorbox
    apt\_model = lm(price ~ area, data = apt)
    summary(apt\_model)
:::

**R code**

::: tcolorbox
    Call:
    lm(formula = price ~ area, data = apt)

    Coefficients:
                Estimate Std. Error t value Pr(>|t|)
    (Intercept)  422.256   74.834     5.643  0.00078 ***
    area         25.690    7.823      3.284  0.01341 *

    Residual standard error: 60.59 on 7 degrees of freedom
    Multiple R-squared: 0.6064,    Adjusted R-squared: 0.5502 
    F-statistic: 10.78 on 1 and 7 DF,  p-value: 0.01341
:::

**Two-sided Test for Slope Coefficient**

By default, R performs a two-sided test:
$$H_0: \beta_1 = 0 \quad \text{vs.} \quad H_a: \beta_1 \neq 0$$

Test statistic:
$$t^* = \frac{\hat{\beta}_1 - 0}{SE(\hat{\beta}_1)} = \frac{25.690 - 0}{7.823} = 3.284$$

$$t^* \sim t_{(n - 2)}  \quad \text{with } df = 9 - 2 = 7$$

**p-value** is the total shaded area in both tails. From R output:
$$\text{p-value} = 0.01341$$
::::::::::::

::: definition
- **Interpolation** is calculating predicted values of $y$ using our
  linear model while working within the range of $x$ in which data was
  available to construct our model.

- **Extrapolation** is calculating predicted values of $y$ using our
  linear model outside the range of $x$ used to obtain the linear model.

- Interpolation is usually safe if we have a good linear model.

- Extrapolation must be performed carefully since extrapolations that
  are done without any foresight can be very inaccurate.
:::

### Residual Plots

Residual plots are used to verify assumptions related to the error terms
in a regression model.

$$Y = \beta_0 + \beta_1 X + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, \sigma^2)$$

The assumption $\varepsilon \sim \mathcal{N}(0, \sigma^2)$ implies:

- Mean of errors is 0

- Constant variance of errors (homoscedasticity)

We plot the residuals: $$e_i = y_i - \hat{y}_i$$ against the fitted
values $\hat{y}_i$ to assess these assumptions.

<!-- ### What to Look for in a Good Residual Plot {#what-to-look-for-in-a-good-residual-plot .unnumbered} -->

If the assumption $\varepsilon \sim \mathcal{N}(0, \sigma^2)$ is
satisfied, the residual plot should have the following features:

1.  **Random scattering:** No obvious pattern in residuals.

    - A pattern (e.g., curve) may indicate a non-linear relationship.

    - Random scattering also suggests independence of errors.

2.  **Constant variance:** Residuals should fall within a horizontal
    band, roughly half above and half below zero.

    - Suggests constant variance (homoscedasticity).

3.  **No influential points or clustering:** The plot should not show
    isolated influential observations or clustering.

<!-- # ```{r residual-stack, fig.width=6, fig.height=16, dev="png", out.width="100%", fig.align="center", message=FALSE, warning=FALSE} -->

```{r fig.cap="Examples of residual plots which satisfy assumpations and which violate assumptions", echo=FALSE, warning=FALSE, fig.align="center", out.width='70%', fig.width=6, fig.height=16}
# Load libraries
library(ggplot2)
library(patchwork)

# Common theme for all plots
theme_clean <- theme_minimal(base_size = 12)

# 1. Satisfies Assumptions
set.seed(1)
x1 <- 1:30
y1 <- 3 + 0.5 * x1 + rnorm(30, 0, 1)
model1 <- lm(y1 ~ x1)
df1 <- data.frame(Fitted = fitted(model1), Residuals = resid(model1))

p1 <- ggplot(df1, aes(Fitted, Residuals)) +
  geom_point(color = "#E68613", size = 2) +
  geom_hline(yintercept = 0, color = "#000000") +
  labs(
    title = "Satisfies Assumptions",
    x     = "Fitted Values",
    y     = "Residuals"
  ) +
  theme_clean

# 2. Non-linear Pattern
x2 <- seq(-2, 2, length.out = 30)
y2 <- x2^2 + rnorm(30, 0, 0.3)
model2 <- lm(y2 ~ x2)
df2 <- data.frame(Fitted = fitted(model2), Residuals = resid(model2))

p2 <- ggplot(df2, aes(Fitted, Residuals)) +
  geom_point(color = "#E68613", size = 2) +
  geom_hline(yintercept = 0, color = "#000000") +
  labs(
    title = "Non-linear Pattern",
    x     = "Fitted Values",
    y     = "Residuals"
  ) +
  theme_clean

# 3. Non-constant Variance
x3 <- 1:30
y3 <- 2 + 0.2 * x3 + rnorm(30, 0, sd = x3 / 10)
model3 <- lm(y3 ~ x3)
df3 <- data.frame(Fitted = fitted(model3), Residuals = resid(model3))

p3 <- ggplot(df3, aes(Fitted, Residuals)) +
  geom_point(color = "#E68613", size = 2) +
  geom_hline(yintercept = 0, color = "#000000") +
  labs(
    title = "Non-constant Variance",
    x     = "Fitted Values",
    y     = "Residuals"
  ) +
  theme_clean

# 4. Clustering
x4 <- c(rnorm(15, 3, 0.2), rnorm(15, 5, 0.2))
y4 <- 2 + 0.5 * x4 + rnorm(30, 0, 1)
model4 <- lm(y4 ~ x4)
df4 <- data.frame(Fitted = fitted(model4), Residuals = resid(model4))

p4 <- ggplot(df4, aes(Fitted, Residuals)) +
  geom_point(color = "#E68613", size = 2) +
  geom_hline(yintercept = 0, color = "#000000") +
  labs(
    title = "Clustering",
    x     = "Fitted Values",
    y     = "Residuals"
  ) +
  theme_clean

# Stack all four plots vertically
final_plot <- p1 / p2 / p3 / p4 +
  plot_layout(ncol = 1)  # one column

# Render the combined plot
print(final_plot)
```


::: tcolorbox
The model is: $Y = \beta_0 + \beta_1 X + \varepsilon$, where
$\varepsilon \sim \mathcal{N}(0, \sigma^2)$

- The relationship between $X$ and $Y$ is linear.

- Residuals:

  - are independent

  - have constant variance

  - are normally distributed

  These assumptions can be verified using residual plots.
:::
